### 2021-10-28 src-latex/main-paper.tex
add bib support

\documentclass[runningheads,svgnames]{llncs}
%
\usepackage[]{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{cleveref}
% \usepackage{tikz}
% \usepackage{nth}
% \usepackage{subcaption}
\usepackage{booktabs}
% \usepackage{array}
\usepackage{orcidlink}

% \usetikzlibrary{shadings,decorations,decorations.pathreplacing}
\renewcommand\UrlFont{\color{blue}\rmfamily}

\newcommand{\mcite}[1]{\text{\cite{#1}}}
\newcommand{\mref}[1]{\text{\ref{#1}}}

% \renewcommand{\baselinestretch}{0.97}

\begin{document}
%
\title{A Benchmark of Named Entity Recognition Approaches in Historical Documents\\
Application to 19$^{th}$ Century French Directories%
\thanks{This work was supported by the French National Research Agency:
Project SoDuCo, grant ANR-18-CE38-0013. ???FIXME need acknowledgments for directory sources???}}
%
\titlerunning{A Benchmark of NER Approaches in Historical Documents}
%
\author{LISTE PROVISOIRE PAR ORDRE ALPHA\\
%
Nathalie Abadie\inst{1}\orcidID{0000-1111-2222-3333} \and
Edwin Carlinet\inst{2,3}\orcidID{1111-2222-3333-4444} \and
Joseph Chazalon\inst{3}\orcidID{2222--3333-4444-5555} \and
Bertrand Duménieu\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{INST, CITY, COUNTRY \and
INST, CITY, COUNTRY\\
\email{example@example.com}\\
\url{https://www.example.com} \and
INST, CITY, COUNTRY\\
\email{example@example.com}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{First Section}
\subsection{A Subsection Sample}
we use spacy \mcite{spacy}

% ---- Bibliography ----
\bibliographystyle{splncs04}
\bibliography{ref}
\end{document}

### 2021-10-28 src-latex/main-paper.tex
Setup paper structure

@@
 Application to 19$^{th}$ Century French Directories%
+% FIXME "19$^{th}$-Century" with dash?
 \thanks{This work was supported by the French National Research Agency:
@@
 \begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-150--250 words.
+\input{parts/00-abstract.tex}
+\keywords{Historical documents \and Natural Language Processing \and Named Entity Recognition \and OCR noise \and Annotation cost.}
+\end{abstract}
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
-\end{abstract}
-%
-%
-%
-\section{First Section}
-\subsection{A Subsection Sample}
-we use spacy \mcite{spacy}
+% ---- Sections ----
+\input{parts/10-introduction.tex}
+\input{parts/20-related-work.tex}
+\input{parts/30-dataset.tex}
+\input{parts/40-ner-approaches.tex}
+\input{parts/50-xp-setup.tex}
+\input{parts/60-results.tex}
+\input{parts/80-conclusion.tex}
 

### 2021-10-28 src-latex/parts/00-abstract.tex
Setup paper structure

% !TeX root = ../main-paper.tex
The paper features a comparison of a selection of named entity recognition (NER) techniques which can be reasonably used in the context of historical document analysis.
Such natural language processing techniques identify which class each text token belong to, like "person name", "location", "number" or "activity".
Introducing a new public dataset built from 19th century French directories, 
we perform a practical assessment of the accuracy of, and the annotation cost (considering various training set sizes) needed by each technique.
We also study the impact of real OCR noise for several OCR systems on the performance of NER.
Experiments suggest that FIXME APPROACH NAME is the most effective approach on the dataset we processed.
All material is available freely at \url{https://github.com/soduco/paper-ner-bench-das22} and the latest version of is paper can be obtained at \url{TODO arxiv URL}.

### 2021-10-28 src-latex/parts/10-introduction.tex
Setup paper structure

--- 2021-10-28 src-latex/parts/00-abstract.tex
@@
 % !TeX root = ../main-paper.tex
-The paper features a comparison of a selection of named entity recognition (NER) techniques which can be reasonably used in the context of historical document analysis.
-Such natural language processing techniques identify which class each text token belong to, like "person name", "location", "number" or "activity".
-Introducing a new public dataset built from 19th century French directories, 
-we perform a practical assessment of the accuracy of, and the annotation cost (considering various training set sizes) needed by each technique.
-We also study the impact of real OCR noise for several OCR systems on the performance of NER.
-Experiments suggest that FIXME APPROACH NAME is the most effective approach on the dataset we processed.
-All material is available freely at \url{https://github.com/soduco/paper-ner-bench-das22} and the latest version of is paper can be obtained at \url{TODO arxiv URL}.
+\section{Introduction}

### 2021-10-28 src-latex/parts/50-xp-setup.tex
Setup paper structure

% !TeX root = ../main-paper.tex
\section{Experimental Setup}

\subsection{Implementation details}
we use spacy \mcite{spacy}

### 2021-10-28 src-latex/parts/00-abstract.tex
Suggestion of structure

@@
 % !TeX root = ../main-paper.tex
-The paper features a comparison of a selection of named entity recognition (NER) techniques which can be reasonably used in the context of historical document analysis.
+The paper features a comparison of a selection (FIXME be more specific/enumerate) of named entity recognition (NER) techniques which can be reasonably used in the context of historical document analysis.
 Such natural language processing techniques identify which class each text token belong to, like "person name", "location", "number" or "activity".
@@
 Experiments suggest that FIXME APPROACH NAME is the most effective approach on the dataset we processed.
-All material is available freely at \url{https://github.com/soduco/paper-ner-bench-das22} and the latest version of is paper can be obtained at \url{TODO arxiv URL}.
+FIXME add numerical results
+Results can be reproduced using data and code freely available at \url{https://github.com/soduco/paper-ner-bench-das22}
+and the latest version of is paper can be obtained at \url{TODO arxiv URL}.

### 2021-10-28 src-latex/parts/10-introduction.tex
Suggestion of structure

@@
 \section{Introduction}
+
+\textbf{TODO sample (simplified) pipeline of the information extraction process}
+\textbf{and/or}
+\textbf{TODO illustration a sample page (full image, large crop, entry crop with highlighted entities)}
+
+% General context + scope limitation
+NER is a fundamental step of the information extraction pipeline when processing historical listing, like directories for instance.
+OCRed text is not sufficient to build a high level, semantic view of the collection under study;
+indeed, being able to properly tag text tokens unlocks the ability to relate entities,
+and provide colleagues from digital humanities with useful databases.
+
+% Problematic
+NER requires training data, and we have little training data for historical documents.
+OCR is also noisy.
+How can we extract reliable elements in this context?
+
+% Contributions
+\begin{enumerate}
+    \item review state of the art regarding NER for historical documents (section 2)
+    \item dataset (section 3)
+    \item list usable techniques (more to evangelize the DAR community) (section 4)
+    \item compare their performance under various training set sizes (thus annotation costs) and OCR noise levels (thus OCR choices/luck/availability) (sec. 5 and 6)
+    \item suggest smart way to leverage side knowledge like lexicons (FIXME iif this works well, maybe a conclusion instead)
+\end{enumerate}

### 2021-10-28 src-latex/parts/30-dataset.tex
Setup paper structure
Suggestion of structure

--- 2021-10-28 src-latex/parts/10-introduction.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Introduction}
+\section{Dataset}
+
+TODO interesting points to detail:
+\begin{itemize}
+    \item origin
+    \item illustration (maybe in first figure)
+    \item various numbers (of original data and of the sample we extracted)
+    \item statistics about entry lengths (histogram of number of char / tokens par entry)
+    \item entities we considered and how we annotated them (how we assessed gold's quality)
+    \item OCR data (automated systems and clean text + how we created the "clean" text + whether/how we assessed clean text's quality)
+    \item \dots
+\end{itemize}

### 2021-10-28 src-latex/parts/40-ner-approaches.tex
Setup paper structure
Suggestion of structure

--- 2021-10-28 src-latex/parts/10-introduction.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Introduction}
+\section{NER Approaches}
+
+TODO describe
+\begin{itemize}
+    \item regexp
+    \item spacy cnn
+    \item spacy cnn + lexicon (warning: how do we build it?)
+    \item transformer
+\end{itemize}
+
+% Other possibilities:
+1. Traditional ML based:
+
+    Conditional Random Fields (CRF) - https://pypi.org/project/sklearn-crfsuite/
+    Maximum-entropy Markov model
+
+2. Neural Networks based:
+
+    LSTMs, bi-LSTM - https://github.com/flairNLP/flair
+    CNNs (SpaCy uses CNN based architecture)
+    Transformers (Spacy has recently launched it) - https://spacy.io/universe/project/spacy-transformers

### 2021-10-28 src-latex/parts/50-xp-setup.tex
Suggestion of structure

@@
 
+\subsection{Metrics}
+\begin{itemize}
+    \item NER quality: accuracy? precision? recall? f-score? (idea: stick with standard spacy/sklearn metrics)
+    \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?)
+\end{itemize}
+
+\subsection{Training procedure}
+train/val/test splits, number of runs, etc.
+
+\subsection{Parameters for each method}
+
 \subsection{Implementation details}
 we use spacy \mcite{spacy}
+will we use FLAIR as well?

### 2021-10-28 src-latex/parts/60-results.tex
Setup paper structure
Suggestion of structure

--- 2021-10-28 src-latex/parts/50-xp-setup.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Experimental Setup}
+\section{Results}
 
-\subsection{Implementation details}
-we use spacy \mcite{spacy}
+\subsection{Experiment 1: Accuracy vs NER approach vs training set size}
+
+Qualitative results
+\textbf{TODO random samples of results + selection of failure cases}
+
+Quantitative results: table + graph ideally (with training set size info)
+
+\begin{verbatim}
+Accuracy
+|                             ..o   
+|                ..o       o/       
+|            ./`          /`          
+|        ./o`      .o   o            
+|     ./`        .`                    One curve per approach
+|    o       o/`                  o  
+|          .`                  ./    
+|       o`                    /      
+|                         .o        
+|                       .`           
+|                      o            
+|                                   
++------------------------------   Training set size
+\end{verbatim}
+
+\subsection{Experiment 2: Accuracy vs NER approach vs OCR noise}
+
+Qualitative results
+\textbf{TODO random samples of results + selection of failure cases}
+
+Quantitative results: table + graph ideally (with OCR noise, same format as previous)
+
+Opt. use synthetic text perturbation as well? Maybe not interesting and too artificial if we have access to 2+ OCR systems.
+(original OCR from BNF, Tesseract 4, Pero OCR…)
+
+\subsection{Discussion}
+Interesting points to discuss:
+\begin{itemize}
+    \item can we train on noisy data? (without manual OCR correction?) => future work? cf Pero OCR training procedure?
+    \item do we need better OCR systems or better post-correction techniques (if NER is reliable enough)?
+    \item Construction of the lexicon and associated cost
+\end{itemize}

### 2021-12-02 material/download.sh
Update json export code (removing extra dict key)

@@ -3,7 +3,7 @@
 set -e
 
 auth_token=$1
-base_url="https://apps.lrde.epita.fr:443/soduco/directory-annotator/directories"
+base_url="https://apps.lrde.epita.fr/soduco/directory-annotator/storage/directories"
 
 download_img() {
     curl -s -H "Authorization: $auth_token" ${view_url}/image > ${file_name}.jpg
@@ -11,7 +11,7 @@ download_img() {
 }
 
 download_json(){
-    curl -s -H "Authorization: $auth_token" ${view_url}/annotation > ${file_name}.json
+    curl -s -H "Authorization: $auth_token" ${view_url}/annotation | jq '.["content"]' > ${file_name}.json
 }
 
 sample() {

### 2022-01-05 src/ner/bert.py
datasets generation + bert and cnn training

@@ -1,18 +1,12 @@
-import nltk
-import csv, json
-from xml.dom.minidom import parseString
-from datasets import Dataset
+import json
+import config as cfg
+import os
+from datasets import load_from_disk
 from transformers import AutoTokenizer
 from transformers import AutoModelForTokenClassification
 from transformers import TrainingArguments, Trainer, DataCollatorForTokenClassification
 from datasets import load_metric
-from sklearn.model_selection import KFold
 import numpy as np
-
-# DATASET
-
-# A text file containing one ner-xml string per line
-data_path = "../data/ner_annotated_full.csv"
 
 label_id_map = {
     "O": 0,
@@ -30,35 +24,18 @@ label2id = label_id_map
 
 id2label = {v: k for k, v in label2id.items()}
 
-
-# MODEL
-base_model = "Jean-Baptiste/camembert-ner"
-
 model = AutoModelForTokenClassification.from_pretrained(
-    base_model,
+    cfg.BERT_BASE_MODEL,
     num_labels=len(label2id),
     ignore_mismatched_sizes=True,
     id2label=id2label,
     label2id=label2id,
 )
 
-tokenizer = AutoTokenizer.from_pretrained(base_model)
+tokenizer = AutoTokenizer.from_pretrained(cfg.BERT_BASE_MODEL)
 
 # TRAINING CONFIGURATION
-
-k_folds = 2
-
-training_args = TrainingArguments(
-    "berties",
-    evaluation_strategy="epoch",
-    learning_rate=1e-4,
-    per_device_train_batch_size=16,
-    per_device_eval_batch_size=16,
-    num_train_epochs=3,
-    weight_decay=1e-5,
-)
-
-metrics_output_path = "../data/ner_bert_metrics.json"
+training_args = TrainingArguments("berties", **cfg.BERT_TRAINER_CONFIG)
 
 
 ###########################
@@ -66,36 +43,24 @@ metrics_output_path = "../data/ner_bert_metrics.json"
 ###########################
 
 
-def train_eval(entries):
+def train_eval(train, test):
     data_collator = DataCollatorForTokenClassification(tokenizer)
 
-    folds = KFold(n_splits=k_folds, shuffle=True, random_state=42).split(entries)
+    trainer = Trainer(
+        model,
+        training_args,
+        train_dataset=train,
+        eval_dataset=test,
+        data_collator=data_collator,
+        tokenizer=tokenizer,
+        compute_metrics=compute_metrics,
+    )
 
-    evaluation_results = []
-    for fold in folds:
[…]

### 2022-01-05 src/ner/cnn.py

@@ -1,0 +1,58 @@
+import spacy
+from spacy.tokens import DocBin, Doc
+from xml.dom.minidom import parseString
+from spacy.training import biluo_tags_to_offsets
+
+nlp = spacy.blank("fr")
+
+
+def create_spacy_dataset(entries):
+    db = DocBin()
+
+    for entry in entries:
+        # Makes the entry a valid XML string, then parses it to a DOM.
+        entry_xml = f"<x>{entry}</x>"
+        x = parseString(entry_xml).getElementsByTagName("x")[0]
+
+        tags = []
+        tokens = []
+        text = ""
+        for el in x.childNodes:
+            el_is_txt = el.nodeName == "#text"
+
+            span = el.nodeValue if el_is_txt else el.childNodes[0].nodeValue
+            text += span
+            span = (
+                span.strip()
+            )  # Remove leading whitespaces or they will be kept as tokens
+
+            inner_doc = nlp.make_doc(span)
+            nertag = el.nodeName.upper()
+
+            # To BILUO
+            inner_tags = []
+            if not el_is_txt:
+                if len(inner_doc) == 1:
+                    inner_tags = [f"U-{nertag}"]
+                elif len(inner_doc) > 1:
+                    inner_tags = (
+                        [f"B-{nertag}"]
+                        + [f"I-{nertag}"] * max(len(inner_doc) - 2, 0)
+                        + [f"L-{nertag}"]
+                    )
+                else:
+                    raise ValueError(f"Empty inner doc in {span}: {entry}")
+                tags.extend(inner_tags)
+            else:
+                tags.extend(["O"] * len(inner_doc))
+
+            tokens.extend([t.text for t in inner_doc])
+
+        words, spaces = spacy.util.get_words_and_spaces(tokens, text)
+        doc = Doc(nlp.vocab, words=words, spaces=spaces)
+
+        offsets = biluo_tags_to_offsets(doc, tags)
+        doc.ents = [doc.char_span(start, end, label=lbl) for start, end, lbl in offsets]
+        db.add(doc)
+
+    return db

### 2022-01-05 src/ner/config.py

@@ -1,0 +1,18 @@
+from pathlib import Path
+
+DATA_DIR = Path("../data")
+INPUT_DATASET = DATA_DIR / "ner_annotated_full.csv"
+
+N_SPLITS = 2
+RANDOM_STATE = 42
+
+BERT_BASE_MODEL = "Jean-Baptiste/camembert-ner"
+
+BERT_TRAINER_CONFIG = {
+    "evaluation_strategy": "epoch",
+    "learning_rate": 1e-4,
+    "per_device_train_batch_size": 16,
+    "per_device_eval_batch_size": 16,
+    "num_train_epochs": 3,
+    "weight_decay": 1e-5,
+}

### 2022-01-05 src/ner/create_train_dev_test.py

@@ -1,0 +1,194 @@
+import spacy
+import csv
+import pickle
+import nltk
+import config as cfg
+from datasets.dataset_dict import DatasetDict
+#from sklearn.model_selection import KFold
+from sklearn.model_selection import train_test_split
+from spacy.tokens import DocBin, Doc
+from xml.dom.minidom import parseString
+from spacy.training import biluo_tags_to_offsets
+from bert import tokenizer, label_id_map
+from datasets import Dataset
+
+
+##########################
+#     ~ HUGGINGFACE ~
+##########################
+
+
+def create_huggingface_dataset(entries):
+    # Creates a Huggingface Dataset from a set of NER-XML entries
+    tokenized_entries = [word_tokens_from_xml(entry) for entry in entries]
+    word_tokens, labels = zip(*tokenized_entries)
+    dataset = Dataset.from_dict({"tokens": word_tokens, "ner_tags": labels})
+    return dataset.map(assign_labels_to_bert_tokens, batched=True)
+
+
+def assign_labels_to_bert_tokens(examples):
+    bert_tokens = tokenizer(
+        examples["tokens"], truncation=True, is_split_into_words=True
+    )
+    prev_word_id = None
+    labels = []
+    for id, label in enumerate(examples["ner_tags"]):
+        labels_ids = []
+        for word_id in bert_tokens.word_ids(batch_index=id):
+            # Set the label only on the first token of each word
+            if word_id in [None, prev_word_id]:
+                labels_ids.append(-100)
+            else:
+                label_id = label_id_map[label[word_id]]
+                labels_ids.append(label_id)
+            prev_word_id = word_id
+
+        labels.append(labels_ids)
+
+    bert_tokens["labels"] = labels
+    return bert_tokens
+
+
+# convenient word tokenizer to create IOB-like data for the BERT models
+nltk.download("punkt")
+
+
+def word_tokens_from_xml(entry):
+    w_tokens = []
+    labels = []
+
+    entry_xml = f"<x>{entry}</x>"
+    x = parseString(entry_xml).getElementsByTagName("x")[0]
+
+    for el in x.childNodes:
+        if el.nodeName == "#text":
+            cat = "O"
+            txt = el.nodeValue
+        else:
+            cat = f"I-{el.nodeName}"
+            txt = el.childNodes[0].nodeValue
+
[…]

### 2022-01-05 src-latex/parts/50-xp-setup.tex
Add list of potential OCR systems

@@
 \begin{itemize}
-    \item NER quality: accuracy? precision? recall? f-score? (idea: stick with standard spacy/sklearn metrics)
-    \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?)
+    \item NER quality: accuracy?
+precision?
+recall?
+f-score?
+(idea: stick with standard spacy/sklearn metrics) \textcolor{blue}{Dans la plupart des articles que j'ai lus, c'est toujours une F-mesure qui est utilisée et parfois la précision et le rappel sont ajoutés.}
+    \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?) \textcolor{blue}{Pour l'instant, j'ai croisé: CER, WER, Levenshtein et Jaccard.
+J'ai encore un état de l'art à lire sur le sujet: il y aura probablement d'autres mesures mentionnées.}
+
+    \textcolor{blue}{Pour la partie Experiment 2: dans la litérature j'ai vu des courbes F-Mesure(CER) ou F-Mesure(WER) et des diagrammes avec F-Mesure calculée sur des sous-ensembles du corpus classés par score de Levenshtein.}
 \end{itemize}
@@
 
+\subsection{OCR systems}
+\begin{itemize}
+    \item original OCR
+    \item Tesseract 4
+    \item Pero
+    \item other?
+\end{itemize}
+
 \subsection{Implementation details}

### 2022-01-06 src-latex/main-paper.tex
Suggestion of structure
Update main-paper.tex
Proposition d'intro: à corriger/compléter/amender...

@@
 \renewcommand\UrlFont{\color{blue}\rmfamily}
+% FIXME proper hyperref setup
 
@@
 %
-Nathalie Abadie\inst{1}\orcidID{0000-1111-2222-3333} \and
+Nathalie Abadie\inst{1}\orcidID{0000-0001-8741-2398} \and
 Edwin Carlinet\inst{2,3}\orcidID{1111-2222-3333-4444} \and
@@
 
+\section*{Thanks}
+Laureencia for annotations, Guillaume Thomas for data processing?
+
 % ---- Bibliography ----

### 2022-01-06 src-latex/parts/10-introduction.tex
Proposition d'intro: à corriger/compléter/amender...

@@
 % General context + scope limitation
-NER is a fundamental step of the information extraction pipeline when processing historical listing, like directories for instance.
-OCRed text is not sufficient to build a high level, semantic view of the collection under study;
-indeed, being able to properly tag text tokens unlocks the ability to relate entities,
-and provide colleagues from digital humanities with useful databases.
 
-% Problematic
-NER requires training data, and we have little training data for historical documents.
-OCR is also noisy.
-How can we extract reliable elements in this context?
+Heritage institutions such as archives or libraries preserve large collections of textual documents, increasingly digitized or natively digital.
+They disseminate these digital documents on the Web to enable users to access them, thus making these knowledge sources available for numerous applications in the fields of social sciences, environmental studies, education, etc.
+Fostering information search and reuse within these collections requires two steps.
+The first step, called Optical Character Recognition (OCR), aims at transforming the images produced during the digitization of the documents into usable digital texts.
+However, OCRed text is not sufficient to build a high level, semantic view of the collection under study.
+A second step thus aims at recognizing in these texts the pieces of information most likely to be searched for by users, such as named entities (people, organisations, dates, places, etc.).
+Indeed, being able to properly tag text tokens unlocks the ability to relate entities, and provide colleagues from digital humanities with useful databases.
 
+In the case of historical textual documents, OCR and Named Entity Recognition (NER) tasks are particularly challenging.
+Indeed, even in the case of printed historical documents, the OCR approaches used for modern documents fail due to many reasons: poor scan quality, irregular printing baseline, printing artifacts, old fonts, complex page layouts, special glyphs, etc.
+\textcolor{green}{Phrases précedentes à reprendre avec les termes adéquats + ajouter une ref.} On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts and from the errors introduced into the text by the OCR step.\textcolor{green}{ajouter une ref}
+
+In this article we focus on a particular type of historical textual sources, as they provide very rich, semi-structured and time-stamped information about people, their activities and their locations: the directories.
+In a recent work to locate potentially polluted urban soils \cite{bell2020automated}, historical directories are leveraged to identify and locate all the gas stations that were established in the city of Providence during the 20th century.
+The proposed information extraction pipeline includes OCR, entries identification and business type and address extraction.
+In the same way, we aim at producing structured spatio-temporal data from the information contained in the 19th century Parisian trade directories.
+These directories are produced by various publishers, organised according to different indexing methods (by name, by activity, by location, etc.), printed in different layouts, with different fonts, etc.
+We therefore investigate various state-of-the-art OCR and NER approaches independently and in combination, to assess their usability on our historical directories corpus.
+Thus, the contributions of this article are the following:
 % Contributions

### 2022-01-06 src-latex/parts/30-dataset.tex
Proposition d'intro: à corriger/compléter/amender...

@@
     \item illustration (maybe in first figure)
-    \item various numbers (of original data and of the sample we extracted)
+    \item various numbers (of original data and of the sample we extracted) : number of entries per pages, variety of directory entries, \dots
     \item statistics about entry lengths (histogram of number of char / tokens par entry)
     \item entities we considered and how we annotated them (how we assessed gold's quality)
+    \item statistics on the distribution of entities across directories and their arrangement within entries
     \item OCR data (automated systems and clean text + how we created the "clean" text + whether/how we assessed clean text's quality)
+
     \item \dots

### 2022-01-06 src-latex/parts/40-ner-approaches.tex
Proposition d'intro: à corriger/compléter/amender...

--- 2021-10-28 src-latex/parts/10-introduction.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Introduction}
+\section{NER Approaches}
+
+\begin{itemize}
+    \item regexp
+    \item Spacy CNN with tok2vec
+    \item CamemBERT finetuned with wikiner-fr + gold dataset (clean texts)
+    \item CamemBERT finetuned with wikiner-fr + gold dataset (clean texts + pretrained on raw texts (Pero ? Kraken ? Both ?)
+\end{itemize}
+
+% Other possibilities:
+1. Traditional ML based:
+    Conditional Random Fields (CRF) - https://pypi.org/project/sklearn-crfsuite/
+    Maximum-entropy Markov model
+
+2. Neural Networks based:
+    LSTMs, bi-LSTM - https://github.com/flairNLP/flair
+    CNNs (SpaCy uses CNN based architecture)
+    Transformers (Spacy has recently launched it) - https://spacy.io/universe/project/spacy-transformers

### 2022-01-06 src-latex/parts/50-xp-setup.tex
Proposition d'intro: à corriger/compléter/amender...

@@
 \begin{itemize}
-    \item NER quality: accuracy?
-precision?
-recall?
-f-score?
-(idea: stick with standard spacy/sklearn metrics) \textcolor{blue}{Dans la plupart des articles que j'ai lus, c'est toujours une F-mesure qui est utilisée et parfois la précision et le rappel sont ajoutés.}
+    \item NER quality: Precision, Recall, F1 score (idea: stick with standard spacy/sklearn metrics) \textcolor{blue}{Dans la plupart des articles que j'ai lus, c'est toujours une F-mesure qui est utilisée et parfois la précision et le rappel sont ajoutés.}
     \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?) \textcolor{blue}{Pour l'instant, j'ai croisé: CER, WER, Levenshtein et Jaccard.
@@
 \subsection{Training procedure}
-train/val/test splits, number of runs, etc.
+5-fold CV for all models.
+\begin{itemize}
+\item Spacy CNN : no limits on the number of epochs, instead use patience=1600.
+\item CamemBERT : 3 epochs.
+\end{itemize}
 

### 2022-01-07 src-latex/parts/60-results.tex
Proposition d'intro: à corriger/compléter/amender...
Début d'état de l'art: à finir aujourd'hui.

@@
 
-\subsection{Experiment 2: Accuracy vs NER approach vs OCR noise}
+\subsection{Experiment 2: sensibility to training on noisy data}
+Idea : Measure the impact of training with noisy data when predicting NER tags on both clean and noisy data.
+
+What's interesting : do models perform better on noisy data when trained on noisy data ? Does it impact predictions on clean data ?
+
+Table : NER NN models VS train on {noisy, clean} datasets VS evaluate on {noisy, clean} datasets
+
+\subsection{Experiment 3: Accuracy vs NER approach vs OCR noise}
 

### 2022-01-07 src-latex/parts/50-xp-setup.tex

@@
 \subsection{Training procedure}
-5-fold CV for all models.
 \begin{itemize}
-\item Spacy CNN : no limits on the number of epochs, instead use patience=1600.
+\item Spacy CNN : do not put a limit on epochs, instead use patience=1600.
 \item CamemBERT : 3 epochs.
 \end{itemize}
+
+Split the full dataset into 80 percent train, 10 percent dev / eval and 10 percent test.
+Use stratified sampling based on the directory name. This way each set contains samples of all directories.
 

### 2022-01-07 src/create_gold.sh
cleaner NER pipeline

@@ -1,0 +1,36 @@
+#!/bin/bash
+
+# Let's get paranoid
+set -eu
+
+######################
+
+WORKDIR=data/gold
+
+mkdir -p $WORKDIR
+
+pushd $WORKDIR > /dev/null
+
+echo "Creating the gold dataset"
+
+# Gather sampled pages numbers & directory name
+declare -A directories
+directories=()
+while IFS= read -r row;
+do
+  arr=($row)
+  directories+=([${arr[0]}]=${arr[1]})
+done < sampled_pages.txt
+
+# Creates the gold dataset from a set of directories jsons and stores it in CSV where:
+# column 1 is a pre-annotated directory entry in XML
+# column 2 is the entry's directory name
+rm -f gold.csv
+for f in $(ls *.json);
+do
+  pagenum=$(echo ${f%.*} | sed 's/^0*//')
+  directory="${directories[$pagenum]}"
+  cat $f | jq --raw-output --arg directory "$directory" '.[] | select(.type=="ENTRY" and .checked==true) | [.ner_xml,$directory] | @csv' >> gold.csv
+done
+
+popd  > /dev/null

### 2022-01-07 src/ner/bert.py

@@ -1,13 +1,16 @@
 import json
 import config as cfg
 import os
-from datasets import load_from_disk
+import numpy as np
+import nltk
+from datasets import Dataset, load_from_disk
 from transformers import AutoTokenizer
 from transformers import AutoModelForTokenClassification
 from transformers import TrainingArguments, Trainer, DataCollatorForTokenClassification
 from datasets import load_metric
-import numpy as np
+from xml.dom.minidom import parseString
 
+# Shared variables
 label_id_map = {
     "O": 0,
     "I-LOC": 1,
@@ -19,45 +22,23 @@ label_id_map = {
     "I-TITRE": 7,
     "I-FT": 8,
 }
-
-label2id = label_id_map
-
-id2label = {v: k for k, v in label2id.items()}
-
-model = AutoModelForTokenClassification.from_pretrained(
-    cfg.BERT_BASE_MODEL,
-    num_labels=len(label2id),
-    ignore_mismatched_sizes=True,
-    id2label=id2label,
-    label2id=label2id,
-)
-
 tokenizer = AutoTokenizer.from_pretrained(cfg.BERT_BASE_MODEL)
-
-# TRAINING CONFIGURATION
 training_args = TrainingArguments("berties", **cfg.BERT_TRAINER_CONFIG)
 
-
-###########################
-# ~ TRAIN & EVAL LOOP ~
-###########################
-
-
-def train_eval(train, test):
+# Training & eval
+def train_eval(train, valid):
     data_collator = DataCollatorForTokenClassification(tokenizer)
 
     trainer = Trainer(
         model,
         training_args,
         train_dataset=train,
-        eval_dataset=test,
+        eval_dataset=valid,
         data_collator=data_collator,
         tokenizer=tokenizer,
         compute_metrics=compute_metrics,
     )
 
-    # TODO : allow to evaluate already trained models instead of training from scratch
-    # at every run.
     trainer.train()
     eval = trainer.evaluate()
     return eval
@@ -87,10 +68,23 @@ def compute_metrics(p):
 
 
 ##########################
-#     ~ ENTRY POINT ~
+#  ~ MAIN ENTRY POINT ~
 ##########################
 
 if __name__ == "__main__":
+
+    label2id = label_id_map
+    id2label = {v: k for k, v in label2id.items()}
+
+    model = AutoModelForTokenClassification.from_pretrained(
+        cfg.BERT_BASE_MODEL,
+        num_labels=len(label2id),
+        ignore_mismatched_sizes=True,
[…]

### 2022-01-07 src/ner/cnn.py

@@ -1,7 +1,12 @@
+import nltk
 import spacy
 from spacy.tokens import DocBin, Doc
 from xml.dom.minidom import parseString
-from spacy.training import biluo_tags_to_offsets
+from spacy.training import tags_to_entities
+
+###########################
+# ~ DATA CONVERSION UTILS ~
+###########################
 
 nlp = spacy.blank("fr")
 
@@ -11,48 +16,47 @@ def create_spacy_dataset(entries):
 
     for entry in entries:
         # Makes the entry a valid XML string, then parses it to a DOM.
-        entry_xml = f"<x>{entry}</x>"
-        x = parseString(entry_xml).getElementsByTagName("x")[0]
+        xml = parseString(f"<xml>{entry}</xml>").getElementsByTagName("xml")[0]
 
         tags = []
-        tokens = []
+        words = []
         text = ""
-        for el in x.childNodes:
+        for el in xml.childNodes:
             el_is_txt = el.nodeName == "#text"
-
             span = el.nodeValue if el_is_txt else el.childNodes[0].nodeValue
             text += span
-            span = (
-                span.strip()
-            )  # Remove leading whitespaces or they will be kept as tokens
 
-            inner_doc = nlp.make_doc(span)
+            if span.isspace():
+                continue
+
+            span_toks = nltk.word_tokenize(span, language="fr", preserve_line=True)
+
+            # BILUO as pivot format
             nertag = el.nodeName.upper()
-
-            # To BILUO
-            inner_tags = []
+            span_biluo = []
             if not el_is_txt:
-                if len(inner_doc) == 1:
-                    inner_tags = [f"U-{nertag}"]
-                elif len(inner_doc) > 1:
-                    inner_tags = (
+                if len(span_toks) == 1:
+                    span_biluo = [f"U-{nertag}"]
+                elif len(span_toks) > 1:
+                    span_biluo = (
                         [f"B-{nertag}"]
-                        + [f"I-{nertag}"] * max(len(inner_doc) - 2, 0)
+                        + [f"I-{nertag}"] * max(len(span_toks) - 2, 0)
                         + [f"L-{nertag}"]
                     )
                 else:
                     raise ValueError(f"Empty inner doc in {span}: {entry}")
-                tags.extend(inner_tags)
+                tags.extend(span_biluo)
             else:
-                tags.extend(["O"] * len(inner_doc))
[…]

### 2022-01-07 src/ner/config.py

@@ -1,10 +1,14 @@
 from pathlib import Path
+import logging
 
-DATA_DIR = Path("../data")
-INPUT_DATASET = DATA_DIR / "ner_annotated_full.csv"
+# Assuming datadir is located in the parent directory of this module
+DATA_DIR = Path(__file__).parents[1] / "data"
+
+GOLD = DATA_DIR / "gold/gold.csv"
 
 N_SPLITS = 2
-RANDOM_STATE = 42
+SEED = 42  # Used to generate train/validation/test datasets
+MIN_TRAINSET_SIZE = 50
 
 BERT_BASE_MODEL = "Jean-Baptiste/camembert-ner"
 
@@ -16,3 +20,7 @@ BERT_TRAINER_CONFIG = {
     "num_train_epochs": 3,
     "weight_decay": 1e-5,
 }
+
+# Set logging configuration for the whole project
+logging.basicConfig()
+logging.root.setLevel(logging.INFO)

### 2022-01-07 src/ner/create_datasets.py

@@ -1,0 +1,119 @@
+import csv
+import logging
+import os
+import config as cfg
+from sklearn.model_selection import train_test_split
+from datasets.dataset_dict import DatasetDict
+from cnn import create_spacy_dataset
+from bert import create_huggingface_dataset
+
+# Helper functions
+def unwrap(list_of_tuples2):
+    return tuple(zip(*list_of_tuples2))
+
+
+# Loads the gold dataset
+with open(cfg.GOLD, encoding="utf-8") as annot:
+    gold_dataset = [(_[0], _[1]) for _ in csv.reader(annot)]
+
+logging.info(f"Gold dataset: {len(gold_dataset)} records")
+
+# Generate Train (80%), Validation|Dev (10%) and Test (10%) datasets
+# with stratified sampling on the directories names.
+
+# Get train and dev+test
+_, classes = unwrap(gold_dataset)
+train_devtest = train_test_split(
+    gold_dataset, train_size=0.8, shuffle=True, random_state=cfg.SEED, stratify=classes
+)
+train = train_devtest[0]
+
+# Split dev+test in half to get dev and test
+devtest = train_devtest[1]
+_, classes = unwrap(devtest)
+dev_test = train_test_split(
+    devtest, train_size=0.5, shuffle=True, random_state=cfg.SEED, stratify=classes
+)
+dev = dev_test[0]
+test = dev_test[1]
+
+logging.info(
+    f"Train ({len(train)} records), Dev ({len(dev)} records), Test ({len(test)} records)"
+)
+
+###
+# Generate smaller train sets for experiment 1
+###
+
+exp_1_trainsets = [train]
+k = len(train)
+while k > cfg.MIN_TRAINSET_SIZE:
+    try:
+        current = exp_1_trainsets[-1]
+        _, classes = unwrap(current)
+        smaller, rest = train_test_split(
+            current,
+            train_size=0.5,
+            shuffle=True,
+            random_state=cfg.SEED,
+            stratify=classes,
+        )
+        exp_1_trainsets.append(smaller)
+        current_train = rest
+        k = len(rest)
+    except ValueError as e:
+        # Stop now if we encounter the error "The least populated class in y has only 1 member".
+        break
+
+logging.info(
+    f"Experiment 1: created {len(exp_1_trainsets)} training sets of sizes {[len(_) for _ in exp_1_trainsets]}"
+)
+
+
+###
[…]

### 2022-01-07 src/ner/create_train_dev_test.py

@@ -1,194 +1,0 @@
-import spacy
-import csv
-import pickle
-import nltk
-import config as cfg
-from datasets.dataset_dict import DatasetDict
-#from sklearn.model_selection import KFold
-from sklearn.model_selection import train_test_split
-from spacy.tokens import DocBin, Doc
-from xml.dom.minidom import parseString
-from spacy.training import biluo_tags_to_offsets
-from bert import tokenizer, label_id_map
-from datasets import Dataset
-
-
-##########################
-#     ~ HUGGINGFACE ~
-##########################
-
-
-def create_huggingface_dataset(entries):
-    # Creates a Huggingface Dataset from a set of NER-XML entries
-    tokenized_entries = [word_tokens_from_xml(entry) for entry in entries]
-    word_tokens, labels = zip(*tokenized_entries)
-    dataset = Dataset.from_dict({"tokens": word_tokens, "ner_tags": labels})
-    return dataset.map(assign_labels_to_bert_tokens, batched=True)
-
-
-def assign_labels_to_bert_tokens(examples):
-    bert_tokens = tokenizer(
-        examples["tokens"], truncation=True, is_split_into_words=True
-    )
-    prev_word_id = None
-    labels = []
-    for id, label in enumerate(examples["ner_tags"]):
-        labels_ids = []
-        for word_id in bert_tokens.word_ids(batch_index=id):
-            # Set the label only on the first token of each word
-            if word_id in [None, prev_word_id]:
-                labels_ids.append(-100)
-            else:
-                label_id = label_id_map[label[word_id]]
-                labels_ids.append(label_id)
-            prev_word_id = word_id
-
-        labels.append(labels_ids)
-
-    bert_tokens["labels"] = labels
-    return bert_tokens
-
-
-# convenient word tokenizer to create IOB-like data for the BERT models
-nltk.download("punkt")
-
-
-def word_tokens_from_xml(entry):
-    w_tokens = []
-    labels = []
-
-    entry_xml = f"<x>{entry}</x>"
-    x = parseString(entry_xml).getElementsByTagName("x")[0]
-
-    for el in x.childNodes:
-        if el.nodeName == "#text":
-            cat = "O"
-            txt = el.nodeValue
-        else:
-            cat = f"I-{el.nodeName}"
-            txt = el.childNodes[0].nodeValue
-
[…]

### 2022-01-07 src/ner/init.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+
+# Let's get paranoid
+set -eu
+
+WORKDIR=$(dirname "$0")
+
+# Build datasets for all experiments
+echo "Initializing training, dev and test datasets"
+python ${WORKDIR}/create_datasets.py

### 2022-01-07 src/ocr/isri_tools_src/Modules/accrpt.c
Add isri binding tools

@@ -1,0 +1,287 @@
+/**********************************************************************
+ *
+ *  accrpt.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include "accrpt.h"
+#include "sort.h"
+
+#define TITLE    "UNLV-ISRI OCR Accuracy Report Version 5.1\n"
+#define DIVIDER  "-----------------------------------------\n"
+
+#define CLASS_OFFSET  29
+#define CONF_OFFSET   20
+
+static char line[100];
+
+/**********************************************************************/
+
+static void update_class(class, count, missed)
+Accclass *class;
+long count, missed;
+{
+    class->count  += count;
+    class->missed += missed;
+}
+/**********************************************************************/
+
+void add_class(accdata, value, count, missed)
+Accdata *accdata;
+Charvalue value;
+long count, missed;
+{
+    update_class(&accdata->large_class[charclass(value)], count, missed);
+    update_class(&accdata->total_class, count, missed);
+    update_class(&accdata->small_class[value], count, missed);
+}
+/**********************************************************************/
+
+void add_conf(accdata, key, errors, marked)
+Accdata *accdata;
+char *key;
+long errors, marked;
+{
+    Conf *conf;
+    conf = table_lookup(&accdata->conftable, key);
+    if (conf)
+    {
+	conf->errors += errors;
+	conf->marked += marked;
+    }
+    else
+    {
[…]

### 2022-01-07 src/ocr/isri_tools_src/Modules/edorpt.c

@@ -1,0 +1,125 @@
+/**********************************************************************
+ *
+ *  edorpt.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include "edorpt.h"
+#include "util.h"
+
+#define TITLE    "UNLV-ISRI Edit Operation Report Version 5.1\n"
+#define DIVIDER  "-------------------------------------------\n"
+
+static char line[100];
+
+/**********************************************************************/
+
+static Boolean read_line(f)
+FILE *f;
+{
+    return(fgets(line, sizeof(line) - 1, f) ? True : False);
+}
+/**********************************************************************/
+
+static Boolean read_value(f, value, sum_value)
+FILE *f;
+long *value, *sum_value;
+{
+    if (read_line(f) && sscanf(line, "%ld", value) == 1)
+    {
+	*sum_value += *value;
+	return(True);
+    }
+    else
+	return(False);
+}
+/**********************************************************************/
+
+static Boolean read_two(f, value1, value2)
+FILE *f;
+long *value1, *value2;
+{
+    return(read_line(f) && sscanf(line, "%ld %ld", value1, value2) == 2 ?
+    True : False);
+}
+/**********************************************************************/
+
+void read_edorpt(edodata, filename)
+Edodata *edodata;
+char *filename;
+{
+    FILE *f;
+    long moves, value1, value2;
+    f = open_file(filename, "r");
+    if (read_line(f) && strncmp(line, TITLE, sizeof(TITLE) - 3) == 0 &&
+    read_line(f) && strcmp(line, DIVIDER) == 0 &&
[…]

### 2022-01-07 src/ocr/isri_tools_src/Modules/list.c

@@ -1,0 +1,148 @@
+/**********************************************************************
+ *
+ *  list.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include "list.h"
+#include "util.h"
+
+BEGIN_ITEM(Item)
+END_ITEM(Item);
+
+BEGIN_LIST_OF(Item)
+END_LIST(List);
+
+/**********************************************************************/
+
+void list_initialize(void *list_)
+{
+  List* list = list_;
+  list->first = list->last = NULL;
+  list->array = NULL;
+  list->count = 0;
+}
+/**********************************************************************/
+
+static void free_array(list)
+List *list;
+{
+    if (list->array)
+    {
+	free(list->array);
+	list->array = NULL;
+    }
+}
+/**********************************************************************/
+
+static void list_insert(list, ref1, item, ref2)
+List *list;
+Item *ref1, *item, *ref2;
+{
+    item->prev = ref1;
+    item->next = ref2;
+    if (ref1)
+	ref1->next = item;
+    else
+	list->first = item;
+    if (ref2)
+	ref2->prev = item;
+    else
+	list->last = item;
+    free_array(list);
+    list->count++;
+}
+/**********************************************************************/
+
+void list_insert_first(list, item)
+List *list;
+Item *item;
+{
+    list_insert(list, NULL, item, list->first);
+}
+/**********************************************************************/
+
+void list_insert_last(list, item)
+List *list;
+Item *item;
+{
+    list_insert(list, list->last, item, NULL);
+}
+/**********************************************************************/
+
+void list_insert_before(list, item, ref)
+List *list;
+Item *item, *ref;
+{
[…]

### 2022-01-07 src/ocr/isri_tools_src/Modules/list.h

@@ -1,0 +1,102 @@
+/**********************************************************************
+ *
+ *  list.h
+ *
+ *  This module provides a general-purpose linked-list capability.  A
+ *  "list" contains zero or more "items".
+ *
+ *  An item is a record structure having the following initial fields:
+ *
+ *	Item *prev;	- pointer to the previous item in the list
+ *	Item *next;	- pointer to the next item in the list
+ *
+ *  This structure is declared by
+ *
+ *  BEGIN_ITEM(Item)
+ * 	<declare other fields here>
+ *  END_ITEM(Item);
+ *
+ *  A list is a record structure having the following initial fields:
+ *
+ *	Item *first;	- pointer to the first item in the list
+ *	Item *last;	- pointer to the last item in the list
+ *	Item *array[];  - array of pointers to the items in the list
+ *	long count;	- number of items in the list
+ *
+ *  This structure is declared by
+ *
+ *  BEGIN_LIST_OF(Item)
+ *	<declare other fields here, if any>
+ *  END_LIST(List);
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#ifndef _LIST_
+#define _LIST_
+
+#define BEGIN_ITEM(Item)		\
+typedef					\
+struct Item				\
+{					\
+    struct Item *prev, *next;
+
+#define END_ITEM(Item)			\
+} Item
+
+#define BEGIN_LIST_OF(Item)		\
+typedef					\
+struct					\
+{					\
+    Item *first, *last, **array;	\
+    long count;
+
+#define END_LIST(List)			\
+} List
+
+void list_initialize(void *list);
+			/* initializes the list; this routine does not need to
+			   be called if the list structure was initialized
+			   statically or dynamically */
+
[…]

### 2022-01-07 src/ocr/isri_tools_src/Modules/text.c

@@ -1,0 +1,238 @@
+/**********************************************************************
+ *
+ *  text.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include "text.h"
+
+#define TITLE    "UNLV-ISRI OCR Vendor-Independent Interface Version 5.1\n"
+#define DIVIDER  "------------------------------------------------------\n"
+
+/**********************************************************************/
+
+void append_char(
+  Text *text,
+  Boolean suspect,
+  Charvalue value)
+{
+    Char *c;
+    c = NEW(Char);
+    c->suspect = suspect;
+    c->value = value;
+    list_insert_last(text, c);
+}
+/**********************************************************************/
+
+static Boolean read_header(f)
+FILE *f;
+{
+    char line[100];
+    short divider_count = 0;
+    if (fgets(line, sizeof(line) - 1, f) &&
+    strncmp(line, TITLE, sizeof(TITLE) - 5) == 0)
+    {
+	while (divider_count < 2 && fgets(line, sizeof(line) - 1, f))
+	    if (strcmp(line, DIVIDER) == 0)
+		divider_count++;
+	return(True);
+    }
+    rewind(f);
+    return(False);
+}
+/**********************************************************************/
+
+static void read_contents(f, text, find_markers, suspect_marker)
+FILE *f;
+Text *text;
+Boolean find_markers;
+int suspect_marker;
+{
+    Boolean suspect = False;
+    int byte;
+    byte = getc(f);
+    while (byte != EOF)
+    {
[…]

### 2022-01-07 src/ocr/isri_tools_src/Modules/text.h

@@ -1,0 +1,108 @@
+/**********************************************************************
+ *
+ *  text.h
+ *
+ *  This module provides definitions and routines to support the reading
+ *  and writing of OCR-generated text.  A "Text" structure is defined to
+ *  be a linked list of "Char" structures, where each "Char" structure
+ *  gives the value of one 16-bit character and indicates whether it
+ *  is suspect.
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#ifndef _TEXT_
+#define _TEXT_
+
+#include "list.h"
+#include "util.h"
+
+#define BLANK             ' '
+#define NEWLINE           '\n'
+#define REJECT_CHARACTER  '~'
+#define SUSPECT_MARKER    '^'
+
+typedef unsigned short Charvalue;
+#define NUM_CHARVALUES  65536
+
+BEGIN_ITEM(Char)
+    Boolean suspect;
+    Charvalue value;
+END_ITEM(Char);
+
+BEGIN_LIST_OF(Char)
+END_LIST(Text);
+
+void append_char(Text *text, Boolean suspect, Charvalue value);
+			/* appends the given character to "text" */
+
+typedef
+struct
+{
+    Boolean find_header;/* if True, a header will be looked for and skipped if
+			   present; if False and a header is present, it will be
+			   stored as text */
+    Boolean find_markers;
+			/* if True, any occurrence of the "suspect_marker"
+			   character will be interpreted as marking the
+			   following character as suspect */
+    int suspect_marker; /* applicable when "find_markers" is True; if zero,
+			   SUSPECT_MARKER will be used */
+    Boolean find_hex_values;
+			/* if True, any occurrence of a hex value of the form
+			   <FF> or <FFFF> will be converted to the corresponding
[…]

### 2022-01-07 src/ocr/isri_tools_src/Modules/unicode.c

@@ -1,0 +1,113 @@
+/**********************************************************************
+ *
+ *  unicode.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include "unicode.h"
+
+/**********************************************************************/
+
+static Boolean read_contents(f, text, find_markers, suspect_marker)
+FILE *f;
+Text *text;
+Boolean find_markers;
+int suspect_marker;
+{
+    Boolean MSB_first = True, suspect = False;
+    int byte1, byte2;
+    Charvalue value;
+    byte1 = getc(f);
+    byte2 = getc(f);
+    if ((byte1 == 0xFE && byte2 == 0xFF) || (byte1 == 0xFF && byte2 == 0xFE))
+    {
+	MSB_first = (byte1 == 0xFE ? True : False);
+	byte1 = getc(f);
+	byte2 = getc(f);
+    }
+    while (byte1 != EOF && byte2 != EOF)
+    {
+	value = (MSB_first ? (byte1 << 8) + byte2 : (byte2 << 8) + byte1);
+	if (find_markers && value == suspect_marker)
+	    suspect = True;
+	else
+	{
+	    append_char(text, suspect, value);
+	    suspect = False;
+	}
+	byte1 = getc(f);
+	byte2 = getc(f);
+    }
+    return(byte1 == EOF ? True : False);
+}
+/**********************************************************************/
+
+static void normalize(text, start)
+Text *text;
+Char *start;
+{
+}
+/**********************************************************************/
+
+void read_unicode(text, filename, unicodeopt)
+Text *text;
[…]

### 2022-01-07 src/ocr/isri_tools_src/Modules/wacrpt.c

@@ -1,0 +1,265 @@
+/**********************************************************************
+ *
+ *  wacrpt.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include <stdlib.h>
+#include "sort.h"
+#include "wacrpt.h"
+
+#define TITLE    "UNLV-ISRI OCR Word Accuracy Report Version 5.1\n"
+#define DIVIDER  "----------------------------------------------\n"
+
+#define TOTAL    " Total\n"
+#define OFFSET   29
+
+static char line[100];
+
+/**********************************************************************/
+
+void increment_wac(wac, count, missed)
+Wac *wac;
+long count, missed;
+{
+    wac->count  += count;
+    wac->missed += missed;
+}
+/**********************************************************************/
+
+void add_term(termtable, key, count, missed)
+Termtable *termtable;
+char *key;
+long count, missed;
+{
+    Term *term;
+    term = table_lookup(termtable, key);
+    if (!term)
+    {
+	term = NEW(Term);
+	term->key = strdup(key);
+	table_insert(termtable, term);
+    }
+    increment_wac(&term->wac, count, missed);
+}
+/**********************************************************************/
+
+static Boolean read_line(f)
+FILE *f;
+{
+    return(fgets(line, sizeof(line) - 1, f) ? True : False);
+}
+/**********************************************************************/
+
[…]

### 2022-01-07 src/ocr/isri_tools_src/Modules/word.c

@@ -1,0 +1,87 @@
+/**********************************************************************
+ *
+ *  word.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include "word.h"
+
+/**********************************************************************/
+
+static Boolean is_wordchar(value)
+Charvalue value;
+{
+    char puncchars[] = " ,.:;!'\"()<>\n\t";
+    char *c;
+    if(!value) {
+        return False;
+    }
+    for(c = puncchars; *c; c++) {
+        if(value == *c) {
+            return False;
+        }
+    }
+    return True;
+}
+/**********************************************************************/
+
+void find_words(wordlist, text)
+Wordlist *wordlist;
+Text *text;
+{
+    char string[MAX_WORDLENGTH + 1];
+    char s[7];
+    long len = 0, i, l;
+    Charvalue value;
+    Word *word;
+    list_in_array(text);
+    string[0] = 0;
+    for (i = 0; i < text->count || len > 0; i++)
+    {
+        value = (i < text->count ? text->array[i]->value : 0);
+        if (is_wordchar(value))
+        {
+            char_to_string(0, value, s, 0);
+            l = strlen(s);
+            if (len + l < MAX_WORDLENGTH) {
+                strncat(string, s, l);
+                len += l;
+            }
+        }
+        else if (len > 0)
+        {
+            word = NEW(Word);
+            string[len] = '\0';
+            word->string = (unsigned char *) strdup(string);
+            list_insert_last(wordlist, word);
+            len = 0;
+            string[0] = 0;
+        }
+    }
+}
+/**********************************************************************/
+
+void free_word(word)
+Word *word;
+{
+    free(word->string);
+    free(word);
+}

### 2022-01-07 src/ocr/isri_tools_src/align.cpp

@@ -1,0 +1,111 @@
+extern "C"
+{
+  #include "Modules/text.h"
+  #include "Modules/sync.h"
+}
+
+#include <string>
+#include <pybind11/pybind11.h>
+#include <pybind11/stl.h>
+
+namespace
+{
+
+  void text_from_string(Text* text, std::wstring a)
+  {
+    for (auto c: a)
+      append_char(text, /* suspect = */false, c);
+  }
+
+  std::wstring text_to_string(Text* text)
+  {
+    std::wstring w;
+    w.reserve(text->count);
+    for (auto* item = text->first; item; item = item->next)
+      w.push_back(item->value);
+    return w;
+  }
+
+  int compute_mapping(Synclist* synclist, int* A, int* B)
+  {
+
+    int i = 0;
+    for (Sync* sync = synclist->first; sync; sync = sync->next)
+    {
+      int ma = sync->substr[0].length;
+      int mb = sync->substr[1].length;
+      int m = std::max(ma, mb);
+      for (int k = 0; k < ma; ++k)
+        A[sync->substr[0].start + k] = i + k;
+      for (int k = 0; k < mb; ++k)
+        B[sync->substr[1].start + k] = i + k;
+      i += m;
+    }
+    return i;
+  }
+
+}
+
+
+int compute_align_map(std::wstring a, std::wstring b, std::vector<int>& map_A, std::vector<int>& map_B)
+{
+  Text A;
+  Text B;
+  list_initialize(&A);
+  list_initialize(&B);
+  text_from_string(&A, a);
+  text_from_string(&B, b);
+
+  map_A.resize(a.size());
+  map_B.resize(b.size());
+
+  int alignment_size;
+  {
+    Text texts[2] = {A, B};
+
+    Synclist synclist;
+    fastukk_sync(&synclist, texts);
+    alignment_size = compute_mapping(&synclist, map_A.data(), map_B.data());
+  }
+
+  list_empty(&A, free);
+  list_empty(&B, free);
+
+  return alignment_size;
+}
+
+std::pair<std::vector<int>, std::vector<int>> align_2(std::wstring a, std::wstring b)
+{
+  std::vector<int> map_A, map_B;
+  int alignment_size = compute_align_map(a, b, map_A, map_B);
+
+  return std::make_pair(map_A, map_B);
+}
+
+
[…]

### 2022-01-07 src/ocr/setup.py

@@ -1,0 +1,27 @@
+from setuptools import Extension, setup
+from pybind11.setup_helpers import Pybind11Extension
+
+sourcefiles = [
+    "isri_tools_src/align.cpp",
+    "isri_tools_src/Modules/accrpt.c",
+    "isri_tools_src/Modules/charclass.c",
+    "isri_tools_src/Modules/ci.c",
+    "isri_tools_src/Modules/dist.c",
+    "isri_tools_src/Modules/edorpt.c",
+    "isri_tools_src/Modules/list.c",
+    "isri_tools_src/Modules/sort.c",
+    "isri_tools_src/Modules/stopword.c",
+    "isri_tools_src/Modules/sync.c",
+    "isri_tools_src/Modules/table.c",
+    "isri_tools_src/Modules/text.c",
+    "isri_tools_src/Modules/unicode.c",
+    "isri_tools_src/Modules/util.c",
+    "isri_tools_src/Modules/wacrpt.c",
+    "isri_tools_src/Modules/word.c",
+    ]
+
+ext = Pybind11Extension("isri_tools", sourcefiles)
+
+setup(
+    ext_modules = [ext]
+)

### 2022-01-07 src/ocr/ner_sync_ref.py
WIP add stub for ner gt sync tool

@@ -1,0 +1,286 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+
+"""
+This program synchronizes NER tagging from reference text to noisy predicted
+text (OCR output).
+
+Sample usage:
+    ner_sync_ref.py reference_utf8.xml [predicted_utf8.xml | predicted_utf8.txt] [alignedref_utf8.xml]
+
+Input restrictions:
+    - Encoding MUST BE UTF-8.
+    - End of line MUST BE either CRLF (Windows), CR (old Macintosh) or LF (OSX,
+      *nix).
+
+Output post-conditions:
+    - Encoding WILL BE UTF-8.
+    - End of line WILL BE LF.
+"""
+
+# Reused code from https://github.com/SmartDOC-MOC/moc_normalization
+
+# ==============================================================================
+# Imports
+import logging
+import argparse
+import sys
+import unicodedata
+
+# ==============================================================================
+# Logging
+logger = logging.getLogger(__name__)
+
+# ==============================================================================
+# Constants
+PROG_VERSION = "1.0"
+PROG_DESCR = (
+    "Synchronizes NER tagging from reference text to noisy predicted text (OCR output)."
+    f" Version: {PROG_VERSION}"
+    )
+PROG_NAME = "ner_sync_ref"
+
+ERRCODE_OK = 0
+ERRCODE_NOFILE = 10
+ERRCODE_EXTRACHAR = 50
+
+# FIXME extract this to a common config file (if we have multiple scripts)
+# FIXME add cyrillic as Pero can output content in this script (actually add entire scripts)
+#       OR project extra char to unused char in GT (Unicode PUA?)
+ALLOWED_INPUT = (
+    """ !"#$%&'()*+,-./0123456789:;<=>?@ABCDEFGHIJKLMNOPQRSTUVWXYZ[\\]^_`abc"""
+    """defghijklmnopqrstuvwxyz{|}~ ¡¢£¤¥¦§¨©ª«¬­®¯°±²³´µ¶·¸¹º»¼½¾¿ÀÁÂÃÄÅÆÇÈÉ"""
+    """ÊËÌÍÎÏÐÑÒÓÔÕÖ×ØÙÚÛÜÝÞßàáâãäåæçèéêëìíîïðñòóôõö÷øùúûüýþÿŒœŠšŸŽžƒˆ˜–—‘’"""
[…]

### 2022-01-09 src/ocr/isri_tools_src/align.cpp
Add comments

@@ -53,8 +53,8 @@ int compute_align_map(std::wstring a, std::wstring b, std::vector<int>& map_A, s
   Text B;
   list_initialize(&A);
   list_initialize(&B);
-  text_from_string(&A, a);
-  text_from_string(&B, b);
+  text_from_string(&A, a);  // 16 bit fixed encoding only for now
+  text_from_string(&B, b);  // 16 bit fixed encoding only for now
 
   map_A.resize(a.size());
   map_B.resize(b.size());
@@ -64,7 +64,7 @@ int compute_align_map(std::wstring a, std::wstring b, std::vector<int>& map_A, s
     Text texts[2] = {A, B};
 
     Synclist synclist;
-    fastukk_sync(&synclist, texts);
+    fastukk_sync(&synclist, texts); // Actual call to original UNLV/ISRI tools
     alignment_size = compute_mapping(&synclist, map_A.data(), map_B.data());
   }
 

### 2022-01-09 src/create_gold.sh
Full code for experiment 1

@@ -1,36 +1,50 @@
-#!/bin/bash
+#!/usr/bin/env bash
 
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# This script creates the gold dataset CSV from a set of JSON pages
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+# ==============================================================================
 # Let's get paranoid
 set -eu
 
-######################
+# Constants
+WORKDIR=$DATADIR/gold
 
-WORKDIR=data/gold
-
-mkdir -p $WORKDIR
-
-pushd $WORKDIR > /dev/null
-
-echo "Creating the gold dataset"
-
-# Gather sampled pages numbers & directory name
+# Globals
 declare -A directories
-directories=()
-while IFS= read -r row;
-do
-  arr=($row)
-  directories+=([${arr[0]}]=${arr[1]})
-done < sampled_pages.txt
+# ==============================================================================
 
-# Creates the gold dataset from a set of directories jsons and stores it in CSV where:
-# column 1 is a pre-annotated directory entry in XML
-# column 2 is the entry's directory name
-rm -f gold.csv
-for f in $(ls *.json);
-do
-  pagenum=$(echo ${f%.*} | sed 's/^0*//')
-  directory="${directories[$pagenum]}"
-  cat $f | jq --raw-output --arg directory "$directory" '.[] | select(.type=="ENTRY" and .checked==true) | [.ner_xml,$directory] | @csv' >> gold.csv
-done
+# ------------------------------------------------------------------------------
+function read_groups() {
+  # Gather sampled pages numbers & directory name
+  directories=()
+  while IFS= read -r row;
+  do
+    arr=($row)
+    directories+=([${arr[0]}]=${arr[1]})
+  done < sampled_pages.txt
+}
 
+# ------------------------------------------------------------------------------
+function create_gold() {
+  # Creates the gold dataset from a set of directories jsons and stores it in CSV where:
+  # column 1 is a pre-annotated directory entry in XML
+  # column 2 is the entry's directory name
+  rm -f gold.csv
+  for f in $(ls *.json);
+  do
+    pagenum=$(echo ${f%.*} | sed 's/^0*//')
[…]

### 2022-01-09 src/ner/camembert.py

@@ -1,16 +1,19 @@
-import json
+import json, argparse, nltk
 import config as cfg
-import os
 import numpy as np
-import nltk
+from pathlib import Path
+from xml.dom.minidom import parseString
 from datasets import Dataset, load_from_disk
+from datasets import load_metric
 from transformers import AutoTokenizer
 from transformers import AutoModelForTokenClassification
 from transformers import TrainingArguments, Trainer, DataCollatorForTokenClassification
-from datasets import load_metric
-from xml.dom.minidom import parseString
 
-# Shared variables
+logger = cfg.logger
+
+# =============================================================================
+# region ~ Model parameters
+
 label_id_map = {
     "O": 0,
     "I-LOC": 1,
@@ -22,32 +25,47 @@ label_id_map = {
     "I-TITRE": 7,
     "I-FT": 8,
 }
-tokenizer = AutoTokenizer.from_pretrained(cfg.BERT_BASE_MODEL)
-training_args = TrainingArguments("berties", **cfg.BERT_TRAINER_CONFIG)
 
-# Training & eval
-def train_eval(train, valid):
+CAMEMBERT_MODEL = "Jean-Baptiste/camembert-ner"
+
+CAMEMBERT_TRAINER_CONFIG = {
+    "evaluation_strategy": "epoch",
+    "learning_rate": 1e-4,
+    "per_device_train_batch_size": 16,
+    "per_device_eval_batch_size": 16,
+    "num_train_epochs": 3,
+    "weight_decay": 1e-5,
+}
+
+tokenizer = AutoTokenizer.from_pretrained(CAMEMBERT_MODEL)
+training_args = TrainingArguments("berties", **CAMEMBERT_TRAINER_CONFIG)
+
+# endregion
+
+# =============================================================================
+# region ~ Train & evaluation
+def train_eval(train, dev, test):
+
     data_collator = DataCollatorForTokenClassification(tokenizer)
 
     trainer = Trainer(
-        model,
+        model,  # Implicit
         training_args,
         train_dataset=train,
-        eval_dataset=valid,
+        eval_dataset=dev,
         data_collator=data_collator,
         tokenizer=tokenizer,
         compute_metrics=compute_metrics,
     )
 
     trainer.train()
-    eval = trainer.evaluate()
-    return eval
+    return trainer.evaluate(test), trainer.evaluate()
 
 
 def compute_metrics(p):
     predictions, labels = p
     predictions = np.argmax(predictions, axis=2)
[…]

### 2022-01-09 src/ner/config.py

@@ -1,26 +1,34 @@
-from pathlib import Path
-import logging
+"""
+ Holds the global configuration used in all python scripts related to NER.
+"""
 
-# Assuming datadir is located in the parent directory of this module
-DATA_DIR = Path(__file__).parents[1] / "data"
+import os
+import logging
+from pathlib import Path
+
+# =============================================================================
+# region ~ Configurations
+
+DATA_DIR = Path(os.environ["DATADIR"])
+
+assert DATA_DIR.exists()
 
 GOLD = DATA_DIR / "gold/gold.csv"
+SEED = 42  # Random seed to generate train/validation/test datasets
+MIN_TRAINSET_SIZE = 30
+DEBUG = os.getenv("DEBUG", "False").lower() in ("true", "1", "t")
+# endregion
 
-N_SPLITS = 2
-SEED = 42  # Used to generate train/validation/test datasets
-MIN_TRAINSET_SIZE = 50
 
-BERT_BASE_MODEL = "Jean-Baptiste/camembert-ner"
-
-BERT_TRAINER_CONFIG = {
-    "evaluation_strategy": "epoch",
-    "learning_rate": 1e-4,
-    "per_device_train_batch_size": 16,
-    "per_device_eval_batch_size": 16,
-    "num_train_epochs": 3,
-    "weight_decay": 1e-5,
-}
-
-# Set logging configuration for the whole project
-logging.basicConfig()
-logging.root.setLevel(logging.INFO)
+# =============================================================================
+# region ~ Logging
+logger = logging.getLogger("nerlogger")
+LOG_LEVEL = logging.getLevelName("DEBUG" if DEBUG else "INFO")
+logger.setLevel(LOG_LEVEL)
+logformatter = logging.Formatter(
+    "%(asctime)s ; %(levelname)s ; %(message)s", datefmt="%d/%m/%Y %I:%M:%S"
+)
+loghandler = logging.StreamHandler()
+loghandler.setFormatter(logformatter)
+logger.addHandler(loghandler)
+# endregion

### 2022-01-09 src/ner/create_datasets.py

@@ -1,119 +1,171 @@
 import csv
-import logging
 import os
+import numpy as np
 import config as cfg
+from sklearn.model_selection import GroupKFold
 from sklearn.model_selection import train_test_split
 from datasets.dataset_dict import DatasetDict
 from cnn import create_spacy_dataset
 from bert import create_huggingface_dataset
 
-# Helper functions
+logger = cfg.logger
+
+# =============================================================================
+# region ~ Helpers
+
+
 def unwrap(list_of_tuples2):
     return tuple(zip(*list_of_tuples2))
 
 
-# Loads the gold dataset
-with open(cfg.GOLD, encoding="utf-8") as annot:
-    gold_dataset = [(_[0], _[1]) for _ in csv.reader(annot)]
+# endregion
 
-logging.info(f"Gold dataset: {len(gold_dataset)} records")
 
-# Generate Train (80%), Validation|Dev (10%) and Test (10%) datasets
-# with stratified sampling on the directories names.
+# =============================================================================
+# region ~ Main processing
 
-# Get train and dev+test
-_, classes = unwrap(gold_dataset)
-train_devtest = train_test_split(
-    gold_dataset, train_size=0.8, shuffle=True, random_state=cfg.SEED, stratify=classes
-)
-train = train_devtest[0]
 
-# Split dev+test in half to get dev and test
-devtest = train_devtest[1]
-_, classes = unwrap(devtest)
-dev_test = train_test_split(
-    devtest, train_size=0.5, shuffle=True, random_state=cfg.SEED, stratify=classes
-)
-dev = dev_test[0]
-test = dev_test[1]
+def main():
+    """Generate all datasets"""
+    with open(cfg.GOLD, encoding="utf-8") as gf:
+        gold_dataset = np.array([(_[0], _[1]) for _ in csv.reader(gf)])
 
-logging.info(
-    f"Train ({len(train)} records), Dev ({len(dev)} records), Test ({len(test)} records)"
-)
+    logger.info(f"GOLD: {len(gold_dataset)} records")
 
-###
-# Generate smaller train sets for experiment 1
-###
+    ##
+    # Create the global Train, Dev and Test datasets
+    ##
+    train, dev, test = make_train_dev_test(gold_dataset)
+    sets_names = ["train", "dev", "test"]
+    zipped = zip([train, dev, test], sets_names)
+    odir = cfg.DATA_DIR / "datasets"
+    export(odir, zipped)
 
-exp_1_trainsets = [train]
[…]

### 2022-01-09 src/ner/experiment_1.sh

@@ -1,0 +1,94 @@
+#!/usr/bin/env bash
+
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# This script trains the CNN, BERT & BERT+PRE models on each subsampled 
+# training set created using create_datasets.py and stored
+# in $DATADIR/experiment_1/datasets
+# 
+# The validation and test sets remain untouched.
+# Precision, recall and F1 scores are measured against the test set
+# and stored in $DATADIR/experiment_1/metrics 
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+# ==============================================================================
+# Let's get paranoid
+set -eu
+
+# Constants
+EXP1=$DATADIR/experiment_1
+METRICS=$EXP1/metrics
+MODEL_PRETRAINED=$DATADIR/pretraining/camembert_best
+
+WORKDIR=`dirname $0`
+# ==============================================================================
+
+
+# ------------------------------------------------------------------------------
+function train_sizes() {
+    # Returns: sizes of the subsampled trainsets in $DATADIR in ascending order
+    find $EXP1/*.spacy -printf "%f\n" |
+    xargs -n1 echo - |
+    sed 's/[^0-9]*//g' |
+    sort -nk1 |
+    uniq
+}
+
+# ------------------------------------------------------------------------------
+function train_eval_cnn() {
+    # Args: trainset sizes
+    for size in $@; do
+        train=$EXP1"/train_$size.spacy"
+        dev=$EXP1"/dev_$size.spacy"
+        test=$EXP1"/test_$size.spacy"
+
+        echo "TRAINING: $train"
+        echo "DEVELOMPENT: $dev"
+        echo "EVALUATION: $test"
+
+        python -m spacy train $WORKDIR/cnn_config.cfg \
+            -o $EXP1 \
+            --paths.train $train \
+            --paths.dev $dev
+
+        python -m spacy evaluate $EXP1/model-best $test \
+            -o "$METRICS/cnn_${size}_test.json" \
+            -dp $EXP1
+
+        python -m spacy evaluate $EXP1/model-best $dev \
+            -o "$METRICS/cnn_${size}_dev.json" \
+            -dp $EXP1
+    done
+}
+
[…]

### 2022-01-09 src/ner/init.sh

@@ -1,10 +1,16 @@
-#!/bin/bash
+#!/usr/bin/env bash
 
+# ==============================================================================
 # Let's get paranoid
 set -eu
 
-WORKDIR=$(dirname "$0")
+WORKDIR=`dirname $0`
+# ==============================================================================
+
+# Make directories
+mkdir -p $DATADIR/experiment_1
+mkdir -p $DATADIR/datasets
 
 # Build datasets for all experiments
 echo "Initializing training, dev and test datasets"
-python ${WORKDIR}/create_datasets.py
+python $WORKDIR/create_datasets.py

### 2022-01-09 material/experiment_1/figures.r
WIP - material for experiment 1

@@ -1,0 +1,53 @@
+library(dplyr)
+library(ggplot2)
+library(xtable)
+library(kableExtra)
+
+# BERT : jq -r ' . | [ .[] ]  | @csv' $(l -1v ./*_test.json)
+# CNN : jq -r 'with_entries(select([.key] | inside( ["ents_p","ents_r","ents_f"] ) ) ) | [ .[] ]  | @csv' $(l -1v ./*_test.json)
+
+
+# Spacy CNN
+cnn = read.csv("cnn/metrics.csv",header=T)
+cnn = cnn %>% arrange(trainsize) %>% group_by(trainsize) %>% filter(dataset == 'test') %>% summarise(across(everything(), mean))
+
+# HF Camembert 
+camembert = read.csv("camembert/metrics.csv",header=T)
+camembert = camembert %>% arrange(trainsize) %>% group_by(trainsize) %>% filter(dataset == 'test') %>% summarise(across(everything(), mean))
+
+# HF Camembert + pretraining
+camembert_pretrained = read.csv("camembert_pretrained/metrics.csv",header=T)
+camembert_pretrained = camembert_pretrained %>% arrange(trainsize) %>% group_by(trainsize) %>% filter(dataset == 'test') %>% summarise(across(everything(), mean))
+
+
+# Reshape for LaTeX export
+tbl = cbind(cnn %>% select(eval_f1), camembert %>% select(eval_f1), camembert_pretrained %>% select(eval_f1))
+colnames(tbl) <- c("Cnn", "Camembert", "Camembert+pretraining")
+
+tbl = t(tbl)
+colnames(tbl) <- camembert[['trainsize']]
+
+
+# TODO : max value of each column in bold
+#fn = function(x) cell_spec(x, bold = T)  
+#tbl[, 2:8] = t(apply(tbl[,2:8], 1, fn))    
+
+tbl %>%
+    kable(booktabs = TRUE, digits=3, caption = "Train size vs F1 score", align = "c", 'latex') %>% 
+    kable_styling(latex_options = c("hold_position")) %>%
+    cat(., file = "table.tex")
+
+
+#png("mygraphic.png",  width=1000)
+#cnn$group = 'Cnn'
+#camembert$group = 'Camembert'
+#camembert_pretrained$group = 'Camembert+pretraining'
+#visuals = rbind(cnn, camembert,camembert_pretrained)
[…]

### 2022-01-09 material/experiment_1/figures.r
Export figure for exp 1

@@ -38,16 +38,28 @@ tbl %>%
     cat(., file = "table.tex")
 
 
-#png("mygraphic.png",  width=1000)
-#cnn$group = 'Cnn'
-#camembert$group = 'Camembert'
-#camembert_pretrained$group = 'Camembert+pretraining'
-#visuals = rbind(cnn, camembert,camembert_pretrained)
-#visuals$vis=c(rep("red",8),rep("blue",8),rep("green",8))
-#ggplot(visuals, aes(y=eval_f1, x=trainsize,group=interaction(group),col=group)) + 
-#   geom_point() + geom_line()
+png("experiment1.png",
+  width     = 20,
+  height    = 10,
+  units     = "cm",
+  res       = 300,
+  pointsize = 4)
 
-#dev.off()
-#browseURL("mygraphic.png") 
+cnn = cnn %>% select(eval_precision, eval_recall, eval_f1, trainsize)
+camembert = camembert %>% select(eval_precision, eval_recall, eval_f1, trainsize)
+camembert_pretrained = camembert_pretrained %>% select(eval_precision, eval_recall, eval_f1, trainsize)
+
+
+cnn$model = 'Cnn'
+camembert$model = 'Camembert'
+camembert_pretrained$model = 'Camembert+pretraining'
+
+visuals = rbind(cnn, camembert,camembert_pretrained)
+
+ggplot(visuals, aes(y=eval_f1, x=trainsize, group=interaction(model), col=model)) + 
+   geom_point(aes(shape = model)) + geom_line(aes(linetype = model)) + theme_light()
+
+dev.off()
+browseURL("experiment1.png") 
 
 

### 2022-01-09 src-latex/parts/20-related-work.tex
Setup paper structure
Suggestion of structure

--- 2021-10-28 src-latex/parts/10-introduction.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Introduction}
+\section{OCR and NER on Historical Texts}
+
+The directory processing pipeline presented in \cite{bell2020automated} to locate old gas stations using mid-20th century directories includes an OCR step, done with Tesseract, and a NER step to identify company names and addresses within the extracted text, performed using rules built as regular expressions.
+This section reviews existing approaches to achieving each of these two tasks.
+It concludes with a review of the works assessing the effects of OCR quality on the NER performance and the proposed solutions to overcome this problem.
+
+\subsection{Optical Character Recognition on Historical Texts}
+
+\subsection{Named Entity Recognition}
+
+Many approaches have been designed to recognize named entities, ranging from handcrafted rules to supervised approaches \cite{nadeau2007}.
+Rule based approaches look for portions of text that match patterns based on word characteristics such as their case, whether they are digits or not, their part-of-speech tags, etc.
+These patterns can be fully handcrafted like in \cite{bell2020automated} or generated by data mining approaches like in \cite{nouvel2011}.
+Rule based approaches can also combine a set of patterns with a set of dictionaries (gazetteers, author lists, etc.) that help recognizing named entities when an exhaustive lexicon of searched entities is available \cite{mansouri2008,maurel2011}.
+For example, the CasEN\footnote{\url{https://tln.lifat.univ-tours.fr/version-francaise/ressources/casen}} transducer cascade leverages syntatic-lexical patterns and dictionnaries to recognize Named Entities in French texts.
+Such kind of approaches achieve very good results when applied to specialized domain corpus - like directories - and when exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
+
+Supervised approaches include both approaches implementing supervised learning algorithms with careful text feature engineering, and deep learning based approaches which automatically build their own features to classify words into one category of named entities or another.
+In recent years, the latter have grown dramatically, yielding state-of-the art performances and establishing new baselines\cite{li2020}.
+A recent survey proposed by \cite{li2020} presents deep learning techniques for NER according to the distributed representation(s) they use as input, their context encoder and their tag decoder.
+It shows that language model embeddings pre-trained using \textit{Transformer} \cite{vaswani2017attention}, like BERT \cite{devlin2018bert}, can not only be used to replace traditional embeddings as input distributed representations but also be fine-tuned for the NER task with one additional output layer while achieving state-of-the-art performances.
+The survey concludes that fine-tuning general-purpose contextualized language models with domain-specific data is very likely to give good performances for use cases with domain-specific texts and few training data.
+
+\subsection{Named Entity Recognition on Historical Texts}
+
+\cite{Labusch2020NamedED} apply a BERT-based NER model on historical texts in German, French and English, produced by OCR.
+It is an of-the-shelf multilingual BERT-based NER model, pre-trained on more than 2 000 000 pages of OCRed historical texts, in the three languages and fine-tuned with historical texts NER ground truth, also in the three languages.
+Eventually, a clear decrease in NER performance on OCRised texts is noted.
+Moreover, the article reports that this decrease is more important for English texts for which the OCR is of poorer quality than for texts in the other two languages.
+
+Recently, several studies have focused on the extent to which the quality of the OCR affects the results produced by a NER model.\cite{van2020assessing} assess the impact of OCR on several NLP downstream tasks, including NER.
+They worked on a corpus published by a post-OCR correction software company, made of many journal articles with different levels of OCR errors and their respective ground-truths.
+For each OCRed article, the Word Error Rate (WER) is computed and the English model \textit{en-core-web-lg} provided by Spacy\footnote{\url{https://spacy.io/}} library is used to perform NER on \textit{Person}, \textit{GPE} and \textit{Date}.
+The performance of the NER model with respect to OCR quality is eventually assessed by calculating the F-measure for each NER class, and each article i.e., each WER value.
+\cite{hamdi2020assessing} performed a similar but more extensive evaluation on four different NER models: CoreNLP using Conditional Random Fields and three deep neural models, BLSTM-CNN, BLSTM-CRF, and BLSTM-CNN-CRF.
+They tested them on two well-known NER benchmark corpora: CoNLL-02 and CoNLL-03.
+They applied four different types of OCR noise to each corpus, with two levels of degradation and computed the WER and Character Error Rate (CER) for each degraded version.
+Finally, they applied each NER model on the progressively degraded versions of the corpora and computed the resulting F-measure.
+Overall, NER F-mesure drops from 90\% to 50\% when the WER increase from 8\% to 50\%.
+Besides, all NER models are sensitive to OCR errors, but models based on deep neural networks seem less prone to errors.
+
+Based on the observation that OCR errors necessarily affect the performance of NER approaches applied to historical texts, \cite{huynh2020use}, \cite{martinek2019hybrid} and \cite{marz2021data} have proposed different approaches to reduce their negative impact.

### 2022-01-09 src-latex/parts/60-results.tex
Adds results table for experiment 1
Updates from Overleaf

@@
 
-\begin{verbatim}
-Accuracy
-|                             ..o   
-|                ..o       o/       
-|            ./`          /`          
-|        ./o`      .o   o            
-|     ./`        .`                    One curve per approach
-|    o       o/`                  o  
-|          .`                  ./    
-|       o`                    /      
-|                         .o        
-|                       .`           
-|                      o            
-|                                   
-+------------------------------   Training set size
-\end{verbatim}
+\begin{table}[!h]
+
+\caption{Models F1 score on unseen data vs trainset size}
+\centering
+\begin{tabular}[t]{lcccccccc}
+\toprule
+Trainset size (\% of the full trainset) & 0.8 & 1.5 & 3.1 & 6.2 & 12.5 & 25 & 50 & 100\\
+\midrule
+Cnn & \textbf{0.855} & \textbf{0.885} & \textbf{0.910} & 0.911 & 0.919 & 0.925 & 0.926 & 0.929\\
+Camembert & 0.236 & 0.626 & 0.856 & \textbf{0.916} & \textbf{0.930} & 0.933 & \textbf{0.948} & 0.953\\
+Camembert+pretraining & 0.190 & 0.685 & 0.889 & 0.913 & \textbf{0.930} & \textbf{0.946} & \textbf{0.948} & \textbf{0.960}\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{figure}[htb!]
+	   \center{\includegraphics[width=\textwidth]
+	       {../material/experiment_1/f1_vs_trainsize.png}}
+	  \vspace{3in}
+	  \caption{\label{fig:f1-vs-trainsize} Models F1 score on unseen data vs trainset size}
+\end{figure}
 

### 2022-01-09 src-latex/parts/50-xp-setup.tex
more details in setup and results

--- 2022-01-06 src-latex/parts/50-xp-setup.tex
@@
 \subsection{Training procedure}
-5-fold CV for all models.
+
 \begin{itemize}
-\item Spacy CNN : no limits on the number of epochs, instead use patience=1600.
-\item CamemBERT : 3 epochs.
+\item All : leave out 3 directories (~20\% of the dataset) as test data.
+Then apply stratified sampling based on the directory of each entry to split the remaining into a train set (~72\%) and a dev set (~8\%).
+\item Spacy CNN : No limit on epochs, instead use patience 1600.
+\item CamemBERT \& CamemBERT pretrained : 3 epochs.
 \end{itemize}
+
+Experiment 1: split the training set in half recursively as long as the train \& dev sets contain at least one entry of each directory in the original trainset.
+On the current dataset with 8k entries this leads to the creations of 8 subsets containing 0.8, 1.5, 3.1, 6.2, 12.5, 25, 50 and 100\% of the full train set.
+Models arerun 5 times on each trainset.
+
+Experiment 2 : 1 run per dataset on the best model, i.e. CamemBERT + pretraining.
+What train/dev/test stragegy ? E.g. trainset 100\% noisy + dev 50\% clean + test 50\% clean ?
 
@@
 \subsection{Implementation details}
-we use spacy \mcite{spacy}
-will we use FLAIR as well?
+we use spacy \mcite{spacy} for the CNN implementation and Huggingface for BERT.
+
+will we use FLAIR as well? $\rightarrow$ No

### 2022-01-09 src-latex/parts/60-results.tex
more details in setup and results

@@
 
-\subsection{Experiment 2: sensibility to training on noisy data}
-Idea : Measure the impact of training with noisy data when predicting NER tags on both clean and noisy data.
-
-What's interesting : do models perform better on noisy data when trained on noisy data ? Does it impact predictions on clean data ?
-
-Table : NER NN models VS train on {noisy, clean} datasets VS evaluate on {noisy, clean} datasets
-
 \subsection{Experiment 3: Accuracy vs NER approach vs OCR noise}
@@
 Quantitative results: table + graph ideally (with OCR noise, same format as previous)
+
+Table : NER NN models VS train on {noisy, clean} datasets VS evaluate on {noisy, clean} datasets
+
+\begin{table}
+\caption{Camembert vs noise}
+\centering
+\begin{tabular}{ll|ll|l}
+                            &                      & \multicolumn{2}{l|}{Training data} &   \\
+                            &                      & noisy & clean                      &   \\ 
+\cline{1-4}
+\multirow{2}{*}{Test data } & noisy                & f1    & f1                         &   \\
+                            & clean                & f1    & f1                         &   \\ 
+\cline{1-4}
+                            & \multicolumn{1}{l}{} &       & \multicolumn{1}{l}{}       &   \\
+                            & \multicolumn{1}{l}{} &       & \multicolumn{1}{l}{}       &  
+\end{tabular}
+\end{table}
 

### 2022-01-10 src/ocr/ner_sync_ref.py
Update ner_sync_ref.py
Fix.

@@ -114,7 +114,7 @@ ANNOT_CODE_MAP = {  # We use unicode private user area when needed (no dedicated
     "\u2013": "-", # –
     "\u2014": "-", # —
     "::SymbM::": "\uE011",
-    ":NC::": "\uE012",
+    "::NC::": "\uE012",
     "::NA::": "\uE013",
     "::TL::": "\uE014",
     "::TG::": "\uE015",

### 2022-01-10 src/build_gold.sh
Cleaner code + doc

@@ -2,6 +2,8 @@
 
 # ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 # This script creates the gold dataset CSV from a set of JSON pages
+# 
+# Requires env. variable $DATADIR.
 # ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
 # ==============================================================================
@@ -26,11 +28,13 @@ function read_groups() {
   done < sampled_pages.txt
 }
 
+
 # ------------------------------------------------------------------------------
 function create_gold() {
-  # Creates the gold dataset from a set of directories jsons and stores it in CSV where:
-  # column 1 is a pre-annotated directory entry in XML
-  # column 2 is the entry's directory name
+  # Creates the gold dataset from a set of directories jsons and stores it 
+  # as a CSV where:
+  # - column 1 is a pre-annotated directory entry in XML
+  # - column 2 is the entry's directory name
   rm -f gold.csv
   for f in $(ls *.json);
   do
@@ -40,11 +44,12 @@ function create_gold() {
   done
 }
 
+
 # ==============================================================================
 # Entry point
 mkdir -p $WORKDIR
 pushd $WORKDIR > /dev/null
 read_groups
-create_gold 
+create_gold
 popd  > /dev/null
 

### 2022-01-10 src/create_datasets.py

@@ -5,8 +5,8 @@ import config as cfg
 from sklearn.model_selection import GroupKFold
 from sklearn.model_selection import train_test_split
 from datasets.dataset_dict import DatasetDict
-from cnn import create_spacy_dataset
-from bert import create_huggingface_dataset
+from spacy_util import create_spacy_dataset
+from camembert import create_huggingface_dataset
 
 logger = cfg.logger
 
@@ -30,14 +30,22 @@ def main():
     with open(cfg.GOLD, encoding="utf-8") as gf:
         gold_dataset = np.array([(_[0], _[1]) for _ in csv.reader(gf)])
 
-    logger.info(f"GOLD: {len(gold_dataset)} records")
+    logger.info("GOLD: %d records" % len(gold_dataset))
 
     ##
-    # Create the global Train, Dev and Test datasets
+    # Create the global Train & Dev
+    # Stratified split 90/10%
     ##
-    train, dev, test = make_train_dev_test(gold_dataset)
-    sets_names = ["train", "dev", "test"]
-    zipped = zip([train, dev, test], sets_names)
+    _, groups = unwrap(gold_dataset)
+    train, dev = train_test_split(
+        gold_dataset,
+        train_size=0.9,
+        shuffle=True,
+        random_state=cfg.SEED,
+        stratify=groups,
+    )
+    logger.info("TRAIN: %s, DEV: %s" % (len(train) ,len(dev)))
+    zipped = zip([train, dev],  ["train", "dev"])
     odir = cfg.DATA_DIR / "datasets"
     export(odir, zipped)
 
@@ -45,12 +53,14 @@ def main():
     # Create datasets for experiment 1
     ##
     odir = cfg.DATA_DIR / "experiment_1"
-
+    
+    train, dev, test = make_train_dev_test(gold_dataset)
     exp1_trainsets = create_experiment_1_trainsets(train)
+    
     for subtrain in exp1_trainsets:
         datasets = [subtrain, dev, test]
         addendum = str(len(subtrain))
-        export(odir, zip(datasets, sets_names), addendum)
+        export(odir, zip(datasets, ["train", "dev", "test"]), addendum)
 
 
 def make_train_dev_test(gold):
@@ -85,9 +95,9 @@ def make_train_dev_test(gold):
     train = train_dev[0]
     dev = train_dev[1]
 
[…]

### 2022-01-10 src/experiment_1.sh

@@ -85,8 +85,8 @@ function train_eval_camembert_pretrained() {
 function main() {
     mkdir -p $METRICS
     train_eval_cnn $(train_sizes)
-    #train_eval_camembert $(train_sizes)
-    #train_eval_camembert_pretrained $(train_sizes)
+    train_eval_camembert $(train_sizes)
+    train_eval_camembert_pretrained $(train_sizes)
 }
 
 # ==============================================================================

### 2022-01-10 src/init_experiments.sh

@@ -1,5 +1,10 @@
 #!/usr/bin/env bash
 
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+# Creates all train, dev & test datasets required in experiments 1 & 2
+# and store them in $DATADIR (env)
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
 # ==============================================================================
 # Let's get paranoid
 set -eu

### 2022-01-10 src/spacy_util.py

@@ -4,9 +4,8 @@ from spacy.tokens import DocBin, Doc
 from xml.dom.minidom import parseString
 from spacy.training import tags_to_entities
 
-###########################
-# ~ DATA CONVERSION UTILS ~
-###########################
+# =============================================================================
+# region ~ Data conversion utils for Spacy
 
 nlp = spacy.blank("fr")
 

### 2022-01-10 src/ocr/ner_sync_ref.py
more tests

@@ -294,7 +294,10 @@ def check_alignment_charset(text_to_align: str, dump_charset: bool) -> bool:
             print(f"Suspicious char@{pos:03d}: {char} ({_unichr2str(char)}"
                   f" -- cat.: {unicodedata.category(char)})")
     if not unicodedata.is_normalized('NFKC', text_to_align):
+        # Note: circled letter symbols do not seem to comply with NFKC. Maybe NFC is enough…
         print("Warning: string is not normalized (compatibility + composed mode).")
+    if not unicodedata.is_normalized('NFC', text_to_align):
+        print("Warning: string is not normalized (composed mode).")
     if dump_charset:
         print("Charset for string:")
         print(f"\t\"{text_to_align}\"")

### 2022-01-10 src-latex/parts/20-related-work.tex

@@
 \subsection{Optical Character Recognition on Historical Texts}
+
+Most of the current state-of-the-art OCR systems, like Tesseract \cite{smith2007overview}, OCRopus \cite{breuel2008ocropus} and PERO OCR \cite{kohut2021ts} are based on a pipeline of convolutional neural networks (CNNs) and long short-term memory networks (LSTM).
+Although this kind of model produces good results on modern texts, it faces several challenges with ancient texts, such as the lack of annotated data for learning, or different transcriptions styles in training data.
+
+\cite{martinek2019hybrid} propose an approach to generate synthetic annotated text for historical OCR training, based on manually collected characters from historical text images.
+This work propose to train an OCR system based on a CNN-
+LSTM network with the synthetic data and then to fine-tune the model with some pages of real historical annotated text. The results show that this approach gives state-of-the-art results.
+
+To overcome the limitations due to different transcription styles in training data, PERO OCR adds a Transcription Style Block layer to a classical model based on a CNN and a Recurrent Neural Network components \cite{kohut2021ts}.
+This block takes the image of the text and a Transcription Style Identifier as inputs and help the network decide what kind of transcription style to use as output.
 
@@
 
-Based on the observation that OCR errors necessarily affect the performance of NER approaches applied to historical texts, \cite{huynh2020use}, \cite{martinek2019hybrid} and \cite{marz2021data} have proposed different approaches to reduce their negative impact.
+Based on the observation that OCR errors necessarily affect the performance of NER approaches applied to historical texts, \cite{huynh2020use} and \cite{marz2021data} have proposed different approaches to reduce their negative impact.

### 2022-01-10 src-latex/parts/20-related-work.tex
Finished Related Work last section (V1)

@@
 Most of the current state-of-the-art OCR systems, like Tesseract \cite{smith2007overview}, OCRopus \cite{breuel2008ocropus} and PERO OCR \cite{kohut2021ts} are based on a pipeline of convolutional neural networks (CNNs) and long short-term memory networks (LSTM).
-Although this kind of model produces good results on modern texts, it faces several challenges with ancient texts, such as the lack of annotated data for learning, or different transcriptions styles in training data.
+Although this kind of model produces good results with modern texts, it faces several challenges with ancient texts, such as the lack of annotated data for learning, or different transcription styles in training data.
 
 \cite{martinek2019hybrid} propose an approach to generate synthetic annotated text for historical OCR training, based on manually collected characters from historical text images.
-This work propose to train an OCR system based on a CNN-
-LSTM network with the synthetic data and then to fine-tune the model with some pages of real historical annotated text. The results show that this approach gives state-of-the-art results.
+This work proposes to train an OCR system based on a CNN-
+LSTM network with synthetic data and then to fine-tune the model with some pages of real historical annotated text. The results show that this approach gives state-of-the-art results.
 
 To overcome the limitations due to different transcription styles in training data, PERO OCR adds a Transcription Style Block layer to a classical model based on a CNN and a Recurrent Neural Network components \cite{kohut2021ts}.
-This block takes the image of the text and a Transcription Style Identifier as inputs and help the network decide what kind of transcription style to use as output.
+This block takes the image of the text and a Transcription Style Identifier as inputs and helps the network decide what kind of transcription style to use as output.
+
+\textcolor{green}{à compléter/retoucher}
 
@@
 Rule based approaches can also combine a set of patterns with a set of dictionaries (gazetteers, author lists, etc.) that help recognizing named entities when an exhaustive lexicon of searched entities is available \cite{mansouri2008,maurel2011}.
-For example, the CasEN\footnote{\url{https://tln.lifat.univ-tours.fr/version-francaise/ressources/casen}} transducer cascade leverages syntatic-lexical patterns and dictionnaries to recognize Named Entities in French texts.
-Such kind of approaches achieve very good results when applied to specialized domain corpus - like directories - and when exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
+For example, the CasEN\footnote{\url{https://tln.lifat.univ-tours.fr/version-francaise/ressources/casen}} transducer cascade leverages syntatic lexical patterns and dictionnaries to recognize Named Entities in French texts.
+Such kind of approaches achieve very good results when applied to specialized domain corpus - like directories - and when an exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
 
 Supervised approaches include both approaches implementing supervised learning algorithms with careful text feature engineering, and deep learning based approaches which automatically build their own features to classify words into one category of named entities or another.
-In recent years, the latter have grown dramatically, yielding state-of-the art performances and establishing new baselines\cite{li2020}.
+In recent years, the latter have grown dramatically, yielding state-of-the-art performances and establishing new baselines\cite{li2020}.
 A recent survey proposed by \cite{li2020} presents deep learning techniques for NER according to the distributed representation(s) they use as input, their context encoder and their tag decoder.
-It shows that language model embeddings pre-trained using \textit{Transformer} \cite{vaswani2017attention}, like BERT \cite{devlin2018bert}, can not only be used to replace traditional embeddings as input distributed representations but also be fine-tuned for the NER task with one additional output layer while achieving state-of-the-art performances.
-The survey concludes that fine-tuning general-purpose contextualized language models with domain-specific data is very likely to give good performances for use cases with domain-specific texts and few training data.
+It shows that language model embeddings pretrained using \textit{Transformer} \cite{vaswani2017attention}, like BERT \cite{devlin2018bert}, can not only be used to replace traditional embeddings as input distributed representations but also be fine-tuned for the NER task with one additional output layer while achieving state-of-the-art performances.
+The survey concludes that fine-tuning general-purpose contextualized language models with domain-specific data is very likely to give good performance for use cases with domain-specific texts and few training data.
 
-\subsection{Named Entity Recognition on Historical Texts}
+\subsection{Named Entity Recognition in Historical Texts}
 
-\cite{Labusch2020NamedED} apply a BERT-based NER model on historical texts in German, French and English, produced by OCR.
-It is an of-the-shelf multilingual BERT-based NER model, pre-trained on more than 2 000 000 pages of OCRed historical texts, in the three languages and fine-tuned with historical texts NER ground truth, also in the three languages.
+\cite{Labusch2020NamedED} applied a BERT-based NER model to historical texts in German, French, and English, produced by OCR.
+It is off-the-shelf multilingual BERT-based NER model, pretrained on more than 2 000 000 pages of OCRed historical texts in three languages and fine-tuned on historical texts NER ground truth, also in three languages.
 Eventually, a clear decrease in NER performance on OCRised texts is noted.
@@
 
-Recently, several studies have focused on the extent to which the quality of the OCR affects the results produced by a NER model.\cite{van2020assessing} assess the impact of OCR on several NLP downstream tasks, including NER.
-They worked on a corpus published by a post-OCR correction software company, made of many journal articles with different levels of OCR errors and their respective ground-truths.
+Several recent studies have focused on the extent to which the quality of the OCR affects the results produced by a NER model.\cite{van2020assessing} assess the impact of OCR on several NLP downstream tasks, including NER.
+They worked on a corpus published by a post-OCR correction software company, made of many journal articles with different levels of OCR errors and their respective groundtruths.
 For each OCRed article, the Word Error Rate (WER) is computed and the English model \textit{en-core-web-lg} provided by Spacy\footnote{\url{https://spacy.io/}} library is used to perform NER on \textit{Person}, \textit{GPE} and \textit{Date}.
@@
 They applied four different types of OCR noise to each corpus, with two levels of degradation and computed the WER and Character Error Rate (CER) for each degraded version.
-Finally, they applied each NER model on the progressively degraded versions of the corpora and computed the resulting F-measure.
+Finally, they applied each NER model to the progressively degraded versions of the corpus and computed the resulting F-measure.
 Overall, NER F-mesure drops from 90\% to 50\% when the WER increase from 8\% to 50\%.
@@
 
-Based on the observation that OCR errors necessarily affect the performance of NER approaches applied to historical texts, \cite{huynh2020use} and \cite{marz2021data} have proposed different approaches to reduce their negative impact.
+Based on the observation that OCR errors necessarily affect the performance of NER models applied to historical texts, \cite{huynh2020use} and \cite{marz2021data} have proposed different approaches to reduce their negative impact.
+The former uses a spelling correction tool on several corpora with variable OCR error rates, to assess whether spelling correction has a beneficial impact on NER performance or not.
+The results showed that as long as OCR errors remain low (CER<2\% and WER<10\%), the F-measure of NER results remains stable.
+It starts to decrease significantly when OCR errors exceed these thresholds.
+The latter work focuses on adapting the training data to facilitate the generalisation of an off-the-shelf NER model from modern texts to historical texts.
+Three different NER models are tested on three historical corpora, in French, English, and Dutch.
+The best results are produced by a model trained on clean modern data, including many  embeddings as input (FastText, character and multilingual Flair embeddings, BERT and historical embeddings computed with Flair on a historical corpus), and fine-tuned on a noisy historical groundtruth.
+
+In conclusion, NER approaches based on deep learning seem to be more suitable for dealing with historical texts as they adapt more easily to OCR errors than rule-based approaches.
+Recent work in this area suggests that the impact of OCR quality on NER can be reduced by using different strategies.
+On the one hand, if the OCR error rate is kept below a certain threshold, the NER models remain little impacted: it is therefore important to reduce OCR errors as much as possible, but a low error rate may remain acceptable.
+On the other hand, it seems beneficial to reuse a NER model trained on modern data and adapt it to historical texts using supervised or unsupervised approaches.

### 2022-01-10 src/ocr/isri_tools_src/Modules/accrpt.h
Add accuracy tools

@@ -78,12 +78,12 @@ struct
 			/* enumeration for each character code */
 } Accdata;
 
-void add_class(/* Accdata *accdata, Charvalue value, long count,
-                  long missed */);
+void add_class(Accdata *accdata, Charvalue value, long count,
+                  long missed);
 			/* adds the given character value to "accdata",
                            updating all relevant class enumerations */
 
-void add_conf(/* Accdata *accdata, char *key, long errors, long marked */);
+void add_conf(Accdata *accdata, char *key, long errors, long marked);
 			/* adds the given confusion to "accdata"; "key" contains
 			   the character string representation of the confusion
 			   that will appear in the accuracy report (including
@@ -96,7 +96,7 @@ void read_accrpt(/* Accdata *accdata, char *filename */);
 			   and quits if unable to open the file, or if the file
 			   does not contain an accuracy report */
 
-void write_accrpt(/* Accdata *accdata, char *filename */);
+void write_accrpt(Accdata *accdata, char *filename);
 			/* writes the contents of "accdata" to the named file
 			   (or stdout if "filename" is NULL); reports an error
 			   and quits if unable to create the file */

### 2022-01-10 src/ocr/isri_tools_src/Modules/text.h

@@ -87,8 +87,8 @@ void read_text(/* Text *text, char *filename, Textopt *textopt */);
 
 #define STRING_SIZE  8
 
-void char_to_string(/* Boolean suspect, Charvalue value, char *string,
-		       Boolean fake_newline */);
+void char_to_string(Boolean suspect, Charvalue value, char *string,
+		       Boolean fake_newline);
 			/* stores a representation of the given character in
 			   "string", which must be at least STRING_SIZE bytes; a
 			   non-printable character is represented by a hex value

### 2022-01-10 src/ocr/isri_tools_src/accuracy.cpp

@@ -1,0 +1,190 @@
+
+/**********************************************************************
+ *
+ *  accuracy.c
+ *
+ *  Author: Stephen V. Rice
+ *
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+extern "C"
+{
+#include "Modules/accrpt.h"
+#include "Modules/sync.h"
+}
+
+#include <pybind11/iostream.h>
+#include <pybind11/pybind11.h>
+#include <string>
+#include <fstream>
+
+namespace py = pybind11;
+
+namespace
+{
+constexpr int MAX_DISPLAY = 24;
+
+void
+make_key (Text *texts, char *key, Sync *sync)
+{
+  long i, j;
+  char buffer[2][MAX_DISPLAY + 4], string[STRING_SIZE];
+  for (i = 0; i < 2; i++)
+    {
+      buffer[i][0] = '\0';
+      for (j = sync->substr[i].start; j <= sync->substr[i].stop; j++)
+        {
+          char_to_string (False, texts[i].array[j]->value, string, True);
+          if (strlen (buffer[i]) + strlen (string) > MAX_DISPLAY)
+            {
+              strcat (buffer[i], "...");
+              break;
+            }
+          strcat (buffer[i], string);
+        }
+    }
+  sprintf (key, "{%s}-{%s}\n", buffer[0], buffer[1]);
+}
+/**********************************************************************/
+
+void
+add_ops (Accops *sum_ops, Accops *ops)
+{
+  sum_ops->ins += ops->ins;
+  sum_ops->subst += ops->subst;
+  sum_ops->del += ops->del;
+  sum_ops->errors += ops->errors;
+}
+
+void
+process_synclist (Text *texts, Synclist *synclist, Accdata *accdata)
+{
+  Sync *sync;
[…]

### 2022-01-10 src/ocr/isri_tools_src/align.cpp

@@ -5,8 +5,10 @@ extern "C"
 }
 
 #include <string>
+#include <vector>
 #include <pybind11/pybind11.h>
 #include <pybind11/stl.h>
+
 
 namespace
 {
@@ -44,9 +46,6 @@ namespace
     return i;
   }
 
-}
-
-
 int compute_align_map(std::wstring a, std::wstring b, std::vector<int>& map_A, std::vector<int>& map_B)
 {
   Text A;
@@ -73,11 +72,12 @@ int compute_align_map(std::wstring a, std::wstring b, std::vector<int>& map_A, s
 
   return alignment_size;
 }
+}
 
-std::pair<std::vector<int>, std::vector<int>> align_2(std::wstring a, std::wstring b)
+std::pair<std::vector<int>, std::vector<int>> get_align_map(std::wstring a, std::wstring b)
 {
   std::vector<int> map_A, map_B;
-  int alignment_size = compute_align_map(a, b, map_A, map_B);
+  compute_align_map(a, b, map_A, map_B);
 
   return std::make_pair(map_A, map_B);
 }
@@ -101,11 +101,10 @@ std::pair<std::wstring, std::wstring> align(std::wstring a, std::wstring b, wcha
   return std::make_pair(A, B);
 }
 
-
-
-
-PYBIND11_MODULE(isri_tools, m) {
-    m.doc() = "ISRI Analytic tools"; // optional module docstring
-    m.def("align", &align, "Align two strings");
-    m.def("get_align_map", &align_2, "Get the mapping between two strings");
+void init_align(pybind11::module* m)
+{
+  m->def("align", &align, "Align two strings");
+  m->def("get_align_map", &get_align_map, "Get the mapping between two strings");
 }
+
+

### 2022-01-10 src/ocr/setup.py

@@ -3,6 +3,8 @@ from pybind11.setup_helpers import Pybind11Extension
 
 sourcefiles = [
     "isri_tools_src/align.cpp",
+    "isri_tools_src/accuracy.cpp",
+    "isri_tools_src/main.cpp",
     "isri_tools_src/Modules/accrpt.c",
     "isri_tools_src/Modules/charclass.c",
     "isri_tools_src/Modules/ci.c",
@@ -19,6 +21,8 @@ sourcefiles = [
     "isri_tools_src/Modules/wacrpt.c",
     "isri_tools_src/Modules/word.c",
     ]
+
+
 
 ext = Pybind11Extension("isri_tools", sourcefiles)
 

### 2022-01-11 src/ocr/text_utils.py
wip new code

@@ -1,0 +1,516 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+
+# ==============================================================================
+# Imports
+from collections import Counter
+import io
+import logging
+import argparse
+import sys
+import unicodedata
+from warnings import warn
+from typing import List, Tuple
+
+import regex
+import numpy as np
+import isri_tools
+
+# ==============================================================================
+# Logging
+logger = logging.getLogger(__name__)
+
+# ==============================================================================
+# Annotation codes projection to Unicode chars
+# Expected input: raw OCR human transcription and annotation (special glyphs w/o public codepoints)
+# Output: OCR reference text with projected annotations (before normalization)
+# Transform type: (multiple chars) -> (1+ chars), can change string length
+
+# From https://hackmd.io/ocGrovCAQZuKAoGs1zTYIg#Liste-des-glyphes-et-de-leur-repr%C3%A9sentation-en-caract%C3%A8res
+ANNOTATION_CODES = [
+    # We use unicode private user area when needed (no dedicated unicode char)
+    ("::LH::", "\uE001"),  # PUA #1 
+    ("::CJ::", "\uE002"),  # PUA #2 
+    ("::CM::", "\uE003"),  # PUA #3
+    ("::AC::", "\uE004"),  # PUA #4
+    ("::PA::", "\uE005"),  # PUA #5
+    ("::CP::", "\uE006"),  # PUA #6
+    ("::CT::", "\uE007"),  # PUA #7
+    ("::CA::", "\uE008"),  # PUA #8
+    ("::MM::", "\uE009"),  # PUA #9
+    ("::MO::", "\u24c4"),  # → "Ⓞ"
+    ("::MA::", "\u24b6"),  # → "Ⓐ"
+    ("::MB::", "\u24b7"),  # → "Ⓑ"
+    ("::MP::", "\u24c5"),  # → "Ⓟ"
+    ("::MV::", "\u24cb"),  # → "Ⓥ"
+    # ("M. H.", ""),
+    # ("C. F.", ""),
+    # ("R. du J. C.", ""),
+    # ("B. I.", ""),
+    # ("Cit.", ""),
+    ("::LG::", "\uE00A"),  # PUA #10
+    ("::LP::", "\uE00B"),  # PUA #11
[…]

### 2022-01-11 src/ocr/text_utils.py
wip rewrite of text preparation tools

@@ -6,15 +6,13 @@
 from collections import Counter
 import io
 import logging
-import argparse
-import sys
+from typing import List, Tuple
 import unicodedata
 from warnings import warn
-from typing import List, Tuple
 
-import regex
-import numpy as np
 import isri_tools
+import numpy as np
+import regex
 
 # ==============================================================================
 # Logging
@@ -57,9 +55,9 @@ ANNOTATION_CODES = [
     ("::UD::", "\uE010"),  # PUA #16
     ("::main::", "\u261E"),  # → "☞"
     ("\U0001F449", "\u261E"),  # "👉" → "☞"  # maybe some extra annotations like this to fix
-    ("::SymbET::", "\u002A"),  # → "*"
+    ("::SymbET::", "\u002A"),  # → "*" (note that \u002A can be represented with 5 or 6 branches in Unicode)
     ("\u2605", "\u002A"), # "★" → "*"  # maybe some extra annotations like this to fix
-    ("::SymbET::", "\u002A"),  # → "*"
+    ("::SymbET::", "\u002A"),  # → "*" (note that \u002A can be represented with 5 or 6 branches in Unicode)
     ("::Tvert::", "\u007C"),  # → "|"
     ("::Thoriz::", "\u002D"),  # → "-"
     ("::SymbM::", "\uE011"),  # PUA #17
@@ -79,109 +77,179 @@ def replace_annotation_codes(text: str) -> str:
 # Unicode charset simplification (mainly for XML compatibility)
 # Expected input: OCR reference text with projected annotations
 # Output: OCR reference text with projected annotations and simplified charset
-# Transform type: (1+ chars) -> (0+ chars), can change string length
+# General transform: (1+ chars) -> (1+ chars) (no deletion for debug purpose)
 
-UNICODE_SIMPLIFICATIONS = [
+# Transform type: (1 char) -> (1+ char), can change string length
[…]

### 2022-01-11 src/ocr/text_utils_tests.py

@@ -1,0 +1,13 @@
+
+from text_utils import UNICODE_SIMPLIFICATIONS_SINGLE_CHAR, simplify_unicode_charset, check_alignment_charset
+
+test_str = ("\U0001F449,AB\u000D\u000Aab "
+     + "".join(chr(x) for x in UNICODE_SIMPLIFICATIONS_SINGLE_CHAR.keys()) 
+     + "\u0001\u0000-\u0009\u000B -\u001F\u007F \u0080-\u009F \uFE00-\uFE0F "
+     + " \uD800-\uDFFF \uFDD0-\uFDEF\uFFFE\uFFFF 123")
+    
+def check_alignment_charset_test():
+    assert(not check_alignment_charset(test_str))
+
+def simplify_unicode_charset_test():
+    assert(check_alignment_charset(simplify_unicode_charset(test_str)))

### 2022-01-11 src/ner/create_datasets.py
Sanity checks: dataset sizes

@@ -1,5 +1,4 @@
 import csv
-import os
 import numpy as np
 import config as cfg
 from sklearn.model_selection import GroupKFold
@@ -20,6 +19,37 @@ def unwrap(list_of_tuples2):
 
 # endregion
 
+# =============================================================================
+# region ~ Sanity checks
+
+
+def check_experiment1_dev_test_sizes(dev, test):
+    actual = len(dev)
+    expected = 668
+    assert actual == expected, "Experiment 1: expected dev set of size %d, got %d." % (
+        actual,
+        expected,
+    )
+    actual = len(test)
+    expected = 1719
+    assert actual == expected, "Experiment 1: expected test set of size %d, got %d." % (
+        actual,
+        expected,
+    )
+
+
+def check_experiment1_training_subsets_sizes(trainsets):
+    actual = len(trainsets)
+    assert actual == 8, "Experiment 1: expected 8 sub-training sets bu got %d" % actual
+    expected = [6012, 3006, 1503, 751, 375, 187, 93, 46]
+    for ix, actual in enumerate(trainsets):
+        actual_len = len(actual)
+        assert (
+            actual_len == expected[ix]
+        ), "Expected sub-training set of size %d, got %d" % (expected[ix], actual_len)
+
+
+# endregion
 
 # =============================================================================
 # region ~ Main processing
@@ -44,8 +74,9 @@ def main():
         random_state=cfg.SEED,
         stratify=groups,
     )
-    logger.info("TRAIN: %s, DEV: %s" % (len(train) ,len(dev)))
-    zipped = zip([train, dev],  ["train", "dev"])
+
+    logger.info("TRAIN: %s, DEV: %s" % (len(train), len(dev)))
+    zipped = zip([train, dev], ["train", "dev"])
     odir = cfg.DATA_DIR / "datasets"
     export(odir, zipped)
 
@@ -53,10 +84,14 @@ def main():
     # Create datasets for experiment 1
     ##
     odir = cfg.DATA_DIR / "experiment_1"
-    
+
     train, dev, test = make_train_dev_test(gold_dataset)
     exp1_trainsets = create_experiment_1_trainsets(train)
-    
+
+    # Sanity checks
+    check_experiment1_dev_test_sizes(dev, test)
+    check_experiment1_training_subsets_sizes(exp1_trainsets)
+
     for subtrain in exp1_trainsets:
[…]

### 2022-01-11 src/ocr/test_text_utils.py
new normalization and alignment functions

@@ -1,0 +1,39 @@
+
+from text_utils import (
+    UNICODE_SIMPLIFICATIONS_SINGLE_CHAR, 
+    simplify_unicode_charset,
+    check_alignment_charset,
+    add_tags_prediction)
+
+test_str = ("\U0001F449,AB\u000D\u000Aab "
+     + "".join(chr(x) for x in UNICODE_SIMPLIFICATIONS_SINGLE_CHAR.keys()) 
+     + "\u0001\u0000-\u0009\u000B -\u001F\u007F \u0080-\u009F \uFE00-\uFE0F "
+     + " \uD800-\uDFFF \uFDD0-\uFDEF\uFFFE\uFFFF 123")
+    
+def test_check_alignment_charset():
+    assert(not check_alignment_charset(test_str))
+
+def test_simplify_unicode_charset():
+    assert(check_alignment_charset(simplify_unicode_charset(test_str)))
+
+def test_str_rstrip_removes_all_spaces():
+    s = "abc\u0009 \u00A0 \u2000 \u2001 \u2002 \u2003 \u2004 \u2005 \u2006 \u2007 \u2008 \u2009 \u200A \u202F \u205F"
+    assert(s.rstrip() == "abc")
+
+def test_str_rstrip_removes_all_newlines():
+    s = "abc \u000A \u000B \u000C \u000D \u0085 \u2028 \u2029"
+    assert(s.rstrip() == "abc")
+
+def test_align_ner_1():
+    A = "<PER>Anthony</PER> \uE001, <FT>fab.</FT> du <ACT>pêche</ACT>, <LOC>Châtelet</LOC>xx"
+    B = "\u261E Mme Antoine, \uE00A fab la p&cheur et \ndu ch>telet et du Faub. St Antoine"
+    R, no_empty_tag = add_tags_prediction(A, B)
+    assert(no_empty_tag)
+    assert(R == "\u261E Mme <PER>Antoine</PER>, \uE00A <FT>fab</FT> la <ACT>p&amp;che</ACT>ur et \u2029du <LOC>ch&gt;telet</LOC> et du Faub. St Antoine")
+
+def test_align_ner_1():
+    A = "<PER>Anthony</PER> \uE001, <FT>fab</FT><FT>.</FT> du <ACT>pêche</ACT>, <LOC>Châtelet</LOC>xx"
[…]

### 2022-01-11 src/ocr/text_utils.py

@@ -18,6 +18,13 @@ import regex
 # Logging
 logger = logging.getLogger(__name__)
 
+
+
+# ==============================================================================
+# Some constants
+TAG_LIST = ["PER", "LOC", "ACT", "CARDINAL", "FT", "TITRE"]
+
+
 # ==============================================================================
 # Annotation codes projection to Unicode chars
 # Expected input: raw OCR human transcription and annotation (special glyphs w/o public codepoints)
@@ -75,6 +82,7 @@ def replace_annotation_codes(text: str) -> str:
 
 # ==============================================================================
 # Unicode charset simplification (mainly for XML compatibility)
+# https://www.w3.org/TR/unicode-xml/#Suitable
 # Expected input: OCR reference text with projected annotations
 # Output: OCR reference text with projected annotations and simplified charset
 # General transform: (1+ chars) -> (1+ chars) (no deletion for debug purpose)
@@ -206,7 +214,7 @@ def simplify_unicode_charset(text: str) -> str:
     # We do not use 'NFKC' because the substitutions may not be the ones we want.
     res = unicodedata.normalize('NFC', text)
 
-    # Replace CRLR to LF
+    # Replace CRLR to LF (must be first)
     res = res.replace("\u000D\u000A", "\u000A")
 
     # Apply all our 1-char substitutions efficiently
@@ -335,95 +343,124 @@ def check_alignment_charset(text_to_align: str, dump_charset: bool=False, show_a
 
 
 # ==============================================================================
-# Fix manual NER XML
+# Fix manual NER XML (raw NER reference)
+# Expected input: NER "XML" produced with annotation tool (contains annotation, 
+# un-normalized charset and lacks XML escaping)
+# Output: Clean NER XML to be used as reference (/!\ newlines encoded as PS)
+# Transformation: can add, remove and change chars
 
-# TODO
+def fix_manual_ner_xml(ner_xml_orig: str) -> str:
+    # Get a clean line
+    fixed = remove_unicode_bom(ner_xml_orig)
+    fixed = fixed.rstrip()  # removes any training whitespace and newline char
+
+    # Sanity check
[…]

### 2022-01-11 src/ocr/text_utils_tests.py

@@ -1,13 +1,0 @@
-
-from text_utils import UNICODE_SIMPLIFICATIONS_SINGLE_CHAR, simplify_unicode_charset, check_alignment_charset
-
-test_str = ("\U0001F449,AB\u000D\u000Aab "
-     + "".join(chr(x) for x in UNICODE_SIMPLIFICATIONS_SINGLE_CHAR.keys()) 
-     + "\u0001\u0000-\u0009\u000B -\u001F\u007F \u0080-\u009F \uFE00-\uFE0F "
-     + " \uD800-\uDFFF \uFDD0-\uFDEF\uFFFE\uFFFF 123")
-    
-def check_alignment_charset_test():
-    assert(not check_alignment_charset(test_str))
-
-def simplify_unicode_charset_test():
-    assert(check_alignment_charset(simplify_unicode_charset(test_str)))

### 2022-01-12 src/ocr/text_utils.py
generate pretraining set

@@ -138,7 +138,7 @@ UNICODE_SIMPLIFICATIONS_SINGLE_CHAR = str.maketrans({
     "\u2011": "\u002D", #  NON-BREAKING HYPHEN
     "\u2012": "\u002D", # ‒ FIGURE DASH
     "\u2013": "\u002D", # – EN DASH
-    "\u2014": "\u002D", # — EM DASH
+    # "\u2014": "\u002D", # — EM DASH  # /!\ kept because we have special chars like this
     "\u2015": "\u002D", # ― HORIZONTAL BAR
     "\u2E3A": "\u002D", # ⸺  two-em dash
     "\u2E3B": "\u002D", # ⸻ THREE-EM DASH
@@ -296,7 +296,7 @@ def check_alignment_charset(text_to_align: str, dump_charset: bool=False, show_a
             or 0xD800 <= codepoint <= 0xDFFF  # surrogates
             or codepoint > 0xFFFF):  # large codepoints
             errors += 1
-            print(f"Invalid char@{pos:03d}: {hex(ord(char))} ({_unichr2str(char)}"
+            print(f"Invalid char@{pos:03d}:   ({hex(ord(char))} {_unichr2str(char)}"
                   f" -- cat.: {unicodedata.category(char)})")
         elif not (
             codepoint == ord("\n")
@@ -304,6 +304,7 @@ def check_alignment_charset(text_to_align: str, dump_charset: bool=False, show_a
             or 0x00A0 <= codepoint <= 0x00FF  # Latin-1_Supplement
             or 0x0100 <= codepoint <= 0x017F  # Latin Extended-A
             or 0x0200 <= codepoint <= 0x026F  # General Punctuation
+            or codepoint == 0x2014  # — EM DASH
             or codepoint in (0x24b6, 0x24b7, 0x24c4, 0x24c5, 0x24cb) # Ⓐ, Ⓑ, Ⓞ, Ⓟ, Ⓥ: Enclosed Alphanumerics (2460–24FF)
             or codepoint in (0x261E, )  # 0x261E:☞ (0x2605: ★): Miscellaneous Symbols (2600–26FF)
             or codepoint in (0x2709, )  # ✉: Dingbats (2700–27BF)
[…]

### 2022-01-12 src/ocr/text_utils.py
more charset adjustments

@@ -305,6 +305,7 @@ def check_alignment_charset(text_to_align: str, dump_charset: bool=False, show_a
             or 0x0100 <= codepoint <= 0x017F  # Latin Extended-A
             or 0x0200 <= codepoint <= 0x026F  # General Punctuation
             or codepoint == 0x2014  # — EM DASH
+            or codepoint == 0x2029 # Paragraph Separator
             or codepoint in (0x24b6, 0x24b7, 0x24c4, 0x24c5, 0x24cb) # Ⓐ, Ⓑ, Ⓞ, Ⓟ, Ⓥ: Enclosed Alphanumerics (2460–24FF)
             or codepoint in (0x261E, )  # 0x261E:☞ (0x2605: ★): Miscellaneous Symbols (2600–26FF)
             or codepoint in (0x2709, )  # ✉: Dingbats (2700–27BF)
@@ -346,7 +347,7 @@ def check_alignment_charset(text_to_align: str, dump_charset: bool=False, show_a
 # ==============================================================================
 # Fix manual NER XML (raw NER reference)
 # Expected input: NER "XML" produced with annotation tool (contains annotation, 
-# un-normalized charset and lacks XML escaping)
+# un-normalized charset but MAY contain XML escaping)
 # Output: Clean NER XML to be used as reference (/!\ newlines encoded as PS)
 # Transformation: can add, remove and change chars
 
@@ -363,7 +364,8 @@ def fix_manual_ner_xml(ner_xml_orig: str) -> str:
     chunks, tags = chop_on_tags(fixed, tag_list=TAG_LIST)
     assert(len(chunks) == len(tags) + 1)
 
-    # Note: We don't unescape XML entities here as the GUI tools doesn't do it.
+    # Unescape XML entities if we have some.
+    chunks = list(map(xml_unescape, chunks))
 
     # Replace annotation codes (e.g. "::LH::") with (private if needed) unicode chars
     chunks = list(map(replace_annotation_codes, chunks))

### 2022-01-12 src/ocr/text_utils.py
improve normalization tools

@@ -783,8 +783,8 @@ def remove_unicode_bom(text: str) -> str:
 
 # 8<----------------------------------
 
-# def remove_xml_tags_and_entities(xml: str) -> str:
-#     # We reuse existing functions here
-#     chunks, _tags = chop_on_tags(xml, tag_list=TAG_LIST)
-#     chunks = list(map(xml_unescape, chunks))
-#     return "".join(chunks)
+def remove_xml_tags_and_entities(xml: str) -> str:
+    # We reuse existing functions here
+    chunks, _tags = chop_on_tags(xml, tag_list=TAG_LIST)
+    chunks = list(map(xml_unescape, chunks))
+    return "".join(chunks)

### 2022-01-12 src/ocr/OCRMetrics.py
Update OCR metrics

@@ -1,0 +1,81 @@
+'''
+Module for Text Alignment and OCR Metrics
+'''
+
+
+
+import isri_tools
+from typing import List, Union
+from dataclasses import dataclass
+
+@dataclass
+class AlignmentStats:
+    """Provide statistics about an aligment
+    """
+    errors: int = 0
+    matches: int = 0
+    reference_length: int = 0
+    x_length: int = 0
+
+    @property
+    def length(self):
+        return self.errors + self.matches
+
+    @property
+    def precision(self) -> float:
+        """Get the alignment recall (ratio match/reference length)
+
+        Returns:
+            float: Precision score
+        """
+        return self.matches / self.x_length
+    
+    @property
+    def recall(self) -> float:
+        '''
+        Get the alignment recall (ratio match/reference length)
+
+        Returns:
+            float: Recall score
+        '''
+        return self.matches / self.reference_length
+
+    @property
+    def accuracy(self) -> float:
+        '''
+        Get the alignment accuracy (number of match over the alignment length)
+        '''
+        return self.matches / (self.matches + self.errors)
+
+
+@dataclass
+class AlignmentReport:
+    '''
+    Provide an alignment report between a reference string and another string
+    '''
+    reference: str
+    x: str
+    aligned_ref: str
+    aligned_x: str
+
+    def __init__(self, reference: str, x: str) -> None:
+        self.reference = reference
+        self.x = x
+        self.aligned_ref, self.aligned_x = isri_tools.align(reference, x, "✗")
+
+        match = 0
+        for a,b in zip(self.aligned_ref, self.aligned_x):
+            match += int(a == b)
+
+        n = len(self.aligned_ref)
+        self._stats = AlignmentStats(matches=match,
+                                    errors=n-match,
+                                    x_length=len(x),
+                                    reference_length=len(reference))
+
+    @property
+    def stats(self) -> AlignmentStats:
+        return self._stats
+
+def align(reference: str, x: str) -> AlignmentReport:
+    return AlignmentReport(reference, x)

### 2022-01-12 src/ocr/isri_tools_src/align.cpp

@@ -107,4 +107,3 @@ void init_align(pybind11::module* m)
   m->def("get_align_map", &get_align_map, "Get the mapping between two strings");
 }
 
-

### 2022-01-12 src/ner/camembert.py
Adapt to ocr-pipeline dataset storage

@@ -98,7 +98,13 @@ def parse_args():
     # parser.add_argument(
     #     "--train_id", type=str, help="Trainset identifier for exp. 1.", required=False
     # )
-    parser.add_argument("--model", type=str, help=f"Name or path to a model. Default is {CAMEMBERT_MODEL}", required=False, default=CAMEMBERT_MODEL)
+    parser.add_argument(
+        "--model",
+        type=str,
+        help=f"Name or path to a model. Default is {CAMEMBERT_MODEL}",
+        required=False,
+        default=CAMEMBERT_MODEL,
+    )
     return parser.parse_args()
 
 

### 2022-01-12 src/ner/config.py

@@ -1,5 +1,9 @@
 """
- Holds the global configuration used in all python scripts related to NER.
+ Holds the global configuration.
+Remarks:
+- DATA_DIR should be the absolute path to paper-ner-bench-das22/dataset
+- WORKDIR is set to the location of this scipt.
+- generated train/dev/test datasets and metrics will be outputed in WORKDIR/computed
 """
 
 import os
@@ -9,14 +13,28 @@ from pathlib import Path
 # =============================================================================
 # region ~ Configurations
 
-DATA_DIR = Path(os.environ["DATADIR"])
-
-assert DATA_DIR.exists()
-
-GOLD = DATA_DIR / "gold/gold.csv"
-SEED = 42  # Random seed to generate train/validation/test datasets
-MIN_TRAINSET_SIZE = 30
 DEBUG = os.getenv("DEBUG", "False").lower() in ("true", "1", "t")
+
+# Paths globals
+DATA_DIR = Path(os.getenv("DATADIR"))
+assert DATA_DIR.exists(), "DATADIR %s does not exist." % DATA_DIR
+
+WORKDIR_DEFAULT = os.path.dirname(__file__)
+WORKDIR = Path(os.getenv("WORKDIR", WORKDIR_DEFAULT))
+assert WORKDIR.exists(), "Workdir: %s is not a valid directory" % WORKDIR
+
+# Training globals
+GOLD = DATA_DIR / "supervised/10-ref-ocr-ner-json/gold.csv"
+SEED = 42  # Ensure the train/dev split to be repeatable
+
+# Experiment 1
+EXPERIMENT_1 = WORKDIR / "experiment_1"
+MIN_TRAINSET_SIZE = 30
+
+# Experiment 2
+EXPERIMENT_2 = WORKDIR / "experiment_2"
+
+
 # endregion
 
 

### 2022-01-12 src/ner/create_datasets.py

@@ -23,30 +23,46 @@ def unwrap(list_of_tuples2):
 # region ~ Sanity checks
 
 
+def check(expected, actual, msg=""):
+    assert actual == expected, msg
+
+
 def check_experiment1_dev_test_sizes(dev, test):
     actual = len(dev)
-    expected = 668
-    assert actual == expected, "Experiment 1: expected dev set of size %d, got %d." % (
-        actual,
+    expected = 709
+    check(
         expected,
+        actual,
+        f"Experiment 1: expected dev set of size {expected}, got {actual}.",
     )
+
     actual = len(test)
-    expected = 1719
-    assert actual == expected, "Experiment 1: expected test set of size %d, got %d." % (
-        actual,
+    expected = 1690
+    check(
         expected,
+        actual,
+        f"Experiment 1: expected dev set of size {expected}, got {actual}.",
     )
 
 
 def check_experiment1_training_subsets_sizes(trainsets):
     actual = len(trainsets)
-    assert actual == 8, "Experiment 1: expected 8 sub-training sets bu got %d" % actual
-    expected = [6012, 3006, 1503, 751, 375, 187, 93, 46]
+    expected = 8
+    check(
+        expected,
+        actual,
+        f"Experiment 1: expected {expected} sub-training sets but got {actual}",
+    )
+
+    expected = [6373, 3186, 1593, 796, 398, 199, 99, 49, 24]
     for ix, actual in enumerate(trainsets):
         actual_len = len(actual)
-        assert (
-            actual_len == expected[ix]
-        ), "Expected sub-training set of size %d, got %d" % (expected[ix], actual_len)
+        expected_len = expected[ix]
+        check(
+            expected_len,
+            actual_len,
+            f"Expected sub-training set of size {expected_len} but got {actual_len}",
+        )
 
 
 # endregion
@@ -75,15 +91,15 @@ def main():
         stratify=groups,
     )
 
-    logger.info("TRAIN: %s, DEV: %s" % (len(train), len(dev)))
-    zipped = zip([train, dev], ["train", "dev"])
-    odir = cfg.DATA_DIR / "datasets"
-    export(odir, zipped)
+    logger.info("Full train: %s, Full dev: %s" % (len(train), len(dev)))
+    # zipped = zip([train, dev], ["train", "dev"])
+    # odir = cfg.DATA_DIR / "datasets"
+    # export(odir, zipped)
 
     ##
[…]

### 2022-01-12 src/ner/experiment_1.sh

@@ -3,11 +3,11 @@
 # ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 # This script trains the CNN, BERT & BERT+PRE models on each subsampled 
 # training set created using create_datasets.py and stored
-# in $DATADIR/experiment_1/datasets
+# in $DATADIR/EXP1eriment_1/datasets
 # 
 # The validation and test sets remain untouched.
 # Precision, recall and F1 scores are measured against the test set
-# and stored in $DATADIR/experiment_1/metrics 
+# and stored in $DATADIR/EXP1eriment_1/metrics 
 # ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
 # ==============================================================================
@@ -15,11 +15,10 @@
 set -eu
 
 # Constants
-EXP1=$DATADIR/experiment_1
+WORKDIR=`dirname $0` # Set WORKDIR to be the script location
+EXP1=$WORKDIR/experiment_1
 METRICS=$EXP1/metrics
-MODEL_PRETRAINED=$DATADIR/pretraining/camembert_best
-
-WORKDIR=`dirname $0`
+MODEL_PRETRAINED="HueyNemud/berties-pretrained-das22"
 # ==============================================================================
 
 
@@ -63,20 +62,18 @@ function train_eval_cnn() {
 # ------------------------------------------------------------------------------
 function train_eval_camembert() {
     # Args: trainset sizes
-    dir=`dirname $0`
     for size in $@; do
         datasetdict="$EXP1/huggingface_$size"
-        python $dir/camembert.py $datasetdict "$METRICS/camembert_${size}"
+        python $WORKDIR/camembert.py $datasetdict "$METRICS/camembert_${size}"
     done
 }
 
 # ------------------------------------------------------------------------------
 function train_eval_camembert_pretrained() {
     # Args: trainset sizes
-    dir=`dirname $0`
     for size in $@; do
         datasetdict="$EXP1/huggingface_$size"
-        python $dir/camembert.py $datasetdict "$METRICS/camembert_pretrained_${size}" \
+        python $WORKDIR/camembert.py $datasetdict "$METRICS/camembert_pretrained_${size}" \
             --model $MODEL_PRETRAINED
     done
 }
[…]

### 2022-01-12 src/ner/init_experiments.sh

@@ -8,13 +8,12 @@
 # ==============================================================================
 # Let's get paranoid
 set -eu
-
-WORKDIR=`dirname $0`
+WORKDIR=`dirname $0` # Set WORKDIR to be the script location
+EXP1=$WORKDIR/experiment_1
 # ==============================================================================
 
 # Make directories
-mkdir -p $DATADIR/experiment_1
-mkdir -p $DATADIR/datasets
+mkdir -p $EXP1
 
 # Build datasets for all experiments
 echo "Initializing training, dev and test datasets"

### 2022-01-12 src/ocr/test_text_utils.py
remove empty and touching tags in NER gold

@@ -3,7 +3,10 @@ from text_utils import (
     UNICODE_SIMPLIFICATIONS_SINGLE_CHAR, 
     simplify_unicode_charset,
     check_alignment_charset,
-    add_tags_prediction)
+    add_tags_prediction,
+    xml_contains_empty_tags,
+    xml_remove_empty_tags,
+    fix_manual_ner_xml)
 
 test_str = ("\U0001F449,AB\u000D\u000Aab "
      + "".join(chr(x) for x in UNICODE_SIMPLIFICATIONS_SINGLE_CHAR.keys()) 
@@ -37,3 +40,59 @@ def test_align_ner_1():
     R, no_empty_tag = add_tags_prediction(A, B)
     assert(not no_empty_tag)
     assert(R == "\u261E Mme <PER>Antoine</PER>, \uE00A <FT>fab</FT><FT></FT> la <ACT>p&amp;che</ACT>ur et \u2029du <LOC>ch&gt;telet</LOC> et du Faub. St Antoine")
+
+def test_xml_contains_empty_tags():
+    XML = "<PER>Lebas (en gros)</PER>,<PER> </PER><LOC>R. et Div. de la Fraternité</LOC>, <CARDINAL>91</CARDINAL>."
+    assert(xml_contains_empty_tags(XML))
+
+    XML = "<PER>Lebas (en gros)</PER><PER> Bob</PER>,<LOC>R. et Div. de la Fraternité</LOC>, <CARDINAL>91</CARDINAL>."
+    assert(not xml_contains_empty_tags(XML))
+
+def test_xml_remove_empty_tags():
+    XML = "<PER>Lebas (en gros)</PER>,<PER>\u2029 </PER><LOC>R. et Div. de la Fraternité</LOC>, <CARDINAL>91</CARDINAL>."
+    XML2 = xml_remove_empty_tags(XML)
+    XML_EXP = "<PER>Lebas (en gros)</PER>,\u2029 <LOC>R. et Div. de la Fraternité</LOC>, <CARDINAL>91</CARDINAL>."
+    assert(XML2 == XML_EXP)
+
+    XML = "<PER>Lebas (en gros)</PER><PER> Bob</PER>,<LOC>R. et Div. de la Fraternité</LOC>, <CARDINAL>91</CARDINAL>."
+    XML2 = xml_remove_empty_tags(XML)
+    assert(XML2 == XML)
+
[…]

### 2022-01-12 src/ocr/text_utils.py

@@ -24,6 +24,17 @@ logger = logging.getLogger(__name__)
 # Some constants
 TAG_LIST = ["PER", "LOC", "ACT", "CARDINAL", "FT", "TITRE"]
 
+_newline_chars = "\u000A\u000B\u000C\u000D\u0085\u2028\u2029"
+_space_chars = "\u0020\u0009\u00A0\u2000\u2001\u2002\u2003\u2004\u2005\u2006\u2007\u2008\u2009\u200A\u202F\u205F"
+# _empty_chars = ""  # Not managed here
+
+REGEX_EMPTY_TAGS = (
+    r"<(\L<tag>)>(["
+         + _newline_chars 
+         + _space_chars 
+         + r"]*)</(?1)>")
+
+WHITESPACE_ONLY_STR_RE = "^[" + _newline_chars + _space_chars + "]*$"
 
 # ==============================================================================
 # Annotation codes projection to Unicode chars
@@ -257,6 +268,7 @@ def simplify_unicode_charset(text: str) -> str:
 
 # ==============================================================================
 # Check OCR reference charset
+# TODO extract codepoint ranges to some contant list
 def check_alignment_charset(text_to_align: str, dump_charset: bool=False, show_all: bool=False) -> bool:
     """Check the string passed as parameter contains only acceptable
     unicode code points for the alignment.
@@ -354,7 +366,7 @@ def check_alignment_charset(text_to_align: str, dump_charset: bool=False, show_a
 def fix_manual_ner_xml(ner_xml_orig: str) -> str:
     # Get a clean line
     fixed = remove_unicode_bom(ner_xml_orig)
-    fixed = fixed.rstrip()  # removes any training whitespace and newline char
+    fixed = fixed.strip() # remove leading and trailing whitespaces
 
     # Sanity check
     if not check_xml(fixed, taglist=TAG_LIST):
@@ -380,10 +392,44 @@ def fix_manual_ner_xml(ner_xml_orig: str) -> str:
     # Recreate string (recombine chunks and tags)
     final_list = []
     tags.append("")
+    has_empty_tags = False
     for c, t in zip(chunks, tags):
+        # If we have an empty chunk, remove the tags
+        removetag_empty = (t.startswith("</")  # on closing tag
+            and regex.match(WHITESPACE_ONLY_STR_RE, c, flags=regex.MULTILINE)  # empty content
+            and len(final_list) > 1  # not at the start of list (should always be true)
+            and len(final_list[-1]) > 2  # should always be true
[…]

### 2022-01-12 src-latex/parts/00-abstract.tex
Updates from Overleaf

@@
 % !TeX root = ../main-paper.tex
-The paper features a comparison of a selection (FIXME be more specific/enumerate) of named entity recognition (NER) techniques which can be reasonably used in the context of historical document analysis.
+This paper features a comparison of four (\bertrand{Ensure it's 4!} named entity recognition (NER) techniques which can be reasonably used in the context of historical document analysis.
 Such natural language processing techniques identify which class each text token belong to, like "person name", "location", "number" or "activity".

### 2022-01-12 src-latex/parts/30-dataset.tex
Updates from Overleaf

@@
 \section{Dataset}
+
+In this work, we  aim  at  producing  structured  spatio-temporal  data  from the information contained in the 19th century Parisian trade directories.
+To do this, it is therefore necessary to extract the text from the scans of the directories to be processed and to identify the entries in the various indexing lists used.
+Then, in each of these entries, the named entities they contain have to be extracted to produce structured spatio-temporal data representing the evolution over time of the people and businesses listed in these directories, their descriptions and their locations.
+Both of these extraction tasks require data to train and evaluate the OCR and NER approaches identified as potentially relevant: Tesseract and PERO OCR for the OCR task and Spacy, and CamemBERT - pre-trained or fine-tuned - neural models for the NER task.
+
+\subsection{The 19th century Parisian trade directories}
+
+\begin{table}[h!]
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{|l|l|l|l|l|}
+\hline
+\multicolumn{1}{|c|}{\textbf{Directory publisher or name}} & \multicolumn{1}{c|}{\textbf{Year}} & \multicolumn{1}{c|}{\textbf{\# Col}} & \multicolumn{1}{c|}{\textbf{Index}} & \multicolumn{1}{c|}{\textbf{Entry structure}}
+\\ \hline
+Favre \& Duchesne & 1798 & 1 & by occupation & {N [O], SN, SNUM — SEC.} 
+\\ \hline
+Duverneuil et La Tynna & 1801 & 1 & \begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & {N, SN, SNUM. SEC.} 
+\\ \hline
+{Notables communaux de la Seine} & 1801 & 1 & by name & {N, [O,] SN[, SNUM].}
+\\ \hline
+Duverneuil et La Tynna & 1805 &	2 &	\begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F)], SN, SNUM.
+— SEC.\\ N [(C)], [(O),] SN, SNUM.
+— SEC.\end{tabular}
+\\ \hline
+Duverneuil et La Tynna & 1806 & 2 & \begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F)], SN, SNUM.\\ N [(C)], SN, SNUM.[ — SEC.]\end{tabular}.
+\\ \hline
+La Tynna & 1813	& 2	& \begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N,[O,] SN[, SNUM].\\ N, [(T)] O, SN[, SNUM] [, D].\end{tabular}.
+\\ \hline
+Panckoucke Comm Dulac & 1820 & 1 & by name & N [(F)], O, [(T),] SN, SNUM[, P] .
+\\ \hline
+Panckoucke hab Dulac & 1820	& 1 & by name & N [(T)], [O,] SN, SNUM [, D].
+\\ \hline
+Bottin1 & 1820 & 2	& \begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F \textbar C)][T], [P], SN, SNUM.\\N [(F \textbar C)][T], O, SN, SNUM.
+CR\end{tabular}.
+\\ \hline
+Bottin1 & 1827 & 2	& \begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], O, SN, SNUM.\end{tabular}.
+\\ \hline
+Deflandre & 1828 & 2 & \begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F \textbar C)][T], [P], SN, SNUM.
+\\N [(F \textbar C)][T], O, SN, SNUM.\end{tabular}.
+\\ \hline
+Deflandre & 1829 & 2 & \begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & .
+\\ \hline
+Bottin1 & 1837 & 2	& \begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & .
+\\ \hline
+Cambon almgene & 1841 & 2 &	\begin{tabular}[c]{@{}l@{}}by occupation\\ by name\end{tabular} & .
+\\ \hline
+Didot & 1841a &	3 &	by name & .
+\\ \hline
+Didot & 1841a &	4 & by occupation & .
+\\ \hline
+Didot & 1851a & 4 & by name & .
+\\ \hline
+Didot & 1851a & 4 &	by occupation & .
+\\ \hline
+Didot & 1851a & 5 & by address & .
+\\ \hline
+Didot & 1854a & 3 &	by name & .
+\\ \hline
+Didot & 1854a & 4 & by name & .
+\\ \hline
+Didot & 1854a &	5 & by address & .
+\\ \hline
+Bottin3 & 1854a & 2 & by name & .
+\\ \hline
+Bottin3 & 1854a & 4 & by address & .
+\\ \hline
+Bottin3 & 1854a	& 3 & by occupation & .
+\\ \hline
+DidotBottin & 1860a	& 3	& by name & .
+\\ \hline
+DidotBottin & 1861a	& 3	& by name & .
+\\ \hline
+\end{tabular}%
+}
+\caption{The directories used to build our dataset.
+Entry structure is described as follows: N - entry name (person or business); F - first name; T - honorary title; C - civility; O - occupation; P - precision; SN - street name; SNUM - street number; SEC - section; D - district; CR - cross-reference}
+\label{tab:my-table}
+\end{table}
+
+\subsection{Groundtruth data extraction}
+
+\subsection{Assessing the groundtruth quality}
+
+Our dataset is thus made of excerpts
 
@@
     \item illustration (maybe in first figure)
-    \item various numbers (of original data and of the sample we extracted) : number of entries per pages, variety of directory entries, \dots
+    \item various numbers (of the original data and of the sample we extracted) : number of entries per pages, variety of directory entries, \dots
     \item statistics about entry lengths (histogram of number of char / tokens par entry)
@@
     \item OCR data (automated systems and clean text + how we created the "clean" text + whether/how we assessed clean text's quality)
-
+    \item Measure inter-annotator agreement ? 
+    \item Explain charset projection
     \item \dots

### 2022-01-12 src-latex/parts/60-results.tex
Updates from Overleaf

@@
 
-\subsection{Experiment 1: Accuracy vs NER approach vs training set size}
+\subsection{Experiment 1: NER sensibility to the number of training samples}
 
@@
 \toprule
+Trainset size (entries) & 49 & 99 & 199 & 398 & 796 & 1593 & 3186 & 6373\\
 Trainset size (\% of the full trainset) & 0.8 & 1.5 & 3.1 & 6.2 & 12.5 & 25 & 50 & 100\\
@@
 
-\subsection{Experiment 3: Accuracy vs NER approach vs OCR noise}
+\subsection{Experiment 2: NER in the presence of noisy OCR texts}
 
@@
 
-\begin{table}
+\begin{table}[h!]
 \caption{Camembert vs noise}
 \centering
-\begin{tabular}{ll|ll|l}
-                            &                      & \multicolumn{2}{l|}{Training data} &   \\
+\begin{tabular}{ll|cc|c}
+ & & \multicolumn{2}{c|}{Training data} & \\
  & & noisy & clean &   \\ 
 \cline{1-4}
-\multirow{2}{*}{Test data } & noisy                & f1    & f1                         &   \\
-                            & clean                & f1    & f1                         &   \\ 
+\multirow{3}{*}{Test data}& noisy gold (Tesseract) & f1 & f1 & \\
+                            & noisy gold (Pero-OCR) & f1 & f1 & \\ 
+                            & reference gold & f1 & f1 & \\ 
 \cline{1-4}

### 2022-01-12 src-latex/parts/10-introduction.tex

--- 2021-10-28 src-latex/parts/10-introduction.tex
@@
 % General context + scope limitation
-NER is a fundamental step of the information extraction pipeline when processing historical listing, like directories for instance.
-OCRed text is not sufficient to build a high level, semantic view of the collection under study;
-indeed, being able to properly tag text tokens unlocks the ability to relate entities,
-and provide colleagues from digital humanities with useful databases.
+The surge of historical documents digitized by cultural heritage institutions and made publicly available on the Web creates a vast knowledge source that can nourish numerous applications in the fields of social sciences, environmental studies or education.
+Most of these documents are texts, either printed or handwritten.
+To make the content of these documents searchable and usable by everyone, the texts engraved within them needs to be extracted through optical character recognition (OCR). 
+However, OCRed texts are not sufficient to build a high level, semantic view of a collection of documents.
+A second step thus aims at recognizing in these texts the pieces of information most likely to be searched for by users, such as named entities (people, organizations, dates, places, etc.).
+Indeed, being able to properly tag text tokens unlocks the ability to relate entities, and provide colleagues from e.g.
+digital humanities with databases ready for exploitation.
 
-% Problematic
-NER requires training data, and we have little training data for historical documents.
-OCR is also noisy.
-How can we extract reliable elements in this context?
+Being active research topics, OCR and named entity recognition (NER) are still difficult tasks when applied to historical text documents.
+OCR approaches used for modern documents are likely to struggle even on printed historical documents due the multiple factors related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and inconsistent page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
+\bertrand{J'ai tenté un premier refactoring, pas plus techniquement précis ceci dit.}
+\textcolor{green}{Phrases précedentes à reprendre avec les termes adéquats + ajouter une ref.}
+On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts and from the errors introduced into the text by the OCR step.\textcolor{green}{ajouter une ref}
 
+In this article we focus on a peculiar category of historical printed books: the city trade directories of Paris from the XIX\textsuperscript{th} century (1801 - 1854).
+They contain hundred-pages long lists of people, along with their activity and their address, so much valuable fine-grained knowledge to study the social dynamics of the city over time.
+Despite containing roughly the same information, the diversity in layouts, information organization and digitizing quality is significant, which hardens makes OCR and NER quite challenging.
+
+In a recent work to locate potentially polluted urban soils \cite{bell2020automated}, trade directories were leveraged to identify and locate all the gas stations in the city of Providence during the XX\textsuperscript{th} century.
+The proposed information extraction pipeline extract texts with OCR, identify directories entries and recognizes business type and address within each entry.
+In the same way, we aim at producing structured spatio-temporal data from the Parisian trade directories.
+These directories originate from several publishers, their content is organized following different indexing methods (by name, by activity or by address), printed in various layouts, use different fonts.
+We therefore investigate several state-of-the-art OCR and NER approaches independently and in combination to assess their usability on the directories corpus.
+The contributions of this article are the following:
 % Contributions

### 2022-01-12 src-latex/parts/40-ner-approaches.tex

@@
 \section{NER Approaches}
+Four approaches, one rules-based, three learning-based.
+
+\paragraph{Regexp}
+::TODO::
+
+\paragraph{Spacy 3 transition-based NER.}
+The \textit{of the shelf} NLP library SpaCY offers a named entity recognition tool based on an ad hoc architecture which has not been yet published but is explained by the developers on their website.
+Although the latest version of SpaCy (v3) leverages BERT models, we could not use them as no NER pipeline is available for the French language at the time of our experiments\footnote{https://spacy.io/models/fr}.
+The vanilla NER pipeline is two-fold.
+Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
+The decision layer is an adaptation of the transition-based model presented in~\cite{lample2016}.
+As words are processed sequentially, their vectors are concatenated with those in the last known entities to encode the nearby predicted semantics as a dynamic attention mechanism.
+The classification layer relies on a finite-state machine whose transition probabilities are learned using a multi-layer perceptron.
+In 2018~\cite{won2018} compared SpaCy's vanilla NER performances on historical texts about places with several other software packages (Stanford-NER, Ner-Tagger, Edinburgh Geopoarser and Polyglot).
+Results showed SpaCy to be average with a F1-score of 0.57.
+The SpaCy developers claims an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
+
+\paragraph{CamemBERT \& CamemBERT+pretraining on historical directories}
+::TODO::
 
@@
     \item regexp
-    \item Spacy CNN with tok2vec
     \item CamemBERT finetuned with wikiner-fr + gold dataset (clean texts)

### 2022-01-12 src-latex/parts/50-xp-setup.tex
info on camembert pretraining

--- 2022-01-06 src-latex/parts/50-xp-setup.tex
@@
 \subsection{Metrics}
+\bertrand{Move after section Cases studies? }
 \begin{itemize}
@@
 
-\subsection{Training procedure}
-5-fold CV for all models.
+\subsection{Cases studies}
+\subsubsection{Experiment 1: NER sensibility to the number of training samples}
+The first experiment evaluates the performances of the three models on training sets of different sizes.
+To do so we split the gold reference into a training set, a development set and a test set.
+The training set is then gradually reduced in size while maintaining the relative frequency of directories within.
+The training and testing procedure is the same for the three models.
+
+As the form an structure of entries varies across directories collections and through time, the models may learn to overfit on a subset of directories with specific features.
+To reduce the evaluation bias we start by leaving out 3 directories (1690 entries, ~20\%) from the gold reference to test each model on unseen directories.
+Then a stratified sampling based on the entry directory name is applied to the remaining set to create a training (6373 entries, ~71\% of the gold reference) and a development set (709 entries, ~8\%).
+This sampling procedure is a convenient way to shape both sets to reflect the diversity of directories within the gold reference.
+
+To generate smaller training sets we start from the initial training set and attractively split it in half using the same stratified sampling strategy.
+We stop if there is only one entry left in a directory or if the current train set contains less than 30 entries.
+Applying this procedure to the initial training set produced 8 training sets containing 49, 99, 199, 398, 796, 1593, 3186 and 6373 entries.
+
+% Move to metrics ?
+All metrics are evaluated on the test set, yet the biased model performances on the development sets are add in the paper material as additional information.
+
+\subsubsection{Experiment 2: named entity recognition in the context of noisy OCR data}
+
 \begin{itemize}
-\item Spacy CNN : no limits on the number of epochs, instead use patience=1600.
-\item CamemBERT : 3 epochs.
+\item At least 1 run per dataset, more if possible.
+\item Only use the best model; should be CamemBERT+pretrained
+\item Train on reference gold, evaluate on reference gold AND ocr-gold (2 datasets)
+\item split train/dev/test the same way than experiment 1. Use the exact same splitting for datasets so the clean test set, pero test set and tesseract test sets contains the same entries.
 \end{itemize}
+
+%\begin{itemize}
+%\item All : leave out 3 directories (~20\% of the dataset) as test data.
+Then apply stratified sampling based on the directory of each entry to split the remaining into a train set (~72\%) and a dev set (~8\%).
+%\item Spacy CNN : No limit on epochs, instead use patience 1600.
+%\item CamemBERT: 3 epochs.
+%\item CamemBERT pretrained: 3 epochs, pretraining on MLM + NSP with ~20000 raw texts extracted with Pero-OCR.
+%\end{itemize}
 
@@
 \subsection{Implementation details}
-we use spacy \mcite{spacy}
-will we use FLAIR as well?
+We use Spacy\mcite{spacy} for the CNN implementation
+and Huggingface for the BERT 
+tok2vec(words embeddings + encoding) + attention layer  +  transition-based model.
+
+\paragraph{Huggingface CamemBERT}
+CamemBERT + linear classifier as decision layer
+
+we use spacy  for the CNN implementation and Huggingface for BERT.
+
+will we use FLAIR as well? $\rightarrow$ No

### 2022-01-12 src-latex/parts/10-introduction.tex
Updates from Overleaf

@@
 Being active research topics, OCR and named entity recognition (NER) are still difficult tasks when applied to historical text documents.
-OCR approaches used for modern documents are likely to struggle even on printed historical documents due the multiple factors related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and inconsistent page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
+OCR approaches used for modern documents are likely to struggle even on printed historical documents due the multiple factors related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
 \bertrand{J'ai tenté un premier refactoring, pas plus techniquement précis ceci dit.}
-\textcolor{green}{Phrases précedentes à reprendre avec les termes adéquats + ajouter une ref.}
-On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts and from the errors introduced into the text by the OCR step.\textcolor{green}{ajouter une ref}
+\nathalie{Phrases précedentes à reprendre avec les termes adéquats + ajouter une ref.}
+On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts and from the errors introduced into the text by the OCR step.\nathalie{ajouter une ref}
 

### 2022-01-12 src-latex/parts/20-related-work.tex
Updates from Overleaf

@@
 % !TeX root = ../main-paper.tex
-\section{OCR and NER on Historical Texts}
+\section{OCR and NER on historical texts}
 
@@
 
-\subsection{Optical Character Recognition on Historical Texts}
+\subsection{Optical Character Recognition on historical texts}
 
@@
 
-\textcolor{green}{à compléter/retoucher}
+\nathalie{à compléter/retoucher}
 
@@
 Rule based approaches can also combine a set of patterns with a set of dictionaries (gazetteers, author lists, etc.) that help recognizing named entities when an exhaustive lexicon of searched entities is available \cite{mansouri2008,maurel2011}.
-For example, the CasEN\footnote{\url{https://tln.lifat.univ-tours.fr/version-francaise/ressources/casen}} transducer cascade leverages syntatic lexical patterns and dictionnaries to recognize Named Entities in French texts.
+For example, the CasEN\footnote{\url{https://tln.lifat.univ-tours.fr/version-francaise/ressources/casen}} transducer cascade leverages syntactic lexical patterns and dictionaries to recognize Named Entities in French texts.
 Such kind of approaches achieve very good results when applied to specialized domain corpus - like directories - and when an exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
@@
 
-\subsection{Named Entity Recognition in Historical Texts}
+\subsection{Named Entity Recognition on historical texts}
 
@@
 It is off-the-shelf multilingual BERT-based NER model, pretrained on more than 2 000 000 pages of OCRed historical texts in three languages and fine-tuned on historical texts NER ground truth, also in three languages.
-Eventually, a clear decrease in NER performance on OCRised texts is noted.
+Eventually, a clear decrease in NER performance on OCRed texts is noted.
 Moreover, the article reports that this decrease is more important for English texts for which the OCR is of poorer quality than for texts in the other two languages.
@@
 Several recent studies have focused on the extent to which the quality of the OCR affects the results produced by a NER model.\cite{van2020assessing} assess the impact of OCR on several NLP downstream tasks, including NER.
-They worked on a corpus published by a post-OCR correction software company, made of many journal articles with different levels of OCR errors and their respective groundtruths.
+They worked on a corpus published by a post-OCR correction software company, made of many journal articles with different levels of OCR errors and their respective ground truths.
 For each OCRed article, the Word Error Rate (WER) is computed and the English model \textit{en-core-web-lg} provided by Spacy\footnote{\url{https://spacy.io/}} library is used to perform NER on \textit{Person}, \textit{GPE} and \textit{Date}.
@@
 Finally, they applied each NER model to the progressively degraded versions of the corpus and computed the resulting F-measure.
-Overall, NER F-mesure drops from 90\% to 50\% when the WER increase from 8\% to 50\%.
+Overall, NER F-score drops from 90\% to 50\% when the WER increase from 8\% to 50\%.
 Besides, all NER models are sensitive to OCR errors, but models based on deep neural networks seem less prone to errors.
@@
 It starts to decrease significantly when OCR errors exceed these thresholds.
-The latter work focuses on adapting the training data to facilitate the generalisation of an off-the-shelf NER model from modern texts to historical texts.
+The latter work focuses on adapting the training data to facilitate the generalization of an off-the-shelf NER model from modern texts to historical texts.
 Three different NER models are tested on three historical corpora, in French, English, and Dutch.
-The best results are produced by a model trained on clean modern data, including many  embeddings as input (FastText, character and multilingual Flair embeddings, BERT and historical embeddings computed with Flair on a historical corpus), and fine-tuned on a noisy historical groundtruth.
+The best results are produced by a model trained on clean modern data, including many  embeddings as input (FastText, character and multilingual Flair embeddings, BERT and historical embeddings computed with Flair on a historical corpus), and fine-tuned on a noisy historical ground truth.
 

### 2022-01-12 src-latex/parts/40-ner-approaches.tex
Updates from Overleaf

@@
 
-\paragraph{Regexp}
+\subsubsection{Regexp}
 ::TODO::
+Needed for experiment 2 ?
 
-\paragraph{Spacy 3 transition-based NER.}
+\subsubsection{Spacy 3 NER pipeline}
 The \textit{of the shelf} NLP library SpaCY offers a named entity recognition tool based on an ad hoc architecture which has not been yet published but is explained by the developers on their website.
@@
 
-\paragraph{CamemBERT \& CamemBERT+pretraining on historical directories}
+\subsubsection{CamemBERT \& CamemBERT+pretrained on raw entries}
 ::TODO::

### 2022-01-12 src-latex/parts/80-conclusion.tex
Setup paper structure
Updates from Overleaf

--- 2021-10-28 src-latex/parts/50-xp-setup.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Experimental Setup}
+\section{Conclusion}
 
-\subsection{Implementation details}
-we use spacy \mcite{spacy}
+Ideas for future work:
+\begin{itemize}
+    \item Evaluate this: fine-tune Spacy NER on a few samples to generated a bootstrap dataset, feed it to BERT (cleaned or raw) to create an good model with minimal efforts ?
+    \item Use ML to infer regexes ?
+    \item As directories entries are always structured the same way (in a given list at least): use NER to identify entries within pages to save the burden of having to rely on the page layout ?
+\end{itemize}

### 2022-01-12 src-latex/parts/50-xp-setup.tex

@@
 \begin{itemize}
-    \item NER quality: Precision, Recall, F1 score (idea: stick with standard spacy/sklearn metrics) \textcolor{blue}{Dans la plupart des articles que j'ai lus, c'est toujours une F-mesure qui est utilisée et parfois la précision et le rappel sont ajoutés.}
-    \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?) \textcolor{blue}{Pour l'instant, j'ai croisé: CER, WER, Levenshtein et Jaccard.
+    \item NER quality: Precision, Recall, F1 score (idea: stick with standard spacy/sklearn metrics) \nathalie{Dans la plupart des articles que j'ai lus, c'est toujours une F-mesure qui est utilisée et parfois la précision et le rappel sont ajoutés.}
+    \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?) \nathalie{Pour l'instant, j'ai croisé: CER, WER, Levenshtein et Jaccard.
 J'ai encore un état de l'art à lire sur le sujet: il y aura probablement d'autres mesures mentionnées.}
 
-    \textcolor{blue}{Pour la partie Experiment 2: dans la litérature j'ai vu des courbes F-Mesure(CER) ou F-Mesure(WER) et des diagrammes avec F-Mesure calculée sur des sous-ensembles du corpus classés par score de Levenshtein.}
+    \nathalie{Pour la partie Experiment 2: dans la litérature j'ai vu des courbes F-Mesure(CER) ou F-Mesure(WER) et des diagrammes avec F-Mesure calculée sur des sous-ensembles du corpus classés par score de Levenshtein.}
 \end{itemize}
@@
 
-\subsubsection{Experiment 2: named entity recognition in the context of noisy OCR data}
+\subsubsection{Experiment 2: NER in the presence of noisy OCR texts}
 

### 2022-01-12 src-latex/parts/60-results.tex

@@
 \midrule
-Cnn & \textbf{0.855} & \textbf{0.885} & \textbf{0.910} & 0.911 & 0.919 & 0.925 & 0.926 & 0.929\\
+Spacy NER & \textbf{0.855} & \textbf{0.885} & \textbf{0.910} & 0.911 & 0.919 & 0.925 & 0.926 & 0.929\\
 Camembert & 0.236 & 0.626 & 0.856 & \textbf{0.916} & \textbf{0.930} & 0.933 & \textbf{0.948} & 0.953\\
@@
                             & reference gold & f1 & f1 & \\ 
+                            & reference gold & f1 & f1 & \\
 \cline{1-4}
-                            & \multicolumn{1}{l}{} &       & \multicolumn{1}{l}{}       &   \\
-                            & \multicolumn{1}{l}{} &       & \multicolumn{1}{l}{}       &  
 \end{tabular}

### 2022-01-12 src-latex/parts/20-related-work.tex
One sentence per line

@@
 \cite{martinek2019hybrid} propose an approach to generate synthetic annotated text for historical OCR training, based on manually collected characters from historical text images.
-This work proposes to train an OCR system based on a CNN-
-LSTM network with synthetic data and then to fine-tune the model with some pages of real historical annotated text. The results show that this approach gives state-of-the-art results.
+This work proposes to train an OCR system based on a CNN-LSTM network with synthetic data and then to fine-tune the model with some pages of real historical annotated text.
+The results show that this approach gives state-of-the-art results.
 
@@
 
-Several recent studies have focused on the extent to which the quality of the OCR affects the results produced by a NER model.\cite{van2020assessing} assess the impact of OCR on several NLP downstream tasks, including NER.
+Several recent studies have focused on the extent to which the quality of the OCR affects the results produced by a NER model.
+\cite{van2020assessing} assess the impact of OCR on several NLP downstream tasks, including NER.
 They worked on a corpus published by a post-OCR correction software company, made of many journal articles with different levels of OCR errors and their respective ground truths.

### 2022-01-12 src/ocr/test_text_utils.py
Add align NER gold for Pero and Tesseract output; and update tools

@@ -27,19 +27,23 @@ def test_str_rstrip_removes_all_newlines():
     s = "abc \u000A \u000B \u000C \u000D \u0085 \u2028 \u2029"
     assert(s.rstrip() == "abc")
 
-def test_align_ner_1():
+def test_add_tags_prediction():
     A = "<PER>Anthony</PER> \uE001, <FT>fab.</FT> du <ACT>pêche</ACT>, <LOC>Châtelet</LOC>xx"
     B = "\u261E Mme Antoine, \uE00A fab la p&cheur et \ndu ch>telet et du Faub. St Antoine"
-    R, no_empty_tag = add_tags_prediction(A, B)
-    assert(no_empty_tag)
+    R = add_tags_prediction(A, B)
     assert(R == "\u261E Mme <PER>Antoine</PER>, \uE00A <FT>fab</FT> la <ACT>p&amp;che</ACT>ur et \u2029du <LOC>ch&gt;telet</LOC> et du Faub. St Antoine")
 
-def test_align_ner_1():
     A = "<PER>Anthony</PER> \uE001, <FT>fab</FT><FT>.</FT> du <ACT>pêche</ACT>, <LOC>Châtelet</LOC>xx"
     B = "\u261E Mme Antoine, \uE00A fab la p&cheur et \ndu ch>telet et du Faub. St Antoine"
-    R, no_empty_tag = add_tags_prediction(A, B)
-    assert(not no_empty_tag)
+    R = add_tags_prediction(A, B)
     assert(R == "\u261E Mme <PER>Antoine</PER>, \uE00A <FT>fab</FT><FT></FT> la <ACT>p&amp;che</ACT>ur et \u2029du <LOC>ch&gt;telet</LOC> et du Faub. St Antoine")
+
+    A = "<PER>Anthony</PER> \uE001, <FT>fab</FT><FT>.</FT> du <ACT>pêche</ACT>, <LOC>Châtelet</LOC>xx"
+    B = ""
+    R = add_tags_prediction(A, B)
+    assert(R == "<PER></PER><FT></FT><FT></FT><ACT></ACT><LOC></LOC>")
+
+
 
 def test_xml_contains_empty_tags():
     XML = "<PER>Lebas (en gros)</PER>,<PER> </PER><LOC>R. et Div. de la Fraternité</LOC>, <CARDINAL>91</CARDINAL>."

### 2022-01-12 src/ocr/text_utils.py

@@ -634,7 +634,7 @@ def simplify_for_alignment(text: str, case_insensitive=True) -> str:
     return simplified
 
 
-def add_tags_prediction(ner_xml_final: str, text_ocr_final: str, debug=False) -> Tuple[str, bool]:
+def add_tags_prediction(ner_xml_final: str, text_ocr_final: str, debug=False) -> str:
     """Align NER tag positions from reference to noisy OCR.
 
     Args:
@@ -649,8 +649,11 @@ def add_tags_prediction(ner_xml_final: str, text_ocr_final: str, debug=False) ->
         ValueError: If OCR text charset is invalid
 
     Returns:
-        Tuple[str, bool]: (xml, no_empty_tag) XML with projected tags and wether this looks correct.
+        xml (str): XML with projected tags.
     """
+    if len(ner_xml_final) == 0:
+        raise ValueError("Alignment from empty ref is undefined.")
+
     # 1. Process ner_xml_final
     # => Assume we already have a clean valid XML content with normalized unicode
     # text without any annotation codes
@@ -688,6 +691,10 @@ def add_tags_prediction(ner_xml_final: str, text_ocr_final: str, debug=False) ->
     if debug:
         print("A", A)
         print("B", B)
+    
+    # Shortcut when B is empty
+    if len(B) == 0:
+        return "".join(A_tags)
 
     # 4. 
     pos_tags = np.cumsum([len(x) for x in A_chunks[:-1]])
@@ -706,13 +713,9 @@ def add_tags_prediction(ner_xml_final: str, text_ocr_final: str, debug=False) ->
     stack = []
     left = 0
     right = left
-    empty_tag = False
     for p, tag in zip(list(pos_tags), list(A_tags)):
         if tag.startswith("</"):
             right = A[p-1] + 1
-            if left+1 == right:
-                print(f"Error: empty tag {tag} after alignment.")
-                empty_tag = True # FIXME check string content instead
         else:
             right = A[p]
         sub = chr_list[left:right]
@@ -728,8 +731,10 @@ def add_tags_prediction(ner_xml_final: str, text_ocr_final: str, debug=False) ->
         print(left, right, sub)
         print("")
     stack.append(xml_escape("".join(sub)))
-
-    return "".join(stack), not empty_tag
+    res = "".join(stack)
+    if debug:
+        print(f"OUT: \t{res}")
+    return res
 
 
 

### 2022-01-13 src-latex/parts/30-dataset.tex
One sentence per line
Added some corrections in the intro, a table in section 3 and detailed explanations about NER models

--- 2022-01-06 src-latex/parts/30-dataset.tex
@@
 \section{Dataset}
+
+In this work, we  aim  at  producing  structured  spatio-temporal  data  from the information contained in the 19th century Parisian trade directories.
+To do this, it is therefore necessary to extract the text from the scans of the directories to be processed and to identify the entries in the various indexing lists used.
+Then, in each of these entries, the named entities they contain have to be extracted to produce structured spatio-temporal data representing the evolution over time of the people and businesses listed in these directories, their descriptions and their locations.
+Both of these extraction tasks require data to train and evaluate the OCR and NER approaches identified as potentially relevant: Tesseract and PERO OCR for the OCR task and Spacy, and CamemBERT - pre-trained or fine-tuned - neural models for the NER task.
+
+\subsection{The 19th century Parisian trade directories}
+
+\begin{table}[h!]
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{|l|l|l|l|l|}
+\hline
+\multicolumn{1}{|c|}{\textbf{Directory publisher or name}} & \multicolumn{1}{c|}{\textbf{Year}} & \multicolumn{1}{c|}{\textbf{\# Col}} & \multicolumn{1}{c|}{\textbf{Index}} & \multicolumn{1}{c|}{\textbf{Entry structure}}
+\\ \hline
+Favre \& Duchesne & 1798 & 1 & by activity & {N [A], SN, SNUM — SEC.} 
+\\ \hline
+Duverneuil et La Tynna & 1801 & 1 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & {N, SN, SNUM. SEC.} 
+\\ \hline
+{Notables communaux de la Seine} & 1801 & 1 & by name & {N, [A,] SN[, SNUM].}
+\\ \hline
+Duverneuil et La Tynna & 1805 &	2 &	\begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F)], SN, SNUM. — SEC.\\ N [(C)], [(A),] SN, SNUM. — SEC.\end{tabular}
+\\ \hline
+Duverneuil et La Tynna & 1806 & 2 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F)], SN, SNUM.\\ N [(C)], SN, SNUM.[ — SEC.]\end{tabular}.
+\\ \hline
+La Tynna & 1813	& 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N,[A,] SN[, SNUM].\\ N, [(T)] A, SN[, SNUM] [, D].\end{tabular}.
+\\ \hline
+Panckoucke Comm Dulac & 1820 & 1 & by name & N [(F)], A, [(T),] SN, SNUM[, P] .
+\\ \hline
+Panckoucke hab Dulac & 1820	& 1 & by name & N [(T)], [A,] SN, SNUM [, D].
+\\ \hline
+Bottin1 & 1820 & 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F \textbar C)][T], [P], SN, SNUM.\\N [(F \textbar C)][T], A, SN, SNUM.
+CR\end{tabular}.
+\\ \hline
+Bottin1 & 1827 & 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], A, SN, SNUM.\end{tabular}.
+\\ \hline
+Deflandre & 1828 & 2 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F \textbar C)][T], [P], SN, SNUM.
+\\N [(F \textbar C)][T], A, SN, SNUM.\end{tabular}.
+\\ \hline
+Deflandre & 1829 & 2 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], A, SN, SNUM.\end{tabular}.
+\\ \hline 
+Bottin1 & 1837 & 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], [A], SN, SNUM.\end{tabular}.
+\\ \hline
+Cambon almgene & 1841 & 2 &	\begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], A, SN, SNUM.\end{tabular}.
+\\ \hline
+Didot & 1841a &	\begin{tabular}[c]{@{}l@{}}3\\ 4\end{tabular} &	\begin{tabular}[c]{@{}l@{}}by activity \\ by name \end{tabular} & 	\begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.
+\\ N [(F \textbar C)][T], [A], SN, SNUM.
+\end{tabular}
+\\ \hline
+Didot & 1851a & \begin{tabular}[c]{@{}l@{}l@{}}4\\ 4\\ 5\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}by activity\\ by name\\ by street name\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], [A], SN, SNUM.\\SNUM N [(F \textbar C)][T], [A].\end{tabular}
+\\ \hline
+Didot & 1854a & \begin{tabular}[c]{@{}l@{}l@{}}4\\ 3\\ 5\end{tabular} &	\begin{tabular}[c]{@{}l@{}l@{}} by activity \\ by name \\ by street name\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}} N [(F \textbar C)], [P], SN, SNUM [,T].\\ N [(F \textbar C)][T], [A], SN, SNUM.\\ SNUM N [(F \textbar C)][T], [A].\end{tabular}
+\\ \hline
+Bottin3 & 1854a & \begin{tabular}[c]{@{}l@{}l@{}}3\\ 2\\ 4\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}} by activity \\ by name \\ by street name\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}N [(F \textbar C)], [P], SN, SNUM [,T].\\ N [(F \textbar C)][T], [A], SN, SNUM.\\ SNUM N [(F \textbar C)], [A] [,T].\end{tabular}
+\\ \hline
+DidotBottin & 1860a	& 3	& by name & N [(F \textbar C)][T], [A], SN, SNUM.
+\\ \hline
+DidotBottin & 1861a	& 3	& by name & N [(F \textbar C)][T], [A], SN, SNUM.
+\\ \hline
+\end{tabular}%
+}
+\caption{The directories used to build our dataset.
+Entry structure is described as follows: N - entry name (person or business); F - first name; T - honorary title; C - civility; A - activity; P - precision; SN - street name; SNUM - street number; SEC - section; D - district; CR - cross-reference}
+\label{tab:my-table}
+\end{table}
+
+\subsection{Groundtruth data extraction}
+
+\subsection{Assessing the groundtruth quality}
+
+Our dataset is thus made of excerpts
 

### 2022-01-13 src/ocr/OCRMetrics.py
Run OCR stats directly

@@ -45,7 +45,25 @@ class AlignmentStats:
         '''
         Get the alignment accuracy (number of match over the alignment length)
         '''
-        return self.matches / (self.matches + self.errors)
+        return self.matches / self.length
+
+    
+    @property
+    def CER(self) -> float:
+        '''
+        Get the alignment error rare (number of errors over the reference length)
+        https://towardsdatascience.com/evaluating-ocr-output-quality-with-character-error-rate-cer-and-word-error-rate-wer-853175297510#5aec
+        '''
+        return self.errors / self.reference_length
+
+    @property
+    def CERnorm(self) -> float:
+        '''
+        Get the alignment error rare (number of errors over the reference length)
+        https://towardsdatascience.com/evaluating-ocr-output-quality-with-character-error-rate-cer-and-word-error-rate-wer-853175297510#5aec
+        https://sites.google.com/site/textdigitisation/qualitymeasures/computingerrorrates
+        '''
+        return self.errors / self.length
 
 
 @dataclass
@@ -61,7 +79,7 @@ class AlignmentReport:
     def __init__(self, reference: str, x: str) -> None:
         self.reference = reference
         self.x = x
-        self.aligned_ref, self.aligned_x = isri_tools.align(reference, x, "✗")
+        self.aligned_ref, self.aligned_x = isri_tools.align(reference, x, "⌴")
 
         match = 0
         for a,b in zip(self.aligned_ref, self.aligned_x):

### 2022-01-13 src/ocr/isri_tools_src/Modules/list.c

@@ -143,6 +143,7 @@ void list_empty(void *list_, void (*process_item)(void*))
     {
 	item = list->first;
 	list_remove(list, item);
-	(*process_item)(item);
+    if (process_item)
+	    (*process_item)(item);
     }
 }

### 2022-01-13 src/ocr/isri_tools_src/accuracy.cpp

@@ -155,8 +155,49 @@ get_accuracy_stats (std::wstring ref, std::wstring prediction)
   auto data = std::make_unique<Accdata> ();
   process_synclist (texts, &synclist, data.get());
 
+  list_empty(&A, nullptr);
+  list_empty(&B, nullptr);
+
   return data;
 }
+
+std::unique_ptr<Accdata>
+get_accuracy_summary(const std::vector<std::wstring>& ref, const std::vector<std::wstring>& prediction)
+{
+
+
+
+  if (ref.size() != prediction.size())
+    throw std::runtime_error("Incompatible list size.");
+
+  std::size_t n = ref.size();
+
+  std::unique_ptr<Accdata> stats = get_accuracy_stats(ref[0], prediction[0]);
+
+  Text A;
+  Text B;
+
+  for (std::size_t i = 1; i < n; ++i)
+  {
+    list_initialize (&A);
+    list_initialize (&B);
+    text_from_string (&A, ref[i]);        // 16 bit fixed encoding only for now
+    text_from_string (&B, prediction[i]); // 16 bit fixed encoding only for now
+    Synclist synclist;
+    Text texts[2] = { A, B };
+    fastukk_sync (&synclist, texts);
+    process_synclist (texts, &synclist, stats.get());
+    list_empty(&synclist, [](void* x) {
+      Sync* e = (Sync*)x;
+      if (e->substr) free(e->substr);
+      if (e->match) free(e->match);
+    });
+    list_empty(&A, nullptr);
+    list_empty(&B, nullptr);
+  }
+  return stats;
+}
+
 
 std::wstring
 print_accurary_report (Accdata *data)
@@ -239,5 +280,5 @@ init_accuracy (py::module &m)
     .def_readonly ("false_marks", &Accdata::false_marks, "number of false marks")
     .def ("__repr__", &print_accurary_report);
   m.def ("compute_accurary_stats", &get_accuracy_stats);
-  m.def ("accurary_summary", &accuracy_summary);
+  m.def ("compute_accurary_summary", &get_accuracy_summary);
 }

### 2022-01-13 src-latex/parts/10-introduction.tex
Completed Table 1 and added some text in section 3

@@
 Being active research topics, OCR and named entity recognition (NER) are still difficult tasks when applied to historical text documents.
-OCR approaches used for modern documents are likely to struggle even on printed historical documents due the multiple factors related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
+OCR approaches used for modern documents are likely to struggle even on printed historical documents due multiple causes related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
 \bertrand{J'ai tenté un premier refactoring, pas plus techniquement précis ceci dit.}

### 2022-01-13 src-latex/parts/30-dataset.tex
Completed Table 1 and added some text in section 3

--- 2022-01-06 src-latex/parts/30-dataset.tex
@@
 \section{Dataset}
+
+%In this work, we  aim  at  producing  structured  spatio-temporal  data  from the information contained in the XIX\textsuperscript{th} century Parisian trade directories.
+%To do this, it is therefore necessary to extract the text from the scans of the directories to be processed and to identify the entries in the various indexing lists used.
+%Then, in each of these entries, the named entities they contain have to be extracted to produce structured spatio-temporal data representing the evolution over time of the people and businesses listed in these directories, their descriptions and their locations.
+%Both of these extraction tasks require data to train and evaluate the OCR and NER approaches identified as potentially relevant: Tesseract and PERO OCR for the OCR task and Spacy and CamemBERT - pre-trained or fine-tuned - neural models for the NER task.
+
+\subsection{The 19th century Parisian trade directories}
+
+The XIX\textsuperscript{th} century directories we have to process are stored in different libraries in Paris\footnote{Citer les bibliothèques sources?}.
+They have been scanned independently from each other and are available in digital format with various levels of quality.
+Moreover, they cover a wide period and ave been produced by different publishers and printers.
+Their contents, indexes, layouts, methods of printing, etc.
+vary thus from one directory to another (see Figure \ref{fig:directories}).
+
+\begin{figure}[htb!]
+	   \center{\includegraphics[width=\textwidth]
+	       {./images/DirectoryExcerpts.png}}
+	  \caption{\label{fig:directories} Examples of directory layouts and contents}
+\end{figure}
+
+In this work, we therefore made a selection among the available directories, trying our best to retain copies representative of the diversity of existing directories and of sufficient quality to produce a reliable training data set.
+The Table \ref{tab:directories} lists the directories selected, their layout, the structure of their entries, and the number of annotated entries according to the indexes considered.
+
+\begin{table}[h!]
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\multicolumn{1}{|c|}{\textbf{Directory publisher or name}} & \multicolumn{1}{c|}{\textbf{Year}} & \multicolumn{1}{c|}{\textbf{\# Col}} & \multicolumn{1}{c|}{\textbf{Index}} & \multicolumn{1}{c|}{\textbf{Entry structure}} & \multicolumn{1}{c|}{\textbf{\# Ann}}
+\\ \hline
+Favre \& Duchesne & 1798 & 1 & by activity & {N [A], SN, SNUM - SEC.} & 233
+\\ \hline
+Duverneuil et La Tynna & 1801 & 1 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & {N, SN, SNUM. SEC.} & \begin{tabular}[c]{@{}l@{}} 136 \\ 103\end{tabular}
+\\ \hline
+{Notables communaux de la Seine} & 1801 & 1 & by name & {N, [A,] SN[, SNUM].} & 92
+\\ \hline
+Duverneuil et La Tynna & 1805 &	2 &	\begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F)], SN, SNUM.
+- SEC.\\ N [(C)], [(A),] SN, SNUM.
+- SEC.\end{tabular} & \begin{tabular}[c]{@{}l@{}} 155 \\ 221 \end{tabular}
+\\ \hline
+Duverneuil et La Tynna & 1806 & 2 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F)], SN, SNUM.\\ N [(C)], SN, SNUM.[ - SEC.]\end{tabular}.
+& \begin{tabular}[c]{@{}l@{}} 186\\ 0\end{tabular}
+\\ \hline
+La Tynna & 1813	& 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N,[A,] SN[, SNUM].\\ N, [(T)] A, SN[, SNUM] [, D].\end{tabular} & \begin{tabular}[c]{@{}l@{}} 135 \\ 177 \end{tabular}
+\\ \hline
+Panckoucke Commerces (Dulac) & 1820 & 1 & by name & N [(F)], A, [(T),] SN, SNUM[, P]. & 0
+\\ \hline
+Panckoucke Habitants (Dulac) & 1820	& 1 & by name & N [(T)], [A,] SN, SNUM [, D]. & 0
+\\ \hline
+Bottin (serie 1: 1819-1838) & 1820 & 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F \textbar C)][T], [P], SN, SNUM.\\N [(F \textbar C)][T], A, SN, SNUM.
+CR\end{tabular} & \begin{tabular}[c]{@{}l@{}} 128 \\ 151 \end{tabular}
+\\ \hline
+Bottin (serie 1: 1819-1838) & 1827 & 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], A, SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}} 228 \\ 285 \end{tabular}
+\\ \hline
+Deflandre & 1828 & 2 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F \textbar C)][T], [P], SN, SNUM.
+\\N [(F \textbar C)][T], A, SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}}115\\ 229\end{tabular}
+\\ \hline
+Deflandre & 1829 & 2 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], A, SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}}177\\ 236\end{tabular}
+\\ \hline 
+Bottin (serie 1: 1819-1838) & 1837 & 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], [A], SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}}158\\ 557\end{tabular}
+\\ \hline
+Cambon - Almanach général & 1841 & 2 &	\begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], A, SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}}182\\ 486\end{tabular}
+\\ \hline
+Didot & 1841a &	\begin{tabular}[c]{@{}l@{}}3\\ 4\end{tabular} &	\begin{tabular}[c]{@{}l@{}}by activity \\ by name \end{tabular} & 	\begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.
+\\ N [(F \textbar C)][T], [A], SN, SNUM.
+\end{tabular} & \begin{tabular}[c]{@{}l@{}}246\\ 843\end{tabular}
+\\ \hline
+Didot & 1851a & \begin{tabular}[c]{@{}l@{}l@{}}4\\ 4\\ 5\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}by activity\\ by name\\ by street name\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], [A], SN, SNUM.\\SNUM N [(F \textbar C)][T], [A].\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}309\\ 961\\ 0\end{tabular}
+\\ \hline
+Didot & 1854a & \begin{tabular}[c]{@{}l@{}l@{}}4\\ 3\\ 5\end{tabular} &	\begin{tabular}[c]{@{}l@{}l@{}} by activity \\ by name \\ by street name\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}} N [(F \textbar C)], [P], SN, SNUM [,T].\\ N [(F \textbar C)][T], [A], SN, SNUM.\\ SNUM N [(F \textbar C)][T], [A].\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}106\\ 362\\ 0\end{tabular}
+\\ \hline
+Bottin (serie 3: 1854-1856) & 1854a & \begin{tabular}[c]{@{}l@{}l@{}}3\\ 2\\ 4\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}} by activity \\ by name \\ by street name\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}N [(F \textbar C)], [P], SN, SNUM [,T].\\ N [(F \textbar C)][T], [A], SN, SNUM.\\ SNUM N [(F \textbar C)], [A] [,T].\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}253\\ 633\\ 0\end{tabular}
+\\ \hline
+Didot-Bottin & 1860a	& 3	& by name & N [(F \textbar C)][T], [A], SN, SNUM. & 378
+\\ \hline
+Didot-Bottin & 1861a	& 3	& by name & N [(F \textbar C)][T], [A], SN, SNUM. & 402
+\\ \hline
+\end{tabular}%
+}
+\caption{The directories used to build our dataset.
+Letters next to year of publication are used to identify the volume of the directory.
+\# Col stands for "Number of columns", and \# Ann stands for "Number of annotated entries".
+Entry structure is described as follows: N - entry name (person or business); F - first name; T - honorary title; C - civility; A - activity; P - precision; SN - street name; SNUM - street number; SEC - section; D - district; CR - cross-reference}
+\label{tab:directories}
+\end{table}
+
+\subsection{Groundtruth data extraction}
+
+\subsection{Assessing the groundtruth quality}
+
+Our dataset is thus made of excerpts
 

### 2022-01-14 src-latex/parts/30-dataset.tex
Continued dataset description

@@
 \begin{figure}[htb!]
-	   \center{\includegraphics[width=\textwidth]
-	       {./images/DirectoryExcerpts.png}}
-	  \caption{\label{fig:directories} Examples of directory layouts and contents}
+	   \center{\includegraphics[width=0.9\textwidth]
+	       {./images/DirectoryExcerpts2.png}}
+	  \caption{\label{fig:directories} Examples of directory layouts and contents: 1) Duverneuil et La Tynna 1806 - index by name; 2) Deflandre 1828 - index by activity ; 3) Bottin 1851 –  index by street name}
 \end{figure}
@@
 The Table \ref{tab:directories} lists the directories selected, their layout, the structure of their entries, and the number of annotated entries according to the indexes considered.
+For example, the Didot directory of 1854 is organised according to three indexes: by name, by activity, and by street name.
+The structure of the following entry is characteristic of its index by name: "Batton (D.-A.) *, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47." This entry begins with a person's name, followed by the initials of their first name in parentheses.The star symbol indicates that this person was awarded the Légion d'Honneur.
+The following part of this entry represents the person's activity,i.e.
+their profession, or social status, here "professor at the Conservatory of music and declamation".
+The street name and number where the person lives or carries out their activity are written at the end of the entry.
+These pieces of information constituting the structure of the entries are the basic data that we want to extract, deduplicate and structure in order to build a spatio-temporal database of XIX\textsuperscript{th} century Parisian inhabitants and businesses.
+With the exception of some potentially wordy activity descriptions, most of these pieces of information correspond to named entities.
+However, as Table \ref{tab:directories} shows, while most entries contain the same types of named entities, the order in which they appear and the way they are written vary from one directory to another.
+This argues for a supervised, rather than rule-based, NER approach.
+This requires annotated data.
 
@@
 \# Col stands for "Number of columns", and \# Ann stands for "Number of annotated entries".
-Entry structure is described as follows: N - entry name (person or business); F - first name; T - honorary title; C - civility; A - activity; P - precision; SN - street name; SNUM - street number; SEC - section; D - district; CR - cross-reference}
+Entry structure is described as follows: N - entry name (person or business); F - first name; T - honorary title; C - civility; A - activity; P - precision; SN - street name; SNUM - street number; SEC - section; D - district; CR - cross-reference; bracketed elements are optional.}
 \label{tab:directories}

### 2022-01-14 src-latex/parts/60-results.tex
Continued dataset description

@@
 \begin{table}[!h]
-
-\caption{Models F1 score on unseen data vs trainset size}
+\caption{Models performances vs trainset size}
 \centering
-\begin{tabular}[t]{lcccccccc}
+\begin{tabular}[t]{lr|cccccccc|}
 \toprule
-Trainset size (entries) & 49 & 99 & 199 & 398 & 796 & 1593 & 3186 & 6373\\
-Trainset size (\% of the full trainset) & 0.8 & 1.5 & 3.1 & 6.2 & 12.5 & 25 & 50 & 100\\
+ & Trainset size & 49 & 99 & 199 & 398 & 796 & 1593 & 3186 & 6373\\
+ & Trainset size (\%) & 0.8 & 1.5 & 3.1 & 6.2 & 12.5 & 25 & 50 & 100\\
+ \hline \hline
+\multirow{3}{*}{F1 score} & SpaCy NER & 0.870 & 0.890 & 0.903 & 0.919 & 0.921 & 0.928 & 0.932 & 0.936\\
+& CamemBERT & 0.894 & 0.903 & 0.927 & 0.934 & 0.943 & 0.953 & 0.947 & 0.955\\
+& CamemBERT+pretraining & 0.900 & 0.914 & 0.929 & 0.934 & 0.942 & 0.945 & 0.948 & 0.954\\
 \midrule
-Spacy NER & \textbf{0.855} & \textbf{0.885} & \textbf{0.910} & 0.911 & 0.919 & 0.925 & 0.926 & 0.929\\
-Camembert & 0.236 & 0.626 & 0.856 & \textbf{0.916} & \textbf{0.930} & 0.933 & \textbf{0.948} & 0.953\\
-Camembert+pretraining & 0.190 & 0.685 & 0.889 & 0.913 & \textbf{0.930} & \textbf{0.946} & \textbf{0.948} & \textbf{0.960}\\
+\multirow{3}{*}{Precision} & SpaCy NER precision & 0.856 & 0.877 & 0.900 & 0.920 & 0.924 & 0.928 & 0.931 & 0.938\\
+& CamemBERT & 0.873 & 0.881 & 0.914 & 0.930 & 0.934 & 0.954 & 0.938 & 0.958\\
+& CamemBERT+pretraining & 0.879 & 0.898 & 0.916 & 0.927 & 0.934 & 0.940 & 0.941 & 0.954\\
+\midrule
+\multirow{3}{*}{Recall} & SpaCy NER & 0.886 & 0.904 & 0.907 & 0.917 & 0.919 & 0.928 & 0.933 & 0.933\\
+& CamemBERT & 0.918 & 0.927 & 0.941 & 0.939 & 0.952 & 0.951 & 0.956 & 0.951\\
+& CamemBERT+pretraining & 0.922 & 0.930 & 0.943 & 0.942 & 0.950 & 0.951 & 0.955 & 0.954\\
 \bottomrule

### 2022-01-14 material/experiment_1/figures.r
Experiment 1 results [WIP]

@@ -1,0 +1,89 @@
+library(dplyr)
+library(gridExtra)
+library(ggplot2)
+library(xtable)
+library(kableExtra)
+library(cowplot)
+
+# BERT : jq -r ' . | [ .[] ]  | @csv' $(l -1v ./*_test.json)
+# CNN : jq -r 'with_entries(select([.key] | inside( ["ents_p","ents_r","ents_f"] ) ) ) | [ .[] ]  | @csv' $(l -1v ./*_test.json)
+
+
+# Spacy CNN
+cnn = read.csv("spacy_ner/metrics.csv",header=T)
+cnn = cnn %>% arrange(trainsize) %>% group_by(trainsize) %>% filter(dataset == 'test') %>% summarise(across(everything(), mean))
+
+# HF Camembert 
+camembert = read.csv("camembert/metrics.csv",header=T)
+camembert = camembert %>% arrange(trainsize) %>% group_by(trainsize) %>% filter(dataset == 'test') %>% summarise(across(everything(), mean))
+
+# HF Camembert + pretraining
+camembert_pretrained = read.csv("camembert_pretrained/metrics.csv",header=T)
+camembert_pretrained = camembert_pretrained %>% arrange(trainsize) %>% group_by(trainsize) %>% filter(dataset == 'test') %>% summarise(across(everything(), mean))
+
+
+
+# EXPORT TABLE 1
+
+# Reshape
+tbl = cbind(cnn %>% select(eval_f1), camembert %>% select(eval_f1), camembert_pretrained %>% select(eval_f1), cnn %>% select(eval_precision), camembert %>% select(eval_precision), camembert_pretrained %>% select(eval_precision), cnn %>% select(eval_recall), camembert %>% select(eval_recall), camembert_pretrained %>% select(eval_recall))
+colnames(tbl) <- c("SpaCy NER F1", "CamemBERT F1", "CamemBERT+pretraining F1", "SpaCy NER precision", "CamemBERT precision", "CamemBERT+pretraining precision", "SpaCy NER recall", "CamemBERT recall", "CamemBERT+pretraining recall")
+
+tbl = t(tbl)
+colnames(tbl) <- camembert[['trainsize']]
+
+
+# TODO : max value of each column in bold
+#fn = function(x) cell_spec(x, bold = T)  
+#tbl[, 2:8] = t(apply(tbl[,2:8], 1, fn))    
+
+tbl %>%
[…]

### 2022-01-14 src-latex/parts/30-dataset.tex

@@
 
-\subsection{The 19th century Parisian trade directories}
+\subsection{The  XIX\textsuperscript{th} century Parisian trade directories}
 
-The XIX\textsuperscript{th} century directories we have to process are stored in different libraries in Paris\footnote{Citer les bibliothèques sources?}.
+The directories we have to process are stored in different libraries in Paris\footnote{Citer les bibliothèques sources?}.
 They have been scanned independently from each other and are available in digital format with various levels of quality.
@@
 However, as Table \ref{tab:directories} shows, while most entries contain the same types of named entities, the order in which they appear and the way they are written vary from one directory to another.
-This argues for a supervised, rather than rule-based, NER approach.
-This requires annotated data.
+This argues for a supervised, rather than rule-based, NER approach, which requires annotated data.
 
@@
 
+Présenter ici l'outil d'annotation et le guide d'annotation suivi.
+
 \subsection{Assessing the groundtruth quality}

### 2022-01-14 src-latex/parts/60-results.tex

--- 2022-01-12 src-latex/parts/60-results.tex
@@
 \begin{table}[!h]
-
-\caption{Models F1 score on unseen data vs trainset size}
+\caption{Experimental results of the NER models performances when trained on varying numbers of examples.}
 \centering
-\begin{tabular}[t]{lcccccccc}
-\toprule
-Trainset size (entries) & 49 & 99 & 199 & 398 & 796 & 1593 & 3186 & 6373\\
-Trainset size (\% of the full trainset) & 0.8 & 1.5 & 3.1 & 6.2 & 12.5 & 25 & 50 & 100\\
-\midrule
-Spacy NER & \textbf{0.855} & \textbf{0.885} & \textbf{0.910} & 0.911 & 0.919 & 0.925 & 0.926 & 0.929\\
-Camembert & 0.236 & 0.626 & 0.856 & \textbf{0.916} & \textbf{0.930} & 0.933 & \textbf{0.948} & 0.953\\
-Camembert+pretraining & 0.190 & 0.685 & 0.889 & 0.913 & \textbf{0.930} & \textbf{0.946} & \textbf{0.948} & \textbf{0.960}\\
-\bottomrule
+\begin{tabular}[t]{lrcccccccc}
+ & Trainset size & 49 & 99 & 199 & 398 & 796 & 1593 & 3186 & 6373\\
+ & \% & 0.8 & 1.5 & 3.1 & 6.2 & 12.5 & 25 & 50 & 100\\
+\midrule[1pt]\bottomrule
+\multirow{3}{*}{\rotatebox{90}{F1}} & SpaCy NER & 0.87 & 0.89 & 0.90 & 0.92 & 0.92 & 0.93 & 0.93 & 0.94\\
+& CamemBERT & 0.89 & 0.90 & 0.93 & 0.93 & 0.94 & 0.95 & 0.95 & 0.95\\
+& CamemBERT.pretrained & 0.90 & 0.91 & 0.93 & 0.93 & 0.94 & 0.95 & 0.95 & 0.95\\
+\cmidrule{1-10}
+\multirow{3}{*}{\rotatebox{90}{Precision}} & SpaCy NER & 0.86 & 0.88 & 0.90 & 0.92 & 0.92 & 0.93 & 0.93 & 0.94\\
+& CamemBERT & 0.87 & 0.88 & 0.91 & 0.93 & 0.93 & 0.95 & 0.94 & 0.96\\
+& CamemBERT.pretrained & 0.88 & 0.90 & 0.92 & 0.93 & 0.93 & 0.94 & 0.94 & 0.95\\
+\cmidrule{1-10}
+\multirow{3}{*}{\rotatebox{90}{Recall}} & SpaCy NER & 0.89 & 0.90 & 0.91 & 0.92 & 0.92 & 0.93 & 0.93 & 0.93\\
+& CamemBERT & 0.92 & 0.93 & 0.94 & 0.94 & 0.95 & 0.95 & 0.96 & 0.95\\
+& CamemBERT.pretrained & 0.92 & 0.93 & 0.94 & 0.94 & 0.95 & 0.95 & 0.95 & 0.95\\
 \end{tabular}

### 2022-01-14 src-latex/parts/60-results.tex
Table 2 metrics

@@
 \midrule[1pt]\bottomrule
-\multirow{3}{*}{\rotatebox{90}{F1}} & SpaCy NER & 0.87 & 0.89 & 0.90 & 0.92 & 0.92 & 0.93 & 0.93 & 0.94\\
-& CamemBERT & 0.89 & 0.90 & 0.93 & 0.93 & 0.94 & 0.95 & 0.95 & 0.95\\
-& CamemBERT.pretrained & 0.90 & 0.91 & 0.93 & 0.93 & 0.94 & 0.95 & 0.95 & 0.95\\
+\multirow{3}{*}{\rotatebox{90}{F1}} & SpaCy NER & 87.0 & 89.0 & 90.3 & 91.9 & 92.1 & 92.8 & 93.2 & 93.6\\
+& CamemBERT & 89.4 & 90.3 & 92.7 & 93.4 & 94.3 & 95.3 & 94.7 & 95.5\\
+& CamemBERT.pretrained & 90.0 & 91.4 & 92.9 & 93.4 & 94.2 & 94.5 & 94.8 & 95.4\\
 \cmidrule{1-10}
-\multirow{3}{*}{\rotatebox{90}{Precision}} & SpaCy NER & 0.86 & 0.88 & 0.90 & 0.92 & 0.92 & 0.93 & 0.93 & 0.94\\
-& CamemBERT & 0.87 & 0.88 & 0.91 & 0.93 & 0.93 & 0.95 & 0.94 & 0.96\\
-& CamemBERT.pretrained & 0.88 & 0.90 & 0.92 & 0.93 & 0.93 & 0.94 & 0.94 & 0.95\\
+\multirow{3}{*}{\rotatebox{90}{Precision}} & SpaCy NER & 85.6 & 87.7 & 90.0 & 92.0 & 92.4 & 92.8 & 93.1 & 93.8\\
+& CamemBERT & 87.3 & 88.1 & 91.4 & 93.0 & 93.4 & 95.4 & 93.8 & 95.8\\
+& CamemBERT.pretrained & 87.9 & 89.8 & 91.6 & 92.7 & 93.4 & 94.0 & 94.1 & 95.4\\
 \cmidrule{1-10}
-\multirow{3}{*}{\rotatebox{90}{Recall}} & SpaCy NER & 0.89 & 0.90 & 0.91 & 0.92 & 0.92 & 0.93 & 0.93 & 0.93\\
-& CamemBERT & 0.92 & 0.93 & 0.94 & 0.94 & 0.95 & 0.95 & 0.96 & 0.95\\
-& CamemBERT.pretrained & 0.92 & 0.93 & 0.94 & 0.94 & 0.95 & 0.95 & 0.95 & 0.95\\
+\multirow{3}{*}{\rotatebox{90}{Recall}} & SpaCy NER & 88.6 & 90.4 & 90.7 & 91.7 & 91.9 & 92.8 & 93.3 & 93.3\\
+& CamemBERT & 91.8 & 92.7 & 94.1 & 93.9 & 95.2 & 95.1 & 95.6 & 95.1\\
+& CamemBERT.pretrained & 92.2 & 93.0 & 94.3 & 94.2 & 95.0 & 95.1 & 95.5 & 95.4\\
 \end{tabular}

### 2022-01-14 src-latex/parts/30-dataset.tex
Finished section 3.2

@@
 
-Présenter ici l'outil d'annotation et le guide d'annotation suivi.
+In order to have a ground truth corpus representative of the diversity of the directories to be processed, we annotated a few pages of each of the major directory series for which we have digitised copies.
+To that purpose, we used an annotation tool developed for our project and whose graphical user interface (GUI) is presented in Figure \ref{fig:annotator}.
+This tool loads images of directory pages.
+Then, it automatically detects each entry in a given page and builds a box around it : these boxes are shown in the left panel of the GUI.
+Then the text of each entry is extracted using an OCR tool and finally the named entities searched are annotated using a NER tool: the results are displayed respectively in the first and second tabs of the right panel of the GUI.
+The entry detection, OCR and NER tasks are performed automatically but as they are prone to errors, their results can be corrected manually through the GUI of the annotation tool.
+In the end, for each page processed, the bounding boxes of the entries that make it up, the exact text contained in each box, and the named entities that are mentioned are stored in a JSON file.
+
+\begin{figure}[htb!]
+	   \center{\includegraphics[width=0.9\textwidth]
+	       {./images/Annot2.png}}
+	  \caption{\label{fig:annotator} GUI for corpus annotation}
+\end{figure}
+
+The annotated named entities are as follows:
+\begin{itemize}
+    \item PER: a person name or a business name, usually referred to as several person names.
+Firts names, initials, or civility mentions are included.
+E.g.: "Alibert (Prosper)", "Allamand frères et Hersent", "Heurtemotte (Vve)", etc.
+    \item TITLE: an honorary title, either written or represented by special glyphs.
+E.g.: "élig.", \scalerel*{\includegraphics{./images/LH.png}}{H} which stand for the Légion d'Honneur, \scalerel*{\includegraphics{./images/GMLondon.png}}{H} which stands for the Great Medal at the London exhibition, .etc.
+    \item ACT: the activity of the person (profession or social status) or the type of business. E.g.: "horlogerie", "export en tous genres", "Conseiller d'Etat", "propriétaire", etc.
+    \item LOC: a street name or a district name. E.g.: "r. de la Jussienne", "Au Marais", etc.
+    \item CARD: a street number, as part of an address. E.g. "16", "5 bis", etc.
+    \item FT: a geographic entity type, used to give more details on a location. E.g.: "boutique", "atelier", "fab.", etc.
+\end{itemize}
+
+Thus, the extracted corpus contains both the bounding boxes of directory entries, their exact text and the named entities they contain.
+It can therefore be used to train and improve the entry detection, OCR and NER tools.
 

### 2022-01-14 src-latex/parts/60-results.tex
Finished section 3.2

@@
     \centering
-\begin{tabular}[t]{lrcccccccc}
- & Trainset size & 49 & 99 & 199 & 398 & 796 & 1593 & 3186 & 6373\\
- & \% & 0.8 & 1.5 & 3.1 & 6.2 & 12.5 & 25 & 50 & 100\\
-\midrule[1pt]\bottomrule
-\multirow{3}{*}{\rotatebox{90}{F1}} & SpaCy NER & 87.0 & 89.0 & 90.3 & 91.9 & 92.1 & 92.8 & 93.2 & 93.6\\
-& CamemBERT & 89.4 & 90.3 & 92.7 & 93.4 & 94.3 & 95.3 & 94.7 & 95.5\\
-& CamemBERT.pretrained & 90.0 & 91.4 & 92.9 & 93.4 & 94.2 & 94.5 & 94.8 & 95.4\\
-\cmidrule{1-10}
-\multirow{3}{*}{\rotatebox{90}{Precision}} & SpaCy NER & 85.6 & 87.7 & 90.0 & 92.0 & 92.4 & 92.8 & 93.1 & 93.8\\
-& CamemBERT & 87.3 & 88.1 & 91.4 & 93.0 & 93.4 & 95.4 & 93.8 & 95.8\\
-& CamemBERT.pretrained & 87.9 & 89.8 & 91.6 & 92.7 & 93.4 & 94.0 & 94.1 & 95.4\\
-\cmidrule{1-10}
-\multirow{3}{*}{\rotatebox{90}{Recall}} & SpaCy NER & 88.6 & 90.4 & 90.7 & 91.7 & 91.9 & 92.8 & 93.3 & 93.3\\
-& CamemBERT & 91.8 & 92.7 & 94.1 & 93.9 & 95.2 & 95.1 & 95.6 & 95.1\\
-& CamemBERT.pretrained & 92.2 & 93.0 & 94.3 & 94.2 & 95.0 & 95.1 & 95.5 & 95.4\\
+    \begin{tabular}{rrrrrrrrrr}
+           & Trainset Size &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
+           & \% & 0.8   & 1.5   & 3.1   & 6.2   & 12.5  & 25.0  & 50.0  & 100.0 \\
+    \midrule\bottomrule
+    \multirow{3}{*}{\rotatebox{90}{F1 score}} & CamemBERT &  89.6 &  90.1 &  92.8 &  93.4 &  94.3 &  95.3 &  94.7 &  95.5 \\
+           & CamemBERT.pretrained &  90.0 &  91.4 &  92.9 &  93.4 &  94.2 &  94.5 &  94.8 &  95.4 \\
+           & SpaCy NER &  87.0 &  89.0 &  90.3 &  91.9 &  92.1 &  92.8 &  93.2 &  93.5 \\
+    \cline{1-10}
+    \multirow{3}{*}{\rotatebox{90}{Precision}} & CamemBERT &  87.8 &  87.8 &  91.7 &  93.0 &  93.4 &  95.4 &  93.8 &  95.8 \\
+           & CamemBERT.pretrained &  87.9 &  89.8 &  91.6 &  92.7 &  93.4 &  94.0 &  94.1 &  95.4 \\
+           & SpaCy NER &  85.6 &  87.7 &  90.0 &  92.0 &  92.4 &  92.8 &  93.1 &  93.7 \\
+    \cline{1-10}
+    \multirow{3}{*}{\rotatebox{90}{Recall}} & CamemBERT &  91.5 &  92.5 &  93.9 &  93.9 &  95.2 &  95.1 &  95.6 &  95.1 \\
+           & CamemBERT.pretrained &  92.2 &  93.0 &  94.3 &  94.2 &  95.0 &  95.1 &  95.5 &  95.4 \\
+           & SpaCy NER &  88.6 &  90.4 &  90.7 &  91.7 &  91.9 &  92.8 &  93.3 &  93.4 \\
     \end{tabular}    
+    \label{tab:my_label}
 \end{table}

### 2022-01-14 src/ocr/isri_tools_src/accuracy.cpp
Fix latin-1/UTF-8 encoding report

@@ -206,14 +206,16 @@ print_accurary_report (Accdata *data)
   int old_desc = dup (fileno (stdout));
   dup2 (fileno(tmp), fileno (stdout));
   write_accrpt (data, nullptr);
+  fflush(stdout);
   dup2 (old_desc, fileno (stdout));
 
+  fflush(tmp);
   rewind(tmp);
   std::wstring result;
+  // The file 'tmp' is encoded in latin-1 -> codepoint are compatible with utf-16 but not utf-8 
   wchar_t c;
-  while ((c = fgetwc(tmp)) != EOF)
+  while ((c = fgetc(tmp)) != EOF)
     result.push_back(c);
-
   fclose(tmp);
   return result;
 }

### 2022-01-15 src/ocr/test_text_utils.py
Add simple function to get charset stats

@@ -1,6 +1,7 @@
 
 from text_utils import (
-    UNICODE_SIMPLIFICATIONS_SINGLE_CHAR, 
+    UNICODE_SIMPLIFICATIONS_SINGLE_CHAR,
+    charset_stats, 
     simplify_unicode_charset,
     check_alignment_charset,
     add_tags_prediction,
@@ -100,3 +101,23 @@ def test_fix_manual_ner_xml():
     XML = "<PER>Dufay</PER>, <ACT>papetier</ACT>, <LOC>r. S.-Martin</LOC>, <CARDINAL>20</CARDINAL>. <CARDINAL>437</CARDINAL>"
     XML2 = fix_manual_ner_xml(XML)
     assert(XML2 == XML)
+
+
+def test_charset_stats():
+    import pandas as pd
+    in1 = ["abc"]
+    res1 = charset_stats(in1)
+    assert(isinstance(res1, pd.DataFrame))
+    assert(set(res1["ord"]) == set(ord(x) for x in in1[0]))
+
+    # check invalid chars
+    from itertools import chain
+    in2 = ["abc", "def\n\uffff\ud800"]
+    res2 = charset_stats(in2)
+    assert(isinstance(res2, pd.DataFrame))
+    assert(set(res2["ord"]) == set(ord(x) for x in chain(*in2)))
+
+    # check generator input
+    res3 = charset_stats(s.upper() for s in in2)
+    assert(isinstance(res3, pd.DataFrame))
+    assert(set(res3["ord"]) == set(ord(x.upper()) for x in chain(*in2)))

### 2022-01-15 src/ocr/text_utils.py

@@ -6,12 +6,13 @@
 from collections import Counter
 import io
 import logging
-from typing import List, Tuple
+from typing import List, Tuple, Iterable
 import unicodedata
 from warnings import warn
 
 import isri_tools
 import numpy as np
+import pandas as pd
 import regex
 
 # ==============================================================================
@@ -269,6 +270,10 @@ def simplify_unicode_charset(text: str) -> str:
 # ==============================================================================
 # Check OCR reference charset
 # TODO extract codepoint ranges to some contant list
+
+def _unichr2str(unichar: str) -> str:
+    return unicodedata.name(unichar, repr(unichar))
+
 def check_alignment_charset(text_to_align: str, dump_charset: bool=False, show_all: bool=False) -> bool:
     """Check the string passed as parameter contains only acceptable
     unicode code points for the alignment.
@@ -285,8 +290,6 @@ def check_alignment_charset(text_to_align: str, dump_charset: bool=False, show_a
     Returns:
         bool: Return True if the text has a valid charset, False otherwise.
     """
-    def _unichr2str(unichar: str) -> str:
-        return unicodedata.name(unichar, repr(unichar))
 
     def _char_str(char_str):
         codepoint = ord(char_str)
@@ -846,3 +849,13 @@ def xml_contains_empty_tags(unescaped_xml: str, tag_list=TAG_LIST) -> bool:
 
 def xml_remove_empty_tags(unescaped_xml: str, tag_list=TAG_LIST) -> bool:
     return regex.sub(REGEX_EMPTY_TAGS, r'\2', unescaped_xml, tag=tag_list, flags=regex.MULTILINE)
+
+def charset_stats(texts: Iterable[str]) -> pd.DataFrame:
+    c = Counter()
+    for t in texts:
+        c.update(t)
+    
+    return pd.DataFrame(
+        ((repr(char), ord(char), _unichr2str(char), unicodedata.category(char), count) for char, count in c.most_common()),
+        columns=("repr", "ord", "name", "category", "count")
+    )

### 2022-01-16 src/ner/camembert_util.py
Full pipeline with notebooks

@@ -28,12 +28,12 @@ LABELS_ID = {
 
 
 # Entry point
-def init_model(model_name, training_config, save_model_path=None):
+def init_model(model_name, training_config):
     logger.info(f"Model {model_name}")
-    tokenizer = AutoTokenizer.from_pretrained(model_name)  # model_name is implicit    
+    tokenizer = AutoTokenizer.from_pretrained(model_name)
     
-    output_path = save_model_path or "/tmp/bert-model"    
-    training_args = TrainingArguments(output_path, **training_config)
+    #output_path = save_model_path or "/tmp/bert-model"    
+    training_args = TrainingArguments(**training_config)
 
     # Load the model
     model = AutoModelForTokenClassification.from_pretrained(
@@ -46,7 +46,7 @@ def init_model(model_name, training_config, save_model_path=None):
     return model, tokenizer, training_args
 
 # Main loop
-def train_eval_loop(model, training_args, tokenizer, train, dev, test):
+def train_eval_loop(model, training_args, tokenizer, train, dev, test, patience=3):
     data_collator = DataCollatorForTokenClassification(tokenizer)
     trainer = Trainer(
         model,
@@ -57,7 +57,7 @@ def train_eval_loop(model, training_args, tokenizer, train, dev, test):
         tokenizer=tokenizer,
         compute_metrics=compute_metrics,
         callbacks=[
-            EarlyStoppingCallback(early_stopping_patience=3)
+            EarlyStoppingCallback(early_stopping_patience=patience)
         ],
     )
     trainer.train()

### 2022-01-17 src-latex/figs/protocol.tex
Add protocol figs

\documentclass[tikz,svgnames]{standalone}
\usetikzlibrary{matrix,positioning}

\begin{document}
\begin{tikzpicture}
  \tikzset{
    title/.style={every node/.style={draw=none, font=\itshape}},
    qa/.style={circle, fill=DarkGreen, text width=5pt}
  };
  \matrix[matrix of nodes, nodes={draw, align=center, text width=3cm}, column sep=2cm, row sep=.5cm, row 1/.style=title] (A)
  {
    & OCR systems &    & NER systems &    \\
    & |(pero)| Pero OCR        &    &    &    \\
    & |(tess)| Tesseract          &   & |(cam)| CamemBERT &   \\
    |(a)| Layout \& Entry extraction &           & |[qa] (b)|  & & |[qa] (c)|  \\
    & |(kraken)| Kraken   &    & |(spacy)| Spacy       &    \\
    & |(gt)| \emph{GroundTruth}   &    &        &    \\
  };

  \node [above=0cm of b, align=center] {OCR\\ QA};
  \node [above=0cm of c, align=center] {NER\\ QA};

  \begin{scope}[-latex]
{\draw (a.east) -- (pero.west); \draw (pero.east)  -- (b); }
{\draw (a.east) -- (tess.west); \draw (tess.east)  -- (b); }
{\draw (a.east) -- (kraken.west); \draw (kraken.east)  -- (b); }
{\draw[dashed] (a.east) -- (gt.west); \draw[dashed] (gt.east)  -- (b); }

{\draw (b) -- (cam.west);  \draw (cam.east) -- (c); }
{\draw (b) -- (spacy.west);  \draw (spacy.east) -- (c); }

\end{scope}
\end{tikzpicture}

\end{document}

### 2022-01-17 src-latex/parts/50-xp-setup.tex
One sentence per line
Add protocol figs

@@
 % !TeX root = ../main-paper.tex
-\section{Experimental Setup}
+\section{Experiment 1: NER sensibility to the number of training samples}
 
-\subsection{Metrics}
+\begin{figure}[tb]
+\includegraphics[width=\linewidth]{figs/protocol.pdf}
+\caption{Scheme of the evaluation protocol.}
+\label{fig.protocol}
+\end{figure}
+
+\subsection{Evaluation protocol \& Metrics}
+
+\Cref{fig.protocol} depicts the evaluation procotol used to assess the OCR and NER systems. Two quality evaluation
+are performed in the pipeline named respectively \emph{OCR Q.A.} and \emph{NER Q.A.}.
+
+First the layout extraction and the entry segmentation of the page are performed with a semi-automated system and
+checked by a human. Afterward, an OCR system runs on the thumbnails of each segmented entry to extract its text. An
+entry might span over multiple text lines but is always single-block. Thus, the most adapted mode is chosen when
+the OCR system allows for detection mode (e.g. the \emph{block} mode for \emph{Tesseract}).
+
+\textbf{OCR Q.A.} is performed after some text normalization of the OCR system outputs. Text normalization consists in
+projecting unicode characters onto the latin-1 charset (whenever it makes sense) and removing extra symbols (hands,
+crosses). Then, the predicted text is aligned with the reference text using standard tools and the Character Error Rates
+(CER) are computed at entry level and at global level.
+
+\begin{align}
+\mathrm{CER} &=  \frac{\#Errors}{\text{Reference Text Length}} & & \mathrm{CER}_\mathrm{norm} =  \frac{\#Errors}{\text{Alignment Length}} 
+\end{align}
+
+In this benchmark, we consider 3 OCR systems well-known for the historical document analysis: Pero OCR, Tesseract and Kraken.
+
+\edwin{FIXME: add ref papier eval OCR + ISRI, eval KRAKEN}
+
+Next, a NER system extracts the named entities from the text of each entry outputed by the OCR and the \textbf{NER Q.A.}
+is performed. The NER system outputs a text with tags that enclose the entities. To assess the quality of the entity
+extraction, we rely on the same technique as for the OCR evaluation. The predicted text is aligned with the reference
+text and the tags are projected in the alignment. Then, the CER is computed between each entity of the alignment and the
+groundtruth as illustrated on \cref{fig.eval-ocr-ner}.
+
 \bertrand{Move after section Cases studies? }
@@
 The first experiment evaluates the performances of the three models on training sets of different sizes.
-To do so we split the gold reference into a training set, a development set and a test set.
+To do so, we split the gold reference into a training set, a development set, and a test set.
 The training set is then gradually reduced in size while maintaining the relative frequency of directories within.
@@
 
-As the form an structure of entries varies across directories collections and through time, the models may learn to overfit on a subset of directories with specific features.
-To reduce the evaluation bias we start by leaving out 3 directories (1690 entries, ~20\%) from the gold reference to test each model on unseen directories.
+As the form and structure of entries varies across directory collections and through time, the model may learn to overfit on a subset of directories with specific features.
+To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, ~20\%) from the gold reference to test each model on unseen directories.
 Then a stratified sampling based on the entry directory name is applied to the remaining set to create a training (6373 entries, ~71\% of the gold reference) and a development set (709 entries, ~8\%).

### 2022-01-17 src-latex/parts/50-xp-setup.tex
Updates from Overleaf

--- 2022-01-12 src-latex/parts/50-xp-setup.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Experimental Setup}
+\section{Experiment 1: NER sensibility to the number of training samples}
 
-\subsection{Metrics}
-\bertrand{Move after section Cases studies? }
-\begin{itemize}
-    \item NER quality: Precision, Recall, F1 score (idea: stick with standard spacy/sklearn metrics) \nathalie{Dans la plupart des articles que j'ai lus, c'est toujours une F-mesure qui est utilisée et parfois la précision et le rappel sont ajoutés.}
-    \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?) \nathalie{Pour l'instant, j'ai croisé: CER, WER, Levenshtein et Jaccard.
-J'ai encore un état de l'art à lire sur le sujet: il y aura probablement d'autres mesures mentionnées.}
-
-    \nathalie{Pour la partie Experiment 2: dans la litérature j'ai vu des courbes F-Mesure(CER) ou F-Mesure(WER) et des diagrammes avec F-Mesure calculée sur des sous-ensembles du corpus classés par score de Levenshtein.}
-\end{itemize}
-
-\subsection{Cases studies}
+\subsection{Case studies}
 \subsubsection{Experiment 1: NER sensibility to the number of training samples}
 The first experiment evaluates the performances of the three models on training sets of different sizes.
-To do so we split the gold reference into a training set, a development set and a test set.
+To do so, we split the gold reference into a training set, a development set, and a test set.
 The training set is then gradually reduced in size while maintaining the relative frequency of directories within.

### 2022-01-17 src-latex/parts/40-ner-approaches.tex
One sentence per line

@@
 \section{NER Approaches}
-Four approaches, one rules-based, three learning-based.
+
+Previous work on the influence of OCR errors on NER models shows that it is preferable to reuse a model trained on modern data and adapt it to OCRised historical texts using supervised or unsupervised approaches.
+This is why, in our tests, we have selected two deep learning-based NER approaches trained on French texts: Spacy 3 NER pipeline and CamemBERT.
 
@@
 Although the latest version of SpaCy (v3) leverages BERT models, we could not use them as no NER pipeline is available for the French language at the time of our experiments\footnote{https://spacy.io/models/fr}.
-The vanilla NER pipeline is two-fold.
-Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
+The vanilla NER pipeline is two-fold. Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
 The decision layer is an adaptation of the transition-based model presented in~\cite{lample2016}.
-As words are processed sequentially, their vectors are concatenated with those in the last known entities to encode the nearby predicted semantics as a dynamic attention mechanism.
-The classification layer relies on a finite-state machine whose transition probabilities are learned using a multi-layer perceptron.
-In 2018~\cite{won2018} compared SpaCy's vanilla NER performances on historical texts about places with several other software packages (Stanford-NER, Ner-Tagger, Edinburgh Geopoarser and Polyglot).
+As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics as a dynamic attention mechanism.
+The classification layer relies on a finite-state machine whose transition probabilities are learned using a multilayer perceptron.
+In 2018~\cite{won2018} compared SpaCy's vanilla NER performance on historical texts about places with several other software packages (Stanford-NER, Ner-Tagger, Edinburgh Geopoarser and Polyglot).
 Results showed SpaCy to be average with a F1-score of 0.57.
-The SpaCy developers claims an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
+The SpaCy developers claimed an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
 
@@
 
-\begin{itemize}
-    \item regexp
-    \item CamemBERT finetuned with wikiner-fr + gold dataset (clean texts)
-    \item CamemBERT finetuned with wikiner-fr + gold dataset (clean texts + pretrained on raw texts (Pero ? Kraken ? Both ?)
-\end{itemize}
+Language model embeddings pretrained using Transformers, like BERT, have become a new paradigm for NER\cite{li2020}.
+Indeed, these embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec and they can be further fine-tuned for NER task by adding an additional output layer.
+Moreover, they can also be further pretrained in an unsupervised way on historical texts for domain adaptation.
+
+As our directories are written in French, we chose the language model CamemBERT \cite{martin-etal-2020-camembert}, a Transformer model trained on a French corpus.
+We started from the CamemBERT-ner model already fine-tuned on WikiNER-fr corpus and published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}}.
+Its output layer is a linear model with a Softmax function.
+
+For our experiments, two models are derived. One is produced by fine-tuning this generic model on our ground truth dataset.
+The second is created by adapting the embeddings of this model using raw directory texts, randomly selected and extracted with PERO OCR.
+The base model is pretrained on these texts for two tasks, namely, Next Sentence Prediction (NSP) and Masked Language Model (MLM).
+Finally, the resulting model is also fine-tuned using our ground truth corpus.
 
 % Other possibilities:
-1. Traditional ML based:
-    Conditional Random Fields (CRF) - https://pypi.org/project/sklearn-crfsuite/
-    Maximum-entropy Markov model
+%1. Traditional ML based:
+%    Conditional Random Fields (CRF) - https://pypi.org/project/sklearn-crfsuite/
+%    Maximum-entropy Markov model
 
-2. Neural Networks based:
-    LSTMs, bi-LSTM - https://github.com/flairNLP/flair
-    CNNs (SpaCy uses CNN based architecture)
-    Transformers (Spacy has recently launched it) - https://spacy.io/universe/project/spacy-transformers
+%2. Neural Networks based:
+%    LSTMs, bi-LSTM - https://github.com/flairNLP/flair
+%    CNNs (SpaCy uses CNN based architecture)
+%    Transformers (Spacy has recently launched it) - %https://spacy.io/universe/project/spacy-transformers

### 2022-01-17 src-latex/parts/50-xp-setup.tex
Changed subsection titles in section 5

--- 2022-01-17 src-latex/parts/50-xp-setup.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Experiment 1: NER sensibility to the number of training samples}
+\section{Experimental Setup}
 
@@
 extraction, we rely on the same technique as for the OCR evaluation. The predicted text is aligned with the reference
-text and the tags are projected in the alignment. Then, the CER is computed between each entity of the alignment and the
+text and the tags are projected in the alignment.
+Then, the CER \nathalie{non ici ce sont la précision, le rappel et la F-mesure qui sont calculées} is computed between each entity of the alignment and the
 groundtruth as illustrated on \cref{fig.eval-ocr-ner}.
 
-\bertrand{Move after section Cases studies? }
-\begin{itemize}
-    \item NER quality: Precision, Recall, F1 score (idea: stick with standard spacy/sklearn metrics) \nathalie{Dans la plupart des articles que j'ai lus, c'est toujours une F-mesure qui est utilisée et parfois la précision et le rappel sont ajoutés.}
-    \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?) \nathalie{Pour l'instant, j'ai croisé: CER, WER, Levenshtein et Jaccard.
-J'ai encore un état de l'art à lire sur le sujet: il y aura probablement d'autres mesures mentionnées.}
-
-    \nathalie{Pour la partie Experiment 2: dans la litérature j'ai vu des courbes F-Mesure(CER) ou F-Mesure(WER) et des diagrammes avec F-Mesure calculée sur des sous-ensembles du corpus classés par score de Levenshtein.}
-\end{itemize}
-
-\subsection{Cases studies}
-\subsubsection{Experiment 1: NER sensibility to the number of training samples}
+\subsection{Experiment 1: NER sensibility to the number of training samples}
 The first experiment evaluates the performances of the three models on training sets of different sizes.
@@
 
-\subsubsection{Experiment 2: NER in the presence of noisy OCR texts}
+\subsection{Experiment 2: NER in the presence of noisy OCR texts}
 

### 2022-01-17 src/ocr/test_text_utils.py
Add OCR charset simplifications for evaluation(s)

@@ -1,7 +1,8 @@
 
 from text_utils import (
     UNICODE_SIMPLIFICATIONS_SINGLE_CHAR,
-    charset_stats, 
+    charset_stats,
+    ocr_simplifications_for_evaluation, 
     simplify_unicode_charset,
     check_alignment_charset,
     add_tags_prediction,
@@ -121,3 +122,33 @@ def test_charset_stats():
     res3 = charset_stats(s.upper() for s in in2)
     assert(isinstance(res3, pd.DataFrame))
     assert(set(res3["ord"]) == set(ord(x.upper()) for x in chain(*in2)))
+
+def test_ocr_simplifications_for_evaluation():
+    text = "Pinchon («J.A.”) [\u00e9l. de Vincent \u043Bet\u2029 \ufb03Augustin}, rue CaumÆœrtin,\nno. —744."
+    expected = "pinchon (\"j.a.\") (el. de vincent \u043Bet  ffiaugustin). rue caumaeoertin. no. -744."
+    output = ocr_simplifications_for_evaluation(
+                text,
+                normalize_spaces=True,
+                normalize_brackets=True,
+                normalize_dashes=True,
+                normalize_punctuation=True,
+                normalize_quotes=True,
+                casefold=True,
+                remove_accents=True,
+                skip_charset_norm=False)
+    assert(output == expected)
+
+    # idempotence
+    expected = "pinchon (\"j.a.\") (el. de vincent \u043Bet  ffiaugustin). rue caumaeoertin. no. -744."
+    output = ocr_simplifications_for_evaluation(
+                expected,
+                normalize_spaces=True,
+                normalize_brackets=True,
+                normalize_dashes=True,
+                normalize_punctuation=True,
+                normalize_quotes=True,
+                casefold=True,
+                remove_accents=True,
+                skip_charset_norm=False)
+    assert(output == expected)
+

### 2022-01-17 src/ocr/text_utils.py

@@ -117,8 +117,8 @@ UNICODE_SIMPLIFICATIONS_SINGLE_CHAR = str.maketrans({
     "\u2044": "/", # ⁄ FRACTION SLASH U+2044
 
     # quotes and brackets
-    "«": "\"", # U+00AB
-    "»": "\"", # U+00BB
+    "\u00AB": "\"", # « U+00AB
+    "\u00BB": "\"", # » U+00BB
     "\u2018": "\'", # ‘ LEFT SINGLE QUOTATION MARK
     "\u2019": "\'", # ’ RIGHT SINGLE QUOTATION MARK
     "\u201A": ",", # ‚ SINGLE LOW-9 QUOTATION MARK
@@ -754,71 +754,95 @@ def add_tags_prediction(ner_xml_final: str, text_ocr_final: str, debug=False) ->
 
 
 # ==============================================================================
-# TODO OCR eval
+# OCR eval
+# Expected input: OCR text with minimal simplifications already performed
+# Output: OCR text with many simplifications (suppressed, changed and expended chars)
+# Transformation: 1 char -> 0+ chars (can change line length)
 
-# TODO OCR equivalences
-# TODO case insensitive or not
-# TODO check charset
-# TODO call isri bindings
-
-
-
-OCR_EQUIVALENCE_MAP = [
-# ¹ U+00B9 to U+0031  1   <---- use NFKC for this?
-# ² U+00B2 to U+0032  2
-# ³ U+00B3 to U+0033  3
-# µ U+00B5 to U+03BC  μ
-    # ("\u00BA", "\u006F"), # º U+00BA to U+006F  o
-    # …   U+2026 to U+002E U+002E U+002E    ...  <-------- ???
-
+OCR_EQUIVALENCE_MAP = str.maketrans({
+    "\u00BA": "\u006F",  # º U+00BA to U+006F  o
+    # "\u2026": "...",  # …   U+2026 to U+002E U+002E U+002E  # Already done
 # ™   U+2122 to U+0054 U+004D   TM
-    # ("\uFB00", "ff"),  # Replace ff, fi, fl, ffi, ffl ligatures with separated 
-    # ("\uFB01", "fi"),  # chars. before Unicode normalization inserts
-    # ("\uFB02", "fl"),  # "200c ZERO WIDTH NON-JOINER"
-    # ("\uFB03", "ffi"), # (actually not wrote to UTF-8 output so not done here)
[…]

### 2022-01-17 src-latex/parts/50-xp-setup.tex
ner models eval : tables & figures

@@
 
-To generate smaller training sets we start from the initial training set and attractively split it in half using the same stratified sampling strategy.
+To generate smaller training sets we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as before.
 We stop if there is only one entry left in a directory or if the current train set contains less than 30 entries.
@@
 % Move to metrics ?
-All metrics are evaluated on the test set, yet the biased model performances on the development sets are add in the paper material as additional information.
+All metrics are evaluated on the test set, yet the biased model performances on the development sets are added in the article material as additional information.
 
-\subsection{Experiment 2: NER in the presence of noisy OCR texts}
+\subsection{Experiment 2: NER in noisy OCRed texts}
+This experiment aims to evaluate the impact of the OCR used on the quality of the NER results.
+To this end, we use directory entries whose text has been extracted, corrected and annotated manually (see Section 3).
+This reference text is aligned with the text produced on the same entries by Tesseract and PERO OCR.
+The named entities annotated in the reference text are then projected into the OCRed texts.
+To make sure that all training, development and test datasets contain the same annotations, only the entries for which the same number of annotations are found in the reference text and in the OCRised texts are kept.Finally, the reference text is divided into three datasets, using the same method as in Experiment 1.
+Two additional test files are generated with exactly the same entries as the test file produced from the reference text: one from the text produced by PERO OCR and one from that produced by Tesseract.
+
+In this experiment, the two best models of Experiment 1 are kept.
+Both of them are fine-tuned on the training dataset produced from the reference text.
+One of them is also pretrained on 845000 entries automatically extracted by PERO OCR.
 

### 2022-01-17 src-latex/parts/60-results.tex
Start OCR eval part
ner models eval : tables & figures

@@
 \section{Results}
+
+\subsection{OCR evaluation}
+
+\begin{figure}
+
+\subcaptionbox{}[.5\linewidth]{
+\begin{tabular}{rll}
+          & CER & CER$_{norm.}$ \\
+\midrule
+Pero      & 3.78\% & 3.76\% \\   
+Tesseract & 6.56\% & 6.45\% \\
+\bottomrule
+\end{tabular}
+
+\bigskip
+
+\includegraphics[width=\linewidth]{images/ocr-eval-2.pdf}
+}
+\subcaptionbox{}[.5\linewidth]{
+\includegraphics[width=\linewidth]{images/ocr-eval-1.pdf}
+}
+\caption{Character error rates at entry-level for Pero OCR and Tesseract. (a) Global CER and distribution of the CER per entry. (b):
+joint plot of the dataset showing that the two systems do not fail on the same entries.}  
+\end{figure}
 
@@
 
-\begin{table}[!h]
-    \caption{Experimental results of the NER models performances when trained on varying numbers of examples.}
+\begin{table}[h!]
 \centering
-    \begin{tabular}{rrrrrrrrrr}
+\caption{Experimental results of the NER models performances when trained on varying numbers of examples}
+\begin{tabular}{llrrrrrrrr}
        & Trainset Size &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
-           & \% & 0.8   & 1.5   & 3.1   & 6.2   & 12.5  & 25.0  & 50.0  & 100.0 \\
+       & \% & 0.8   & 1.6   & 3.1   & 6.2   & 12.5  & 25.0  & 50.0  & 100.0 \\
 \midrule\bottomrule
-    \multirow{3}{*}{\rotatebox{90}{F1 score}} & CamemBERT &  89.6 &  90.1 &  92.8 &  93.4 &  94.3 &  95.3 &  94.7 &  95.5 \\
-           & CamemBERT.pretrained &  90.0 &  91.4 &  92.9 &  93.4 &  94.2 &  94.5 &  94.8 &  95.4 \\
+\multirow{3}{*}{\rotatebox{90}{F1 score}} & Camembert &  89.5 &  90.5 &  92.7 &  93.3 &  94.1 &  94.9 &  94.6 &  95.1 \\
+       & Camembert-pretrained &  92.4 &  92.9 &  93.7 &  93.8 &  93.8 &  94.0 &  94.6 &  94.4 \\
        & SpaCy NER &  87.0 &  89.0 &  90.3 &  91.9 &  92.1 &  92.8 &  93.2 &  93.5 \\
 \cline{1-10}
-    \multirow{3}{*}{\rotatebox{90}{Precision}} & CamemBERT &  87.8 &  87.8 &  91.7 &  93.0 &  93.4 &  95.4 &  93.8 &  95.8 \\
-           & CamemBERT.pretrained &  87.9 &  89.8 &  91.6 &  92.7 &  93.4 &  94.0 &  94.1 &  95.4 \\
+\multirow{3}{*}{\rotatebox{90}{Precision}} & Camembert &  87.4 &  88.7 &  91.5 &  92.7 &  93.3 &  94.9 &  93.9 &  95.1 \\
+       & Camembert-pretrained &  91.2 &  91.9 &  92.9 &  93.0 &  93.0 &  93.2 &  94.2 &  93.9 \\
        & SpaCy NER &  85.6 &  87.7 &  90.0 &  92.0 &  92.4 &  92.8 &  93.1 &  93.7 \\
 \cline{1-10}
-    \multirow{3}{*}{\rotatebox{90}{Recall}} & CamemBERT &  91.5 &  92.5 &  93.9 &  93.9 &  95.2 &  95.1 &  95.6 &  95.1 \\
-           & CamemBERT.pretrained &  92.2 &  93.0 &  94.3 &  94.2 &  95.0 &  95.1 &  95.5 &  95.4 \\
+\multirow{3}{*}{\rotatebox{90}{Recall}} & Camembert &  91.6 &  92.5 &  93.9 &  93.9 &  94.9 &  94.9 &  95.4 &  95.1 \\
+       & Camembert-pretrained &  93.7 &  94.0 &  94.4 &  94.7 &  94.6 &  94.8 &  95.0 &  94.9 \\
        & SpaCy NER &  88.6 &  90.4 &  90.7 &  91.7 &  91.9 &  92.8 &  93.3 &  93.4 \\
 \end{tabular}
-    \label{tab:my_label}
+\label{tab:experiment-1-models-performances}
 \end{table}
 
-\begin{figure}[htb!]
+\begin{figure}[h!]
 	   \center{\includegraphics[width=\textwidth]
-	       {../material/experiment_1/f1_vs_trainsize.png}}
+	       {images/experiment-1-models-performances.pdf}}
 	  \vspace{3in}
@@
 Table : NER NN models VS train on {noisy, clean} datasets VS evaluate on {noisy, clean} datasets
+
+\begin{tabular}{lrrrrrr}
+{} & \multicolumn{3}{c}{Camembert.pretrained} & \multicolumn{3}{c}{Camembert} \\
+{} &                  ref & pero-ocr & tesseract &       ref & pero-ocr & tesseract \\
+\midrule
+Precision &                 96.5 &     93.3 &      90.3 &      96.1 &     92.4 &      89.7 \\
+Recall    &                 96.5 &     92.6 &      91.3 &      96.6 &     92.5 &      91.2 \\
+F1 score        &                 96.5 &     92.9 &      90.8 &      96.4 &     92.5 &      90.5 \\
+\bottomrule
+\end{tabular}
 

### 2022-01-17 src-latex/figs/eval-ocr-ner.tex
Add evaluation illutration

\documentclass[tikz,svgnames]{standalone}
\usepackage{newunicodechar}
\usepackage{amssymb,pifont}

\usetikzlibrary{matrix,positioning,calc}
\newunicodechar{⎵}{\verbvisiblespace}
\newunicodechar{❌}{{\color{red}\scriptsize\ding{53}}}

\begin{document}
\Pifont{pcr}

\begin{tikzpicture}
  \tikzset{alignment/.style={
      matrix of nodes, nodes={font=\tt, inner sep=0pt, minimum height=15pt}, column sep=0
    },
    taggedalignment/.style={alignment,
      column 1/.style={nodes={fill=red!20}},
      column 3/.style={nodes={fill=green!20}},
      column 5/.style={nodes={fill=blue!20}},
      column 7/.style={nodes={fill=yellow!20}},
    }
  }

  \node (A) { \verb|Alubert (Prosper),horlogerie,r, de la Ius-sienne,78| };

  \matrix[taggedalignment] (B) [right=3cm of A]
  {
    Alibert (Prosper)&,~ &horlogerie&,~ &r.~de la Jus- sienne&,~ &16&.\\
  };

  \matrix[alignment, row 2/.style={nodes={fill=none, gray!20}}] (C) [below=1.5cm of A]
  {
    Alibert (Prosper)&,~ &horlogerie&,~ &r.~de la Jus- sienne&,~ &16&.\\
    {||❌||||||||||||||}&{|❌}&{||||||||||}&{|❌}&{|❌❌||||||❌|||❌||||||}&{|❌}&{❌❌}&{❌}\\
    Alubert (Prosper)&,⎵ &horlogerie&,⎵  &r, de la Ius-⎵sienne&,7&8⎵ & ⎵\\
  };

  \matrix[taggedalignment, row 2/.style={nodes={fill=none, gray!20}}] (D) at ($ (C.east)!(B)!(C.west) $)
  {
    Alibert (Prosper)&,~ &horlogerie&,~ &r.~de la Jus- sienne&,~ &16&.\\
    {||❌||||||||||||||}&{|❌}&{||||||||||}&{|❌}&{|❌❌||||||❌|||❌||||||}&{|❌}&{❌❌}&{❌}\\
    Alubert (Prosper)&,⎵ &horlogerie&,⎵  &r, de la Ius-⎵sienne&,7&8⎵ & ⎵\\
  };

  \matrix[taggedalignment] (E1) [below=of D]
  {
    Alubert (Prosper)&,⎵ &horlogerie&,⎵  &r, de la Ius-⎵sienne&,7&8⎵ & ⎵\\
  };

  \matrix[taggedalignment] (E2) [below=of E1.west, anchor=west]
  {
    Alubert& (Prosper),⎵ &horlogerie&,⎵  &r, de la Ius-⎵sienne&,&78&⎵⎵\\
  };

  \node [above left=0cm of A, font=\it] {OCR Prediction};
  \node [above left=0cm of B, font=\it] {GOLD};

  \draw[-latex] (A.west) -- ++ (-1,0) |- (C-3-1) node[pos=0.25,left] { Alignment };

  \draw[-latex] (B) -- (D) node[midway,right] { Tag projection };
  \draw[-latex] (B.west) -- ++ (-1,0) |- (C-1-8);
  \draw[-latex] (C) -- (D);

  \draw[-latex] (C.south) -- ++(0,-1) node[midway,right,font=\small] {OCR Q.A.} node[below] {CER = 24\%};
  \draw[-latex] (D.south) -- ++(0,-1) node[midway,right,font=\small] {NER gt.};
  \draw[-latex] (C.south east) |- (E2) node[pos=0.75,below, font=\it] {NER Prediction};
  \draw[-latex] (E2.south) -- ++(0,-1) node[midway,right,font=\small] {NER Q.A.} node[below] {FIXME\%};
\end{tikzpicture}

\end{document}

### 2022-01-17 src-latex/parts/50-xp-setup.tex
Add evaluation illutration

@@
 
-\subsection{Experiment 1: NER sensibility to the number of training samples}
+\begin{figure}[tb]
+    \includegraphics[width=\linewidth]{figs/eval-ocr-ner.pdf}
+    \label{fig.eval-ocr-ner}
+    \caption{OCR and NER evaluation protocol example.}
+\end{figure}
+
+\bertrand{Move after section Cases studies? }
+\begin{itemize}
+    \item NER quality: Precision, Recall, F1 score (idea: stick with standard spacy/sklearn metrics) \nathalie{Dans la plupart des articles que j'ai lus, c'est toujours une F-mesure qui est utilisée et parfois la précision et le rappel sont ajoutés.}
+    \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?) \nathalie{Pour l'instant, j'ai croisé: CER, WER, Levenshtein et Jaccard.
+J'ai encore un état de l'art à lire sur le sujet: il y aura probablement d'autres mesures mentionnées.}
+
+    \nathalie{Pour la partie Experiment 2: dans la litérature j'ai vu des courbes F-Mesure(CER) ou F-Mesure(WER) et des diagrammes avec F-Mesure calculée sur des sous-ensembles du corpus classés par score de Levenshtein.}
+\end{itemize}
+
+\subsection{Cases studies}
+\subsubsection{Experiment 1: NER sensibility to the number of training samples}
 The first experiment evaluates the performances of the three models on training sets of different sizes.

### 2022-01-17 src-latex/figs/eval-ocr-ner.tex
Update eval with precision/recall/fscore

@@
   {
-    Alubert& (Prosper),⎵ &horlogerie&,⎵  &r, de la Ius-⎵sienne&,&78&⎵⎵\\
+    & Alubert (Prosper),⎵ &horlogerie&,⎵  &r, de la Ius-⎵sienne&,&78&⎵⎵\\
   };
@@
   \draw[-latex] (C.south) -- ++(0,-1) node[midway,right,font=\small] {OCR Q.A.} node[below] {CER = 24\%};
-  \draw[-latex] (D.south) -- ++(0,-1) node[midway,right,font=\small] {NER gt.};
+  \draw[-latex] (D.south) -- ++(0,-1) node[midway,right,font=\small] {NER reference.};
   \draw[-latex] (C.south east) |- (E2) node[pos=0.75,below, font=\it] {NER Prediction};
-  \draw[-latex] (E2.south) -- ++(0,-1) node[midway,right,font=\small] {NER Q.A.} node[below] {FIXME\%};
+  \draw[-latex] (E2.south) -- ++(0,-1) node[midway,right,font=\small] {NER Q.A.}
+  node[below] {Precision = 33\%, Recall = 25\%, F-score = 0.4};
 \end{tikzpicture}

### 2022-01-17 src-latex/parts/50-xp-setup.tex
Update eval with precision/recall/fscore

@@
 
-Next, a NER system extracts the named entities from the text of each entry outputed by the OCR and the \textbf{NER Q.A.}
+Next, a NER system extracts the named entities from the text of each entry output by the OCR and the \textbf{NER Q.A.}
 is performed. The NER system outputs a text with tags that enclose the entities. To assess the quality of the entity
-extraction, we rely on the same technique as for the OCR evaluation. The predicted text is aligned with the reference
-text and the tags are projected in the alignment.
-Then, the CER \nathalie{non ici ce sont la précision, le rappel et la F-mesure qui sont calculées} is computed between each entity of the alignment and the
-groundtruth as illustrated on \cref{fig.eval-ocr-ner}.
+extraction, we rely on a technique similar as for the OCR evaluation. The predicted text is aligned with the reference
+text and the tags are projected in the alignment. Then, the precision, recall and F-measure are computed considering
+only the exact matches between entities.
+
+\begin{align}
+    \mathrm{precision} &= \frac{\text{\#exact matches}}{\text{\#entries in prediction}} & \mathrm{recall} &= \frac{\text{\#exact matches}}{\text{\#entries in reference}}
+\end{align}
+
+The whole process is illustrated on~\cref{fig.eval-ocr-ner}. The OCR text and the GOLD text are first aligned to
+evaluate the OCR accuracy. As there are 11 mismatches over 56 aligned characters, the CER is thus 24\%. This alignment
+is then used to back-project the GOLD tags to build the tagged NER reference string. Finally, the NER system runs on the
+OCR text; its output is compared to the NER reference string. There is only 2 over 3 tags matching in the prediction (precision),
+while only 2 over 4 entities are matching in the reference (recall). It follows an overall f-score of 0.4.
 

### 2022-01-17 src-latex/main-paper.tex
Updates from Overleaf
more details in setup and results
Finished section 3.2
Start OCR eval part
Add protocol figs

@@
 \usepackage{orcidlink}
+\usepackage{subcaption} % Addition
+\usepackage{multirow} % Addition
+\usepackage{tabularx} % Addition
+\usepackage{booktabs} % Addition
+\usepackage{scalerel}
+\usepackage{cleveref}
 
@@
 % \renewcommand{\baselinestretch}{0.97}
+
+% Comment 
+% FIXME : delete before submission :)
+\newcommand{\bertrand}[1]{\textcolor{violet}{Bertrand:~#1}}
+\newcommand{\nathalie}[1]{\textcolor{purple}{Nathalie:~#1}}
+\newcommand{\edwin}[1]{\textcolor{pink}{Edwin:~#1}}
+
+% Customize table rendering
+\setlength{\tabcolsep}{5pt}
+\renewcommand{\arraystretch}{1.4}
 
@@
 %
-\institute{INST, CITY, COUNTRY \and
+\institute{LASTIG, Univ. Gustave Eiffel, IGN-ENSG, F-94160 Saint-Mandé, France\\
+\email{nathalie-f.abadie@ign.fr}\\
+\url{https://www.umr-lastig.fr/nathalie-abadie/} \and
 INST, CITY, COUNTRY\\
@@
 \url{https://www.example.com} \and
-INST, CITY, COUNTRY\\
-\email{example@example.com}}
+CRH-EHESS, Paris, France\\
+\email{bertrand.dumenieu@ehess.fr}\\ \url{http://crh.ehess.fr/index.php?5206}}
 %
@@
 
+\section*{Acknowledment}
+This work is supported by the French National Research Agency (ANR), as part of the SODUCO project, under Grant ANR-18-CE38-0013.
+
 % ---- Bibliography ----

### 2022-01-17 src-latex/parts/50-xp-setup.tex
Updates from Overleaf

--- 2022-01-17 src-latex/parts/50-xp-setup.tex
@@
 % Move to metrics ?
-All metrics are evaluated on the test set, yet the biased model performances on the development sets are added in the article material as additional information.
+All metrics are evaluated on the test set, yet the biased model performances on the development sets are added in the article material for information.
 
@@
 The named entities annotated in the reference text are then projected into the OCRed texts.
-To make sure that all training, development and test datasets contain the same annotations, only the entries for which the same number of annotations are found in the reference text and in the OCRised texts are kept.Finally, the reference text is divided into three datasets, using the same method as in Experiment 1.
+To make sure that all training, development and test datasets contain the same annotations, only the entries for which the same number of annotations are found in the reference text and in the OCRised texts are kept.
+Finally, the reference text is divided into three datasets, using the same method as in Experiment 1.
 Two additional test files are generated with exactly the same entries as the test file produced from the reference text: one from the text produced by PERO OCR and one from that produced by Tesseract.
 
-In this experiment, the two best models of Experiment 1 are kept.
-Both of them are fine-tuned on the training dataset produced from the reference text.
-One of them is also pretrained on 845000 entries automatically extracted by PERO OCR.
-
-\begin{itemize}
-\item At least 1 run per dataset, more if possible.
-\item Only use the best model; should be CamemBERT+pretrained
-\item Train on reference gold, evaluate on reference gold AND ocr-gold (2 datasets)
-\item split train/dev/test the same way than experiment 1. Use the exact same splitting for datasets so the clean test set, pero test set and tesseract test sets contains the same entries.
-\end{itemize}
+%\begin{itemize}
+%\item At least 1 run per dataset, more if possible.
+%\item Only use the best model; should be CamemBERT+pretrained
+%\item Train on reference gold, evaluate on reference gold AND ocr-gold (2 datasets)
+%\item split train/dev/test the same way than experiment 1. Use the exact same splitting %for datasets so the clean test set, pero test set and tesseract test sets contains the %same entries.
+%\end{itemize}
 
@@
 
-\subsection{Parameters for each method}
+%\subsection{Parameters for each method}
 
-\subsection{OCR systems}
+\subsection{OCR systems parameters}
 \begin{itemize}
@@
 
-we use spacy  for the CNN implementation and Huggingface for BERT.
+EX1: Pour chaque modèle train sur les 8 datasets * 5 runs, avec mêmes paramètres pour l’entraînement: early stopping  (patience = 3)
 
-will we use FLAIR as well? $\rightarrow$ No
+Ex 2: On garde CamemBERT et CamemBERT pretrained fine-tuned sur le train clean (patience = 5), validé avec dev clean. Teste les deux modèles sur les trois datasets de test.

### 2022-01-17 src-latex/parts/60-results.tex
Updates from Overleaf

@@
 \centering
-\caption{Experimental results of the NER models performances when trained on varying numbers of examples}
+\caption{NER performances at entry-level for SpaCy NER, CamemBERT and CamemBERT-pretrained when fine-tuned with an increasing number of examples.}
 \begin{tabular}{llrrrrrrrr}
-       & Trainset Size &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
+       & Number of training examples &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
        & \% & 0.8   & 1.6   & 3.1   & 6.2   & 12.5  & 25.0  & 50.0  & 100.0 \\
 \midrule\bottomrule
-\multirow{3}{*}{\rotatebox{90}{F1 score}} & Camembert &  89.5 &  90.5 &  92.7 &  93.3 &  94.1 &  94.9 &  94.6 &  95.1 \\
-       & Camembert-pretrained &  92.4 &  92.9 &  93.7 &  93.8 &  93.8 &  94.0 &  94.6 &  94.4 \\
+\multirow{3}{*}{\rotatebox{90}{F1 score}} & Camembert &  89.5 &  90.5 &  92.7 &  93.3 &  \textbf{94.1} &  \textbf{94.9} &  \textbf{94.6} &  \textbf{95.1} \\
+       & Camembert-pretrained &  \textbf{92.4} &  \textbf{92.9} & \textbf{93.7} &  \textbf{93.8} &  93.8 &  94.0 &  \textbf{94.6} &  94.4 \\
        & SpaCy NER &  87.0 &  89.0 &  90.3 &  91.9 &  92.1 &  92.8 &  93.2 &  93.5 \\
@@
 	       {images/experiment-1-models-performances.pdf}}
-	  \vspace{3in}
-	  \caption{\label{fig:f1-vs-trainsize} Models F1 score on unseen data vs trainset size}
+	  \caption{\label{fig:f1-vs-trainsize} Precision, recall and f1 score from table~\ref{tab:experiment-1-models-performances}}.
 \end{figure}
@@
 
+\begin{table}
+\caption{CamemBERT and CamemBERT-pretrained NER performances at entry-level on manually corrected and noisy OCRerd entries when fine-tuned on manually corrected entries.}
 \begin{tabular}{lrrrrrr}
@@
 \bottomrule
-\end{tabular}
-
-\begin{table}[h!]
-\caption{Camembert vs noise}
-\centering
-\begin{tabular}{ll|cc|c}
- & & \multicolumn{2}{c|}{Training data} & \\
- & & noisy & clean &   \\ 
-\cline{1-4}
-\multirow{3}{*}{Test data}& noisy gold (Tesseract) & f1 & f1 & \\
-                            & noisy gold (Pero-OCR) & f1 & f1 & \\ 
-                            & reference gold & f1 & f1 & \\ 
-                            & reference gold & f1 & f1 & \\
-\cline{1-4}
 \end{tabular}

### 2022-01-17 src-latex/parts/40-ner-approaches.tex
ner models eval : tables & figures

@@
 This is why, in our tests, we have selected two deep learning-based NER approaches trained on French texts: Spacy 3 NER pipeline and CamemBERT.
-
-\subsubsection{Regexp}
-::TODO::
-Needed for experiment 2 ?
 
@@
 
-\subsubsection{CamemBERT \& CamemBERT+pretrained on raw entries}
-::TODO::
+\subsubsection{CamemBERT \& CamemBERT pretrained on raw entries}
 
@@
 For our experiments, two models are derived. One is produced by fine-tuning this generic model on our ground truth dataset.
-The second is created by adapting the embeddings of this model using raw directory texts, randomly selected and extracted with PERO OCR.
+The second is created by adapting the embeddings of this model using 845 000 raw directory entries, randomly selected and extracted with PERO OCR.
 The base model is pretrained on these texts for two tasks, namely, Next Sentence Prediction (NSP) and Masked Language Model (MLM).

### 2022-01-17 src-latex/parts/60-results.tex
adds metrics for the 5th (last) bert-pretrained run

@@
 \multirow{3}{*}{\rotatebox{90}{F1 score}} & Camembert &  89.5 &  90.5 &  92.7 &  93.3 &  \textbf{94.1} &  \textbf{94.9} &  \textbf{94.6} &  \textbf{95.1} \\
-       & Camembert-pretrained &  \textbf{92.4} &  \textbf{92.9} & \textbf{93.7} &  \textbf{93.8} &  93.8 &  94.0 &  \textbf{94.6} &  94.4 \\
+       & Camembert-pretrained &  \textbf{92.2} &  \textbf{92.9} &  \textbf{93.6} &  \textbf{93.8} &  93.8 &  94.1 &  \textbf{94.6} &  94.4 \\
        & SpaCy NER &  87.0 &  89.0 &  90.3 &  91.9 &  92.1 &  92.8 &  93.2 &  93.5 \\
@@
 \multirow{3}{*}{\rotatebox{90}{Precision}} & Camembert &  87.4 &  88.7 &  91.5 &  92.7 &  93.3 &  94.9 &  93.9 &  95.1 \\
-       & Camembert-pretrained &  91.2 &  91.9 &  92.9 &  93.0 &  93.0 &  93.2 &  94.2 &  93.9 \\
+       & Camembert-pretrained &  90.8 &  91.8 &  92.9 &  93.0 &  93.0 &  93.4 &  94.1 &  93.9 \\
        & SpaCy NER &  85.6 &  87.7 &  90.0 &  92.0 &  92.4 &  92.8 &  93.1 &  93.7 \\
@@
 \multirow{3}{*}{\rotatebox{90}{Recall}} & Camembert &  91.6 &  92.5 &  93.9 &  93.9 &  94.9 &  94.9 &  95.4 &  95.1 \\
-       & Camembert-pretrained &  93.7 &  94.0 &  94.4 &  94.7 &  94.6 &  94.8 &  95.0 &  94.9 \\
+       & Camembert-pretrained &  93.6 &  94.0 &  94.4 &  94.6 &  94.6 &  94.8 &  95.0 &  94.9 \\
        & SpaCy NER &  88.6 &  90.4 &  90.7 &  91.7 &  91.9 &  92.8 &  93.3 &  93.4 \\

### 2022-01-18 src-latex/parts/30-dataset.tex
Updates from Overleaf
Started to correct and reduce section 3

@@
 
-The directories we have to process are stored in different libraries in Paris\footnote{Citer les bibliothèques sources?}.
-They have been scanned independently from each other and are available in digital format with various levels of quality.
-Moreover, they cover a wide period and ave been produced by different publishers and printers.
-Their contents, indexes, layouts, methods of printing, etc.
-vary thus from one directory to another (see Figure \ref{fig:directories}).
+The directories to be processed are stored in different libraries in Paris\footnote{\textcolor{green}{Citer les bibliothèques sources?}}.
+They have been scanned independently and with various levels of quality.
+Moreover, they cover a wide period and have been produced by different publishers and printers.
+Therefore, their contents, indexes, layouts, methods of printing, etc.
+may vary a lot from one directory to another (see Figure \ref{fig:directories}).
 
@@
 
-In this work, we therefore made a selection among the available directories, trying our best to retain copies representative of the diversity of existing directories and of sufficient quality to produce a reliable training data set.
+In this work, we made a selection among the available directories, trying our best to retain copies representative of the diversity of existing directories and of sufficient quality to produce a reliable training data set.
 The Table \ref{tab:directories} lists the directories selected, their layout, the structure of their entries, and the number of annotated entries according to the indexes considered.
@@
 
-In order to have a ground truth corpus representative of the diversity of the directories to be processed, we annotated a few pages of each of the major directory series for which we have digitised copies.
-To that purpose, we used an annotation tool developed for our project and whose graphical user interface (GUI) is presented in Figure \ref{fig:annotator}.
-This tool loads images of directory pages.
-Then, it automatically detects each entry in a given page and builds a box around it : these boxes are shown in the left panel of the GUI.
-Then the text of each entry is extracted using an OCR tool and finally the named entities searched are annotated using a NER tool: the results are displayed respectively in the first and second tabs of the right panel of the GUI.
-The entry detection, OCR and NER tasks are performed automatically but as they are prone to errors, their results can be corrected manually through the GUI of the annotation tool.
-In the end, for each page processed, the bounding boxes of the entries that make it up, the exact text contained in each box, and the named entities that are mentioned are stored in a JSON file.
+To build our ground truth corpus, we annotated a few pages of each of the major directory series for which we have digitised copies, using a tool developed for our project (see Figure \ref{fig:annotator}).
+First, the image of a given directory page is loaded.
+Second, entries are automatically detected and boxes are drawn around them as shown in the left panel of the Graphical User Interface (GUI).
+Then the text of each entry is extracted by an OCR tool and finally named entities are annotated by a NER tool: the results are displayed respectively in the first and second tabs of the right panel of the GUI.
+The entry detection, OCR and NER tasks are performed automatically.
+But as they are prone to errors, their results can be corrected manually through the GUI.
+In the end, the entry bounding boxes, their text, and the named entities are stored in a JSON file.
 
@@
 \begin{itemize}
-    \item origin
-    \item illustration (maybe in first figure)
-    \item various numbers (of the original data and of the sample we extracted) : number of entries per pages, variety of directory entries, \dots
     \item statistics about entry lengths (histogram of number of char / tokens par entry)

### 2022-01-18 src-latex/parts/50-xp-setup.tex
Started to correct and reduce section 3

@@
 extraction, we rely on a technique similar as for the OCR evaluation. The predicted text is aligned with the reference
-text and the tags are projected in the alignment. Then, the precision, recall and F-measure are computed considering
-only the exact matches between entities.
+text and the tags are projected in the alignment. Then, the precision, recall and f-measure (or f-score) are computed considering
+only the exact matches between entities. Precision and recall are defined by equation (2); the f-measure in their harmonic mean.
 

### 2022-01-18 src-latex/parts/60-results.tex
Started to correct and reduce section 3

@@
 \begin{tabular}{llrrrrrrrr}
-       & Number of training examples &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
+       & Training examples &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
        & \% & 0.8   & 1.6   & 3.1   & 6.2   & 12.5  & 25.0  & 50.0  & 100.0 \\
@@
 \caption{CamemBERT and CamemBERT-pretrained NER performances at entry-level on manually corrected and noisy OCRerd entries when fine-tuned on manually corrected entries.}
-\begin{tabular}{lrrrrrr}
-{} & \multicolumn{3}{c}{Camembert.pretrained} & \multicolumn{3}{c}{Camembert} \\
-{} &                  ref & pero-ocr & tesseract &       ref & pero-ocr & tesseract \\
+\begin{tabular}{lrrrrrrrrr}
+Model & \multicolumn{3}{c}{Camembert.pretrained} & \multicolumn{3}{c}{Camembert} & \multicolumn{3}{c}{Camembert.pretrained} \\
+Train set & \multicolumn{3}{c}{ref} & \multicolumn{3}{c}{ref} & \multicolumn{3}{c}{pero-ocr} \\
+Test set &                  ref & pero-ocr & tesseract &       ref & pero-ocr & tesseract &                  ref & pero-ocr & tesseract \\
 \midrule
-Precision &                 96.5 &     93.3 &      90.3 &      96.1 &     92.4 &      89.7 \\
-Recall    &                 96.5 &     92.6 &      91.3 &      96.6 &     92.5 &      91.2 \\
-F1 score        &                 96.5 &     92.9 &      90.8 &      96.4 &     92.5 &      90.5 \\
-\bottomrule
+Precision &                 96.5 &     93.3 &      90.3 &      96.1 &     92.4 &      89.7 &                 94.3 &     93.6 &      89.9 \\
+Recall    &                 96.5 &     92.6 &      91.3 &      96.6 &     92.5 &      91.2 &                 96.1 &     94.5 &      92.5 \\
+F1 score        &                 96.5 &     92.9 &      90.8 &      96.4 &     92.5 &      90.5 &                 95.2 &     94.1 &      91.2 \\
 \end{tabular}

### 2022-01-18 src-latex/parts/20-related-work.tex
Reduced sections 2 and 3

@@
 
-The directory processing pipeline presented in \cite{bell2020automated} to locate old gas stations using mid-20th century directories includes an OCR step, done with Tesseract, and a NER step to identify company names and addresses within the extracted text, performed using rules built as regular expressions.
-This section reviews existing approaches to achieving each of these two tasks.
-It concludes with a review of the works assessing the effects of OCR quality on the NER performance and the proposed solutions to overcome this problem.
+The directory processing pipeline presented in \cite{bell2020automated} includes an OCR step, done with Tesseract, and a NER step to identify company names and addresses, performed using regular expressions.
+This section reviews existing OCR and NER approaches with historical texts and presents some works assessing the effects of OCR quality on the NER performance and the proposed solutions.
 
@@
 
-\nathalie{à compléter/retoucher}
+\nathalie{à compléter/retoucher?}
 
@@
 Many approaches have been designed to recognize named entities, ranging from handcrafted rules to supervised approaches \cite{nadeau2007}.
-Rule based approaches look for portions of text that match patterns based on word characteristics such as their case, whether they are digits or not, their part-of-speech tags, etc.
+Rule based approaches look for portions of text that match patterns based on word characteristics such as their case, their part-of-speech tags, etc.
 These patterns can be fully handcrafted like in \cite{bell2020automated} or generated by data mining approaches like in \cite{nouvel2011}.
 Rule based approaches can also combine a set of patterns with a set of dictionaries (gazetteers, author lists, etc.) that help recognizing named entities when an exhaustive lexicon of searched entities is available \cite{mansouri2008,maurel2011}.
-For example, the CasEN\footnote{\url{https://tln.lifat.univ-tours.fr/version-francaise/ressources/casen}} transducer cascade leverages syntactic lexical patterns and dictionaries to recognize Named Entities in French texts.
+%For example, the CasEN\footnote{\url{https://tln.lifat.univ-tours.fr/version-francaise/ressources/casen}} transducer cascade leverages syntactic lexical patterns and dictionaries to recognize Named Entities in French texts.
 Such kind of approaches achieve very good results when applied to specialized domain corpus - like directories - and when an exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
 
-Supervised approaches include both approaches implementing supervised learning algorithms with careful text feature engineering, and deep learning based approaches which automatically build their own features to classify words into one category of named entities or another.
+Supervised approaches include both approaches implementing supervised learning algorithms with careful text feature engineering, and deep learning based approaches which automatically build their own features to classify tokens into named entity categories.
 In recent years, the latter have grown dramatically, yielding state-of-the-art performances and establishing new baselines\cite{li2020}.
@@
 
-\cite{Labusch2020NamedED} applied a BERT-based NER model to historical texts in German, French, and English, produced by OCR.
-It is off-the-shelf multilingual BERT-based NER model, pretrained on more than 2 000 000 pages of OCRed historical texts in three languages and fine-tuned on historical texts NER ground truth, also in three languages.
-Eventually, a clear decrease in NER performance on OCRed texts is noted.
-Moreover, the article reports that this decrease is more important for English texts for which the OCR is of poorer quality than for texts in the other two languages.
+\cite{Labusch2020NamedED} pretrained an off-the-shelf multilingual BERT-based NER model on more than 2 000 000 pages of OCRed historical texts in German, French, and English.
+They fine-tuned it on historical texts NER ground truth, also in these three languages and tested it on OCRed historical texts in each language.
+Eventually, a clear decrease in NER performance on OCRed texts is noted, especially for the English texts for which the OCR is of poorer quality.
 
@@
 They worked on a corpus published by a post-OCR correction software company, made of many journal articles with different levels of OCR errors and their respective ground truths.
-For each OCRed article, the Word Error Rate (WER) is computed and the English model \textit{en-core-web-lg} provided by Spacy\footnote{\url{https://spacy.io/}} library is used to perform NER on \textit{Person}, \textit{GPE} and \textit{Date}.
-The performance of the NER model with respect to OCR quality is eventually assessed by calculating the F-measure for each NER class, and each article i.e., each WER value.
+For each OCRed article, the Word Error Rate (WER) is computed and the English model \textit{en-core-web-lg} provided by Spacy\footnote{\url{https://spacy.io/}} library is used to perform NER on \textit{Person}, \textit{GPE}\footnote{Geopolitical Entity} and \textit{Date}.
+The performance of the NER model with respect to OCR quality is eventually assessed by computing the F-measure for each NER class, and each article i.e., each WER value.
 \cite{hamdi2020assessing} performed a similar but more extensive evaluation on four different NER models: CoreNLP using Conditional Random Fields and three deep neural models, BLSTM-CNN, BLSTM-CRF, and BLSTM-CNN-CRF.
@@
 Finally, they applied each NER model to the progressively degraded versions of the corpus and computed the resulting F-measure.
-Overall, NER F-score drops from 90\% to 50\% when the WER increase from 8\% to 50\%.
-Besides, all NER models are sensitive to OCR errors, but models based on deep neural networks seem less prone to errors.
+Overall, NER F-measure drops from 90\% to 50\% when the WER increase from 8\% to 50\%. However, models based on deep neural networks seem less sensitive to OCR errors.
 
-Based on the observation that OCR errors necessarily affect the performance of NER models applied to historical texts, \cite{huynh2020use} and \cite{marz2021data} have proposed different approaches to reduce their negative impact.
-The former uses a spelling correction tool on several corpora with variable OCR error rates, to assess whether spelling correction has a beneficial impact on NER performance or not.
-The results showed that as long as OCR errors remain low (CER<2\% and WER<10\%), the F-measure of NER results remains stable.
+\cite{huynh2020use} and \cite{marz2021data} have proposed different approaches to reduce the negative impact of OCR errors on NER performance with historical texts.
+The former uses a spelling correction tool on several corpora with variable OCR error rates, to assess whether NER performance benefit from spelling corrections or not.
+As long as OCR errors remain low (CER<2\% and WER<10\%), the F-measure of NER results remains stable.
 It starts to decrease significantly when OCR errors exceed these thresholds.
@@
 Three different NER models are tested on three historical corpora, in French, English, and Dutch.
-The best results are produced by a model trained on clean modern data, including many  embeddings as input (FastText, character and multilingual Flair embeddings, BERT and historical embeddings computed with Flair on a historical corpus), and fine-tuned on a noisy historical ground truth.
+The best results are produced by a model trained on clean modern data, including embeddings computed with Flair on a historical corpus, and fine-tuned on a noisy historical ground truth.
 
@@
 On the one hand, if the OCR error rate is kept below a certain threshold, the NER models remain little impacted: it is therefore important to reduce OCR errors as much as possible, but a low error rate may remain acceptable.
-On the other hand, it seems beneficial to reuse a NER model trained on modern data and adapt it to historical texts using supervised or unsupervised approaches.
+On the other hand, reusing a NER model trained on modern data and adapted to historical texts using supervised or unsupervised approaches seems a good strategy.

### 2022-01-18 src-latex/parts/30-dataset.tex
Reduced sections 2 and 3

@@
 
-In this work, we made a selection among the available directories, trying our best to retain copies representative of the diversity of existing directories and of sufficient quality to produce a reliable training data set.
-The Table \ref{tab:directories} lists the directories selected, their layout, the structure of their entries, and the number of annotated entries according to the indexes considered.
+For training as testing purposes, our groundtruth dataset must be as representative as possible of the diversity of the available directories.
+The Table \ref{tab:directories} lists the directories selected, their layout, the structure of their entries, and the number of annotated entries according to the considered index.
+
 For example, the Didot directory of 1854 is organised according to three indexes: by name, by activity, and by street name.
-The structure of the following entry is characteristic of its index by name: "Batton (D.-A.) *, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47." This entry begins with a person's name, followed by the initials of their first name in parentheses.The star symbol indicates that this person was awarded the Légion d'Honneur.
+The structure of the following entry is characteristic of its index by name: "Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47." This entry begins with a person's name, followed by the initials of their first name in parentheses.The symbol indicates that this person was awarded the Légion d'Honneur.
 The following part of this entry represents the person's activity,i.e.
@@
 The street name and number where the person lives or carries out their activity are written at the end of the entry.
+
 These pieces of information constituting the structure of the entries are the basic data that we want to extract, deduplicate and structure in order to build a spatio-temporal database of XIX\textsuperscript{th} century Parisian inhabitants and businesses.
-With the exception of some potentially wordy activity descriptions, most of these pieces of information correspond to named entities.
-However, as Table \ref{tab:directories} shows, while most entries contain the same types of named entities, the order in which they appear and the way they are written vary from one directory to another.
-This argues for a supervised, rather than rule-based, NER approach, which requires annotated data.
+With the exception of some potentially wordy activity descriptions, they correspond to named entities.
+However, as Table \ref{tab:directories} shows, while most entries contain the same types of named entities, their order and the way they are written vary from one directory or index to another.
+To provide examples of each entry structure, pages from each type have thus been annotated\footnote{In this work, the street name index entries have not been included as they are very different from the others.
+The strategy to be adopted to process them without loss of quality for the NER model will be the subject of future work.}.
 
@@
 \subsection{Assessing the groundtruth quality}
+\nathalie{On va utiliser cette sous-section ou pas?}
+Our dataset is thus made of excerpts TODO interesting points to detail:
 
-Our dataset is thus made of excerpts
-
-TODO interesting points to detail:
 \begin{itemize}

### 2022-01-18 src-latex/parts/40-ner-approaches.tex
Reduced sections 2 and 3

@@
 
-Previous work on the influence of OCR errors on NER models shows that it is preferable to reuse a model trained on modern data and adapt it to OCRised historical texts using supervised or unsupervised approaches.
-This is why, in our tests, we have selected two deep learning-based NER approaches trained on French texts: Spacy 3 NER pipeline and CamemBERT.
+In our tests, we have selected two deep learning-based NER models trained on French texts: Spacy 3 NER pipeline and CamemBERT. They are adapted to  historical texts as follows.
 

### 2022-01-18 src-latex/parts/10-introduction.tex
Reorganised end of section 5 to save space

@@
 OCR approaches used for modern documents are likely to struggle even on printed historical documents due multiple causes related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
-\bertrand{J'ai tenté un premier refactoring, pas plus techniquement précis ceci dit.}
-\nathalie{Phrases précedentes à reprendre avec les termes adéquats + ajouter une ref.}
+\bertrand{ajouter une ref.}
 On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts and from the errors introduced into the text by the OCR step.\nathalie{ajouter une ref}
@@
     \item compare their performance under various training set sizes (thus annotation costs) and OCR noise levels (thus OCR choices/luck/availability) (sec. 5 and 6)
-    \item suggest smart way to leverage side knowledge like lexicons (FIXME iif this works well, maybe a conclusion instead)
+    \item suggest smart way to leverage side knowledge like lexicons (FIXME iif this works well, maybe a conclusion instead) \nathalie{pas traité dans nos tests... à mettre en perspective}
 \end{enumerate}

### 2022-01-18 src-latex/parts/50-xp-setup.tex
Reorganised end of section 5 to save space

@@
 
-\bertrand{Move after section Cases studies? }
-\begin{itemize}
-    \item NER quality: Precision, Recall, F1 score (idea: stick with standard spacy/sklearn metrics) \nathalie{Dans la plupart des articles que j'ai lus, c'est toujours une F-mesure qui est utilisée et parfois la précision et le rappel sont ajoutés.}
-    \item OCR accuracy: outputs from standard tools (beware of unknown glyphs: do we need any special care?) \nathalie{Pour l'instant, j'ai croisé: CER, WER, Levenshtein et Jaccard.
-J'ai encore un état de l'art à lire sur le sujet: il y aura probablement d'autres mesures mentionnées.}
-
-    \nathalie{Pour la partie Experiment 2: dans la litérature j'ai vu des courbes F-Mesure(CER) ou F-Mesure(WER) et des diagrammes avec F-Mesure calculée sur des sous-ensembles du corpus classés par score de Levenshtein.}
-\end{itemize}
-
-\subsection{Cases studies}
-\subsubsection{Experiment 1: NER sensibility to the number of training samples}
+\subsection{NER sensibility to the number of training samples}
 The first experiment evaluates the performances of the three models on training sets of different sizes.
@@
 
-\subsection{Experiment 2: NER in noisy OCRed texts}
-This experiment aims to evaluate the impact of the OCR used on the quality of the NER results.
+\subsection{NER in noisy OCRed texts}
+The second experiment aims to evaluate the impact of the OCR used on the quality of the NER results.
 To this end, we use directory entries whose text has been extracted, corrected and annotated manually (see Section 3).
@@
 
-%\subsection{Parameters for each method}
+\subsection{Implementation details}
 
-\subsection{OCR systems parameters}
+\subsubsection{OCR systems parameters}
 \begin{itemize}
@@
 
-\subsection{Implementation details}
+\subsubsection{Spacy}
 We use Spacy\mcite{spacy} for the CNN implementation
@@
 
+\subsubsection{Huggingface CamemBERT}
 \paragraph{Huggingface CamemBERT}

### 2022-01-18 src/ocr/test_text_utils.py
OCR eval: more stupid rules

@@ -152,3 +152,63 @@ def test_ocr_simplifications_for_evaluation():
                 skip_charset_norm=False)
     assert(output == expected)
 
+    # special casefold case: casefold changes string length
+    text = "aA\u0130bBéÉ"  # U+0130 LATIN CAPITAL LETTER I WITH DOT ABOVE
+    output = ocr_simplifications_for_evaluation( text, casefold=True, remove_accents=False)
+    expected = "aa\u0069\u0307bbéé"
+    assert(output == expected)
+
+    text = "aA\u0130bBéÉ"  # U+0130 LATIN CAPITAL LETTER I WITH DOT ABOVE
+    output = ocr_simplifications_for_evaluation( text, casefold=False, remove_accents=True)
+    expected = "aAIbBeE"
+    assert(output == expected)
+
+    text = "aA\u0130bBéÉ"  # U+0130 LATIN CAPITAL LETTER I WITH DOT ABOVE
+    output = ocr_simplifications_for_evaluation( text, casefold=True, remove_accents=True)
+    expected = "aaibbee"
+    assert(output == expected)
+
+    # More space normalization
+    text = " aaa  aa   ( aaaaa a 2 )\u2029\n   aaa aa.\n  * "
+    output = ocr_simplifications_for_evaluation(text, 
+                normalize_spaces=False,
+                deduplicate_spaces=False,
+                apply_french_spacing_rules=False,
+                skip_charset_norm=True)
+    # no change requested but newline normalization is not optional yet
+    expected = " aaa  aa   ( aaaaa a 2 )\n\n   aaa aa.\n  * "
+    assert(output == expected)
+
+    output = ocr_simplifications_for_evaluation(text, 
+                normalize_spaces=False,
+                deduplicate_spaces=True,
+                apply_french_spacing_rules=False,
+                skip_charset_norm=True)
+    expected = "aaa aa ( aaaaa a 2 )\naaa aa.\n*"
+    assert(output == expected)
+
+    output = ocr_simplifications_for_evaluation(text, 
+                normalize_spaces=False,
+                deduplicate_spaces=False,
+                apply_french_spacing_rules=True,
+                skip_charset_norm=True)
+    expected = " aaa  aa (aaaaa a 2) aaa aa.\n  * "
+    assert(output == expected)
+
+    output = ocr_simplifications_for_evaluation(text, 
+                normalize_spaces=False,
+                deduplicate_spaces=True,
[…]

### 2022-01-18 src/ocr/text_utils.py

@@ -789,6 +789,8 @@ def ocr_simplifications_for_evaluation(
         normalize_punctuation: bool=False,
         remove_accents: bool=False,
         casefold: bool=False,
+        deduplicate_spaces: bool=False,
+        apply_french_spacing_rules: bool=False,
         skip_charset_norm=False,
         ) -> str:
     # Make sure the original GT simplifications are applied
@@ -801,8 +803,8 @@ def ocr_simplifications_for_evaluation(
         text = NFD(text)
     if casefold:
         text = text.casefold()
-    if remove_accents: # Once again (may not be necessary with Latin scripts)
-        text = NFD(text)
+        if remove_accents: # Once again (may not be necessary with Latin scripts)
+            text = NFD(text)
     categories = [unicodedata.category(c) for c in text]
     res = []
     # https://www.unicode.org/reports/tr44/#General_Category_Values
@@ -835,7 +837,34 @@ def ocr_simplifications_for_evaluation(
             raise ValueError(f"Unsupported category: {cat}.")
         #else: char = char
         res.append(char)
-    return "".join(res)
+    res = "".join(res)
+    # We perform whitespace de-duplication and removal once the text is clean
+    if deduplicate_spaces:
+        # keep only first space
+        res = regex.sub(r"(\s)\s*", r"\1", res)
+    if apply_french_spacing_rules:
+        # 1 space before "double punctuation"
+        res = regex.sub(r"(\w)\s*([;:!?])", r"\1 \2", res)
+        # no space before "single punctuation"
+        res = regex.sub(r"(\w)\s*([.,])", r"\1\2", res)
+        # 1 space after punctuation
+        res = regex.sub(r"([.,;:!?])\s*(\w)", r"\1 \2", res)
+        # opening/initial punctuation marks: 1 space before, no space after
+        res = regex.sub(r"(\p{Open_Punctuation}|p{Initial_Punctuation})\s*(\w)", r"\1\2", res)
+        res = regex.sub(r"(\w)\s*(\p{Open_Punctuation}|p{Initial_Punctuation})", r"\1 \2", res)
+        # closing/final punctuation marks: no space before, 1 space after
+        res = regex.sub(r"(\p{Close_Punctuation}|p{Final_Punctuation})\s*(\w)", r"\1 \2", res)
[…]

### 2022-01-18 src-latex/parts/60-results.tex
Add fig. on ner vs ocr noise

@@
 
-\begin{table}
-\caption{CamemBERT and CamemBERT-pretrained NER performances at entry-level on manually corrected and noisy OCRerd entries when fine-tuned on manually corrected entries.}
-\begin{tabular}{lrrrrrrrrr}
-Model & \multicolumn{3}{c}{Camembert.pretrained} & \multicolumn{3}{c}{Camembert} & \multicolumn{3}{c}{Camembert.pretrained} \\
-Train set & \multicolumn{3}{c}{ref} & \multicolumn{3}{c}{ref} & \multicolumn{3}{c}{pero-ocr} \\
-Test set &                  ref & pero-ocr & tesseract &       ref & pero-ocr & tesseract &                  ref & pero-ocr & tesseract \\
-\midrule
-Precision &                 96.5 &     93.3 &      90.3 &      96.1 &     92.4 &      89.7 &                 94.3 &     93.6 &      89.9 \\
-Recall    &                 96.5 &     92.6 &      91.3 &      96.6 &     92.5 &      91.2 &                 96.1 &     94.5 &      92.5 \\
-F1 score        &                 96.5 &     92.9 &      90.8 &      96.4 &     92.5 &      90.5 &                 95.2 &     94.1 &      91.2 \\
-\end{tabular}
-\end{table}
-
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{src-latex/figs/eval-ner-exp2.pdf}
+    \caption{F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing data, either manually corrected or OCRed with Pero-OCR and Tesseract.
+The type of examples used to train the NER task is noted in indice after the model name (e.g.
+Bert$_{reference}$).
+Results show that best performances on OCRed entries are obtained when the BERT model has been pretrained and fine-tuned for NER on examples affected with similar OCR errors.}
+    \label{fig:exp_2_eval_ner}
+\end{figure}
 Opt. use synthetic text perturbation as well? Maybe not interesting and too artificial if we have access to 2+ OCR systems.

### 2022-01-18 src-latex/main-paper.tex
WIP: section 4

@@
 
-\section*{Thanks}
-Laureencia for annotations, Guillaume Thomas for data processing?
+\section*{Acknowledments and Thanks}
+This work is supported by the French National Research Agency (ANR), as part of the SODUCO project, under Grant ANR-18-CE38-0013.
 
-\section*{Acknowledment}
-This work is supported by the French National Research Agency (ANR), as part of the SODUCO project, under Grant ANR-18-CE38-0013.
+The authors want to thank \textcolor{green}{Laureencia for annotations, Guillaume Thomas for data processing?
+Stéphane Bacciochi and Pascal Cristofoli for helping to create the gold reference dataset ?}
 

### 2022-01-18 src-latex/parts/30-dataset.tex
WIP: section 4

@@
 
-To build our ground truth corpus, we annotated a few pages of each of the major directory series for which we have digitised copies, using a tool developed for our project (see Figure \ref{fig:annotator}).
+To build our ground truth corpus, we annotate a few pages of each of the major directory series for which we have digitised copies, using a tool developed for our project (see Figure \ref{fig:annotator}).
 First, the image of a given directory page is loaded.
@@
     \item statistics about entry lengths (histogram of number of char / tokens par entry)
-    \item entities we considered and how we annotated them (how we assessed gold's quality)
+    \item entities we consider and how we annotated them (how we assessed gold's quality)
     \item statistics on the distribution of entities across directories and their arrangement within entries
-    \item OCR data (automated systems and clean text + how we created the "clean" text + whether/how we assessed clean text's quality)
+    \item OCR data (automated systems and clean text + how we created the "clean" text + whether/how we assess clean text's quality)
     \item Measure inter-annotator agreement ? 

### 2022-01-18 src-latex/parts/40-ner-approaches.tex
Reorganised end of section 5 to save space
WIP: section 4

@@
 \section{NER Approaches}
+We select twe deep-learning based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
+In the case of CamemBERT, we also create an additional model pre-trained on a collection of unannotated directories entries extracted with Pero-OCR.
+The three models are then fine-tuned for named entity recognition in historical directories following a procedure detailed in sections~\ref{subsection:experiment-1} and ~\ref{subsection:experiment-2}.
 
-In our tests, we have selected two deep learning-based NER models trained on French texts: Spacy 3 NER pipeline and CamemBERT. They are adapted to  historical texts as follows.
+\subsubsection{SpaCy NER pipeline}
+Spacy is a software library that offers NLP components assembled in modular pipelines specialised by language.
+Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer.
+Hence we rely on SpaCy's ad hoc pipelines trained on French corpora and capable of named entity recognition.
+% deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.
+The global architecture of these pipelines have not been yet published but is explained by the developers on their website.
+Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
+The decision layer is an adaptation of the transition-based model presented in~\cite{lample2016}.
+As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics.
+The classification layer relies on a finite-state machine whose transition probabilities are learned using a multilayer perceptron.
+\bertrand{Drop the next sentences if we need space}
+In 2018~\cite{won2018} evaluated SpaCy's NER ability to detect place names in five corpora of ancient letters written in English.
+They measured an average F1 score of 0.57.
+The SpaCy developers claim an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
+%For our experiments, the French model \textit{fr\_core\_news\_lg} is fine-tuned using our ground truth corpus.
 
-\subsubsection{Spacy 3 NER pipeline}
-The \textit{of the shelf} NLP library SpaCY offers a named entity recognition tool based on an ad hoc architecture which has not been yet published but is explained by the developers on their website.
-Although the latest version of SpaCy (v3) leverages BERT models, we could not use them as no NER pipeline is available for the French language at the time of our experiments\footnote{https://spacy.io/models/fr}.
-The vanilla NER pipeline is two-fold. Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
-The decision layer is an adaptation of the transition-based model presented in~\cite{lample2016}.
-As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics as a dynamic attention mechanism.
-The classification layer relies on a finite-state machine whose transition probabilities are learned using a multilayer perceptron.
-In 2018~\cite{won2018} compared SpaCy's vanilla NER performance on historical texts about places with several other software packages (Stanford-NER, Ner-Tagger, Edinburgh Geopoarser and Polyglot).
-Results showed SpaCy to be average with a F1-score of 0.57.
-The SpaCy developers claimed an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
-
-\subsubsection{CamemBERT \& CamemBERT pretrained on raw entries}
-
-Language model embeddings pretrained using Transformers, like BERT, have become a new paradigm for NER\cite{li2020}.
-Indeed, these embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec and they can be further fine-tuned for NER task by adding an additional output layer.
-Moreover, they can also be further pretrained in an unsupervised way on historical texts for domain adaptation.
+\subsubsection{CamemBERT}
+Language model based on Transformers, like BERT, have become a new paradigm for NER\cite{li2020}. 
+The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer.
+They can also be pretrained in an unsupervised way on historical texts for domain adaptation.
 
 As our directories are written in French, we chose the language model CamemBERT \cite{martin-etal-2020-camembert}, a Transformer model trained on a French corpus.
-We started from the CamemBERT-ner model already fine-tuned on WikiNER-fr corpus and published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}}.
-Its output layer is a linear model with a Softmax function.
+\bertrand{WIP}
 

### 2022-01-18 src-latex/parts/50-xp-setup.tex
WIP: section 4

@@
 \subsection{NER sensibility to the number of training samples}
+\label{subsection:experiment-1}
 The first experiment evaluates the performances of the three models on training sets of different sizes.
@@
 \subsection{NER in noisy OCRed texts}
-The second experiment aims to evaluate the impact of the OCR used on the quality of the NER results.
+\label{subsection:experiment-2}
+The second experiment assesses the impact of the OCR used on the quality of the NER results.
 To this end, we use directory entries whose text has been extracted, corrected and annotated manually (see Section 3).
@@
 \subsubsection{OCR systems parameters}
+\nathalie{Y a-t-il des précisions à ajouter sur la façon dont ont été utilisés les OCR? Ou bien ce sont les versions "sur étagère" qui ont servi?}
 \begin{itemize}
-    \item original OCR
     \item Tesseract 4
     \item Pero
-    \item other?
 \end{itemize}
@@
 \subsubsection{Spacy}
-We use Spacy\mcite{spacy} for the CNN implementation
-and Huggingface for the BERT 
-tok2vec(words embeddings + encoding) + attention layer  +  transition-based model.
+We used the French model \textit{fr\_core\_news\_lg} provided by Spacy\mcite{spacy} library and leveraged the library utilities to fine-tune it with various training and validation datasets derived from our ground truth dataset (see section 5.2 for details).
+The base model is fine-tuned in turn with each of the 8 generated training data sets, using the same training parameters.
+Early stopping is activated with patience parameters set at 1600.
+For each size of training dataset, 5 runs are performed, always with the same parameters, and the mean value of each performance score is computed.
+
+%and Huggingface for the BERT 
+%tok2vec(words embeddings + encoding) + attention layer  +  transition-based %model.
 
 \subsubsection{Huggingface CamemBERT}
-\paragraph{Huggingface CamemBERT}
-CamemBERT + linear classifier as decision layer
+We started from the CamemBERT-ner model already fine-tuned on WikiNER-fr corpus and published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}}.
+Its output layer is a linear model with a Softmax function.
+Two NER models are derived from this base model.
 
-EX1: Pour chaque modèle train sur les 8 datasets * 5 runs, avec mêmes paramètres pour l’entraînement: early stopping  (patience = 3)
+\textcolor{green}{Ex 2: On garde CamemBERT et CamemBERT pretrained fine-tuned sur le train clean (patience = 5), validé avec dev clean. Teste les deux modèles sur les trois datasets de test.}
 
-Ex 2: On garde CamemBERT et CamemBERT pretrained fine-tuned sur le train clean (patience = 5), validé avec dev clean. Teste les deux modèles sur les trois datasets de test.
+The base model is pretrained on 845 000 entries extracted with PERO OCR for the Next  Sentence  Prediction  (NSP) and  Masked  Language  Model(MLM) tasks.
+Then it is fine-tuned with the same parameters and on the same training datasets as the first model.
+
+The source code of all our experiments is publicly available on the following repository \url{lien github}.

### 2022-01-18 src-latex/main-paper.tex
updated section 5

@@
 
-\section*{Acknowledments and Thanks}
+\section*{Acknowledments}
 This work is supported by the French National Research Agency (ANR), as part of the SODUCO project, under Grant ANR-18-CE38-0013.

### 2022-01-18 src-latex/parts/50-xp-setup.tex
updated section 5

@@
 \subsubsection{Spacy}
-We used the French model \textit{fr\_core\_news\_lg} provided by Spacy\mcite{spacy} library and leveraged the library utilities to fine-tune it with various training and validation datasets derived from our ground truth dataset (see section 5.2 for details).
-The base model is fine-tuned in turn with each of the 8 generated training data sets, using the same training parameters.
+We use the French model \textit{fr\_core\_news\_lg} provided by Spacy\mcite{spacy} library and leverage the library utilities to fine-tune it with various training and validation datasets derived from our ground truth dataset (see section 5.2 for details).
+The base model is fine-tuned in turn with each of the 8 generated training datasets, using the same training parameters.
 Early stopping is activated with patience parameters set at 1600.
@@
 \subsubsection{Huggingface CamemBERT}
-We started from the CamemBERT-ner model already fine-tuned on WikiNER-fr corpus and published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}}.
+We start from the CamemBERT-ner model already fine-tuned on WikiNER-fr corpus and published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}}.
 Its output layer is a linear model with a Softmax function.
-Two NER models are derived from this base model.
+Four NER models are derived from this base model:
 
-\textcolor{green}{Ex 2: On garde CamemBERT et CamemBERT pretrained fine-tuned sur le train clean (patience = 5), validé avec dev clean. Teste les deux modèles sur les trois datasets de test.}
+\textit{BERTreference} The base model is fine-tuned in turn on each of the training datasets generated from the groundtruth corpus.
+Training parameters are set depending on the considered experiment: early stopping is activated with patience parameter set at 3 for experiment 1 and at 5 for experiment 2.
+\nathalie{<-- bertrand tu peux vérifier si c'est OK stp?}
 
-The base model is pretrained on 845 000 entries extracted with PERO OCR for the Next  Sentence  Prediction  (NSP) and  Masked  Language  Model(MLM) tasks.
-Then it is fine-tuned with the same parameters and on the same training datasets as the first model.
+\textit{BERTpero-ocr} This model is used for experiement 2 only.
+The base model is fine-tuned on the training dataset generated by projecting the named entities annotation on the text extracted by PERO OCR, using the same parameters as for the \textit{BERTreference} model.
+
+\textit{BERTptrn-reference} The base model is pretrained on 845 000 entries extracted with PERO OCR for the Next  Sentence  Prediction  (NSP) and  Masked  Language  Model(MLM) tasks.
+Then it is fine-tuned with the parameters and on the same training datasets as the \textit{BERTreference} model.
+
+\textit{BERTptrn-pero-ocr} The base model is pretrained like the \textit{BERTptrn-reference} model.
+Then it is fine-tuned with the parameters and on the same training datasets as the \textit{BERTpero-ocr} model.
 

### 2022-01-18 src-latex/parts/60-results.tex
updated section 5

@@
     \includegraphics[width=\textwidth]{src-latex/figs/eval-ner-exp2.pdf}
-    \caption{F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing data, either manually corrected or OCRed with Pero-OCR and Tesseract.
+    \caption{F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing examples, either manually corrected or raw from Pero-OCR and Tesseract.
 The type of examples used to train the NER task is noted in indice after the model name (e.g.

### 2022-01-18 src-latex/main-paper.tex
minor changes

@@
 \newcommand{\edwin}[1]{\textcolor{pink}{Edwin:~#1}}
+\newcommand{\joseph}[1]{\textcolor{red}{Joseph:~#1}}
 
@@
 %
-\author{LISTE PROVISOIRE PAR ORDRE ALPHA\\
+\author{%
+Nathalie Abadie\inst{1}\orcidID{0000-0001-8741-2398} \and
+Edwin Carlinet\inst{2}\orcidID{0000-0001-5737-5266} \and
+Joseph Chazalon\inst{2}\orcidID{0000-0002-3757-074X} \and
+Bertrand Duménieu\inst{3}\orcidID{0000-0002-2517-2058}\\
+{\footnotesize \emph{all authors contributed equally}}%
+}
 %
-Nathalie Abadie\inst{1}\orcidID{0000-0001-8741-2398} \and
-Edwin Carlinet\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-Joseph Chazalon\inst{3}\orcidID{2222--3333-4444-5555} \and
-Bertrand Duménieu\inst{3}\orcidID{2222--3333-4444-5555}}
+\authorrunning{N. Abadie et al.}
 %
-\authorrunning{F. Author et al.}
-% First names are abbreviated in the running head.
-% If there are more than two authors, 'et al.' is used.
-%
-\institute{LASTIG, Univ. Gustave Eiffel, IGN-ENSG, F-94160 Saint-Mandé, France\\
+\institute{%
+% 1
+LASTIG, Univ. Gustave Eiffel, IGN-ENSG, F-94160 Saint-Mandé, France\\
 \email{nathalie-f.abadie@ign.fr}\\
 \url{https://www.umr-lastig.fr/nathalie-abadie/} \and
-INST, CITY, COUNTRY\\
-\email{example@example.com}\\
-\url{https://www.example.com} \and
+% 2
+EPITA Research \& Development Laboratory (LRDE), Le Kremlin-Bicêtre, France\\
+\email{\{edwin.carlinet,joseph.chazalon\}@lrde.epita.fr}\\
+\url{https://lrde.epita.fr} \and
+% 3
 CRH-EHESS, Paris, France\\
-\email{bertrand.dumenieu@ehess.fr}\\ \url{http://crh.ehess.fr/index.php?5206}}
+\email{bertrand.dumenieu@ehess.fr}\\
+\url{http://crh.ehess.fr/index.php?5206}%
+}
 %

### 2022-01-18 src-latex/parts/50-xp-setup.tex
minor changes

@@
 \nathalie{Y a-t-il des précisions à ajouter sur la façon dont ont été utilisés les OCR? Ou bien ce sont les versions "sur étagère" qui ont servi?}
+\joseph{Totalement sur l'étagère. On a normalisé un peu les données après, c'est tout.}
 \begin{itemize}
-    \item Tesseract 4
-    \item Pero
+    \item Tesseract 4.1.1 (newest version: 5, released Nov. 2021)
+    \item Pero OCR version from git repo, master branch updated on Sep 15, 2021 % 8b20f29    
+    \item other?  % kraken if times allows
 \end{itemize}
+
+Pero OCR uses ParseNet \joseph{TODO ref} as internal layout parser for line detection.
+Pero OCR uses an LSTM engine \joseph{TODO check + ref} for text line recognition.
+Pero OCR generally works very well, as long as the bounding boxes of the regions to recognize are not too tightly adjusted.
+Trained on newspapers in European languages by its development team.
+Can output results in Latin, Greek and Cyrillic scripts, as well as some commonly-used typographic symbols.
+
+% PERO refs
+% O Kodym, M Hradiš: Page Layout Analysis System for Unconstrained Historic Documents. ICDAR, 2021.
+% M Kišš, K Beneš, M Hradiš: AT-ST: Self-Training Adaptation Strategy for OCR in Domains with Limited Transcriptions. ICDAR, 2021.
+% J Kohút, M Hradiš: TS-Net: OCR Trained to Switch Between Text Transcription Styles. ICDAR, 2021.
 

### 2022-01-18 src-latex/main-paper.tex
Compress page 1

@@
 % FIXME "19$^{th}$-Century" with dash?
-\thanks{This work was supported by the French National Research Agency:
-Project SoDuCo, grant ANR-18-CE38-0013. ???FIXME need acknowledgments for directory sources???}}
+% \thanks{%
+% This work was supported by the French National Research Agency:
+% Project SoDuCo, grant ANR-18-CE38-0013.% ???FIXME need acknowledgments for directory sources???
+% }
+}
 %
@@
 \author{%
-Nathalie Abadie\inst{1}\orcidID{0000-0001-8741-2398} \and
-Edwin Carlinet\inst{2}\orcidID{0000-0001-5737-5266} \and
-Joseph Chazalon\inst{2}\orcidID{0000-0002-3757-074X} \and
-Bertrand Duménieu\inst{3}\orcidID{0000-0002-2517-2058}\\
+N. Abadie\inst{1}\orcidlink{0000-0001-8741-2398} \and
+E. Carlinet\inst{2}\orcidlink{0000-0001-5737-5266} \and
+J. Chazalon\inst{2}\orcidlink{0000-0002-3757-074X} \and
+B. Duménieu\inst{3}\orcidlink{0000-0002-2517-2058}\\
 {\footnotesize \emph{all authors contributed equally}}%
@@
 LASTIG, Univ. Gustave Eiffel, IGN-ENSG, F-94160 Saint-Mandé, France\\
-\email{nathalie-f.abadie@ign.fr}\\
-\url{https://www.umr-lastig.fr/nathalie-abadie/} \and
+\email{nathalie-f.abadie@ign.fr}%\\
+% \url{https://www.umr-lastig.fr/nathalie-abadie/}%
+\and
 % 2
 EPITA Research \& Development Laboratory (LRDE), Le Kremlin-Bicêtre, France\\
-\email{\{edwin.carlinet,joseph.chazalon\}@lrde.epita.fr}\\
-\url{https://lrde.epita.fr} \and
+\email{\{edwin.carlinet,joseph.chazalon\}@lrde.epita.fr}%\\
+%\url{https://lrde.epita.fr}
+\and
 % 3
 CRH-EHESS, Paris, France\\
-\email{bertrand.dumenieu@ehess.fr}\\
-\url{http://crh.ehess.fr/index.php?5206}%
+\email{bertrand.dumenieu@ehess.fr}%\\
+% \url{http://crh.ehess.fr/index.php?5206}\\
+\joseph{Je propose de supprimer les URLs.}%
 }
@@
 
-\section*{Acknowledments}
+\section*{Acknowledgments}
 This work is supported by the French National Research Agency (ANR), as part of the SODUCO project, under Grant ANR-18-CE38-0013.
 
-The authors want to thank \textcolor{green}{Laureencia for annotations, Guillaume Thomas for data processing?
-Stéphane Bacciochi and Pascal Cristofoli for helping to create the gold reference dataset ?}
+The authors want to thank Stéphane Bacciochi and Pascal Cristofoli for helping to create the gold reference dataset, 
+Laureencia Maurice for annotating data,
+as well as Guillaume Thomas, Paul Abi Saad, Raphaël Lelièvre, Dorian Mignon, Thibault Cavaciuti and Pierre Sadki for contributing to the annotation platform.
+\joseph{On peut essayer de gagner de la place en déplaçant ça en footnote de 1ere page.}
 

### 2022-01-18 src-latex/parts/30-dataset.tex
Add some comments

@@
 
-The directories to be processed are stored in different libraries in Paris\footnote{\textcolor{green}{Citer les bibliothèques sources?}}.
+The directories to be processed are stored in different libraries in Paris\footnote{\textcolor{green}{Citer les bibliothèques sources?} \joseph{est-ce nécessaire ici ?
+Est-ce que ça peut être dans une annexe ?}}.
 They have been scanned independently and with various levels of quality.
@@
 
-To build our ground truth corpus, we annotate a few pages of each of the major directory series for which we have digitised copies, using a tool developed for our project (see Figure \ref{fig:annotator}).
+To build our ground truth corpus, we annotate 78 pages randomly sampled from each of the major directory series for which we have digitised copies, using a tool developed for our project (see Figure \ref{fig:annotator}).
 First, the image of a given directory page is loaded.
@@
 	       {./images/Annot2.png}}
-	  \caption{\label{fig:annotator} GUI for corpus annotation}
+	  \caption{\label{fig:annotator} GUI for corpus annotation \joseph{remove GUI-related details?}}
 \end{figure}
@@
 It can therefore be used to train and improve the entry detection, OCR and NER tools.
+% book, page, box, text_ocr, ner_xml (+uid within a page)
 
@@
 \nathalie{On va utiliser cette sous-section ou pas?}
+\joseph{Pas vraiment car on n'a pas eu le temps de faire ça, mais on peut parler de l'étage d'extraction d'entrées qui est validé manuellement, et des quelques règles de simplification du charset pour homogénéiser entre les annotateurs / pallier des ambiguïtés dans les règles d'annotation (qu'on pourrait encore améliorer)}
 Our dataset is thus made of excerpts TODO interesting points to detail:

### 2022-01-18 src-latex/parts/40-ner-approaches.tex
review of section 4

@@
 \section{NER Approaches}
-We select twe deep-learning based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
-In the case of CamemBERT, we also create an additional model pre-trained on a collection of unannotated directories entries extracted with Pero-OCR.
-The three models are then fine-tuned for named entity recognition in historical directories following a procedure detailed in sections~\ref{subsection:experiment-1} and ~\ref{subsection:experiment-2}.
+We select two deep-learning based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
+We also create an additional CamemBERT model pre-trained on a collection of unannotated directories entries extracted with Pero-OCR.
+The NER layer of each three models is then fine-tuned on historical directories as detailed in sections~\ref{subsection:experiment-1} and ~\ref{subsection:experiment-2}.
 
-\subsubsection{SpaCy NER pipeline}
+\paragraph{Naming conventions} The short name of each model, indicated in square brackets, is used to refer to it from now on.
+
+\subsubsection{SpaCy NER pipeline~[SpaCy NER]}
 Spacy is a software library that offers NLP components assembled in modular pipelines specialised by language.
-Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer.
+Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (jan. 2022).
 Hence we rely on SpaCy's ad hoc pipelines trained on French corpora and capable of named entity recognition.
 % deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.
-The global architecture of these pipelines have not been yet published but is explained by the developers on their website.
+The global architecture of these pipelines have not been yet published but are explained by the developers on their website.
 Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
@@
 
-\subsubsection{CamemBERT}
+\subsubsection{CamemBERT~[CmBERT]}
 Language model based on Transformers, like BERT, have become a new paradigm for NER\cite{li2020}. 
@@
 They can also be pretrained in an unsupervised way on historical texts for domain adaptation.
-
-As our directories are written in French, we chose the language model CamemBERT \cite{martin-etal-2020-camembert}, a Transformer model trained on a French corpus.
+As the directories are written in French, we chose the language model CamemBERT \cite{martin-etal-2020-camembert}, a Transformer model trained on a French corpus.
 \bertrand{WIP}
-
-For our experiments, two models are derived. One is produced by fine-tuning this generic model on our ground truth dataset.
-The second is created by adapting the embeddings of this model using 845 000 raw directory entries, randomly selected and extracted with PERO OCR.
-The base model is pretrained on these texts for two tasks, namely, Next Sentence Prediction (NSP) and Masked Language Model (MLM).
-Finally, the resulting model is also fine-tuned using our ground truth corpus.
+\subsubsection{CamemBERT pretrained on directories~[CmBERT+ptrn]}
+In order to evaluate the benefits of pre-training the language model on OCR texts from domain-related documents, we adapt the CmBERT embeddings using \num{845.000} entries randomly selected wihtin the collection of directories and extracted with Pero-OCR.
+We achieve this by training the CmBERT model for 3 epochs on two unsupervised tasks: next sentence prediction and masked language model.
 

### 2022-01-18 src-latex/parts/50-xp-setup.tex
Add some comments
review of section 4

--- 2022-01-18 src-latex/parts/50-xp-setup.tex
@@
 
-\Cref{fig.protocol} depicts the evaluation procotol used to assess the OCR and NER systems. Two quality evaluation
-are performed in the pipeline named respectively \emph{OCR Q.A.} and \emph{NER Q.A.}.
+The Figure \Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems. Two quality evaluations
+are performed in the pipeline, respectively named \emph{OCR Q.A.} and \emph{NER Q.A.}.
 
-First the layout extraction and the entry segmentation of the page are performed with a semi-automated system and
+First, the layout extraction and entry segmentation of the page are performed with a semi-automated system and
 checked by a human. Afterward, an OCR system runs on the thumbnails of each segmented entry to extract its text. An
-entry might span over multiple text lines but is always single-block. Thus, the most adapted mode is chosen when
-the OCR system allows for detection mode (e.g. the \emph{block} mode for \emph{Tesseract}).
+entry might span over multiple text lines but is always a single block. Thus, the most adapted mode is chosen when
+the OCR system allows for the detection mode (e.g. the \emph{block} mode for \emph{Tesseract}).
 
 \textbf{OCR Q.A.} is performed after some text normalization of the OCR system outputs. Text normalization consists in
-projecting unicode characters onto the latin-1 charset (whenever it makes sense) and removing extra symbols (hands,
+projecting Unicode characters onto the latin-1 charset (whenever it makes sense) and removing extra symbols (hands,
 crosses). Then, the predicted text is aligned with the reference text using standard tools and the Character Error Rates
-(CER) are computed at entry level and at global level.
+(CER) are computed at the entry level and at the global level.
 
@@
 
-In this benchmark, we consider 3 OCR systems well-known for the historical document analysis: Pero OCR, Tesseract and Kraken.
+In this benchmark, we consider 3 OCR systems well known for historical document analysis: Pero OCR, Tesseract, and Kraken.
 
@@
 extraction, we rely on a technique similar as for the OCR evaluation. The predicted text is aligned with the reference
-text and the tags are projected in the alignment. Then, the precision, recall and f-measure (or f-score) are computed considering
-only the exact matches between entities. Precision and recall are defined by equation (2); the f-measure in their harmonic mean.
+text and the tags are projected in the alignment. Then, the precision, recall, and f-measure (or f-score) are computed considering
+only the exact matches between entities. Precision and recall are defined by equation (2); the f-measure is their harmonic mean.
 
@@
 OCR text; its output is compared to the NER reference string. There is only 2 over 3 tags matching in the prediction (precision),
-while only 2 over 4 entities are matching in the reference (recall). It follows an overall f-score of 0.4.
+while only 2 over 4 entities are matched in the reference (recall). It follows an overall f-score of 0.4.
 
@@
 
-\subsection{NER sensibility to the number of training samples}
+\subsection{Experiment 1: NER sensibility to the number of training examples}
 \label{subsection:experiment-1}
@@
 
-To generate smaller training sets we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as before.
+To generate smaller training sets, we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as before.
 We stop if there is only one entry left in a directory or if the current train set contains less than 30 entries.
-Applying this procedure to the initial training set produced 8 training sets containing 49, 99, 199, 398, 796, 1593, 3186 and 6373 entries.
+Applying this procedure to the initial training set produced 8 training sets containing 49, 99, 199, 398, 796, 1593, 3186, and 6373 entries.
 
@@
 
-\subsection{NER in noisy OCRed texts}
+\subsection{Experiment 1: NER on texts with OCR noise}
 \label{subsection:experiment-2}
-The second experiment assesses the impact of the OCR used on the quality of the NER results.
-To this end, we use directory entries whose text has been extracted, corrected and annotated manually (see Section 3).
+The second experiment assesses the impact of the OCR on the quality of the NER results.
+To this end, we use directory entries whose text has been extracted, corrected, and annotated manually (see Section 3).
 This reference text is aligned with the text produced on the same entries by Tesseract and PERO OCR.
 The named entities annotated in the reference text are then projected into the OCRed texts.
-To make sure that all training, development and test datasets contain the same annotations, only the entries for which the same number of annotations are found in the reference text and in the OCRised texts are kept.
-Finally, the reference text is divided into three datasets, using the same method as in Experiment 1.
-Two additional test files are generated with exactly the same entries as the test file produced from the reference text: one from the text produced by PERO OCR and one from that produced by Tesseract.
+To make sure that all training, development, and test datasets contain the same annotations, only the entries for which the same number of annotations are found in the reference text and in the OCRised texts are kept.
+Finally, the reference text and the PERO OCR text are divided into three datasets, using the same method as in Experiment 1.
+An additional test file is generated from the text produced by Tesseract with the same entries as the test files produced from the reference and PERO OCR texts.
 
@@
 
-\subsubsection{OCR systems parameters}
+\subsubsection{OCR system parameters}
 \nathalie{Y a-t-il des précisions à ajouter sur la façon dont ont été utilisés les OCR? Ou bien ce sont les versions "sur étagère" qui ont servi?}
@@
 \subsubsection{Spacy}
-We used the French model \textit{fr\_core\_news\_lg} provided by Spacy\mcite{spacy} library and leveraged the library utilities to fine-tune it with various training and validation datasets derived from our ground truth dataset (see section 5.2 for details).
-The base model is fine-tuned in turn with each of the 8 generated training data sets, using the same training parameters.
+We use the French model \textit{fr\_core\_news\_lg} provided by Spacy\mcite{spacy} library and leverage the library utilities to fine-tune it on various training and validation datasets derived from our ground truth dataset (see Section 5.2 for details).
+The base model is fine-tuned in turn on each of the 8 generated training datasets, using the same training parameters.
 Early stopping is activated with patience parameters set at 1600.
-For each size of training dataset, 5 runs are performed, always with the same parameters, and the mean value of each performance score is computed.
+For each size of the training dataset, 5 runs are performed, always with the same parameters, and the mean value of each performance score is computed.
 
@@
 \subsubsection{Huggingface CamemBERT}
-We started from the CamemBERT-ner model already fine-tuned on WikiNER-fr corpus and published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}}.
+We start from the CamemBERT-ner model already fine-tuned on WikiNER-fr corpus and published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}}.
 Its output layer is a linear model with a Softmax function.
-Two NER models are derived from this base model.
+Four NER models are derived from this base model:
 
-\textcolor{green}{Ex 2: On garde CamemBERT et CamemBERT pretrained fine-tuned sur le train clean (patience = 5), validé avec dev clean. Teste les deux modèles sur les trois datasets de test.}
+$BERT_{reference}$ The base model is fine-tuned in turn on each of the training datasets generated from the groundtruth corpus.
+Training parameters are set depending on the considered experiment: early stopping is activated with the patience parameter set at 3 for experiment 1 and at 5 for experiment 2.
 
-The base model is pretrained on 845 000 entries extracted with PERO OCR for the Next  Sentence  Prediction  (NSP) and  Masked  Language  Model(MLM) tasks.
-Then it is fine-tuned with the same parameters and on the same training datasets as the first model.
+$BERT_{pero-ocr}$ This model is used for experiment 2 only.
+The base model is fine-tuned on the training dataset generated by projecting the named entity annotations on the text extracted by PERO OCR, using the same parameters as for the $BERT_{reference}$ model.
+
+$BERT_{ptrn-reference}$ The base model is pretrained using 845000 entries extracted by PERO OCR with the Next  Sentence  Prediction  (NSP) and  Masked  Language  Model(MLM) tasks.
+Then it is fine-tuned with the same parameters and on the same training datasets generated from the groundtruth corpus as the $BERT_{reference}$ model.
+
+$BERT_{ptrn-pero-ocr}$ This model is used for experiment 2 only.
+The base model is pretrained like the $BERT_{ptrn-reference}$ model and it is fine-tuned with the same parameters and on the same training datasets as the $BERT_{pero-ocr}$ model.
 

### 2022-01-18 src-latex/parts/50-xp-setup.tex

@@
 \includegraphics[width=\linewidth]{figs/protocol.pdf}
-\caption{Scheme of the evaluation protocol.}
+\caption{Scheme of the evaluation protocol. \joseph{Missing "Ground truth" for NER stage?}}
 \label{fig.protocol}

### 2022-01-19 src-latex/parts/50-xp-setup.tex
Shortened 5.2

@@
 \label{subsection:experiment-1}
-The first experiment evaluates the performances of the three models on training sets of different sizes.
-To do so, we split the gold reference into a training set, a development set, and a test set.
-The training set is then gradually reduced in size while maintaining the relative frequency of directories within.
-The training and testing procedure is the same for the three models.
+\bertrand{Find a better name ?}
+The constitution of annotated example sets to train a NER model on a target domain a is a critical preliminary step.
+Often done manually, possibly from bootstrapped annotations, this task is tedious, time consuming and error prone.
+The ability of a model to perform well even with few training examples is a pratical criteria to consider.
+In this first experiment, we investigate the performance of the three models when fine-tuned for NER on training sets of decreasing sizes.
 
-As the form and structure of entries varies across directory collections and through time, the model may learn to overfit on a subset of directories with specific features.
-To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, ~20\%) from the gold reference to test each model on unseen directories.
-Then a stratified sampling based on the entry directory name is applied to the remaining set to create a training (6373 entries, ~71\% of the gold reference) and a development set (709 entries, ~8\%).
-This sampling procedure is a convenient way to shape both sets to reflect the diversity of directories within the gold reference.
+To do so, we split the gold reference of manually annotated entries into a training set, a development set, and a test set. 
+The training set is then gradually reduced in size while maintaining the relative frequency of directories within it.
 
+As the organisation and structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features.
+To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, $\approx 19\%$) from the gold reference to test each model on unseen directories.
+Then, a stratified sampling based on the source directory of each entry is run on the remaining set to create a training (6373 entries, $\approx 73\%$ of the gold reference) and a development set (709 entries, $\approx 8\%$).
+This sampling procedure is a convenient way to shape both sets so they reflect the diversity of directories within the gold reference.
+The development set is used to evaluate the models performance during the training phase.
 To generate smaller training sets, we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as before.
-We stop if there is only one entry left in a directory or if the current train set contains less than 30 entries.
-Applying this procedure to the initial training set produced 8 training sets containing 49, 99, 199, 398, 796, 1593, 3186, and 6373 entries.
+We stop if a directory has only one entry left, or if the current training subset contains less than 30 entries.
+Applying this procedure to the initial training set produced 8 training subsets containing 49, 99, 199, 398, 796, 1593, 3186, and 6373 entries.
 
-% Move to metrics ?
-All metrics are evaluated on the test set, yet the biased model performances on the development sets are added in the article material for information.
+All tree models are fine-tuned for NER against every training subset.
+The performance metrics are measured against the test set after each training.
 
-\subsection{Experiment 1: NER on texts with OCR noise}
+% Move to results
+%All metrics are evaluated on the test set, yet the biased model performances on the development sets are added in the article material for information.
+
+\subsection{Experiment 2: NER on entries with OCR noise}
 \label{subsection:experiment-2}

### 2022-01-19 src-latex/parts/30-dataset.tex
Add annotated example

@@
 
+\bertrand{Show an example for each entity ? Or even a full entry with its entities highlighted ?}
 The annotated named entities are as follows:
@@
 E.g.: "Alibert (Prosper)", "Allamand frères et Hersent", "Heurtemotte (Vve)", etc.
-    \item TITLE: an honorary title, either written or represented by special glyphs.
-E.g.: "élig.", \scalerel*{\includegraphics{./images/LH.png}}{H} which stand for the Légion d'Honneur, \scalerel*{\includegraphics{./images/GMLondon.png}}{H} which stands for the Great Medal at the London exhibition, .etc.
-    \item ACT: the activity of the person (profession or social status) or the type of business. E.g.: "horlogerie", "export en tous genres", "Conseiller d'Etat", "propriétaire", etc.
-    \item LOC: a street name or a district name. E.g.: "r. de la Jussienne", "Au Marais", etc.
-    \item CARD: a street number, as part of an address. E.g. "16", "5 bis", etc.
-    \item FT: a geographic entity type, used to give more details on a location. E.g.: "boutique", "atelier", "fab.", etc.
+    \item TITLE: an honorary title, either text, glyphs or a combination of the two.
+E.g: "O.
+\scalerel*{\includegraphics{./images/LH.png}}{H}" for "Officier de la Légion d'Honneur" or "\scalerel*{\includegraphics{./images/GMLondon.png}}{H}" to represent the Great Medal at the London exhibition.
+    \item ACT: the profession or social status of a person summarized in the single concept of activity. E.g.: "horlogerie", "export en tous genres", "Conseiller d'Etat", "propriétaire".
+    \item LOC: mostly street names ("r. de la Jussienne"), but may also be neighborhoods ("Marais du Temple") or any indirect spatial references ("au Palais Royal").
+    \item CARDINAL: a street number, as part of an address ("16", "5 bis"), or a range of numbers (e.g. "23-25" or "5 à 9").
+    \item FT: a geographic entity type, used to give more details on a location, e.g. "boutique", "atelier", "fab." or "dépôt".
 \end{itemize}
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{images/examples_labeled_entries.pdf}
+    \caption{Two manually annotated dictionary entries showing the different labels used in the corpus.}
+    \label{fig:my_label}
+\end{figure}
 Thus, the extracted corpus contains both the bounding boxes of directory entries, their exact text and the named entities they contain.

### 2022-01-19 src-latex/parts/40-ner-approaches.tex
Add annotated example

@@
 We also create an additional CamemBERT model pre-trained on a collection of unannotated directories entries extracted with Pero-OCR.
-The NER layer of each three models is then fine-tuned on historical directories as detailed in sections~\ref{subsection:experiment-1} and ~\ref{subsection:experiment-2}.
+The NER layer of each three models is then fine-tuned on historical directories as detailed in sections~\ref{subsection:experiment-1-setup} and ~\ref{subsection:experiment-2-setup}.
 

### 2022-01-19 src-latex/parts/50-xp-setup.tex
Add annotated example

@@
 \subsection{Experiment 1: NER sensibility to the number of training examples}
-\label{subsection:experiment-1}
+\label{subsection:experiment-1-setup}
 \bertrand{Find a better name ?}
@@
 
-All tree models are fine-tuned for NER against every training subset.
+All tree models are fine-tuned for NER on each training subset.
 The performance metrics are measured against the test set after each training.
@@
 
-\subsection{Experiment 2: NER on entries with OCR noise}
-\label{subsection:experiment-2}
+\subsection{Experiment 2: Impact of OCR nosie on named entity recognition}
+\label{subsection:experiment-2-setup}
+Noise introduced by OCR is known to impede the named entity recognition process (see section~\ref{subsection:stoa-ner-on-historical-texts}).
+In this second experiment, we are interested in two questions:
+\begin{itemize}
+    \item which model is the most robust to OCR noise ?
+    \item what is most appropriate strategy to fine-tune the NER layer if the texts to label are noisy
+\end{itemize}
+
 The second experiment assesses the impact of the OCR on the quality of the NER results.

### 2022-01-19 src-latex/parts/30-dataset.tex
5.3 v1

@@
 
-\bertrand{Show an example for each entity ? Or even a full entry with its entities highlighted ?}
 The annotated named entities are as follows:
+\bertrand{Add a ref to Fig. \ref{fig:annotated-example}}
+
 \begin{itemize}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
     \centering
     \includegraphics[width=\textwidth]{images/examples_labeled_entries.pdf}
-    \caption{Two manually annotated dictionary entries showing the different labels used in the corpus.}
-    \label{fig:my_label}
+    \caption{\label{fig:annotated-example}Two annotated entries showing the different labels used in the corpus.}
 \end{figure}

### 2022-01-19 src-latex/parts/40-ner-approaches.tex
5.3 v1

--- 2022-01-18 src-latex/parts/40-ner-approaches.tex
@@
 \section{NER Approaches}
-We select two deep-learning based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
-We also create an additional CamemBERT model pre-trained on a collection of unannotated directories entries extracted with Pero-OCR.
-The NER layer of each three models is then fine-tuned on historical directories as detailed in sections~\ref{subsection:experiment-1} and ~\ref{subsection:experiment-2}.
+We select two deep-learning-based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
+We also create an additional CamemBERT model pretrained on a collection of unannotated directory entries extracted with Pero-OCR.
+The NER layer of each three models is then fine-tuned on historical directories as detailed in\cref{subsection:experiment-1-setup} and\cref{subsection:experiment-2-setup}.
 

### 2022-01-19 src-latex/parts/50-xp-setup.tex
5.3 v1

--- 2022-01-19 src-latex/parts/50-xp-setup.tex
@@
 % Move to results
-%All metrics are evaluated on the test set, yet the biased model performances on the development sets are added in the article material for information.
+%All metrics are evaluated on the test set, yet the biased metrics on the development set are added in the article material as additional information.
 
-\subsection{Experiment 2: NER on entries with OCR noise}
-\label{subsection:experiment-2}
-The second experiment assesses the impact of the OCR on the quality of the NER results.
-To this end, we use directory entries whose text has been extracted, corrected, and annotated manually (see Section 3).
-This reference text is aligned with the text produced on the same entries by Tesseract and PERO OCR.
-The named entities annotated in the reference text are then projected into the OCRed texts.
-To make sure that all training, development, and test datasets contain the same annotations, only the entries for which the same number of annotations are found in the reference text and in the OCRised texts are kept.
-Finally, the reference text and the PERO OCR text are divided into three datasets, using the same method as in Experiment 1.
-An additional test file is generated from the text produced by Tesseract with the same entries as the test files produced from the reference and PERO OCR texts.
+\subsection{Experiment 2: Impact of OCR nosie on named entity recognition}
+\label{subsection:experiment-2-setup}
+\bertrand{Assume that we use Kraken too. If not, adapt in consequence.}
+Noise introduced by OCR is known to impede the named entity recognition process (see section~\ref{subsection:stoa-ner-on-historical-texts}).
+%The second experiment assesses the impact of the OCR on the quality of the NER results.
+The second experiment addresses two questions (1) which model is the most robust to OCR noise and (2) what is the most appropriate strategy to fine-tune the NER layer if the input texts to label are noisy ?
+To this end, we leverage the labeled OCR datasets produced following to the method explained in Section 5, in addition to the gold reference.
 
-%\begin{itemize}
-%\item At least 1 run per dataset, more if possible.
-%\item Only use the best model; should be CamemBERT+pretrained
-%\item Train on reference gold, evaluate on reference gold AND ocr-gold (2 datasets)
-%\item split train/dev/test the same way than experiment 1. Use the exact same splitting %for datasets so the clean test set, pero test set and tesseract test sets contains the %same entries.
-%\end{itemize}
+First and foremost, to make sure that the datasets contain the same entries a preprocessing step must be performed.
+Indeed, entries where the OCR produced an empty string and those for which no entity could be projected from the gold reference text are ignored.
+To overcome this problem we filter each dataset by keeping only the entries common to all.
+At the cost of losing  5\% of the gold references entries, we thus have four new annotated datasets containing the exact same entries.
+They are denoted hereafter {reference, pero, tesseract, kraken}-gold.
+Finally each is is split into a training, a dev and a test subset following the procedure presented in\cref{subsection:experiment-1-setup}.
 
-%\begin{itemize}
-%\item All : leave out 3 directories (~20\% of the dataset) as test data.
-Then apply stratified sampling based on the directory of each entry to split the remaining into a train set (~72\%) and a dev set (~8\%).
-%\item Spacy CNN : No limit on epochs, instead use patience 1600.
-%\item CamemBERT: 3 epochs.
-%\item CamemBERT pretrained: 3 epochs, pretraining on MLM + NSP with ~20000 raw texts extracted with Pero-OCR.
-%\end{itemize}
+We fine-tune CmBERT and CmBERT+ptrn on reference-gold and pero-gold to create two versions of each model: the former trained on manually corrected entries and the latter on OCR entries.
+SpaCy NER is left aside as results from experiment 1 show that it is outperformed by BERT models.
+Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold, tesseract-gold and kraken-gold.
 

### 2022-01-19 src-latex/parts/50-xp-setup.tex
v1 section 5

@@
 \label{subsection:experiment-2-setup}
-\bertrand{Assume that we use Kraken too. If not, adapt in consequence.}
 Noise introduced by OCR is known to impede the named entity recognition process (see section~\ref{subsection:stoa-ner-on-historical-texts}).
@@
 At the cost of losing  5\% of the gold references entries, we thus have four new annotated datasets containing the exact same entries.
-They are denoted hereafter {reference, pero, tesseract, kraken}-gold.
+They are denoted hereafter {reference, pero, tesseract}-gold.
 Finally each is is split into a training, a dev and a test subset following the procedure presented in\cref{subsection:experiment-1-setup}.
@@
 SpaCy NER is left aside as results from experiment 1 show that it is outperformed by BERT models.
-Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold, tesseract-gold and kraken-gold.
+Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold and tesseract-gold.
 
@@
 \subsubsection{Spacy}
-We use the French model \textit{fr\_core\_news\_lg} provided by Spacy\mcite{spacy} library and leverage the library utilities to fine-tune it on various training and validation datasets derived from our ground truth dataset (see Section 5.2 for details).
+We use the French pipeline \textit{fr\_core\_news\_lg} provided by SpaCy\mcite{spacy}, already trained for NER on the deep-sequoia and wikiner-fr corpora.
+The default NER labels known to the pipeline are PER, ORG, MISC and LOC.
+In experiment 1 the base model is fine-tuned on every of the 8 training sets using the same training parameters.
+Early stopping is activated with a patience value of 1600 training steps without improvement of the f1 score.
+
+% \bertrand{Move to results}
+%For each size of the training dataset, 5 runs are performed, always with the same parameters, and the mean value of each performance score is computed.
+
+%library and leverage the library utilities to fine-tune it on various training and validation datasets derived from our ground truth dataset (see Section 5.2 for details).
 The base model is fine-tuned in turn on each of the 8 generated training datasets, using the same training parameters.
@@
 \subsubsection{Huggingface CamemBERT}
-We start from the CamemBERT-ner model already fine-tuned on WikiNER-fr corpus and published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}}.
-Its output layer is a linear model with a Softmax function.
-Four NER models are derived from this base model:
+Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface.
+Our baseline CmBERT model is CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
+Its NER head is a linear model with a Softmax function.
+CmBERT and CmBERT+ptrn are always fine-tuned using the same parameters, with at most 5000 training steps and an early stopping condition set to 3 evaluations in experiment 1 (5 in experiment 2) without improvement of the f1 score.
+Evaluations are run every 100 steps.
 
-$BERT_{reference}$ The base model is fine-tuned in turn on each of the training datasets generated from the groundtruth corpus.
+Details can be found for both the SpaCy NER and the CamemBERT NER models in the source code publicly available on the GitHub repository \bertrand{LIEN GITHUB}.
+
+%$BERT_{reference}$ The base model is fine-tuned in turn on each of the training datasets generated from the groundtruth corpus.
 Training parameters are set depending on the considered experiment: early stopping is activated with the patience parameter set at 3 for experiment 1 and at 5 for experiment 2.
 
-$BERT_{pero-ocr}$ This model is used for experiment 2 only.
+%$BERT_{pero-ocr}$ This model is used for experiment 2 only.
 The base model is fine-tuned on the training dataset generated by projecting the named entity annotations on the text extracted by PERO OCR, using the same parameters as for the $BERT_{reference}$ model.
 
-$BERT_{ptrn-reference}$ The base model is pretrained using 845000 entries extracted by PERO OCR with the Next  Sentence  Prediction  (NSP) and  Masked  Language  Model(MLM) tasks.
+%$BERT_{ptrn-reference}$ The base model is pretrained using 845000 entries extracted by PERO OCR with the Next  Sentence  Prediction  (NSP) and  Masked  Language  Model(MLM) tasks.
 Then it is fine-tuned with the same parameters and on the same training datasets generated from the groundtruth corpus as the $BERT_{reference}$ model.
 
-$BERT_{ptrn-pero-ocr}$ This model is used for experiment 2 only.
+%$BERT_{ptrn-pero-ocr}$ This model is used for experiment 2 only.
 The base model is pretrained like the $BERT_{ptrn-reference}$ model and it is fine-tuned with the same parameters and on the same training datasets as the $BERT_{pero-ocr}$ model.
-
-The source code of all our experiments is publicly available on the following repository \url{lien github}.

### 2022-01-19 src-latex/parts/50-xp-setup.tex
adds spacy & hf libs versions

@@
 \subsubsection{Spacy}
-We use the French pipeline \textit{fr\_core\_news\_lg} provided by SpaCy\mcite{spacy}, already trained for NER on the deep-sequoia and wikiner-fr corpora.
+We use the French pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the deep-sequoia and wikiner-fr corpora.
 The default NER labels known to the pipeline are PER, ORG, MISC and LOC.
@@
 \subsubsection{Huggingface CamemBERT}
-Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface.
+Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
 Our baseline CmBERT model is CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
@@
 CmBERT and CmBERT+ptrn are always fine-tuned using the same parameters, with at most 5000 training steps and an early stopping condition set to 3 evaluations in experiment 1 (5 in experiment 2) without improvement of the f1 score.
-Evaluations are run every 100 steps.
+Evaluations are performed every 100 steps.
 

### 2022-01-19 src-latex/parts/50-xp-setup.tex
Update OCR results with kraken
Updates from Overleaf

@@
 
-\subsubsection{Spacy}
+\subsubsection{SpaCy}
 We use the French pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the deep-sequoia and wikiner-fr corpora.
@@
 
-Details can be found for both the SpaCy NER and the CamemBERT NER models in the source code publicly available on the GitHub repository \bertrand{LIEN GITHUB}.
+Details can be found for all models in the source code publicly available on the GitHub repository \bertrand{LIEN GITHUB}.
 

### 2022-01-19 src-latex/parts/60-results.tex
Fix figure path for build to work
adds spacy & hf libs versions
Update OCR results with kraken

@@
 \begin{tabular}{rll}
+\toprule
           & CER & CER$_{norm.}$ \\
@@
 Tesseract & 6.56\% & 6.45\% \\
+Kraken    & 15.72\% & 15.46\% \\  
 \bottomrule
@@
 \centering
-\caption{NER performances at entry-level for SpaCy NER, CamemBERT and CamemBERT-pretrained when fine-tuned with an increasing number of examples.}
+\caption{F1 score, precision and recall measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER in experiment 1 for 8 training sets of increasing sizes.}
 \begin{tabular}{llrrrrrrrr}
@@
 \midrule\bottomrule
-\multirow{3}{*}{\rotatebox{90}{F1 score}} & Camembert &  89.5 &  90.5 &  92.7 &  93.3 &  \textbf{94.1} &  \textbf{94.9} &  \textbf{94.6} &  \textbf{95.1} \\
-       & Camembert-pretrained &  \textbf{92.2} &  \textbf{92.9} &  \textbf{93.6} &  \textbf{93.8} &  93.8 &  94.1 &  \textbf{94.6} &  94.4 \\
+\multirow{3}{*}{\rotatebox{90}{F1 score}} & CmBERT &  89.5 &  90.5 &  92.7 &  93.3 &  \textbf{94.1} &  \textbf{94.9} &  \textbf{94.6} &  \textbf{95.1} \\
+       & CmBERT-ptrn &  \textbf{92.2} &  \textbf{92.9} &  \textbf{93.6} &  \textbf{93.8} &  93.8 &  94.1 &  \textbf{94.6} &  94.4 \\
        & SpaCy NER &  87.0 &  89.0 &  90.3 &  91.9 &  92.1 &  92.8 &  93.2 &  93.5 \\
 \cline{1-10}
-\multirow{3}{*}{\rotatebox{90}{Precision}} & Camembert &  87.4 &  88.7 &  91.5 &  92.7 &  93.3 &  94.9 &  93.9 &  95.1 \\
-       & Camembert-pretrained &  90.8 &  91.8 &  92.9 &  93.0 &  93.0 &  93.4 &  94.1 &  93.9 \\
+\multirow{3}{*}{\rotatebox{90}{Precision}} & CmBERT &  87.4 &  88.7 &  91.5 &  92.7 &  93.3 &  94.9 &  93.9 &  95.1 \\
+       & CmBERT-ptrn &  90.8 &  91.8 &  92.9 &  93.0 &  93.0 &  93.4 &  94.1 &  93.9 \\
        & SpaCy NER &  85.6 &  87.7 &  90.0 &  92.0 &  92.4 &  92.8 &  93.1 &  93.7 \\
 \cline{1-10}
-\multirow{3}{*}{\rotatebox{90}{Recall}} & Camembert &  91.6 &  92.5 &  93.9 &  93.9 &  94.9 &  94.9 &  95.4 &  95.1 \\
-       & Camembert-pretrained &  93.6 &  94.0 &  94.4 &  94.6 &  94.6 &  94.8 &  95.0 &  94.9 \\
+\multirow{3}{*}{\rotatebox{90}{Recall}} & CmBERT &  91.6 &  92.5 &  93.9 &  93.9 &  94.9 &  94.9 &  95.4 &  95.1 \\
+       & CmBERT-ptrn &  93.6 &  94.0 &  94.4 &  94.6 &  94.6 &  94.8 &  95.0 &  94.9 \\
        & SpaCy NER &  88.6 &  90.4 &  90.7 &  91.7 &  91.9 &  92.8 &  93.3 &  93.4 \\
@@
     \centering
-    \includegraphics[width=\textwidth]{src-latex/figs/eval-ner-exp2.pdf}
+    \includegraphics[width=\textwidth]{figs/eval-ner-exp2.pdf}
     \caption{F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing examples, either manually corrected or raw from Pero-OCR and Tesseract.

### 2022-01-19 src-latex/parts/00-abstract.tex
Updates from Overleaf

@@
 % !TeX root = ../main-paper.tex
-This paper features a comparison of four (\bertrand{Ensure it's 4!} named entity recognition (NER) techniques which can be reasonably used in the context of historical document analysis.
+This paper features a comparison of three \bertrand{not four anymore ?}named entity recognition (NER) techniques which can be reasonably used in the context of historical document analysis.
 Such natural language processing techniques identify which class each text token belong to, like "person name", "location", "number" or "activity".

### 2022-01-19 src-latex/parts/60-results.tex
Updates from Overleaf
6.2 and 6.3 v1

@@
 
-\subsection{Experiment 1: NER sensibility to the number of training samples}
+\subsection{Experiment 1: NER sensibility to the number of training examples}
+Table \ref{tab:experiment-1-models-performances} and Figure \ref{fig:f1-vs-trainsize} show the mean F1 score, precision, and recall computed on the test dataset, for 5 fine tunings of each model and for each of the 8 training subsets.
+CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.
+The F1 score increases by 4.6 points between 49 and 796 examples for CmBERT (resp. 1.6 for CmBERT+ptrn and 5.1 for SpaCy NER) but only by 1 point between 796 and 6373 examples (resp. 0.6 and 1.4).
+The models derived from CamemBERT always outperform the Spacy model.
 
-Qualitative results
-\textbf{TODO random samples of results + selection of failure cases}
-
-Quantitative results: table + graph ideally (with training set size info)
+It appears that pretraining the CamemBERT model on OCRed texts seems worth it only when a the training set used to fine-tune the NER layer is small.
+This effect might be due to the differences in nature between the training subsets, whose texts are manually corrected, and the noisy OCR texts used to pretrain CamemBERT.
+Indeed, the learned embeddings from the pretraining are specialized on noisy texts and therefore less adapted to clean texts.
+The pretraining aims at adapting the model to the vocabulary of the domain and to the errors caused by the OCR, which reveals not helpful and even counter-productive when the texts do not contain these types of errors.
 
@@
 
-\subsection{Experiment 2: NER in the presence of noisy OCR texts}
+\subsection{Experiment 2: NER in the presence of noisy OCRed texts}
+CmBERT and CmBERT+ptrn are fine-tuned on the training sets from reference-gold (manually corrected entries) and pero-gold to produce 4 different models.
+Their F1 score measured against the test tests from reference-gold, pero-gold and tesseract-gold are given in \cref{fig:exp_2_eval_ner}.
+The results clearly show that models perform best when both the pre-training and the NER fine-tuning share the same characteristics (here, OCR noise) as the texts to be processed.
 
-Qualitative results
-\textbf{TODO random samples of results + selection of failure cases}
-
-Quantitative results: table + graph ideally (with OCR noise, same format as previous)
-
-Table : NER NN models VS train on {noisy, clean} datasets VS evaluate on {noisy, clean} datasets
+In our tests, pre-training the model brings a slight gain in performance ($\approx 0.5\%$).
+We did not pre-train or fine-tune with texts extracted with Tesseract.
+However, despite a loss of performance the model pre-trained and fine-tuned with pero-gold still gives the best results.
+This is probably due to the fact that the texts produced by Pero-OCR feature characteristics intermediary between human transcriptions and Tesseract.
 
@@
 \end{figure}
-Opt. use synthetic text perturbation as well? Maybe not interesting and too artificial if we have access to 2+ OCR systems.
-(original OCR from BNF, Tesseract 4, Pero OCR…)
 

### 2022-01-20 src-latex/main-paper.tex
review of section 4
Shrink acknowledgments

@@
 \usepackage{booktabs} % Addition
+\usepackage{siunitx} % Addition
 \usepackage{scalerel}
@@
 This work is supported by the French National Research Agency (ANR), as part of the SODUCO project, under Grant ANR-18-CE38-0013.
-
-The authors want to thank Stéphane Bacciochi and Pascal Cristofoli for helping to create the gold reference dataset, 
-Laureencia Maurice for annotating data,
-as well as Guillaume Thomas, Paul Abi Saad, Raphaël Lelièvre, Dorian Mignon, Thibault Cavaciuti and Pierre Sadki for contributing to the annotation platform.
-\joseph{On peut essayer de gagner de la place en déplaçant ça en footnote de 1ere page.}
+%
+The authors want to thank S. Bacciochi and P. Cristofoli for helping to create the reference dataset, 
+L. Morice for annotating data,
+as well as G. Thomas, P. Abi Saad, R. Lelièvre, D. Mignon, T. Cavaciuti and P. Sadki for contributing to the annotation platform.
+%
+% Full names
+% Stéphane Bacciochi
+% Pascal Cristofoli
+% Laureencia Morice
+% Guillaume Thomas
+% Paul Abi Saad
+% Raphaël Lelièvre
+% Dorian Mignon
+% Thibault Cavaciuti
+% Pierre Sadki
 

### 2022-01-20 src-latex/parts/10-introduction.tex
Shrink abstract and put fig in 1st page

@@
 % !TeX root = ../main-paper.tex
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=.9\textwidth]{figs/overview-intro.pdf}
+    \caption{%
+    Overview of the pipeline under study.
+    From previously-extracted images of directory entries, 
+    we perform OCR and named entity recognition (NER) using different techniques.
+    We aim at answering the following questions:
+    \emph{How noisy are modern, out-of-the-box OCR systems?}
+    \emph{What is the behavior of NER when OCR is noisy?}
+    \emph{Can NER be made more robust to OCR noise?}
+    }
+    \label{<label>}
+\end{figure}
+\clearpage% force float flush!
+
 \section{Introduction}
-
-\textbf{TODO sample (simplified) pipeline of the information extraction process}
-\textbf{and/or}
-\textbf{TODO illustration a sample page (full image, large crop, entry crop with highlighted entities)}
 

### 2022-01-20 src-latex/parts/50-xp-setup.tex
Text pour l'eval OCR + fix missing ref dans le protocol.

@@
 
-The Figure \Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems. Two quality evaluations
+\Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems. Two quality evaluations
 are performed in the pipeline, respectively named \emph{OCR Q.A.} and \emph{NER Q.A.}.
@@
 \textbf{OCR Q.A.} is performed after some text normalization of the OCR system outputs. Text normalization consists in
-projecting Unicode characters onto the latin-1 charset (whenever it makes sense) and removing extra symbols (hands,
-crosses). Then, the predicted text is aligned with the reference text using standard tools and the Character Error Rates
-(CER) are computed at the entry level and at the global level.
+projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes sense) and removing extra symbols (hands,
+crosses). Then, the predicted text is aligned with the reference text using standard tools from Stephen V. Rice's
+thesis~\cite{santos.2019.wcmel,neudecker.2021.whdip} and the Character Error Rates (CER) are computed at the entry level
+and at the global level.
 
@@
 In this benchmark, we consider 3 OCR systems well known for historical document analysis: Pero OCR, Tesseract, and Kraken.
-
-\edwin{FIXME: add ref papier eval OCR + ISRI, eval KRAKEN}
 
@@
     \includegraphics[width=\linewidth]{figs/eval-ocr-ner.pdf}
+    \caption{OCR and NER evaluation protocol example.}
     \label{fig.eval-ocr-ner}
-    \caption{OCR and NER evaluation protocol example.}
 \end{figure}
@@
 \bertrand{Find a better name ?}
-The constitution of annotated example sets to train a NER model on a target domain a is a critical preliminary step.
-Often done manually, possibly from bootstrapped annotations, this task is tedious, time consuming and error prone.
-The ability of a model to perform well even with few training examples is a pratical criteria to consider.
+The constitution of annotated example sets to train a NER model on a target domain is a critical preliminary step.
+Often done manually, possibly from bootstrapped annotations, this task is tedious, time-consuming and error-prone.
+The ability of a model to perform well even with few training examples is a practical criteria to consider.
 In this first experiment, we investigate the performance of the three models when fine-tuned for NER on training sets of decreasing sizes.
@@
 
-As the organisation and structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features.
+As the organization and structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features.
 To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, $\approx 19\%$) from the gold reference to test each model on unseen directories.
 Then, a stratified sampling based on the source directory of each entry is run on the remaining set to create a training (6373 entries, $\approx 73\%$ of the gold reference) and a development set (709 entries, $\approx 8\%$).
-This sampling procedure is a convenient way to shape both sets so they reflect the diversity of directories within the gold reference.
-The development set is used to evaluate the models performance during the training phase.
+This sampling procedure is a convenient way to shape both sets, so they reflect the diversity of directories within the gold reference.
+The development set is used to evaluate the model performance during the training phase.
 To generate smaller training sets, we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as before.

### 2022-01-20 src-latex/parts/10-introduction.tex
Shrink intro

@@
 % General context + scope limitation
-The surge of historical documents digitized by cultural heritage institutions and made publicly available on the Web creates a vast knowledge source that can nourish numerous applications in the fields of social sciences, environmental studies or education.
-Most of these documents are texts, either printed or handwritten.
-To make the content of these documents searchable and usable by everyone, the texts engraved within them needs to be extracted through optical character recognition (OCR). 
-However, OCRed texts are not sufficient to build a high level, semantic view of a collection of documents.
-A second step thus aims at recognizing in these texts the pieces of information most likely to be searched for by users, such as named entities (people, organizations, dates, places, etc.).
-Indeed, being able to properly tag text tokens unlocks the ability to relate entities, and provide colleagues from e.g.
-digital humanities with databases ready for exploitation.
+OCRed texts are generally not sufficient to build a high level, semantic view of a collection of historical documents.
+A subsequent stage is often needed to extract the pieces of information most likely to be searched for by users, such as named entities: people, organizations, dates, places, etc.
+Indeed, being able to properly tag text tokens unlocks the ability to relate entities, and provide colleagues from other fields with databases ready for exploitation.
 
@@
 OCR approaches used for modern documents are likely to struggle even on printed historical documents due multiple causes related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
-\bertrand{ajouter une ref.}
-On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts and from the errors introduced into the text by the OCR step.\nathalie{ajouter une ref}
+On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts, and also from the errors introduced into the text by the OCR step.
 
-In this article we focus on a peculiar category of historical printed books: the city trade directories of Paris from the XIX\textsuperscript{th} century (1801 - 1854).
-They contain hundred-pages long lists of people, along with their activity and their address, so much valuable fine-grained knowledge to study the social dynamics of the city over time.
-Despite containing roughly the same information, the diversity in layouts, information organization and digitizing quality is significant, which hardens makes OCR and NER quite challenging.
+In this article we focus on a peculiar category of historical printed books: city trade directories of Paris from the XIX\textsuperscript{th} century (1801 - 1854).
+They contain hundred-pages long lists of people, along with their activity and their address; a much valuable fine-grained knowledge to study the social dynamics of the city over time.
+Despite containing roughly the same information, the diversity in layouts, information organization and digitizing quality is challenging both for OCR and NER.
 
@@
 These directories originate from several publishers, their content is organized following different indexing methods (by name, by activity or by address), printed in various layouts, use different fonts.
-We therefore investigate several state-of-the-art OCR and NER approaches independently and in combination to assess their usability on the directories corpus.
-The contributions of this article are the following:
+We therefore investigate several state-of-the-art OCR and NER approaches independently and in combination to assess their usability on the corpus.
+
+The contributions of this article are the following. \joseph{FIXME recheck le plan à la fin}
 % Contributions
-\begin{enumerate}
-    \item review state of the art regarding NER for historical documents (section 2)
-    \item dataset (section 3)
-    \item list usable techniques (more to evangelize the DAR community) (section 4)
-    \item compare their performance under various training set sizes (thus annotation costs) and OCR noise levels (thus OCR choices/luck/availability) (sec. 5 and 6)
-    \item suggest smart way to leverage side knowledge like lexicons (FIXME iif this works well, maybe a conclusion instead) \nathalie{pas traité dans nos tests... à mettre en perspective}
-\end{enumerate}
+\begin{enumerate*}[(i)]
+    \item We review state of the art OCR and NER systems for historical documents (\cref{sec:related-work}).
+    \item We introduce a new dataset suitable for OCR and NER evaluation (\cref{sec:dataset}).
+    \item We measure the performance of three modern OCR systems on real data (\cref{sec:TODO}).
+    \item We introduce modern NER approaches and study their requirements in terms of training data, as well as the effects of pre-training (\cref{sec:TODO}).
+    \item We show that Transformed-based NER can benefit from pre-training and fine-tuning to improve its performance on noisy OCR (\cref{sec:TODO}).
+\end{enumerate*}

### 2022-01-20 src-latex/main-supplementary-material.tex
New paper structure with 1 section for each experiment

--- 2022-01-19 src-latex/parts/30-dataset.tex
@@
-% !TeX root = ../main-paper.tex
+\documentclass[runningheads,svgnames]{llncs}
+%
+\usepackage[]{graphicx}
+\usepackage{amsmath}
+\usepackage{hyperref}
+\usepackage{cleveref}
+\usepackage{booktabs}
+\usepackage{orcidlink}
+\usepackage{subcaption} % Addition
+\usepackage{multirow} % Addition
+\usepackage{tabularx} % Addition
+\usepackage{siunitx} % Addition
+\usepackage{scalerel}
+\usepackage[inline,shortlabels]{enumitem}
+
+% \usetikzlibrary{shadings,decorations,decorations.pathreplacing}
+\renewcommand\UrlFont{\color{blue}\rmfamily}
+% FIXME proper hyperref setup
+
+\newcommand{\mcite}[1]{\text{\cite{#1}}}
+\newcommand{\mref}[1]{\text{\ref{#1}}}
+
+% Customize table rendering
+\setlength{\tabcolsep}{5pt}
+\renewcommand{\arraystretch}{1.4}
+
+\begin{document}
+%
+\title{%
+Supplementary material for\\
+A Benchmark of Named Entity Recognition Approaches in Historical Documents\\
+Application to 19$^{th}$ Century French Directories%
+}
+%
+\titlerunning{Supplementary material}
+%
+\author{%
+N. Abadie\inst{1}\orcidlink{0000-0001-8741-2398} \and
+E. Carlinet\inst{2}\orcidlink{0000-0001-5737-5266} \and
+J. Chazalon\inst{2}\orcidlink{0000-0002-3757-074X} \and
+B. Duménieu\inst{3}\orcidlink{0000-0002-2517-2058}\\
+{\footnotesize \emph{all authors contributed equally}}%
+}
+%
+\authorrunning{N. Abadie et al.}
+%
+\institute{%
+% 1
+LASTIG, Univ. Gustave Eiffel, IGN-ENSG, F-94160 Saint-Mandé, France\\
+\email{nathalie-f.abadie@ign.fr}%\\
+% \url{https://www.umr-lastig.fr/nathalie-abadie/}%
+\and
+% 2
+EPITA Research \& Development Laboratory (LRDE), Le Kremlin-Bicêtre, France\\
+\email{\{edwin.carlinet,joseph.chazalon\}@lrde.epita.fr}%\\
+%\url{https://lrde.epita.fr}
+\and
+% 3
+CRH-EHESS, Paris, France\\
+\email{bertrand.dumenieu@ehess.fr}%\\
+% \url{http://crh.ehess.fr/index.php?5206}\\
+}
+%
+\maketitle
+%
+\begin{abstract}
+This document contains supplementary material for the paper ``A Benchmark of Named Entity Recognition Approaches in Historical Documents ---
+Application to 19$^{th}$ Century French Directories''.
+\end{abstract}
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% #####     ##     #####    ##     ####   ######   #####
+% #    #   #  #      #     #  #   #       #          #
+% #    #  #    #     #    #    #   ####   #####      #
+% #    #  ######     #    ######       #  #          #
+% #    #  #    #     #    #    #  #    #  #          #
+% #####   #    #     #    #    #   ####   ######     #
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Dataset}
 
-%In this work, we  aim  at  producing  structured  spatio-temporal  data  from the information contained in the XIX\textsuperscript{th} century Parisian trade directories.
-%To do this, it is therefore necessary to extract the text from the scans of the directories to be processed and to identify the entries in the various indexing lists used.
-%Then, in each of these entries, the named entities they contain have to be extracted to produce structured spatio-temporal data representing the evolution over time of the people and businesses listed in these directories, their descriptions and their locations.
-%Both of these extraction tasks require data to train and evaluate the OCR and NER approaches identified as potentially relevant: Tesseract and PERO OCR for the OCR task and Spacy and CamemBERT - pre-trained or fine-tuned - neural models for the NER task.
-
-\subsection{The  XIX\textsuperscript{th} century Parisian trade directories}
-
-The directories to be processed are stored in different libraries in Paris\footnote{\textcolor{green}{Citer les bibliothèques sources?} \joseph{est-ce nécessaire ici ?
-Est-ce que ça peut être dans une annexe ?}}.
-They have been scanned independently and with various levels of quality.
-Moreover, they cover a wide period and have been produced by different publishers and printers.
-Therefore, their contents, indexes, layouts, methods of printing, etc.
-may vary a lot from one directory to another (see Figure \ref{fig:directories}).
-
-\begin{figure}[htb!]
-	   \center{\includegraphics[width=0.9\textwidth]
-	       {./images/DirectoryExcerpts2.png}}
-	  \caption{\label{fig:directories} Examples of directory layouts and contents: 1) Duverneuil et La Tynna 1806 - index by name; 2) Deflandre 1828 - index by activity ; 3) Bottin 1851 –  index by street name}
-\end{figure}
-
-For training as testing purposes, our groundtruth dataset must be as representative as possible of the diversity of the available directories.
-The Table \ref{tab:directories} lists the directories selected, their layout, the structure of their entries, and the number of annotated entries according to the considered index.
-
-For example, the Didot directory of 1854 is organised according to three indexes: by name, by activity, and by street name.
-The structure of the following entry is characteristic of its index by name: "Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47." This entry begins with a person's name, followed by the initials of their first name in parentheses.The symbol indicates that this person was awarded the Légion d'Honneur.
-The following part of this entry represents the person's activity,i.e.
-their profession, or social status, here "professor at the Conservatory of music and declamation".
-The street name and number where the person lives or carries out their activity are written at the end of the entry.
-
-These pieces of information constituting the structure of the entries are the basic data that we want to extract, deduplicate and structure in order to build a spatio-temporal database of XIX\textsuperscript{th} century Parisian inhabitants and businesses.
-With the exception of some potentially wordy activity descriptions, they correspond to named entities.
-However, as Table \ref{tab:directories} shows, while most entries contain the same types of named entities, their order and the way they are written vary from one directory or index to another.
-To provide examples of each entry structure, pages from each type have thus been annotated\footnote{In this work, the street name index entries have not been included as they are very different from the others.
-The strategy to be adopted to process them without loss of quality for the NER model will be the subject of future work.}.
-
+\subsection{Content}
 \begin{table}[h!]
@@
 
-\subsection{Groundtruth data extraction}
-
+\subsubsection{Annotation tools}
 To build our ground truth corpus, we annotate 78 pages randomly sampled from each of the major directory series for which we have digitised copies, using a tool developed for our project (see Figure \ref{fig:annotator}).
@@
 	       {./images/Annot2.png}}
-	  \caption{\label{fig:annotator} GUI for corpus annotation \joseph{remove GUI-related details?}}
+	  \caption{\label{fig:annotator} GUI for corpus annotation}
 \end{figure}
-
-The annotated named entities are as follows:
-\bertrand{Add a ref to Fig. \ref{fig:annotated-example}}
-
-\begin{itemize}
-    \item PER: a person name or a business name, usually referred to as several person names.
-Firts names, initials, or civility mentions are included.
-E.g.: "Alibert (Prosper)", "Allamand frères et Hersent", "Heurtemotte (Vve)", etc.
-    \item TITLE: an honorary title, either text, glyphs or a combination of the two.
-E.g: "O.
-\scalerel*{\includegraphics{./images/LH.png}}{H}" for "Officier de la Légion d'Honneur" or "\scalerel*{\includegraphics{./images/GMLondon.png}}{H}" to represent the Great Medal at the London exhibition.
-    \item ACT: the profession or social status of a person summarized in the single concept of activity. E.g.: "horlogerie", "export en tous genres", "Conseiller d'Etat", "propriétaire".
-    \item LOC: mostly street names ("r. de la Jussienne"), but may also be neighborhoods ("Marais du Temple") or any indirect spatial references ("au Palais Royal").
-    \item CARDINAL: a street number, as part of an address ("16", "5 bis"), or a range of numbers (e.g. "23-25" or "5 à 9").
-    \item FT: a geographic entity type, used to give more details on a location, e.g. "boutique", "atelier", "fab." or "dépôt".
-\end{itemize}
 
@@
 \end{figure}
-Thus, the extracted corpus contains both the bounding boxes of directory entries, their exact text and the named entities they contain.
-It can therefore be used to train and improve the entry detection, OCR and NER tools.
-% book, page, box, text_ocr, ner_xml (+uid within a page)
 
-\subsection{Assessing the groundtruth quality}
-\nathalie{On va utiliser cette sous-section ou pas?}
-\joseph{Pas vraiment car on n'a pas eu le temps de faire ça, mais on peut parler de l'étage d'extraction d'entrées qui est validé manuellement, et des quelques règles de simplification du charset pour homogénéiser entre les annotateurs / pallier des ambiguïtés dans les règles d'annotation (qu'on pourrait encore améliorer)}
-Our dataset is thus made of excerpts TODO interesting points to detail:
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% #####   ######   ####   #    #  #        #####   ####
+% #    #  #       #       #    #  #          #    #
+% #    #  #####    ####   #    #  #          #     ####
+% #####   #            #  #    #  #          #         #
+% #   #   #       #    #  #    #  #          #    #    #
+% #    #  ######   ####    ####   ######     #     ####
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results}
 
-\begin{itemize}
-    \item statistics about entry lengths (histogram of number of char / tokens par entry)
-    \item entities we consider and how we annotated them (how we assessed gold's quality)
-    \item statistics on the distribution of entities across directories and their arrangement within entries
-    \item OCR data (automated systems and clean text + how we created the "clean" text + whether/how we assess clean text's quality)
-    \item Measure inter-annotator agreement ? 
-    \item Explain charset projection
-    \item \dots
-\end{itemize}
+% ---- Bibliography ----
+% \bibliographystyle{splncs04}
+% \bibliography{ref,ref-supp}
+\end{document}

### 2022-01-20 src-latex/parts/20-related-work.tex
Add annotated example
Shrink intro
New paper structure with 1 section for each experiment

@@
 \section{OCR and NER on historical texts}
+\label{sec:related-work}
 
@@
 
-\nathalie{à compléter/retoucher?}
+\begin{itemize}
+    \item Tesseract 4.1.1 (newest version: 5, released Nov. 2021)
+    \item Pero OCR version from git repo, master branch updated on Sep 15, 2021 % 8b20f29    
+    \item other?  % kraken if times allows
+\end{itemize}
+Pero OCR uses ParseNet \joseph{TODO ref} as internal layout parser for line detection.
+Pero OCR uses an LSTM engine \joseph{TODO check + ref} for text line recognition.
+Pero OCR generally works very well, as long as the bounding boxes of the regions to recognize are not too tightly adjusted.
+Trained on newspapers in European languages by its development team.
+Can output results in Latin, Greek and Cyrillic scripts, as well as some commonly-used typographic symbols.
+
+% PERO refs
+% O Kodym, M Hradiš: Page Layout Analysis System for Unconstrained Historic Documents. ICDAR, 2021.
+% M Kišš, K Beneš, M Hradiš: AT-ST: Self-Training Adaptation Strategy for OCR in Domains with Limited Transcriptions. ICDAR, 2021.
+% J Kohút, M Hradiš: TS-Net: OCR Trained to Switch Between Text Transcription Styles. ICDAR, 2021.
 
@@
 \subsection{Named Entity Recognition on historical texts}
-
+\label{subsection:stoa-ner-on-historical-texts}
 \cite{Labusch2020NamedED} pretrained an off-the-shelf multilingual BERT-based NER model on more than 2 000 000 pages of OCRed historical texts in German, French, and English.
@@
 On the other hand, reusing a NER model trained on modern data and adapted to historical texts using supervised or unsupervised approaches seems a good strategy.
+
+\subsection{== À réintégrer depuis ancienne section 4 ==}
+We select two deep-learning-based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
+
+\subsubsection{spacy}
+Spacy is a software library that offers NLP components assembled in modular pipelines specialised by language.
+Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (jan. 2022).
+Hence, we rely on SpaCy's ad hoc pipeline trained on French corpora and capable of named entity recognition.
+% deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.
+The global architecture of these pipelines have not been yet published but are explained by the developers on their website.
+Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
+The decision layer is an adaptation of the transition-based model presented in~\cite{lample2016}.
+As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics.
+The classification layer relies on a finite-state machine whose transition probabilities are learned using a multilayer perceptron.
+% \bertrand{Drop the next sentences if we need space}
+% In 2018~\cite{won2018} evaluated SpaCy's NER ability to detect place names in five corpora of ancient letters written in English.
+% They measured an average F1 score of 0.57.
+% The SpaCy developers claim an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
+%For our experiments, the French model \textit{fr\_core\_news\_lg} is fine-tuned using our ground truth corpus.
+
+% Other possibilities:
+%1. Traditional ML based:
+%    Conditional Random Fields (CRF) - https://pypi.org/project/sklearn-crfsuite/
+%    Maximum-entropy Markov model
+
+%2. Neural Networks based:
+%    LSTMs, bi-LSTM - https://github.com/flairNLP/flair
+%    CNNs (SpaCy uses CNN based architecture)
+%    Transformers (Spacy has recently launched it) - %https://spacy.io/universe/project/spacy-transformers
+
+\subsubsection{Transformers / Bert / Hugging face}
+Language model based on Transformers, like BERT, have become a new paradigm for NER\cite{li2020}. 
+The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer.
+They can also be pretrained in an unsupervised way on historical texts for domain adaptation.
+As the directories are written in French, we chose the language model CamemBERT \cite{martin-etal-2020-camembert}, a Transformer model trained on a French corpus.
+\joseph{plutôt tourner ça pour dire qu'il existe des modèles FR / qu'on est limités par la dispo de modèles FR pré-entraînés ?}
+
+\subsection{Pipeline summary}
+
+\Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems.
+
+\begin{figure}[tb]
+    \includegraphics[width=\linewidth]{figs/protocol.pdf}
+    \caption{Scheme of the evaluation protocol. \joseph{Missing "Ground truth" for NER stage?}}
+    \label{fig.protocol}
+    \end{figure}

### 2022-01-20 src-latex/parts/30-dataset.tex
Shrink intro
New paper structure with 1 section for each experiment

@@
 \section{Dataset}
+\label{sec:dataset}
 
@@
 	       {./images/DirectoryExcerpts2.png}}
-	  \caption{\label{fig:directories} Examples of directory layouts and contents: 1) Duverneuil et La Tynna 1806 - index by name; 2) Deflandre 1828 - index by activity ; 3) Bottin 1851 –  index by street name}
+	  \caption{\label{fig:directories} Examples of directory layouts and contents: 1) Duverneuil et La Tynna 1806 - index by name; 2) Deflandre 1828 - index by activity ; 3) Bottin 1851 - index by street name}
 \end{figure}
@@
 
-\begin{table}[h!]
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{|l|l|l|l|l|l|}
-\hline
-\multicolumn{1}{|c|}{\textbf{Directory publisher or name}} & \multicolumn{1}{c|}{\textbf{Year}} & \multicolumn{1}{c|}{\textbf{\# Col}} & \multicolumn{1}{c|}{\textbf{Index}} & \multicolumn{1}{c|}{\textbf{Entry structure}} & \multicolumn{1}{c|}{\textbf{\# Ann}}
-\\ \hline
-Favre \& Duchesne & 1798 & 1 & by activity & {N [A], SN, SNUM - SEC.} & 233
-\\ \hline
-Duverneuil et La Tynna & 1801 & 1 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & {N, SN, SNUM. SEC.} & \begin{tabular}[c]{@{}l@{}} 136 \\ 103\end{tabular}
-\\ \hline
-{Notables communaux de la Seine} & 1801 & 1 & by name & {N, [A,] SN[, SNUM].} & 92
-\\ \hline
-Duverneuil et La Tynna & 1805 &	2 &	\begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F)], SN, SNUM.
-- SEC.\\ N [(C)], [(A),] SN, SNUM.
-- SEC.\end{tabular} & \begin{tabular}[c]{@{}l@{}} 155 \\ 221 \end{tabular}
-\\ \hline
-Duverneuil et La Tynna & 1806 & 2 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F)], SN, SNUM.\\ N [(C)], SN, SNUM.[ - SEC.]\end{tabular}.
-& \begin{tabular}[c]{@{}l@{}} 186\\ 0\end{tabular}
-\\ \hline
-La Tynna & 1813	& 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N,[A,] SN[, SNUM].\\ N, [(T)] A, SN[, SNUM] [, D].\end{tabular} & \begin{tabular}[c]{@{}l@{}} 135 \\ 177 \end{tabular}
-\\ \hline
-Panckoucke Commerces (Dulac) & 1820 & 1 & by name & N [(F)], A, [(T),] SN, SNUM[, P]. & 0
-\\ \hline
-Panckoucke Habitants (Dulac) & 1820	& 1 & by name & N [(T)], [A,] SN, SNUM [, D]. & 0
-\\ \hline
-Bottin (serie 1: 1819-1838) & 1820 & 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F \textbar C)][T], [P], SN, SNUM.\\N [(F \textbar C)][T], A, SN, SNUM.
-CR\end{tabular} & \begin{tabular}[c]{@{}l@{}} 128 \\ 151 \end{tabular}
-\\ \hline
-Bottin (serie 1: 1819-1838) & 1827 & 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], A, SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}} 228 \\ 285 \end{tabular}
-\\ \hline
-Deflandre & 1828 & 2 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}} N [(F \textbar C)][T], [P], SN, SNUM.
-\\N [(F \textbar C)][T], A, SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}}115\\ 229\end{tabular}
-\\ \hline
-Deflandre & 1829 & 2 & \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], A, SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}}177\\ 236\end{tabular}
-\\ \hline 
-Bottin (serie 1: 1819-1838) & 1837 & 2	& \begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], [A], SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}}158\\ 557\end{tabular}
-\\ \hline
-Cambon - Almanach général & 1841 & 2 &	\begin{tabular}[c]{@{}l@{}}by activity\\ by name\end{tabular} & \begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], A, SN, SNUM.\end{tabular} & \begin{tabular}[c]{@{}l@{}}182\\ 486\end{tabular}
-\\ \hline
-Didot & 1841a &	\begin{tabular}[c]{@{}l@{}}3\\ 4\end{tabular} &	\begin{tabular}[c]{@{}l@{}}by activity \\ by name \end{tabular} & 	\begin{tabular}[c]{@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.
-\\ N [(F \textbar C)][T], [A], SN, SNUM.
-\end{tabular} & \begin{tabular}[c]{@{}l@{}}246\\ 843\end{tabular}
-\\ \hline
-Didot & 1851a & \begin{tabular}[c]{@{}l@{}l@{}}4\\ 4\\ 5\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}by activity\\ by name\\ by street name\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}N [(F \textbar C)][T], [P], SN, SNUM.\\ N [(F \textbar C)][T], [A], SN, SNUM.\\SNUM N [(F \textbar C)][T], [A].\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}309\\ 961\\ 0\end{tabular}
-\\ \hline
-Didot & 1854a & \begin{tabular}[c]{@{}l@{}l@{}}4\\ 3\\ 5\end{tabular} &	\begin{tabular}[c]{@{}l@{}l@{}} by activity \\ by name \\ by street name\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}} N [(F \textbar C)], [P], SN, SNUM [,T].\\ N [(F \textbar C)][T], [A], SN, SNUM.\\ SNUM N [(F \textbar C)][T], [A].\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}106\\ 362\\ 0\end{tabular}
-\\ \hline
-Bottin (serie 3: 1854-1856) & 1854a & \begin{tabular}[c]{@{}l@{}l@{}}3\\ 2\\ 4\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}} by activity \\ by name \\ by street name\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}N [(F \textbar C)], [P], SN, SNUM [,T].\\ N [(F \textbar C)][T], [A], SN, SNUM.\\ SNUM N [(F \textbar C)], [A] [,T].\end{tabular} & \begin{tabular}[c]{@{}l@{}l@{}}253\\ 633\\ 0\end{tabular}
-\\ \hline
-Didot-Bottin & 1860a	& 3	& by name & N [(F \textbar C)][T], [A], SN, SNUM. & 378
-\\ \hline
-Didot-Bottin & 1861a	& 3	& by name & N [(F \textbar C)][T], [A], SN, SNUM. & 402
-\\ \hline
-\end{tabular}%
-}
-\caption{The directories used to build our dataset.
-Letters next to year of publication are used to identify the volume of the directory.
-\# Col stands for "Number of columns", and \# Ann stands for "Number of annotated entries".
-Entry structure is described as follows: N - entry name (person or business); F - first name; T - honorary title; C - civility; A - activity; P - precision; SN - street name; SNUM - street number; SEC - section; D - district; CR - cross-reference; bracketed elements are optional.}
-\label{tab:directories}
-\end{table}
+\subsection{Groundtruth data extraction}
+\joseph{TODO: separate entry detection, OCR and NER stages explicitely}
 
-\subsection{Groundtruth data extraction}
+First, the layout extraction and entry segmentation of the page are performed with a semi-automated system and
+checked by a human. Afterward, an OCR system runs on the thumbnails of each segmented entry to extract its text. An
+entry might span over multiple text lines but is always a single block. Thus, the most adapted mode is chosen when
+the OCR system allows for the detection mode (e.g. the \emph{block} mode for \emph{Tesseract}).
 
-To build our ground truth corpus, we annotate 78 pages randomly sampled from each of the major directory series for which we have digitised copies, using a tool developed for our project (see Figure \ref{fig:annotator}).
-First, the image of a given directory page is loaded.
-Second, entries are automatically detected and boxes are drawn around them as shown in the left panel of the Graphical User Interface (GUI).
-Then the text of each entry is extracted by an OCR tool and finally named entities are annotated by a NER tool: the results are displayed respectively in the first and second tabs of the right panel of the GUI.
-The entry detection, OCR and NER tasks are performed automatically.
-But as they are prone to errors, their results can be corrected manually through the GUI.
-In the end, the entry bounding boxes, their text, and the named entities are stored in a JSON file.
-
-\begin{figure}[htb!]
-	   \center{\includegraphics[width=0.9\textwidth]
-	       {./images/Annot2.png}}
-	  \caption{\label{fig:annotator} GUI for corpus annotation \joseph{remove GUI-related details?}}
-\end{figure}
+OCR: some basic charset normalization is performed after human annotation:
+- remove annotation codes for glyphs which have no character in unicode
+- alleviate some ambiguities in annotation rules (n° -> no, --- -> -, homogenize star and hand symbols allowed)
 
 The annotated named entities are as follows:
-\bertrand{Add a ref to Fig. \ref{fig:annotated-example}}
 
@@
 
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=\textwidth]{images/examples_labeled_entries.pdf}
-    \caption{\label{fig:annotated-example}Two annotated entries showing the different labels used in the corpus.}
-\end{figure}
 Thus, the extracted corpus contains both the bounding boxes of directory entries, their exact text and the named entities they contain.
@@
 
-\subsection{Assessing the groundtruth quality}
-\nathalie{On va utiliser cette sous-section ou pas?}
-\joseph{Pas vraiment car on n'a pas eu le temps de faire ça, mais on peut parler de l'étage d'extraction d'entrées qui est validé manuellement, et des quelques règles de simplification du charset pour homogénéiser entre les annotateurs / pallier des ambiguïtés dans les règles d'annotation (qu'on pourrait encore améliorer)}
-Our dataset is thus made of excerpts TODO interesting points to detail:
+\subsection{Summary}
+\joseph{TODO intégrer dans les sections avant ?}
 
-\begin{itemize}
-    \item statistics about entry lengths (histogram of number of char / tokens par entry)
-    \item entities we consider and how we annotated them (how we assessed gold's quality)
-    \item statistics on the distribution of entities across directories and their arrangement within entries
-    \item OCR data (automated systems and clean text + how we created the "clean" text + whether/how we assess clean text's quality)
-    \item Measure inter-annotator agreement ? 
-    \item Explain charset projection
-    \item \dots
-\end{itemize}
+\begin{verbatim}
+[('<LOC>', 9709),
+('<PER>', 8788),
+('<CARDINAL>', 8747),
+('<ACT>', 6472),
+('<TITRE>', 483),
+('<FT>', 43)]
+\end{verbatim}
+
+ valid ocr: 8765 entries
+
+ valid ner (ref \& pero \& tess): 8341
+
+\joseph{parler de l'étage d'extraction d'entrées qui est validé manuellement, et des quelques règles de simplification du charset pour homogénéiser entre les annotateurs / pallier des ambiguïtés dans les règles d'annotation (qu'on pourrait encore améliorer)}
+
+TODO a word about the charset, and existence of glyphs unknown to the OCR
+
+TODO average entry len (in char)
+
+TODO average number of tags/entities per entry
+
+TODO lists actual dataset content: set of entries. For each entry:
+\begin{itemize*}
+    \item ref to original book, page, region
+    \item human transcription (OCR ref)
+    \item human NER (NER ref)
+    \item OCR produced by Pero OCR, Tesseract v4, Kraken (precise model used)
+\end{itemize*}

### 2022-01-20 src-latex/parts/40-ner-approaches-BACKUP.tex
New paper structure with 1 section for each experiment

--- 2022-01-19 src-latex/parts/40-ner-approaches.tex
@@
 We also create an additional CamemBERT model pretrained on a collection of unannotated directory entries extracted with Pero-OCR.
-The NER layer of each three models is then fine-tuned on historical directories as detailed in\cref{subsection:experiment-1-setup} and\cref{subsection:experiment-2-setup}.
+The NER layer of each three models is then fine-tuned on historical directories as detailed in \cref{subsection:experiment-1-setup} and \cref{subsection:experiment-2-setup}.
 
@@
 Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (jan. 2022).
-Hence we rely on SpaCy's ad hoc pipelines trained on French corpora and capable of named entity recognition.
+Hence, we rely on SpaCy's ad hoc pipeline trained on French corpora and capable of named entity recognition.
 % deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.

### 2022-01-20 src-latex/parts/40-ocr-bench.tex
New paper structure with 1 section for each experiment

% !TeX root = ../main-paper.tex
\section{OCR benchmark}

\subsection{Dataset}
input: image crops for each entry
expected output: human transcription

8765 entries used, \joseph{FIXME number of chars}

\subsection{Metrics}
\textbf{OCR Q.A.} is performed after some text normalization of the OCR system outputs. Text normalization consists in
projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes sense) and removing extra symbols (hands,
crosses). Then, the predicted text is aligned with the reference text using standard tools from Stephen V. Rice's
thesis~\cite{santos.2019.wcmel,neudecker.2021.whdip} and the Character Error Rates (CER) are computed at the entry level
and at the global level.

\begin{align}
\mathrm{CER} &=  \frac{\#Errors}{\text{Reference Text Length}} & & \mathrm{CER}_\mathrm{norm} =  \frac{\#Errors}{\text{Alignment Length}} 
\end{align}

\subsection{Systems/Variants under test}
In this benchmark, we consider 3 OCR systems well known for historical document analysis: Pero OCR, Tesseract, and Kraken.

\subsection{Protocol}

\subsection{Results and discussion}

\begin{figure}

\subcaptionbox{}[.5\linewidth]{
\begin{tabular}{rll}
\toprule
            & CER & CER$_{norm.}$ \\
\midrule
Pero-OCR  & 3.78\% & 3.76\% \\   
Tesseract & 6.56\% & 6.45\% \\
Kraken    & 15.72\% & 15.46\% \\  
\bottomrule
\end{tabular}

\bigskip

\includegraphics[width=\linewidth]{images/ocr-eval-2.pdf}
}
\subcaptionbox{}[.5\linewidth]{
\includegraphics[width=\linewidth]{images/ocr-eval-1.pdf}
}
\caption{Character error rates at entry-level for Pero-OCR, Kraken and Tesseract. (a) Global CER and distribution of the CER per entry. (b)
joint plot of the per-entry error rate showing that Pero-OCR and Tesseract do not fail on the same entries.}  
\label{fig.ocr-results}
\end{figure}

\Cref{fig.ocr-results} compares the performance of the OCR systems on our dataset. We can see Kraken's performance are
not as good as the two first OCR. This is partially due to the fact that the closest available model is for English text
and so it misses French specific symbols. On the over hand, even when using a French model trained on French 19th
documents, performance does not increase (and relaxing the character match rules does not help either). Tesseract and
Pero-OCR are performing better on this dataset ``out-of-the-box''. With no fine-tuning, Pero-OCR gets the best accuracy
with less than 4\% character errors. Many of them are even due to a bad line detection in case of multi-lines entries
and are not related to the OCR system itself. \Cref{fig.ocr-results} (b) shows that errors from the two best OCR are not
committed on the same entries (if so, all points would be on the diagonal line) and that combining the outputs of
Pero-OCR and Tesseract would improve the overall recognition quality. Also, we will not consider Kraken in the following
NER experiments because of its recognition rate is already low and cannot be artificially degraded in a controlled setup.

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
New paper structure with 1 section for each experiment

--- 2022-01-20 src-latex/parts/50-xp-setup.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Experimental Setup}
+\section{Experiment 1: NER sensibility to the number of training examples}
+\label{sec:ner-xp1}
 
-\begin{figure}[tb]
-\includegraphics[width=\linewidth]{figs/protocol.pdf}
-\caption{Scheme of the evaluation protocol. \joseph{Missing "Ground truth" for NER stage?}}
-\label{fig.protocol}
-\end{figure}
+The constitution of annotated example sets to train a NER model on a target domain is a critical preliminary step.
+Often done manually, possibly from bootstrapped annotations, this task is tedious, time-consuming and error-prone.
+The ability of a model to perform well even with few training examples is a practical criteria to consider.
+In this first experiment, we investigate the performance of the three models when fine-tuned for NER on training sets of decreasing sizes.
 
-\subsection{Evaluation protocol \& Metrics}
+\subsection{Dataset}
+input: ocr ref (human transcription)
+expected output: ner ref (human tagging)
 
-\Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems. Two quality evaluations
-are performed in the pipeline, respectively named \emph{OCR Q.A.} and \emph{NER Q.A.}.
+8765 entries, \joseph{FIXME XXXX entities to detect}
 
-First, the layout extraction and entry segmentation of the page are performed with a semi-automated system and
-checked by a human. Afterward, an OCR system runs on the thumbnails of each segmented entry to extract its text. An
-entry might span over multiple text lines but is always a single block. Thus, the most adapted mode is chosen when
-the OCR system allows for the detection mode (e.g. the \emph{block} mode for \emph{Tesseract}).
-
-\textbf{OCR Q.A.} is performed after some text normalization of the OCR system outputs. Text normalization consists in
-projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes sense) and removing extra symbols (hands,
-crosses). Then, the predicted text is aligned with the reference text using standard tools from Stephen V. Rice's
-thesis~\cite{santos.2019.wcmel,neudecker.2021.whdip} and the Character Error Rates (CER) are computed at the entry level
-and at the global level.
-
-\begin{align}
-\mathrm{CER} &=  \frac{\#Errors}{\text{Reference Text Length}} & & \mathrm{CER}_\mathrm{norm} =  \frac{\#Errors}{\text{Alignment Length}} 
-\end{align}
-
-In this benchmark, we consider 3 OCR systems well known for historical document analysis: Pero OCR, Tesseract, and Kraken.
-
+\subsection{Metrics}
 Next, a NER system extracts the named entities from the text of each entry output by the OCR and the \textbf{NER Q.A.}
@@
 
-The whole process is illustrated on~\cref{fig.eval-ocr-ner}. The OCR text and the GOLD text are first aligned to
-evaluate the OCR accuracy. As there are 11 mismatches over 56 aligned characters, the CER is thus 24\%. This alignment
-is then used to back-project the GOLD tags to build the tagged NER reference string. Finally, the NER system runs on the
-OCR text; its output is compared to the NER reference string. There is only 2 over 3 tags matching in the prediction (precision),
-while only 2 over 4 entities are matched in the reference (recall). It follows an overall f-score of 0.4.
+\subsection{Systems/Variants under test}
+We select two deep-learning-based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
+We also create an additional CamemBERT model pretrained on a collection of unannotated directory entries extracted with Pero-OCR.
+The NER layer of each three models is then fine-tuned on historical directories as detailed in \cref{subsection:experiment-1-setup} and \cref{subsection:experiment-2-setup}
 
-\begin{figure}[tb]
-    \includegraphics[width=\linewidth]{figs/eval-ocr-ner.pdf}
-    \caption{OCR and NER evaluation protocol example.}
-    \label{fig.eval-ocr-ner}
-\end{figure}
+\subsubsection{SpaCy}
+We use the French pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the deep-sequoia and wikiner-fr corpora.
+The default NER labels known to the pipeline are PER, ORG, MISC and LOC.
+In experiment 1 the base model is fine-tuned on every of the 8 training sets using the same training parameters.
+Early stopping is activated with a patience value of 1600 training steps without improvement of the f1 score.
 
-\subsection{Experiment 1: NER sensibility to the number of training examples}
-\label{subsection:experiment-1-setup}
-\bertrand{Find a better name ?}
-The constitution of annotated example sets to train a NER model on a target domain is a critical preliminary step.
-Often done manually, possibly from bootstrapped annotations, this task is tedious, time-consuming and error-prone.
-The ability of a model to perform well even with few training examples is a practical criteria to consider.
-In this first experiment, we investigate the performance of the three models when fine-tuned for NER on training sets of decreasing sizes.
+\subsubsection{CamemBERT pretrained on directories~[CmBERT+ptrn]}
+In order to evaluate the benefits of pre-training the language model on OCR texts from domain-related documents, we adapt the CmBERT embeddings using \num{845.000} entries randomly selected wihtin the collection of directories and extracted with Pero-OCR.
+We achieve this by training the CmBERT model for 3 epochs on two unsupervised tasks: next sentence prediction and masked language model.
 
+\subsubsection{Huggingface CamemBERT}
+\label{sub:ner-xp1-sysytems-huggingcam}
+Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
+Our baseline CmBERT model is CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
+Its NER head is a linear model with a Softmax function.
+CmBERT and CmBERT+ptrn are always fine-tuned using the same parameters, with at most 5000 training steps and an early stopping condition set to 3 evaluations in experiment 1 (5 in experiment 2) without improvement of the f1 score.
+Evaluations are performed every 100 steps.
+
+\subsection{Protocol}
 To do so, we split the gold reference of manually annotated entries into a training set, a development set, and a test set. 
@@
 
-% Move to results
-%All metrics are evaluated on the test set, yet the biased metrics on the development set are added in the article material as additional information.
+\subsection{Results and discussion}
 
-\subsection{Experiment 2: Impact of OCR nosie on named entity recognition}
-\label{subsection:experiment-2-setup}
-Noise introduced by OCR is known to impede the named entity recognition process (see section~\ref{subsection:stoa-ner-on-historical-texts}).
-%The second experiment assesses the impact of the OCR on the quality of the NER results.
-The second experiment addresses two questions (1) which model is the most robust to OCR noise and (2) what is the most appropriate strategy to fine-tune the NER layer if the input texts to label are noisy ?
-To this end, we leverage the labeled OCR datasets produced following to the method explained in Section 5, in addition to the gold reference.
+Table \ref{tab:experiment-1-models-performances} and Figure \ref{fig:f1-vs-trainsize} show the mean F1 score, precision, and recall computed on the test dataset, for 5 fine tunings of each model and for each of the 8 training subsets.
+CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.
+The F1 score increases by 4.6 points between 49 and 796 examples for CmBERT (resp. 1.6 for CmBERT+ptrn and 5.1 for SpaCy NER) but only by 1 point between 796 and 6373 examples (resp. 0.6 and 1.4).
+The models derived from CamemBERT always outperform the Spacy model.
 
-First and foremost, to make sure that the datasets contain the same entries a preprocessing step must be performed.
-Indeed, entries where the OCR produced an empty string and those for which no entity could be projected from the gold reference text are ignored.
-To overcome this problem we filter each dataset by keeping only the entries common to all.
-At the cost of losing  5\% of the gold references entries, we thus have four new annotated datasets containing the exact same entries.
-They are denoted hereafter {reference, pero, tesseract}-gold.
-Finally each is is split into a training, a dev and a test subset following the procedure presented in\cref{subsection:experiment-1-setup}.
+It appears that pretraining the CamemBERT model on OCRed texts seems worth it only when a the training set used to fine-tune the NER layer is small.
+This effect might be due to the differences in nature between the training subsets, whose texts are manually corrected, and the noisy OCR texts used to pretrain CamemBERT.
+Indeed, the learned embeddings from the pretraining are specialized on noisy texts and therefore less adapted to clean texts.
+The pretraining aims at adapting the model to the vocabulary of the domain and to the errors caused by the OCR, which reveals not helpful and even counter-productive when the texts do not contain these types of errors.
 
-We fine-tune CmBERT and CmBERT+ptrn on reference-gold and pero-gold to create two versions of each model: the former trained on manually corrected entries and the latter on OCR entries.
-SpaCy NER is left aside as results from experiment 1 show that it is outperformed by BERT models.
-Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold and tesseract-gold.
+\begin{table}[h!]
+\centering
+\caption{F1 score, precision and recall measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER in experiment 1 for 8 training sets of increasing sizes.}
+\begin{tabular}{llrrrrrrrr}
+       & Training examples &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
+       & \% & 0.8   & 1.6   & 3.1   & 6.2   & 12.5  & 25.0  & 50.0  & 100.0 \\
+\midrule\bottomrule
+\multirow{3}{*}{\rotatebox{90}{F1 score}} & CmBERT &  89.5 &  90.5 &  92.7 &  93.3 &  \textbf{94.1} &  \textbf{94.9} &  \textbf{94.6} &  \textbf{95.1} \\
+       & CmBERT-ptrn &  \textbf{92.2} &  \textbf{92.9} &  \textbf{93.6} &  \textbf{93.8} &  93.8 &  94.1 &  \textbf{94.6} &  94.4 \\
+       & SpaCy NER &  87.0 &  89.0 &  90.3 &  91.9 &  92.1 &  92.8 &  93.2 &  93.5 \\
+\cline{1-10}
+\multirow{3}{*}{\rotatebox{90}{Precision}} & CmBERT &  87.4 &  88.7 &  91.5 &  92.7 &  93.3 &  94.9 &  93.9 &  95.1 \\
+       & CmBERT-ptrn &  90.8 &  91.8 &  92.9 &  93.0 &  93.0 &  93.4 &  94.1 &  93.9 \\
+       & SpaCy NER &  85.6 &  87.7 &  90.0 &  92.0 &  92.4 &  92.8 &  93.1 &  93.7 \\
+\cline{1-10}
+\multirow{3}{*}{\rotatebox{90}{Recall}} & CmBERT &  91.6 &  92.5 &  93.9 &  93.9 &  94.9 &  94.9 &  95.4 &  95.1 \\
+       & CmBERT-ptrn &  93.6 &  94.0 &  94.4 &  94.6 &  94.6 &  94.8 &  95.0 &  94.9 \\
+       & SpaCy NER &  88.6 &  90.4 &  90.7 &  91.7 &  91.9 &  92.8 &  93.3 &  93.4 \\
+\end{tabular}
+\label{tab:experiment-1-models-performances}
+\end{table}
 
-\subsection{Implementation details}
-
-\subsubsection{OCR system parameters}
-\nathalie{Y a-t-il des précisions à ajouter sur la façon dont ont été utilisés les OCR? Ou bien ce sont les versions "sur étagère" qui ont servi?}
-\joseph{Totalement sur l'étagère. On a normalisé un peu les données après, c'est tout.}
-\begin{itemize}
-    \item Tesseract 4.1.1 (newest version: 5, released Nov. 2021)
-    \item Pero OCR version from git repo, master branch updated on Sep 15, 2021 % 8b20f29    
-    \item other?  % kraken if times allows
-\end{itemize}
-
-Pero OCR uses ParseNet \joseph{TODO ref} as internal layout parser for line detection.
-Pero OCR uses an LSTM engine \joseph{TODO check + ref} for text line recognition.
-Pero OCR generally works very well, as long as the bounding boxes of the regions to recognize are not too tightly adjusted.
-Trained on newspapers in European languages by its development team.
-Can output results in Latin, Greek and Cyrillic scripts, as well as some commonly-used typographic symbols.
-
-% PERO refs
-% O Kodym, M Hradiš: Page Layout Analysis System for Unconstrained Historic Documents. ICDAR, 2021.
-% M Kišš, K Beneš, M Hradiš: AT-ST: Self-Training Adaptation Strategy for OCR in Domains with Limited Transcriptions. ICDAR, 2021.
-% J Kohút, M Hradiš: TS-Net: OCR Trained to Switch Between Text Transcription Styles. ICDAR, 2021.
-
-\subsubsection{SpaCy}
-We use the French pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the deep-sequoia and wikiner-fr corpora.
-The default NER labels known to the pipeline are PER, ORG, MISC and LOC.
-In experiment 1 the base model is fine-tuned on every of the 8 training sets using the same training parameters.
-Early stopping is activated with a patience value of 1600 training steps without improvement of the f1 score.
-
-% \bertrand{Move to results}
-%For each size of the training dataset, 5 runs are performed, always with the same parameters, and the mean value of each performance score is computed.
-
-%library and leverage the library utilities to fine-tune it on various training and validation datasets derived from our ground truth dataset (see Section 5.2 for details).
-The base model is fine-tuned in turn on each of the 8 generated training datasets, using the same training parameters.
-Early stopping is activated with patience parameters set at 1600.
-For each size of the training dataset, 5 runs are performed, always with the same parameters, and the mean value of each performance score is computed.
-
-%and Huggingface for the BERT 
-%tok2vec(words embeddings + encoding) + attention layer  +  transition-based %model.
-
-\subsubsection{Huggingface CamemBERT}
-Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
-Our baseline CmBERT model is CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
-Its NER head is a linear model with a Softmax function.
-CmBERT and CmBERT+ptrn are always fine-tuned using the same parameters, with at most 5000 training steps and an early stopping condition set to 3 evaluations in experiment 1 (5 in experiment 2) without improvement of the f1 score.
-Evaluations are performed every 100 steps.
-
-Details can be found for all models in the source code publicly available on the GitHub repository \bertrand{LIEN GITHUB}.
-
-%$BERT_{reference}$ The base model is fine-tuned in turn on each of the training datasets generated from the groundtruth corpus.
-Training parameters are set depending on the considered experiment: early stopping is activated with the patience parameter set at 3 for experiment 1 and at 5 for experiment 2.
-
-%$BERT_{pero-ocr}$ This model is used for experiment 2 only.
-The base model is fine-tuned on the training dataset generated by projecting the named entity annotations on the text extracted by PERO OCR, using the same parameters as for the $BERT_{reference}$ model.
-
-%$BERT_{ptrn-reference}$ The base model is pretrained using 845000 entries extracted by PERO OCR with the Next  Sentence  Prediction  (NSP) and  Masked  Language  Model(MLM) tasks.
-Then it is fine-tuned with the same parameters and on the same training datasets generated from the groundtruth corpus as the $BERT_{reference}$ model.
-
-%$BERT_{ptrn-pero-ocr}$ This model is used for experiment 2 only.
-The base model is pretrained like the $BERT_{ptrn-reference}$ model and it is fine-tuned with the same parameters and on the same training datasets as the $BERT_{pero-ocr}$ model.
+\begin{figure}[h!]
+	   \center{\includegraphics[width=\textwidth]
+	       {images/experiment-1-models-performances.pdf}}
+	  \caption{\label{fig:f1-vs-trainsize} Precision, recall and f1 score from table~\ref{tab:experiment-1-models-performances}}.
+\end{figure}

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
New paper structure with 1 section for each experiment

--- 2022-01-19 src-latex/parts/50-xp-setup.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Experimental Setup}
+\section{Experiment 2: Impact of OCR nosie on named entity recognition}
+\label{sec:ner-xp2}
+Noise introduced by OCR is known to impede the named entity recognition process (see section~\ref{subsection:stoa-ner-on-historical-texts}).
+%The second experiment assesses the impact of the OCR on the quality of the NER results.
+The second experiment addresses two questions (1) which model is the most robust to OCR noise and (2) what is the most appropriate strategy to fine-tune the NER layer if the input texts to label are noisy ?
 
-\begin{figure}[tb]
-\includegraphics[width=\linewidth]{figs/protocol.pdf}
-\caption{Scheme of the evaluation protocol. \joseph{Missing "Ground truth" for NER stage?}}
-\label{fig.protocol}
-\end{figure}
+\subsection{Dataset}
 
-\subsection{Evaluation protocol \& Metrics}
+To this end, we leverage the labeled OCR datasets produced following to the method explained in Section 5, in addition to the gold reference.
 
-The Figure \Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems. Two quality evaluations
-are performed in the pipeline, respectively named \emph{OCR Q.A.} and \emph{NER Q.A.}.
+First and foremost, to make sure that the datasets contain the same entries a preprocessing step must be performed.
+Indeed, entries where the OCR produced an empty string and those for which no entity could be projected from the gold reference text are ignored.
+To overcome this problem we filter each dataset by keeping only the entries common to all.
+At the cost of losing  5\% of the gold references entries, we thus have four new annotated datasets containing the exact same entries.
+They are denoted hereafter {reference, pero, tesseract}-gold.
+Finally each is is split into a training, a dev and a test subset following the procedure presented in\cref{subsection:experiment-1-setup}.
 
-First, the layout extraction and entry segmentation of the page are performed with a semi-automated system and
-checked by a human. Afterward, an OCR system runs on the thumbnails of each segmented entry to extract its text. An
-entry might span over multiple text lines but is always a single block. Thus, the most adapted mode is chosen when
-the OCR system allows for the detection mode (e.g. the \emph{block} mode for \emph{Tesseract}).
-
-\textbf{OCR Q.A.} is performed after some text normalization of the OCR system outputs. Text normalization consists in
-projecting Unicode characters onto the latin-1 charset (whenever it makes sense) and removing extra symbols (hands,
-crosses). Then, the predicted text is aligned with the reference text using standard tools and the Character Error Rates
-(CER) are computed at the entry level and at the global level.
-
-\begin{align}
-\mathrm{CER} &=  \frac{\#Errors}{\text{Reference Text Length}} & & \mathrm{CER}_\mathrm{norm} =  \frac{\#Errors}{\text{Alignment Length}} 
-\end{align}
-
-In this benchmark, we consider 3 OCR systems well known for historical document analysis: Pero OCR, Tesseract, and Kraken.
-
-\edwin{FIXME: add ref papier eval OCR + ISRI, eval KRAKEN}
-
-Next, a NER system extracts the named entities from the text of each entry output by the OCR and the \textbf{NER Q.A.}
-is performed. The NER system outputs a text with tags that enclose the entities. To assess the quality of the entity
-extraction, we rely on a technique similar as for the OCR evaluation. The predicted text is aligned with the reference
-text and the tags are projected in the alignment. Then, the precision, recall, and f-measure (or f-score) are computed considering
-only the exact matches between entities. Precision and recall are defined by equation (2); the f-measure is their harmonic mean.
-
-\begin{align}
-    \mathrm{precision} &= \frac{\text{\#exact matches}}{\text{\#entries in prediction}} & \mathrm{recall} &= \frac{\text{\#exact matches}}{\text{\#entries in reference}}
-\end{align}
+\subsection{Metrics}
+prec/recall/F1 as in \ref{sec:ner-xp1}
 
@@
     \includegraphics[width=\linewidth]{figs/eval-ocr-ner.pdf}
+    \caption{OCR and NER evaluation protocol example.}
     \label{fig.eval-ocr-ner}
-    \caption{OCR and NER evaluation protocol example.}
 \end{figure}
 
-\subsection{Experiment 1: NER sensibility to the number of training examples}
-\label{subsection:experiment-1-setup}
-\bertrand{Find a better name ?}
-The constitution of annotated example sets to train a NER model on a target domain a is a critical preliminary step.
-Often done manually, possibly from bootstrapped annotations, this task is tedious, time consuming and error prone.
-The ability of a model to perform well even with few training examples is a pratical criteria to consider.
-In this first experiment, we investigate the performance of the three models when fine-tuned for NER on training sets of decreasing sizes.
-
-To do so, we split the gold reference of manually annotated entries into a training set, a development set, and a test set. 
-The training set is then gradually reduced in size while maintaining the relative frequency of directories within it.
-
-As the organisation and structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features.
-To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, $\approx 19\%$) from the gold reference to test each model on unseen directories.
-Then, a stratified sampling based on the source directory of each entry is run on the remaining set to create a training (6373 entries, $\approx 73\%$ of the gold reference) and a development set (709 entries, $\approx 8\%$).
-This sampling procedure is a convenient way to shape both sets so they reflect the diversity of directories within the gold reference.
-The development set is used to evaluate the models performance during the training phase.
-To generate smaller training sets, we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as before.
-We stop if a directory has only one entry left, or if the current training subset contains less than 30 entries.
-Applying this procedure to the initial training set produced 8 training subsets containing 49, 99, 199, 398, 796, 1593, 3186, and 6373 entries.
-
-All tree models are fine-tuned for NER on each training subset.
-The performance metrics are measured against the test set after each training.
-
-% Move to results
-%All metrics are evaluated on the test set, yet the biased metrics on the development set are added in the article material as additional information.
-
-\subsection{Experiment 2: Impact of OCR nosie on named entity recognition}
-\label{subsection:experiment-2-setup}
-Noise introduced by OCR is known to impede the named entity recognition process (see section~\ref{subsection:stoa-ner-on-historical-texts}).
-%The second experiment assesses the impact of the OCR on the quality of the NER results.
-The second experiment addresses two questions (1) which model is the most robust to OCR noise and (2) what is the most appropriate strategy to fine-tune the NER layer if the input texts to label are noisy ?
-To this end, we leverage the labeled OCR datasets produced following to the method explained in Section 5, in addition to the gold reference.
-
-First and foremost, to make sure that the datasets contain the same entries a preprocessing step must be performed.
-Indeed, entries where the OCR produced an empty string and those for which no entity could be projected from the gold reference text are ignored.
-To overcome this problem we filter each dataset by keeping only the entries common to all.
-At the cost of losing  5\% of the gold references entries, we thus have four new annotated datasets containing the exact same entries.
-They are denoted hereafter {reference, pero, tesseract}-gold.
-Finally each is is split into a training, a dev and a test subset following the procedure presented in\cref{subsection:experiment-1-setup}.
-
-We fine-tune CmBERT and CmBERT+ptrn on reference-gold and pero-gold to create two versions of each model: the former trained on manually corrected entries and the latter on OCR entries.
-SpaCy NER is left aside as results from experiment 1 show that it is outperformed by BERT models.
-Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold and tesseract-gold.
-
-\subsection{Implementation details}
-
-\subsubsection{OCR system parameters}
-\nathalie{Y a-t-il des précisions à ajouter sur la façon dont ont été utilisés les OCR? Ou bien ce sont les versions "sur étagère" qui ont servi?}
-\joseph{Totalement sur l'étagère. On a normalisé un peu les données après, c'est tout.}
-\begin{itemize}
-    \item Tesseract 4.1.1 (newest version: 5, released Nov. 2021)
-    \item Pero OCR version from git repo, master branch updated on Sep 15, 2021 % 8b20f29    
-    \item other?  % kraken if times allows
-\end{itemize}
-
-Pero OCR uses ParseNet \joseph{TODO ref} as internal layout parser for line detection.
-Pero OCR uses an LSTM engine \joseph{TODO check + ref} for text line recognition.
-Pero OCR generally works very well, as long as the bounding boxes of the regions to recognize are not too tightly adjusted.
-Trained on newspapers in European languages by its development team.
-Can output results in Latin, Greek and Cyrillic scripts, as well as some commonly-used typographic symbols.
-
-% PERO refs
-% O Kodym, M Hradiš: Page Layout Analysis System for Unconstrained Historic Documents. ICDAR, 2021.
-% M Kišš, K Beneš, M Hradiš: AT-ST: Self-Training Adaptation Strategy for OCR in Domains with Limited Transcriptions. ICDAR, 2021.
-% J Kohút, M Hradiš: TS-Net: OCR Trained to Switch Between Text Transcription Styles. ICDAR, 2021.
-
-\subsubsection{SpaCy}
-We use the French pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the deep-sequoia and wikiner-fr corpora.
-The default NER labels known to the pipeline are PER, ORG, MISC and LOC.
-In experiment 1 the base model is fine-tuned on every of the 8 training sets using the same training parameters.
-Early stopping is activated with a patience value of 1600 training steps without improvement of the f1 score.
-
-% \bertrand{Move to results}
-%For each size of the training dataset, 5 runs are performed, always with the same parameters, and the mean value of each performance score is computed.
-
-%library and leverage the library utilities to fine-tune it on various training and validation datasets derived from our ground truth dataset (see Section 5.2 for details).
-The base model is fine-tuned in turn on each of the 8 generated training datasets, using the same training parameters.
-Early stopping is activated with patience parameters set at 1600.
-For each size of the training dataset, 5 runs are performed, always with the same parameters, and the mean value of each performance score is computed.
-
-%and Huggingface for the BERT 
-%tok2vec(words embeddings + encoding) + attention layer  +  transition-based %model.
+\subsection{Systems/Variants under test}
 
 \subsubsection{Huggingface CamemBERT}
+\joseph{Copie de \cref{sub:ner-xp1-sysytems-huggingcam}. Je suggère de rédiger la section de l'XP 1 en premier et apporter précision nécessaires ici.}
+
 Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
@@
 
-Details can be found for all models in the source code publicly available on the GitHub repository \bertrand{LIEN GITHUB}.
+\subsection{Protocol}
 
-%$BERT_{reference}$ The base model is fine-tuned in turn on each of the training datasets generated from the groundtruth corpus.
-Training parameters are set depending on the considered experiment: early stopping is activated with the patience parameter set at 3 for experiment 1 and at 5 for experiment 2.
+We fine-tune CmBERT and CmBERT+ptrn on reference-gold and pero-gold to create two versions of each model: the former trained on manually corrected entries and the latter on OCR entries.
+SpaCy NER is left aside as results from experiment 1 show that it is outperformed by BERT models.
+Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold and tesseract-gold.
 
-%$BERT_{pero-ocr}$ This model is used for experiment 2 only.
-The base model is fine-tuned on the training dataset generated by projecting the named entity annotations on the text extracted by PERO OCR, using the same parameters as for the $BERT_{reference}$ model.
+\subsection{Results and discussion}
 
-%$BERT_{ptrn-reference}$ The base model is pretrained using 845000 entries extracted by PERO OCR with the Next  Sentence  Prediction  (NSP) and  Masked  Language  Model(MLM) tasks.
-Then it is fine-tuned with the same parameters and on the same training datasets generated from the groundtruth corpus as the $BERT_{reference}$ model.
+CmBERT and CmBERT+ptrn are fine-tuned on the training sets from reference-gold (manually corrected entries) and pero-gold to produce 4 different models.
+Their F1 score measured against the test tests from reference-gold, pero-gold and tesseract-gold are given in \cref{fig:exp_2_eval_ner}.
+The results clearly show that models perform best when both the pre-training and the NER fine-tuning share the same characteristics (here, OCR noise) as the texts to be processed.
 
-%$BERT_{ptrn-pero-ocr}$ This model is used for experiment 2 only.
-The base model is pretrained like the $BERT_{ptrn-reference}$ model and it is fine-tuned with the same parameters and on the same training datasets as the $BERT_{pero-ocr}$ model.
+In our tests, pre-training the model brings a slight gain in performance ($\approx 0.5\%$).
+We did not pre-train or fine-tune with texts extracted with Tesseract.
+However, despite a loss of performance the model pre-trained and fine-tuned with pero-gold still gives the best results.
+This is probably due to the fact that the texts produced by Pero-OCR feature characteristics intermediary between human transcriptions and Tesseract.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figs/eval-ner-exp2.pdf}
+    \caption{F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing examples, either manually corrected or raw from Pero-OCR and Tesseract.
+The type of examples used to train the NER task is noted in indice after the model name (e.g.
+Bert$_{reference}$).
+Results show that best performances on OCRed entries are obtained when the BERT model has been pretrained and fine-tuned for NER on examples affected with similar OCR errors.}
+    \label{fig:exp_2_eval_ner}
+\end{figure}

### 2022-01-20 src-latex/parts/60-results-BACKUP.tex
New paper structure with 1 section for each experiment

--- 2022-01-19 src-latex/parts/60-results.tex
@@
 \midrule
-Pero      & 3.78\% & 3.76\% \\   
+Pero-OCR  & 3.78\% & 3.76\% \\   
 Tesseract & 6.56\% & 6.45\% \\
@@
 }
-\caption{Character error rates at entry-level for Pero OCR and Tesseract. (a) Global CER and distribution of the CER per entry. (b):
-joint plot of the dataset showing that the two systems do not fail on the same entries.}  
+\caption{Character error rates at entry-level for Pero-OCR, Kraken and Tesseract. (a) Global CER and distribution of the CER per entry. (b)
+joint plot of the per-entry error rate showing that Pero-OCR and Tesseract do not fail on the same entries.}  
+\label{fig.ocr-results}
 \end{figure}
+
+\Cref{fig.ocr-results} compares the performance of the OCR systems on our dataset. We can see Kraken's performance are
+not as good as the two first OCR. This is partially due to the fact that the closest available model is for English text
+and so it misses French specific symbols. On the over hand, even when using a French model trained on French 19th
+documents, performance does not increase (and relaxing the character match rules does not help either). Tesseract and
+Pero-OCR are performing better on this dataset ``out-of-the-box''. With no fine-tuning, Pero-OCR gets the best accuracy
+with less than 4\% character errors. Many of them are even due to a bad line detection in case of multi-lines entries
+and are not related to the OCR system itself. \Cref{fig.ocr-results} (b) shows that errors from the two best OCR are not
+committed on the same entries (if so, all points would be on the diagonal line) and that combining the outputs of
+Pero-OCR and Tesseract would improve the overall recognition quality. Also, we will not consider Kraken in the following
+NER experiments because of its recognition rate is already low and cannot be artificially degraded in a controlled setup.
 

### 2022-01-20 src-latex/main-supplementary-material.tex
Move content from related work to supp. material

@@
 \renewcommand{\arraystretch}{1.4}
+
+% Comment 
+% FIXME : delete before submission :)
+\newcommand{\bertrand}[1]{\textcolor{violet}{Bertrand:~#1}}
+\newcommand{\nathalie}[1]{\textcolor{purple}{Nathalie:~#1}}
+\newcommand{\edwin}[1]{\textcolor{pink}{Edwin:~#1}}
+\newcommand{\joseph}[1]{\textcolor{red}{Joseph:~#1}}
 
@@
 
+\subsection{Entry segmentation}
+
+\subsection{Line detection}
+Pero OCR uses ParseNet as internal layout parser for line detection.
+Pero OCR uses an LSTM engine for text line recognition.
+Pero OCR generally works very well, as long as the bounding boxes of the regions to recognize are not too tightly adjusted.
+Trained on newspapers in European languages by its development team.
+Can output results in Latin, Greek and Cyrillic scripts, as well as some commonly-used typographic symbols.
+
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-01-20 src-latex/parts/20-related-work.tex
Move content from related work to supp. material

@@
 \label{sec:related-work}
+
+\joseph{Suggestion pour réduire la présentation de l'état de l'art : l'orienter exclusivement vers une justification des systèmes que nous avons retenus pour cet article.}
 
@@
 
+\joseph{Move to implementation details for XP OCR or to pipeline summary}
 \begin{itemize}
@@
 \end{itemize}
-Pero OCR uses ParseNet \joseph{TODO ref} as internal layout parser for line detection.
-Pero OCR uses an LSTM engine \joseph{TODO check + ref} for text line recognition.
-Pero OCR generally works very well, as long as the bounding boxes of the regions to recognize are not too tightly adjusted.
-Trained on newspapers in European languages by its development team.
-Can output results in Latin, Greek and Cyrillic scripts, as well as some commonly-used typographic symbols.
 

### 2022-01-20 src-latex/parts/10-introduction.tex
Fix contrib wording
Fix refs in intro/contrib

@@
     \item We introduce a new dataset suitable for OCR and NER evaluation (\cref{sec:dataset}).
-    \item We measure the performance of three modern OCR systems on real data (\cref{sec:TODO}).
-    \item We introduce modern NER approaches and study their requirements in terms of training data, as well as the effects of pre-training (\cref{sec:TODO}).
-    \item We show that Transformed-based NER can benefit from pre-training and fine-tuning to improve its performance on noisy OCR (\cref{sec:TODO}).
+    \item We measure the performance of three modern OCR systems on real data (\cref{sec:ocr-xp}).
+    \item We evaluate modern NER approaches: their requirements in terms of training data, and the effects of pre-training (\cref{sec:ner-xp1}).
+    \item We show that Transformed-based NER can benefit from pre-training and fine-tuning to improve its performance on noisy OCR (\cref{sec:ner-xp2}).
 \end{enumerate*}

### 2022-01-20 src-latex/parts/20-related-work.tex
Updates from Overleaf

@@
 Many approaches have been designed to recognize named entities, ranging from handcrafted rules to supervised approaches \cite{nadeau2007}.
-Rule based approaches look for portions of text that match patterns based on word characteristics such as their case, their part-of-speech tags, etc.
-These patterns can be fully handcrafted like in \cite{bell2020automated} or generated by data mining approaches like in \cite{nouvel2011}.
-Rule based approaches can also combine a set of patterns with a set of dictionaries (gazetteers, author lists, etc.) that help recognizing named entities when an exhaustive lexicon of searched entities is available \cite{mansouri2008,maurel2011}.
-%For example, the CasEN\footnote{\url{https://tln.lifat.univ-tours.fr/version-francaise/ressources/casen}} transducer cascade leverages syntactic lexical patterns and dictionaries to recognize Named Entities in French texts.
-Such kind of approaches achieve very good results when applied to specialized domain corpus - like directories - and when an exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
+Rule based approaches look for portions of the text that match patterns like in \cite{bell2020automated,nouvel2011} or dictionary (gazetteers, author lists, etc.) entries like in \cite{mansouri2008,maurel2011}.
+Such kind of approaches achieve very good results when applied to a specialized domain corpus and when an exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
 
 Supervised approaches include both approaches implementing supervised learning algorithms with careful text feature engineering, and deep learning based approaches which automatically build their own features to classify tokens into named entity categories.
-In recent years, the latter have grown dramatically, yielding state-of-the-art performances and establishing new baselines\cite{li2020}.
-A recent survey proposed by \cite{li2020} presents deep learning techniques for NER according to the distributed representation(s) they use as input, their context encoder and their tag decoder.
-It shows that language model embeddings pretrained using \textit{Transformer} \cite{vaswani2017attention}, like BERT \cite{devlin2018bert}, can not only be used to replace traditional embeddings as input distributed representations but also be fine-tuned for the NER task with one additional output layer while achieving state-of-the-art performances.
-The survey concludes that fine-tuning general-purpose contextualized language models with domain-specific data is very likely to give good performance for use cases with domain-specific texts and few training data.
+In recent years, the latter have grown dramatically, yielding state-of-the-art performances as shown in the recent survey proposed by \cite{li2020}.
+This survey concludes that fine-tuning general-purpose contextualized language models with domain-specific data is very likely to give good performance for use cases with domain-specific texts and few training data.
+This strategy has been adopted by \cite{Labusch2020NamedED} to extract named entities in OCRed historical texts in German, French, and English.
+However, the NER performance decreased dramatically with OCRed texts is noted, especially for the English texts for which the OCR is of poorer quality.
 
-\subsection{Named Entity Recognition on historical texts}
+\subsection{Improving Named Entity Recognition in historical texts}
 \label{subsection:stoa-ner-on-historical-texts}
-\cite{Labusch2020NamedED} pretrained an off-the-shelf multilingual BERT-based NER model on more than 2 000 000 pages of OCRed historical texts in German, French, and English.
-They fine-tuned it on historical texts NER ground truth, also in these three languages and tested it on OCRed historical texts in each language.
-Eventually, a clear decrease in NER performance on OCRed texts is noted, especially for the English texts for which the OCR is of poorer quality.
 
@@
 \subsubsection{Transformers / Bert / Hugging face}
-Language model based on Transformers, like BERT, have become a new paradigm for NER\cite{li2020}. 
+Language model based on \textit{Transformers} \cite{vaswani2017attention}, like BERT \cite{devlin2018bert}, have become a new paradigm for NER\cite{li2020}. 
 The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer.

### 2022-01-20 src-latex/parts/40-ocr-bench.tex
Fix refs in intro/contrib
Update OCR text/eval top keep one CER metric

@@
 \section{OCR benchmark}
+\label{sec:ocr-xp}
 
@@
 \textbf{OCR Q.A.} is performed after some text normalization of the OCR system outputs. Text normalization consists in
-projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes sense) and removing extra symbols (hands,
-crosses). Then, the predicted text is aligned with the reference text using standard tools from Stephen V. Rice's
-thesis~\cite{santos.2019.wcmel,neudecker.2021.whdip} and the Character Error Rates (CER) are computed at the entry level
-and at the global level.
-
-\begin{align}
-\mathrm{CER} &=  \frac{\#Errors}{\text{Reference Text Length}} & & \mathrm{CER}_\mathrm{norm} =  \frac{\#Errors}{\text{Alignment Length}} 
-\end{align}
+projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes sense) and removing extra symbols
+(hands, crosses). Then, the predicted text is aligned with the reference text using standard tools from Stephen V.
+Rice's thesis~\cite{santos.2019.wcmel,neudecker.2021.whdip}. The Character Error Rate (CER) is computed at the entry
+level and at the global level defined as the ratio between the number of errors (insertions/deletions/substitutions)
+over the reference text length.
 
@@
 \subcaptionbox{}[.5\linewidth]{
-\begin{tabular}{rll}
+\begin{tabular}{rlll}
 \toprule
-            & CER & CER$_{norm.}$ \\
+ & Pero-OCR  & Tesseract & Kraken \\    
 \midrule
-Pero-OCR  & 3.78\% & 3.76\% \\   
-Tesseract & 6.56\% & 6.45\% \\
-Kraken    & 15.72\% & 15.46\% \\  
+CER & 3.78\% & 6.56\% & 15.72\% \\  
 \bottomrule
 \end{tabular}
-
 \bigskip
@@
 \includegraphics[width=\linewidth]{images/ocr-eval-2.pdf}
+\medskip
 }

### 2022-01-20 src-latex/parts/20-related-work.tex

@@
 
-\subsection{Named Entity Recognition}
+\subsection{Named Entity Recognition in historical texts}
+\label{subsection:stoa-ner-on-historical-texts}
 
@@
 This strategy has been adopted by \cite{Labusch2020NamedED} to extract named entities in OCRed historical texts in German, French, and English.
-However, the NER performance decreased dramatically with OCRed texts is noted, especially for the English texts for which the OCR is of poorer quality.
-
-\subsection{Improving Named Entity Recognition in historical texts}
-\label{subsection:stoa-ner-on-historical-texts}
-
+However, the NER performance decreased dramatically for OCRed texts compared to the groundtruth dataset.
+This decrease went hand in hand with the drop of the OCR quality.
 Several recent studies have focused on the extent to which the quality of the OCR affects the results produced by a NER model.

### 2022-01-20 src-latex/parts/20-related-work.tex
Changed title of 3.1

@@
 \subsection{== À réintégrer depuis ancienne section 4 ==}
+
+A introduire en disant quo'n chercher à utiliser au plus des modèles simples à reprendre: déjà entraînés, dispos pour du français et adaptables
 We select two deep-learning-based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.

### 2022-01-20 src-latex/parts/30-dataset.tex
Changed title of 3.1
Move evaluation-related content to dataset part

@@
 \label{sec:dataset}
+
+% Conventions nommage
+% Dataset annoté à la main = "ground truth"
+% ground truth aligné = "ner-reference"
+% entrées pero aligné = "ner-pero"
+% entrées tesseract alignées = "ner-tesseract"
 
@@
 
-\subsection{The  XIX\textsuperscript{th} century Parisian trade directories}
+\subsection{A selection of Paris trade directories from 1798 to 1854}
 
@@
 \end{itemize*}
+
+\subsection{Metrics proposed}
+
+\Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems. Two quality evaluations
+are performed in the pipeline, respectively named \emph{OCR Q.A.} and \emph{NER Q.A.}.
+
+First, the layout extraction and entry segmentation of the page are performed with a semi-automated system and
+checked by a human. Afterward, an OCR system runs on the thumbnails of each segmented entry to extract its text. An
+entry might span over multiple text lines but is always a single block. Thus, the most adapted mode is chosen when
+the OCR system allows for the detection mode (e.g. the \emph{block} mode for \emph{Tesseract}).
+
+\textbf{OCR Q.A.} is performed after some text normalization of the OCR system outputs. Text normalization consists in
+projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes sense) and removing extra symbols (hands,
+crosses). Then, the predicted text is aligned with the reference text using standard tools from Stephen V. Rice's
+thesis~\cite{santos.2019.wcmel,neudecker.2021.whdip} and the Character Error Rates (CER) are computed at the entry level
+and at the global level.
+
+\begin{align}
+\mathrm{CER} &=  \frac{\#Errors}{\text{Reference Text Length}} & & \mathrm{CER}_\mathrm{norm} =  \frac{\#Errors}{\text{Alignment Length}} 
+\end{align}
+
+In this benchmark, we consider 3 OCR systems well known for historical document analysis: Pero OCR, Tesseract, and Kraken.
+
+Next, a NER system extracts the named entities from the text of each entry output by the OCR and the \textbf{NER Q.A.}
+is performed. The NER system outputs a text with tags that enclose the entities. To assess the quality of the entity
+extraction, we rely on a technique similar as for the OCR evaluation. The predicted text is aligned with the reference
+text and the tags are projected in the alignment. Then, the precision, recall, and f-measure (or f-score) are computed considering
+only the exact matches between entities. Precision and recall are defined by equation (2); the f-measure is their harmonic mean.
+
+\begin{align}
+    \mathrm{precision} &= \frac{\text{\#exact matches}}{\text{\#entries in prediction}} & \mathrm{recall} &= \frac{\text{\#exact matches}}{\text{\#entries in reference}}
+\end{align}
+
+The whole process is illustrated on~\cref{fig.eval-ocr-ner}. The OCR text and the GOLD text are first aligned to
+evaluate the OCR accuracy. As there are 11 mismatches over 56 aligned characters, the CER is thus 24\%. This alignment
+is then used to back-project the GOLD tags to build the tagged NER reference string. Finally, the NER system runs on the
+OCR text; its output is compared to the NER reference string. There is only 2 over 3 tags matching in the prediction (precision),
+while only 2 over 4 entities are matched in the reference (recall). It follows an overall f-score of 0.4.
+
+\begin{figure}[tb]
+    \includegraphics[width=\linewidth]{figs/eval-ocr-ner.pdf}
+    \caption{OCR and NER evaluation protocol example.}
+    \label{fig.eval-ocr-ner}
+\end{figure}
+
+\joseph{TODO expliquer qu'il faut faire la fusion des sous-datasets NER}
+
+First and foremost, to make sure that the datasets contain the same entries a preprocessing step must be performed.
+Indeed, entries where the OCR produced an empty string and those for which no entity could be projected from the gold reference text are ignored.
+To overcome this problem we filter each dataset by keeping only the entries common to all.
+At the cost of losing  5\% of the gold references entries, we thus have four new annotated datasets containing the exact same entries.

### 2022-01-20 src-latex/parts/40-ocr-bench.tex
Move evaluation-related content to dataset part

@@
 
-\subsection{Dataset}
-input: image crops for each entry
-expected output: human transcription
+\subsection{Protocol}
 
-8765 entries used, \joseph{FIXME number of chars}
+Dataset:
+- input: image crops for each entry
+- expected output: human transcription
+- 8765 entries used, \joseph{FIXME number of chars}
 
-\subsection{Metrics}
-\textbf{OCR Q.A.} is performed after some text normalization of the OCR system outputs. Text normalization consists in
-projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes sense) and removing extra symbols
-(hands, crosses). Then, the predicted text is aligned with the reference text using standard tools from Stephen V.
-Rice's thesis~\cite{santos.2019.wcmel,neudecker.2021.whdip}. The Character Error Rate (CER) is computed at the entry
-level and at the global level defined as the ratio between the number of errors (insertions/deletions/substitutions)
-over the reference text length.
-
-\subsection{Systems/Variants under test}
+Systems
 In this benchmark, we consider 3 OCR systems well known for historical document analysis: Pero OCR, Tesseract, and Kraken.
-
-\subsection{Protocol}
 

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Move evaluation-related content to dataset part

@@
 \subsection{Metrics}
-Next, a NER system extracts the named entities from the text of each entry output by the OCR and the \textbf{NER Q.A.}
-is performed. The NER system outputs a text with tags that enclose the entities. To assess the quality of the entity
-extraction, we rely on a technique similar as for the OCR evaluation. The predicted text is aligned with the reference
-text and the tags are projected in the alignment. Then, the precision, recall, and f-measure (or f-score) are computed considering
-only the exact matches between entities. Precision and recall are defined by equation (2); the f-measure is their harmonic mean.
-
-\begin{align}
-    \mathrm{precision} &= \frac{\text{\#exact matches}}{\text{\#entries in prediction}} & \mathrm{recall} &= \frac{\text{\#exact matches}}{\text{\#entries in reference}}
-\end{align}
 

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Update xp2 figure with bar value.
Move evaluation-related content to dataset part

@@
 
-\subsection{Dataset}
+\subsection{Protocol}
+
+Dataset
 
@@
 
-First and foremost, to make sure that the datasets contain the same entries a preprocessing step must be performed.
-Indeed, entries where the OCR produced an empty string and those for which no entity could be projected from the gold reference text are ignored.
-To overcome this problem we filter each dataset by keeping only the entries common to all.
-At the cost of losing  5\% of the gold references entries, we thus have four new annotated datasets containing the exact same entries.
 They are denoted hereafter {reference, pero, tesseract}-gold.
@@
 
-\subsection{Metrics}
-prec/recall/F1 as in \ref{sec:ner-xp1}
-
-The whole process is illustrated on~\cref{fig.eval-ocr-ner}. The OCR text and the GOLD text are first aligned to
-evaluate the OCR accuracy. As there are 11 mismatches over 56 aligned characters, the CER is thus 24\%. This alignment
-is then used to back-project the GOLD tags to build the tagged NER reference string. Finally, the NER system runs on the
-OCR text; its output is compared to the NER reference string. There is only 2 over 3 tags matching in the prediction (precision),
-while only 2 over 4 entities are matched in the reference (recall). It follows an overall f-score of 0.4.
-
-\begin{figure}[tb]
-    \includegraphics[width=\linewidth]{figs/eval-ocr-ner.pdf}
-    \caption{OCR and NER evaluation protocol example.}
-    \label{fig.eval-ocr-ner}
-\end{figure}
-
-\subsection{Systems/Variants under test}
+\joseph{préciser qu'on ne prend que les données qui sont valides pour pero et tesseract (soit 8341 entrées).
+On ne considère pas kraken car pas le temps, on perd 500 entrées dans l'intersection, la qualité est pourrie}
 
@@
     \centering
-    \includegraphics[width=\textwidth]{figs/eval-ner-exp2.pdf}
+    \includegraphics[width=0.75\textwidth]{images/experiment_2_f1_with_noise_graph.pdf}
     \caption{F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing examples, either manually corrected or raw from Pero-OCR and Tesseract.

### 2022-01-20 src-latex/parts/20-related-work.tex

@@
 This decrease went hand in hand with the drop of the OCR quality.
-Several recent studies have focused on the extent to which the quality of the OCR affects the results produced by a NER model.
-\cite{van2020assessing} assess the impact of OCR on several NLP downstream tasks, including NER.
-They worked on a corpus published by a post-OCR correction software company, made of many journal articles with different levels of OCR errors and their respective ground truths.
-For each OCRed article, the Word Error Rate (WER) is computed and the English model \textit{en-core-web-lg} provided by Spacy\footnote{\url{https://spacy.io/}} library is used to perform NER on \textit{Person}, \textit{GPE}\footnote{Geopolitical Entity} and \textit{Date}.
-The performance of the NER model with respect to OCR quality is eventually assessed by computing the F-measure for each NER class, and each article i.e., each WER value.
+
+Several recent studies have focused on the impact of OCR quality on NER results.
+\cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by Spacy\footnote{\url{https://spacy.io/}} library to perform NER on \textit{Person}, \textit{GPE}\footnote{Geopolitical Entity} and \textit{Date} on a corpus made of many journal articles with different levels of OCR errors.
+For each OCRed article, a ground truth text is available so that the Word Error Rate (WER) can be computed.
+The performance of the NER model with respect to OCR quality is eventually assessed by computing the F1-score for each NER class and each article, i.e., each WER value.
 \cite{hamdi2020assessing} performed a similar but more extensive evaluation on four different NER models: CoreNLP using Conditional Random Fields and three deep neural models, BLSTM-CNN, BLSTM-CRF, and BLSTM-CNN-CRF.

### 2022-01-20 src-latex/parts/30-dataset.tex
Updates from Overleaf
Partially reapply lost commit 1ea831260218d31bd46713eeba862c80782e024a

@@
 The directories to be processed are stored in different libraries in Paris\footnote{\textcolor{green}{Citer les bibliothèques sources?} \joseph{est-ce nécessaire ici ?
-Est-ce que ça peut être dans une annexe ?}}.
+Est-ce que ça peut être dans une annexe ?}\nathalie{Oui totalement...}}.
 They have been scanned independently and with various levels of quality.

### 2022-01-20 src-latex/parts/30-dataset.tex
Proposition plan section dataset
Move some ref to supp. material

--- 2022-01-20 src-latex/parts/30-dataset.tex
@@
 
-\subsection{The  XIX\textsuperscript{th} century Parisian trade directories}
+\subsection{A selection of Paris trade directories from 1798 to 1854}
 
 The directories to be processed are stored in different libraries in Paris\footnote{\textcolor{green}{Citer les bibliothèques sources?} \joseph{est-ce nécessaire ici ?
-Est-ce que ça peut être dans une annexe ?}}.
+Est-ce que ça peut être dans une annexe ?}\nathalie{Oui totalement...}}.
 They have been scanned independently and with various levels of quality.
@@
 For training as testing purposes, our groundtruth dataset must be as representative as possible of the diversity of the available directories.
-The Table \ref{tab:directories} lists the directories selected, their layout, the structure of their entries, and the number of annotated entries according to the considered index.
 
@@
 
-\subsection{Groundtruth data extraction}
-\joseph{TODO: separate entry detection, OCR and NER stages explicitely}
+\subsection{A dataset for OCR and NER evaluation}
 
-First, the layout extraction and entry segmentation of the page are performed with a semi-automated system and
-checked by a human. Afterward, an OCR system runs on the thumbnails of each segmented entry to extract its text. An
-entry might span over multiple text lines but is always a single block. Thus, the most adapted mode is chosen when
-the OCR system allows for the detection mode (e.g. the \emph{block} mode for \emph{Tesseract}).
+\begin{figure}[tb]
+    \includegraphics[width=\linewidth]{figs/protocol.pdf}
+    \caption{Scheme of the data extraction pipeline with two quality control checkpoints. \joseph{Missing "Ground truth" for NER stage?}}
+    \label{fig.pipeline}
+\end{figure}
 
-OCR: some basic charset normalization is performed after human annotation:
-- remove annotation codes for glyphs which have no character in unicode
-- alleviate some ambiguities in annotation rules (n° -> no, --- -> -, homogenize star and hand symbols allowed)
+%\joseph{TODO: separate entry detection, OCR and NER stages explicitely}
 
-The annotated named entities are as follows:
+The simplified data extraction pipeline depicted in~\cref{fig.pipeline} processes the documents presented
+in~\cref{sec:dataset}. First, the layout extraction and entry segmentation of the page are performed with a
+semi-automated system and checked by a human. The resulting images, representing 8765 entries, are the inputs of a customizable two-stage pipeline.
 
-\begin{itemize}
-    \item PER: a person name or a business name, usually referred to as several person names.
-Firts names, initials, or civility mentions are included.
+%1. Entrée/Sortie de l'OCR
+\textbf{OCR stage.} The first stage of the pipeline aims at extracting the raw text from the images. An OCR system runs
+on the thumbnails of each segmented entry to extract its text. An entry might span over multiple text lines but is
+always a single block. Thus, the most adapted mode is chosen when the OCR system allows for the detection mode (e.g. the
+\emph{block} mode for \emph{Tesseract}). The glyphs used in this dataset might be unknown to the OCR. Some of them, like
+\scalerel*{\includegraphics{./images/LH.png}}{H}, do not even have a unicode codepoint. Text normalization is thus required
+on the OCR outputs and consists in projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes
+sense), homogenize extra symbols (star and hand symbols) and alleviate some ambiguities (e.g. \emph{no} and \emph{n°}).
+
+At the end of this first stage, an OCR Quality Assessment (Q.A.) between the \emph{normalized} groundtruth text and the
+OCR output is performed.
+
+%\joseph{[DONE] a word about the charset, and existence of glyphs unknown to the OCR}
+%\joseph{[DONE] parler de l'étage d'extraction d'entrées qui est validé manuellement, et des quelques règles de simplification du charset pour homogénéiser entre les annotateurs / pallier des ambiguïtés dans les règles d'annotation (qu'on pourrait encore améliorer)}
+
+%2. Entrée/Sortie
+\textbf{NER Stage.} The second stage of the pipeline aims an extracting the named entities from a text with a NER
+system. This text originates from the OCR outputs in a real-word scenario but might be groundtruth text in order to
+evaluate the NER performance standalone. There are 5 type of entity to detect~(see.~\cref{tab.entities}).
+The NER system has to classify non-overlapping parts of the text into one of these entity (or none of them). A second
+Q.A. (namely NER Q.A.) is performed at the end of this stage between the groundtruth and the NER output.
+
+Note that the dataset used to assess the NER stage is subset of the entries. Indeed, we need to ensure that the datasets
+contain the same entries whichever the OCR used in the previous stage. Entries where the OCR produced an empty string
+and those for which no entity could be projected from the groundtruth have to be ignored. We filter the set of entries by
+keeping only the entries that are always valid at the end of the stage 1. At the cost of losing  5\% of the references
+entries, we thus have 8341 valid entries at the input of the stage 2.
+
+\begin{table}
+\small
+\begin{tabular}{llp{8cm}}
+Entity & Count & Description\\    
+PER & 8788 & a person name or a business name, usually referred to as several person names.
+First names, initials, or civility mentions are included.
 E.g.: "Alibert (Prosper)", "Allamand frères et Hersent", "Heurtemotte (Vve)", etc.
-    \item TITLE: an honorary title, either text, glyphs or a combination of the two.
+\\
+TITLE & 483 & an honorary title, either text, glyphs or a combination of the two.
 E.g: "O.
-\scalerel*{\includegraphics{./images/LH.png}}{H}" for "Officier de la Légion d'Honneur" or "\scalerel*{\includegraphics{./images/GMLondon.png}}{H}" to represent the Great Medal at the London exhibition.
-    \item ACT: the profession or social status of a person summarized in the single concept of activity. E.g.: "horlogerie", "export en tous genres", "Conseiller d'Etat", "propriétaire".
-    \item LOC: mostly street names ("r. de la Jussienne"), but may also be neighborhoods ("Marais du Temple") or any indirect spatial references ("au Palais Royal").
-    \item CARDINAL: a street number, as part of an address ("16", "5 bis"), or a range of numbers (e.g. "23-25" or "5 à 9").
-    \item FT: a geographic entity type, used to give more details on a location, e.g. "boutique", "atelier", "fab." or "dépôt".
-\end{itemize}
+\scalerel*{\includegraphics{./images/LH.png}} {H}" for "Officier de la Légion d'Honneur" or "\scalerel*{\includegraphics{./images/GMLondon.png}}{H}" to represent the Great Medal at the London exhibition.\\
+ACT & 6472 & the profession or social status of a person summarized in the single concept of activity. E.g.: "horlogerie", "export en tous genres", "Conseiller d'Etat", "propriétaire".\\
+LOC & 9709 & mostly street names ("r. de la Jussienne"), but may also be neighborhoods ("Marais du Temple") or any indirect spatial references ("au Palais Royal").\\
+CARDINAL & 8747 & a street number, as part of an address ("16", "5 bis"), or a range of numbers (e.g. "23-25" or "5 à 9").\\
+FT & 43 & a geographic entity type, used to give more details on a location, e.g. "boutique", "atelier", "fab." or "dépôt". \\
+\end{tabular}
+\caption{Entities to recognize in the dataset.}
+\label{tab.entities}
+\end{table}
 
-Thus, the extracted corpus contains both the bounding boxes of directory entries, their exact text and the named entities they contain.
-It can therefore be used to train and improve the entry detection, OCR and NER tools.
+% It can therefore be used to train and improve the entry detection, OCR and NER tools.
 % book, page, box, text_ocr, ner_xml (+uid within a page)
 
-\subsection{Summary}
-\joseph{TODO intégrer dans les sections avant ?}
+%\joseph{TODO intégrer dans les sections avant ?}
+%
+%
+% valid ner (ref \& pero \& tess): 8341
+%
+%TODO average entry len (in char)
+%TODO average number of tags/entities per entry
+%TODO lists actual dataset content: set of entries. For each entry:
+%\begin{itemize*}
+%    \item ref to original book, page, region
+%    \item human transcription (OCR GT)
+%    \item human NER (NER GT)
+%    \item [GENERATED] OCR produced by Pero OCR, Tesseract v4, Kraken (precise model used)
+%    \item [GENERATED] NER GT aligned on text predictions from Pero OCR, Tesseract v4, Kraken
+%\end{itemize*}
 
-\begin{verbatim}
-[('<LOC>', 9709),
-('<PER>', 8788),
-('<CARDINAL>', 8747),
-('<ACT>', 6472),
-('<TITRE>', 483),
-('<FT>', 43)]
-\end{verbatim}
+%\joseph{TODO expliquer qu'il faut faire la fusion des sous-datasets NER}
 
- valid ocr: 8765 entries
+%\subsection{Focus on how NER GT was aligned on noisy OCR}
 
- valid ner (ref \& pero \& tess): 8341
+\subsection{Metrics for OCR and NER Quality Assessment}
 
-\joseph{parler de l'étage d'extraction d'entrées qui est validé manuellement, et des quelques règles de simplification du charset pour homogénéiser entre les annotateurs / pallier des ambiguïtés dans les règles d'annotation (qu'on pourrait encore améliorer)}
+\begin{figure}[tb]
+    \includegraphics[width=\linewidth]{figs/eval-ocr-ner.pdf}
+    \caption{OCR and NER evaluation protocol example.}
+    \label{fig.eval-ocr-ner}
+\end{figure}
 
-TODO a word about the charset, and existence of glyphs unknown to the OCR
+\textbf{OCR Q.A.}  The predicted text by the OCR system is aligned with the groundtruth text using standard tools from
+Stephen V. Rice's thesis~\cite{santos.2019.wcmel,neudecker.2021.whdip}. The Character Error Rate (CER) is computed at
+the entry level and at the global level defined as the ratio between the number of errors
+(insertions/deletions/substitutions) over the reference text length.
 
-TODO average entry len (in char)
+%In this benchmark, we consider 3 OCR systems well known for historical document analysis: Pero OCR, Tesseract, and Kraken.
 
-TODO average number of tags/entities per entry
+\textbf{NER Q.A.}. The NER system outputs a text with tags that enclose the entities. To assess the quality of the
+entity extraction, we rely on a technique similar as for the OCR evaluation to build the \emph{NER-target}. The
+\emph{NER-target} is different from the groundtruth because it should not involve the errors committed during the
+previous stages. The OCR text is first aligned with the groundtruth text to form the NER-\emph{input} (where
+\emph{input} is a placeholder for \emph{pero} if the input text is from PERO, NER-\emph{tesseract},
+NER-\emph{reference}\ldots). The tags of the groundtruth are then projected in the alignment on NER-\emph{input} to
+provide the \emph{NER-target}. The NER system then runs on NER-\emph{input} and outputs the \emph{NER-prediction}. The
+precision, recall, and f-measure (or f-score) are computed considering only the exact matches between entities of the
+\emph{NER-target} and those from the \emph{NER-prediction}. Precision and recall are defined by \cref{eqn.prec-recall};
+the f-measure is their harmonic mean.
 
-TODO lists actual dataset content: set of entries. For each entry:
-\begin{itemize*}
-    \item ref to original book, page, region
-    \item human transcription (OCR ref)
-    \item human NER (NER ref)
-    \item OCR produced by Pero OCR, Tesseract v4, Kraken (precise model used)
-\end{itemize*}
+\begin{align}
+    \mathrm{precision} &= \frac{\text{\#exact matches}}{\text{\#entries in prediction}} & \mathrm{recall} &= \frac{\text{\#exact matches}}{\text{\#entries in reference}}
+    \label{eqn.prec-recall}
+\end{align}
+
+The evaluation metrics are illustrated on~\cref{fig.eval-ocr-ner}. The OCR text and the groundtruth are first aligned to
+evaluate the OCR accuracy. As there are 11 mismatches over 56 aligned characters, the CER is thus 24\%. This alignment
+is then used to back-project the groundtruth tags to build the tagged NER-\emph{target}. Finally, the NER system runs on the
+OCR text; its output is compared to the NER groundtruth. There is only 2 over 3 tags matching in the prediction (precision),
+while only 2 over 4 entities are matched in the reference (recall). It follows an overall f-score of 0.4.

### 2022-01-20 src-latex/parts/10-introduction.tex
Fix typos in intro

@@
 
-In this article we focus on a peculiar category of historical printed books: city trade directories of Paris from the XIX\textsuperscript{th} century (1801 - 1854).
-They contain hundred-pages long lists of people, along with their activity and their address; a much valuable fine-grained knowledge to study the social dynamics of the city over time.
+In this article, we focus on a peculiar category of historical printed books: city trade directories of Paris from the XIX\textsuperscript{th} century (1801 - 1854).
+They contain hundred-pages long lists of people, along with their activity and their address, and provide valuable fine-grained knowledge to study the social dynamics of the city over time.
 Despite containing roughly the same information, the diversity in layouts, information organization and digitizing quality is challenging both for OCR and NER.
@@
 In a recent work to locate potentially polluted urban soils \cite{bell2020automated}, trade directories were leveraged to identify and locate all the gas stations in the city of Providence during the XX\textsuperscript{th} century.
-The proposed information extraction pipeline extract texts with OCR, identify directories entries and recognizes business type and address within each entry.
-In the same way, we aim at producing structured spatio-temporal data from the Parisian trade directories.
+The proposed information extraction pipeline extracts texts with OCR, identifies directories entries and recognizes business type and address within each entry.
+In the same way, we aim at producing structured spatio-temporal data from Parisian trade directories.
 These directories originate from several publishers, their content is organized following different indexing methods (by name, by activity or by address), printed in various layouts, use different fonts.
-We therefore investigate several state-of-the-art OCR and NER approaches independently and in combination to assess their usability on the corpus.
+We therefore investigate several state-of-the-art OCR and NER approaches to assess their usability on the corpus.
 
-The contributions of this article are the following. \joseph{FIXME recheck le plan à la fin}
+The contributions of this article are the following.
 % Contributions

### 2022-01-20 src-latex/parts/10-introduction.tex
More typo fixes in intro

@@
 Being active research topics, OCR and named entity recognition (NER) are still difficult tasks when applied to historical text documents.
-OCR approaches used for modern documents are likely to struggle even on printed historical documents due multiple causes related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
+OCR approaches used for modern documents are likely to struggle even on printed historical documents due to multiple causes related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
 On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts, and also from the errors introduced into the text by the OCR step.
@@
 In the same way, we aim at producing structured spatio-temporal data from Parisian trade directories.
-These directories originate from several publishers, their content is organized following different indexing methods (by name, by activity or by address), printed in various layouts, use different fonts.
+Because they originate from several publishers, their content is organized following different indexing methods (by name, by activity or by address), is printed in various layouts, and uses different fonts.
 We therefore investigate several state-of-the-art OCR and NER approaches to assess their usability on the corpus.

### 2022-01-20 src-latex/parts/20-related-work.tex
Updates from Overleaf

@@
 Several recent studies have focused on the impact of OCR quality on NER results.
+Most of the time, they have evaluated NER approaches based on deep learning architectures as they seem to adapt more easily to OCR errors than rule-based or more classical supervised approaches.
 \cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by Spacy\footnote{\url{https://spacy.io/}} library to perform NER on \textit{Person}, \textit{GPE}\footnote{Geopolitical Entity} and \textit{Date} on a corpus made of many journal articles with different levels of OCR errors.
@@
 The performance of the NER model with respect to OCR quality is eventually assessed by computing the F1-score for each NER class and each article, i.e., each WER value.
-\cite{hamdi2020assessing} performed a similar but more extensive evaluation on four different NER models: CoreNLP using Conditional Random Fields and three deep neural models, BLSTM-CNN, BLSTM-CRF, and BLSTM-CNN-CRF.
-They tested them on two well-known NER benchmark corpora: CoNLL-02 and CoNLL-03.
-They applied four different types of OCR noise to each corpus, with two levels of degradation and computed the WER and Character Error Rate (CER) for each degraded version.
-Finally, they applied each NER model to the progressively degraded versions of the corpus and computed the resulting F-measure.
-Overall, NER F-measure drops from 90\% to 50\% when the WER increase from 8\% to 50\%. However, models based on deep neural networks seem less sensitive to OCR errors.
+\cite{hamdi2020assessing} performed a similar but more extensive evaluation on four supervised NER models: CoreNLP using Conditional Random Fields and three deep neural models, BLSTM-CNN, BLSTM-CRF, and BLSTM-CNN-CRF.
+They tested them on CoNLL-02 and CoNLL-03 NER benchmark corpora, degraded by applying four different types of OCR noise.
+Overall, NER F-measure drops from 90\% to 50\% when the Word Error Rate increases from 8\% to 50\%.
+However, models based on deep neural networks seem less sensitive to OCR errors.
+Two approaches have been proposed by \cite{huynh2020use} and \cite{marz2021data}  to reduce the negative impact of OCR errors on NER performance on historical texts.
+The former applies a spelling correction tool to several corpora with variable OCR error rates. 
+As long as OCR errors remain low ($CER<2\%$ and $WER<10\%$), this strategy makes it possible to maintain good NER results.
+However, the F1-score starts to decrease significantly when OCR errors exceed these thresholds.
+The latter work focuses on adapting the training data to facilitate the generalization of an off-the-shelf NER model from modern texts to historical texts.
+Finally, reusing a model trained on clean modern data, including embeddings computed on a historical corpus, and fine-tuned on a noisy historical ground truth has proven to be the most effective strategy.
 
-\cite{huynh2020use} and \cite{marz2021data} have proposed different approaches to reduce the negative impact of OCR errors on NER performance with historical texts.
-The former uses a spelling correction tool on several corpora with variable OCR error rates, to assess whether NER performance benefit from spelling corrections or not.
-As long as OCR errors remain low (CER<2\% and WER<10\%), the F-measure of NER results remains stable.
-It starts to decrease significantly when OCR errors exceed these thresholds.
-The latter work focuses on adapting the training data to facilitate the generalization of an off-the-shelf NER model from modern texts to historical texts.
-Three different NER models are tested on three historical corpora, in French, English, and Dutch.
-The best results are produced by a model trained on clean modern data, including embeddings computed with Flair on a historical corpus, and fine-tuned on a noisy historical ground truth.
-
-In conclusion, NER approaches based on deep learning seem to be more suitable for dealing with historical texts as they adapt more easily to OCR errors than rule-based approaches.
-Recent work in this area suggests that the impact of OCR quality on NER can be reduced by using different strategies.
-On the one hand, if the OCR error rate is kept below a certain threshold, the NER models remain little impacted: it is therefore important to reduce OCR errors as much as possible, but a low error rate may remain acceptable.
-On the other hand, reusing a NER model trained on modern data and adapted to historical texts using supervised or unsupervised approaches seems a good strategy.
+Based on those works,
 

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Updates from Overleaf

@@
 \label{sec:ner-xp1}
-
 The constitution of annotated example sets to train a NER model on a target domain is a critical preliminary step.
@@
 The ability of a model to perform well even with few training examples is a practical criteria to consider.
-In this first experiment, we investigate the performance of the three models when fine-tuned for NER on training sets of decreasing sizes.
+In this first experiment, we investigate the NER performance of SpaCy and CamemBERT when fine-tuned with a decreasing number of training examples.
 
-\subsection{Dataset}
-input: ocr ref (human transcription)
-expected output: ner ref (human tagging)
+\subsection{Training and evaluation protocol}
+The following three models form our baseline for both NER experiments.
+Their short names written in square brackets will be used to reference them from now on.
+\begin{itemize}
+    \item \textbf{SpaCy NER pipeline for French~[SpaCy NER]}: We use the pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the french corpora deep-sequoia and wikiner-fr.
+    We stress again that we use the CNN version of this pipeline, not the transformer-based available in SpaCy v3.
+    \item \textbf{Huggingface CamemBERT~[CmBERT]}: We rely on the implementation of BERT models provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
+We chose to reuse a CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
+\item \textbf{CamemBERT pretrained on Paris directories~[CmBERT+ptrn]}: 
+In order to evaluate whether adapting the model to the domain lexicon brings a performance gain, we perform an unsupervised pre-training of CmBERT for next sentence prediciton and masked language modeling, using \num{845000} directories entries extracted from random pages with Pero-OCR.
+The model is trained for 3 epochs and is available online\footnote{\url{https://huggingface.co/HueyNemud/das22-10-camembert_pretrained}}.
+\end{itemize}
 
-8765 entries, \joseph{FIXME XXXX entities to detect}
+Each model is then fine-tuned on subsets of the ground truth of increasing size.
+The NER metrics are eventually measured against a common test set.
+The procedure for creating these sets is as follows.
 
-\subsection{Metrics}
+%First, we split the ground truth into a training set, a development set, and a test set. 
+%The training set is then gradually reduced in size while maintaining the relative frequency of directories within it.
+As the organisation and structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features. 
+To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, $\approx 19\%$) from the ground truth as a test set containing entries from unseen directories.
 
-\subsection{Systems/Variants under test}
-We select two deep-learning-based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
-We also create an additional CamemBERT model pretrained on a collection of unannotated directory entries extracted with Pero-OCR.
-The NER layer of each three models is then fine-tuned on historical directories as detailed in \cref{subsection:experiment-1-setup} and \cref{subsection:experiment-2-setup}
+Then, a stratified sampling based on the source directory of each entry is run on the remainder to create a training set (6373 entries, $\approx 73\%$ of the gold reference) and a development set (709 entries, $\approx 8\%$).
+The development set is used to evaluate the model performance during the training phase.
+This resampling procedure is a convenient way to shape both sets, so they reflect the diversity of directories within the ground truth.
 
-\subsubsection{SpaCy}
-We use the French pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the deep-sequoia and wikiner-fr corpora.
-The default NER labels known to the pipeline are PER, ORG, MISC and LOC.
-In experiment 1 the base model is fine-tuned on every of the 8 training sets using the same training parameters.
-Early stopping is activated with a patience value of 1600 training steps without improvement of the f1 score.
+To generate smaller training sets, we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as for the train/dev split to maintain the relative frequency of directories.
+We stop if a directory has only one entry left, or if the current training subset contains less than 30 entries maintaining the relative frequency of directories within it
+Applying this procedure to the initial training set produced 8 training subsets containing 49, 99, 199, 398, 796, 1593, 3186, and 6373 entries.
 
-\subsubsection{CamemBERT pretrained on directories~[CmBERT+ptrn]}
-In order to evaluate the benefits of pre-training the language model on OCR texts from domain-related documents, we adapt the CmBERT embeddings using \num{845.000} entries randomly selected wihtin the collection of directories and extracted with Pero-OCR.
-We achieve this by training the CmBERT model for 3 epochs on two unsupervised tasks: next sentence prediction and masked language model.
+The three models are trained 5 times on each of the 8 training subsets using an early stopping criterion based on the number of training steps without improvement of the F1-score. 
+This patience threshold is set at 1600 steps for SpaCY NER, and 3 evaluations (1 evaluation every 100 steps) for CmBERT and CmBERT+ptrn.
+The metrics are measured for the 24 resulting NER models on the common test and averaged over the runs.
 
-\subsubsection{Huggingface CamemBERT}
-\label{sub:ner-xp1-sysytems-huggingcam}
-Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
-Our baseline CmBERT model is CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
-Its NER head is a linear model with a Softmax function.
-CmBERT and CmBERT+ptrn are always fine-tuned using the same parameters, with at most 5000 training steps and an early stopping condition set to 3 evaluations in experiment 1 (5 in experiment 2) without improvement of the f1 score.
+%\subsection{Dataset}
+%input: ocr ref (human transcription)
+%expected output: ner ref (human tagging)
+
+%8765 entries, \joseph{FIXME XXXX entities to detect}
+
+%\subsection{Metrics}
+
+%\subsection{Systems/Variants under test}
+%We select two deep-learning-based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
+%We also create an additional CamemBERT model pretrained on a collection of unannotated directory entries extracted with Pero-OCR.
+%The NER layer of each three models is then fine-tuned on historical directories as detailed in \cref{subsection:experiment-1-setup} and \cref{subsection:experiment-2-setup}
+
+%\subsubsection{SpaCy}
+%We use the French pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the deep-sequoia and wikiner-fr corpora.
+%The default NER labels known to the pipeline are PER, ORG, MISC and LOC.
+%In experiment 1 the base model is fine-tuned on every of the 8 training sets using the same training parameters.
+%Early stopping is activated with a patience value of 1600 training steps without improvement of the f1 score.
+
+%\subsubsection{CamemBERT pretrained on directories~[CmBERT+ptrn]}
+%In order to evaluate the benefits of pre-training the language model on OCR texts from domain-related documents, we adapt the CmBERT embeddings using \num{845.000} entries randomly selected wihtin the collection of directories and extracted with Pero-OCR.
+%We achieve this by training the CmBERT model for 3 epochs on two unsupervised tasks: next sentence prediction and masked language model.
+
+%\subsubsection{Huggingface CamemBERT}
+%\label{sub:ner-xp1-sysytems-huggingcam}
+%Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
+%Our baseline CmBERT model is CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
+%Its NER head is a linear model with a Softmax function.
+%CmBERT and CmBERT+ptrn are always fine-tuned using the same parameters, with at most 5000 training steps and an early stopping condition set to 3 evaluations in experiment 1 (5 in experiment 2) without improvement of the f1 score.
 Evaluations are performed every 100 steps.
 
-\subsection{Protocol}
-To do so, we split the gold reference of manually annotated entries into a training set, a development set, and a test set. 
-The training set is then gradually reduced in size while maintaining the relative frequency of directories within it.
+%\subsection{Protocol}
+%To do so, we split the gold reference of manually annotated entries into a training set, a development set, and a test set. 
+%The training set is then gradually reduced in size while maintaining the relative frequency of directories within it.
 
-As the organization and structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features.
-To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, $\approx 19\%$) from the gold reference to test each model on unseen directories.
-Then, a stratified sampling based on the source directory of each entry is run on the remaining set to create a training (6373 entries, $\approx 73\%$ of the gold reference) and a development set (709 entries, $\approx 8\%$).
-This sampling procedure is a convenient way to shape both sets, so they reflect the diversity of directories within the gold reference.
-The development set is used to evaluate the model performance during the training phase.
-To generate smaller training sets, we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as before.
-We stop if a directory has only one entry left, or if the current training subset contains less than 30 entries.
-Applying this procedure to the initial training set produced 8 training subsets containing 49, 99, 199, 398, 796, 1593, 3186, and 6373 entries.
-
-All tree models are fine-tuned for NER on each training subset.
-The performance metrics are measured against the test set after each training.
+%As the organization and structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features.
+%To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, $\approx 19\%$) from the gold reference to test each model on unseen directories.
+%Then, a stratified sampling based on the source directory of each entry is run on the remaining set to create a training (6373 entries, $\approx 73\%$ of the gold reference) and a development set (709 entries, $\approx 8\%$).
+%This sampling procedure is a convenient way to shape both sets, so they reflect the diversity of directories within the gold reference.
+%The development set is used to evaluate the model performance during the training phase.
+%To generate smaller training sets, we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as before.
+%We stop if a directory has only one entry left, or if the current training subset contains less than 30 entries.
+%Applying this procedure to the initial training set produced 8 training subsets containing 49, 99, 199, %398, 796, 1593, 3186, and 6373 entries.
 
 \subsection{Results and discussion}
-
-Table \ref{tab:experiment-1-models-performances} and Figure \ref{fig:f1-vs-trainsize} show the mean F1 score, precision, and recall computed on the test dataset, for 5 fine tunings of each model and for each of the 8 training subsets.
+Table \ref{tab:experiment-1-models-performances} and Figure \ref{fig:f1-vs-trainsize} display the averaged precision, recall and F1-score for all models on the 8 subsets of ground-truth examples.
 CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.
@@
 
-It appears that pretraining the CamemBERT model on OCRed texts seems worth it only when a the training set used to fine-tune the NER layer is small.
+It appears that pretraining the CamemBERT model on OCR texts seems worth it only when a the training set used to fine-tune the NER layer is small.
 This effect might be due to the differences in nature between the training subsets, whose texts are manually corrected, and the noisy OCR texts used to pretrain CamemBERT.
@@
 	       {images/experiment-1-models-performances.pdf}}
-	  \caption{\label{fig:f1-vs-trainsize} Precision, recall and f1 score from table~\ref{tab:experiment-1-models-performances}}.
+	  \caption{\label{fig:f1-vs-trainsize} Graph view of the metrics displayed in ~\ref{tab:experiment-1-models-performances}.
+As SpaCy NER is outperformed by BERT models from the start, CmBERT slightly surpasses the pre-trained version when the number of training samples grows.}.
 \end{figure}

### 2022-01-20 src-latex/parts/20-related-work.tex
Fixed quirks in section 5

@@
 
-Based on those works,
-
-\subsection{== À réintégrer depuis ancienne section 4 ==}
-
-A introduire en disant quo'n chercher à utiliser au plus des modèles simples à reprendre: déjà entraînés, dispos pour du français et adaptables
-We select two deep-learning-based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
+Based on those works, we adopt two French deep-learning-based language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
 

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Fixed quirks in section 5

@@
 Then, a stratified sampling based on the source directory of each entry is run on the remainder to create a training set (6373 entries, $\approx 73\%$ of the gold reference) and a development set (709 entries, $\approx 8\%$).
-The development set is used to evaluate the model performance during the training phase.
+The development set is used to evaluate the model during the training phase.
 This resampling procedure is a convenient way to shape both sets, so they reflect the diversity of directories within the ground truth.
@@
 
-The three models are trained 5 times on each of the 8 training subsets using an early stopping criterion based on the number of training steps without improvement of the F1-score. 
-This patience threshold is set at 1600 steps for SpaCY NER, and 3 evaluations (1 evaluation every 100 steps) for CmBERT and CmBERT+ptrn.
-The metrics are measured for the 24 resulting NER models on the common test and averaged over the runs.
+The three models are fine-tuned on the NER task 5 times using each of the 8 training subsets, with an early stopping criterion based on the number of training steps without improvement of the F1-score.
+This patience threshold is set to 1600 steps for SpaCY NER and 3 evaluations (1 evaluation every 100 steps) for CmBERT and CmBERT+ptrn.
+The metrics are measured for the 24 resulting NER models on the common test set and averaged over the runs.
 
@@
 \centering
-\caption{F1 score, precision and recall measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER in experiment 1 for 8 training sets of increasing sizes.}
+\caption{\label{tab:experiment-1-models-performances} F1 score, precision and recall measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER in experiment 1 for 8 training sets of increasing sizes.}
 \begin{tabular}{llrrrrrrrr}
@@
 \end{tabular}
-\label{tab:experiment-1-models-performances}
 \end{table}
@@
 	       {images/experiment-1-models-performances.pdf}}
-	  \caption{\label{fig:f1-vs-trainsize} Graph view of the metrics displayed in ~\ref{tab:experiment-1-models-performances}.
-As SpaCy NER is outperformed by BERT models from the start, CmBERT slightly surpasses the pre-trained version when the number of training samples grows.}.
+	  \caption{\label{fig:f1-vs-trainsize} Graph view of the metrics displayed in~\cref{tab:experiment-1-models-performances}.
+As SpaCy NER is outperformed by both BERT models from the start, CmBERT slightly surpasses the pre-trained version when the number of training examples becomes large.}.
 \end{figure}

### 2022-01-20 src-latex/figs/eval-ocr-ner.tex
Update OCR results with kraken
MAJ Part 3

@@
 
-  \node [above left=0cm of A, font=\it] {OCR Prediction};
+  \node [above left=0cm of A, font=\it] {OCR-\emph{prediction}};
   \node [above left=0cm of B, font=\it] {GOLD};
@@
   \draw[-latex] (C.south) -- ++(0,-1) node[midway,right,font=\small] {OCR Q.A.} node[below] {CER = 24\%};
-  \draw[-latex] (D.south) -- ++(0,-1) node[midway,right,font=\small] {NER reference.};
-  \draw[-latex] (C.south east) |- (E2) node[pos=0.75,below, font=\it] {NER Prediction};
+  \draw[-latex] (D.south) -- ++(0,-1) node[midway,right,font=\small] {NER-\emph{target}};
+  \draw[-latex] (C.south east) |- (E2) node[pos=0.75,below, font=\it] {NER-\emph{prediction}};
   \draw[-latex] (E2.south) -- ++(0,-1) node[midway,right,font=\small] {NER Q.A.}
-  node[below] {Precision = 33\%, Recall = 25\%, F-score = 0.4};
+  node[below] {Precision = 66\%, Recall = 50\%, F-score = 0.4};
 \end{tikzpicture}

### 2022-01-20 src-latex/parts/20-related-work.tex
MAJ Part 3
Updates from Overleaf

@@
 \joseph{Move to implementation details for XP OCR or to pipeline summary}
-\begin{itemize}
-    \item Tesseract 4.1.1 (newest version: 5, released Nov. 2021)
-    \item Pero OCR version from git repo, master branch updated on Sep 15, 2021 % 8b20f29    
-    \item other?  % kraken if times allows
-\end{itemize}
+\nathalie{Pour l'instant je les ai déplacés dans le 2.3 pipeline summary}
 
@@
 This strategy has been adopted by \cite{Labusch2020NamedED} to extract named entities in OCRed historical texts in German, French, and English.
-However, the NER performance decreased dramatically for OCRed texts compared to the groundtruth dataset.
-This decrease went hand in hand with the drop of the OCR quality.
+However, the NER performance drops significantly as the quality of the OCR decreases and is correlated with its decrease.
 
@@
 
-Based on those works, we adopt two French deep-learning-based language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
+\subsection{Pipeline summary}
 
-\subsubsection{spacy}
-Spacy is a software library that offers NLP components assembled in modular pipelines specialised by language.
+Based on those works, we chose to test three OCR systems, namely, Tesseract, Pero OCR, and Kraken.
+We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
+
+\textbf{Tesseract} 4.1.1 (newest version: 5, released Nov. 2021)
+\textbf{Pero OCR} version from git repo, master branch updated on Sep 15, 2021
+\textbf{Kraken}
+
+\textbf{Spacy} is a software library that offers NLP components assembled in modular pipelines specialised by language.
 Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (jan. 2022).
@@
 
-% Other possibilities:
-%1. Traditional ML based:
-%    Conditional Random Fields (CRF) - https://pypi.org/project/sklearn-crfsuite/
-%    Maximum-entropy Markov model
-
-%2. Neural Networks based:
-%    LSTMs, bi-LSTM - https://github.com/flairNLP/flair
-%    CNNs (SpaCy uses CNN based architecture)
-%    Transformers (Spacy has recently launched it) - %https://spacy.io/universe/project/spacy-transformers
-
-\subsubsection{Transformers / Bert / Hugging face}
+\textbf{Transformers / Bert / Hugging face}
 Language model based on \textit{Transformers} \cite{vaswani2017attention}, like BERT \cite{devlin2018bert}, have become a new paradigm for NER\cite{li2020}. 
@@
 \joseph{plutôt tourner ça pour dire qu'il existe des modèles FR / qu'on est limités par la dispo de modèles FR pré-entraînés ?}
-
-\subsection{Pipeline summary}
 

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Updates from Overleaf

@@
 The ability of a model to perform well even with few training examples is a practical criteria to consider.
-In this first experiment, we investigate the NER performance of SpaCy and CamemBERT when fine-tuned with a decreasing number of training examples.
+In this first experiment, we investigate the NER performance of SpaCy and CamemBERT when fine-tuned with an increasing number of training examples.
 
 \subsection{Training and evaluation protocol}
+\label{sec:ner-xp1-protocol}
 The following three models form our baseline for both NER experiments.
@@
 \item \textbf{CamemBERT pretrained on Paris directories~[CmBERT+ptrn]}: 
-In order to evaluate whether adapting the model to the domain lexicon brings a performance gain, we perform an unsupervised pre-training of CmBERT for next sentence prediciton and masked language modeling, using \num{845000} directories entries extracted from random pages with Pero-OCR.
+In order to evaluate whether adapting CamemBERT to the domain increases its performances, we do an unsupervised pre-training of CmBERT for next sentence prediction and masked language modeling, using \num{845000} entries randomly sampled and OCRed with Pero.
 The model is trained for 3 epochs and is available online\footnote{\url{https://huggingface.co/HueyNemud/das22-10-camembert_pretrained}}.

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex

@@
 % !TeX root = ../main-paper.tex
-\section{Experiment 2: Impact of OCR nosie on named entity recognition}
+\section{Experiment 2: Impact of OCR noise on named entity recognition}
 \label{sec:ner-xp2}
-Noise introduced by OCR is known to impede the named entity recognition process (see section~\ref{subsection:stoa-ner-on-historical-texts}).
-%The second experiment assesses the impact of the OCR on the quality of the NER results.
-The second experiment addresses two questions (1) which model is the most robust to OCR noise and (2) what is the most appropriate strategy to fine-tune the NER layer if the input texts to label are noisy ?
+Noise introduced by OCR is known to have a negative impact on named entity recognition, because it alters the structure and lexicon of the input texts, moving them away from the language model known to the NER process.
+In real-life situations the models used are often trained on texts without such noise, even though the texts to be annotated are extracted with OCR.
+This raises the question of the most appropriate strategy to build a NER model tolerant to OCR noise.
+In this second experiment we address this question based on our benchmark.
 
-\subsection{Protocol}
+\subsection{Training and evaluation protocol}
+Only CmBERT and CmBERT+ptrn are considered since experiment 1 shows that SpaCy NER is systematically outperformed by these two models.
 
-Dataset
+We leverage the labeled datasets NER-reference, NER-pero and NER-tesseract created as explained in \cref{sec:dataset}.
+Kraken is left aside as it produces poor results that would require removing 500 entries from all datasets to keep the same set of valid entries.
 
-To this end, we leverage the labeled OCR datasets produced following to the method explained in Section 5, in addition to the gold reference.
+Each dataset is split into a training, a development and a test set following the exact same protocol as described in \cref{sec:ner-xp1-protocol}, except this time we do not need to create smaller training sets.
+The train/dev/test sets are the same size for each NER set as they contain the same entries:
 
-They are denoted hereafter {reference, pero, tesseract}-gold.
+CmBERT and CmBERT+ptrn are fine-tuneon reference-gold and pero-gold to create two versions of each model: the former trained on manually corrected entries and the latter on OCR entries.
+SpaCy NER is left aside as results from experiment 1 show that it is outperformed by BERT models.
+Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold and tesseract-gold.
+
 Finally each is is split into a training, a dev and a test subset following the procedure presented in\cref{subsection:experiment-1-setup}.
-
-\joseph{préciser qu'on ne prend que les données qui sont valides pour pero et tesseract (soit 8341 entrées).
-On ne considère pas kraken car pas le temps, on perd 500 entrées dans l'intersection, la qualité est pourrie}
 

### 2022-01-20 src-latex/parts/20-related-work.tex
Small changes to state of the art / OCR

@@
 \label{sec:related-work}
-
-\joseph{Suggestion pour réduire la présentation de l'état de l'art : l'orienter exclusivement vers une justification des systèmes que nous avons retenus pour cet article.}
 
@@
 \subsection{Optical Character Recognition on historical texts}
+Among the large amount of OCR solutions, being either open, free or paid software, available as libraries, Python packages, binaries or cloud API, not all options seem suitable for historical document processing.
+We chose to avoid in our study paid and closed-source solutions.
+This notably discards Transkribus~\cite{transkribus}, which relies on the commercial system ABBYY's Finereader as well as on two handwritten transcription engines, to process text.
+% for Transkribus HTR engines, there is one closed source and one apparently open source
+% we should test the open source one: https://github.com/jpuigcerver/pylaia
 
-Most of the current state-of-the-art OCR systems, like Tesseract \cite{smith2007overview}, OCRopus \cite{breuel2008ocropus} and PERO OCR \cite{kohut2021ts} are based on a pipeline of convolutional neural networks (CNNs) and long short-term memory networks (LSTM).
+Most of the current state-of-the-art open-source OCR systems, like Tesseract \cite{smith2007overview}, OCRopus \cite{breuel2008ocropus}, and the recent Kraken~\cite{kraken}, Calamari~\cite{wick_calamari_2020} and \peroocr \cite{kohut2021ts} are based on a pipeline of convolutional neural networks (CNNs) and long short-term memory networks (LSTM).
 Although this kind of model produces good results with modern texts, it faces several challenges with ancient texts, such as the lack of annotated data for learning, or different transcription styles in training data.
+%
+% \cite{martinek2019hybrid} propose an approach to generate synthetic annotated text for historical OCR training, based on manually collected characters from historical text images.
+% This work proposes to train an OCR system based on a CNN-LSTM network with synthetic data and then to fine-tune the model with some pages of real historical annotated text.
+% The results show that this approach gives state-of-the-art results.
 
-\cite{martinek2019hybrid} propose an approach to generate synthetic annotated text for historical OCR training, based on manually collected characters from historical text images.
-This work proposes to train an OCR system based on a CNN-LSTM network with synthetic data and then to fine-tune the model with some pages of real historical annotated text.
-The results show that this approach gives state-of-the-art results.
-
-To overcome the limitations due to different transcription styles in training data, PERO OCR adds a Transcription Style Block layer to a classical model based on a CNN and a Recurrent Neural Network components \cite{kohut2021ts}.
+To overcome the limitations due to different transcription styles in training data, \peroocr adds a Transcription Style Block layer to a classical model based on a CNN and a Recurrent Neural Network components \cite{kohut2021ts}.
 This block takes the image of the text and a Transcription Style Identifier as inputs and helps the network decide what kind of transcription style to use as output.
 
-\joseph{Move to implementation details for XP OCR or to pipeline summary}
-\nathalie{Pour l'instant je les ai déplacés dans le 2.3 pipeline summary}
+In this study, we consider Tesseract version 4, Kraken and \peroocr.
+They were selected because of their ease of use (in the case of Tesseract and Kraken)
+or because of their performance (in the case of \peroocr),
+but further work should include more systems from our initial selection.
+
+% Finally don't mention OCR-D as it relies on either Tesseract, Kraken, Ocropus or Calamari
 
@@
 \Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems.
-
-\begin{figure}[tb]
-    \includegraphics[width=\linewidth]{figs/protocol.pdf}
-    \caption{Scheme of the evaluation protocol. \joseph{Missing "Ground truth" for NER stage?}}
-    \label{fig.protocol}
-    \end{figure}

### 2022-01-20 src-latex/parts/20-related-work.tex
Typo titre?

@@
 
-\subsection{Optical Character Recognition on historical texts}
+\subsection{Optical Character Recognition of historical texts}
 Among the large amount of OCR solutions, being either open, free or paid software, available as libraries, Python packages, binaries or cloud API, not all options seem suitable for historical document processing.

### 2022-01-20 src-latex/parts/20-related-work.tex
Updates from Overleaf

--- 2022-01-20 src-latex/parts/20-related-work.tex
@@
 \textbf{Tesseract} 4.1.1 (newest version: 5, released Nov. 2021)
+
 \textbf{Pero OCR} version from git repo, master branch updated on Sep 15, 2021
+
 \textbf{Kraken}
 
-\textbf{Spacy} is a software library that offers NLP components assembled in modular pipelines specialised by language.
+\textbf{SpaCy} is a software library that offers NLP components assembled in modular pipelines specialised by language.
 Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (jan. 2022).
@@
 
-\textbf{Transformers / Bert / Hugging face}
+\textbf{CamemBERT}
 Language model based on \textit{Transformers} \cite{vaswani2017attention}, like BERT \cite{devlin2018bert}, have become a new paradigm for NER\cite{li2020}. 
@@
 They can also be pretrained in an unsupervised way on historical texts for domain adaptation.
-As the directories are written in French, we chose the language model CamemBERT \cite{martin-etal-2020-camembert}, a Transformer model trained on a French corpus.
-\joseph{plutôt tourner ça pour dire qu'il existe des modèles FR / qu'on est limités par la dispo de modèles FR pré-entraînés ?}
+CamemBERT \cite{martin-etal-2020-camembert}, is a weel-known French language model derived from BERT, of which there are already trained versions for the NER task on the Hugging Face library.
 
-\Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems.
-
-\begin{figure}[tb]
-    \includegraphics[width=\linewidth]{figs/protocol.pdf}
-    \caption{Scheme of the evaluation protocol. \joseph{Missing "Ground truth" for NER stage?}}
-    \label{fig.protocol}
-    \end{figure}
+The Figure \ref{fig.pipeline} depicts the evaluation protocol used to assess the combined performance of these OCR and NER systems.

### 2022-01-20 src-latex/parts/30-dataset.tex
MAJ Part 3
Updates from Overleaf

@@
 
-The directories to be processed are stored in different libraries in Paris\footnote{\textcolor{green}{Citer les bibliothèques sources?} \joseph{est-ce nécessaire ici ?
-Est-ce que ça peut être dans une annexe ?}\nathalie{Oui totalement...}}.
+The directories to be processed are stored in different libraries in Paris.
 They have been scanned independently and with various levels of quality.
@@
 For training as testing purposes, our groundtruth dataset must be as representative as possible of the diversity of the available directories.
+The Table \ref{tab:directories} lists the directories selected, their layout, the structure of their entries, and the number of annotated entries according to the considered index.
 

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Updates from Overleaf

@@
 Each dataset is split into a training, a development and a test set following the exact same protocol as described in \cref{sec:ner-xp1-protocol}, except this time we do not need to create smaller training sets.
-The train/dev/test sets are the same size for each NER set as they contain the same entries:
+As the NER sets contain 8341 entries, the producing train sets (resp. development and test) contain 6004 entries - 72\% of the total (resp. 668 - 8\% and 1669 - 20\%).
 
-CmBERT and CmBERT+ptrn are fine-tuneon reference-gold and pero-gold to create two versions of each model: the former trained on manually corrected entries and the latter on OCR entries.
-SpaCy NER is left aside as results from experiment 1 show that it is outperformed by BERT models.
-Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold and tesseract-gold.
+CmBERT and CmBERT+ptrn are fine-tuned on the training sets built from NER-reference and NER-pero.
+Finally the metrics are measured against the three tests sets.
 
-Finally each is is split into a training, a dev and a test subset following the procedure presented in\cref{subsection:experiment-1-setup}.
+%\subsubsection{Huggingface CamemBERT}
+%\joseph{Copie de \cref{sub:ner-xp1-sysytems-huggingcam}. Je suggère de rédiger la section de l'XP 1 en premier et apporter précision nécessaires ici.}
 
-\subsubsection{Huggingface CamemBERT}
-\joseph{Copie de \cref{sub:ner-xp1-sysytems-huggingcam}. Je suggère de rédiger la section de l'XP 1 en premier et apporter précision nécessaires ici.}
-
-Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
-Our baseline CmBERT model is CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
-Its NER head is a linear model with a Softmax function.
-CmBERT and CmBERT+ptrn are always fine-tuned using the same parameters, with at most 5000 training steps and an early stopping condition set to 3 evaluations in experiment 1 (5 in experiment 2) without improvement of the f1 score.
+%Experiments 1 and 2 rely on the implementation of transformers provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
+%Our baseline CmBERT model is CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
+%Its NER head is a linear model with a Softmax function.
+%CmBERT and CmBERT+ptrn are always fine-tuned using the same parameters, with at most 5000 training steps and an early stopping condition set to 3 evaluations in experiment 1 (5 in experiment 2) without improvement of the f1 score.
 Evaluations are performed every 100 steps.
 
-\subsection{Protocol}
+%\subsection{Protocol}
 
-We fine-tune CmBERT and CmBERT+ptrn on reference-gold and pero-gold to create two versions of each model: the former trained on manually corrected entries and the latter on OCR entries.
-SpaCy NER is left aside as results from experiment 1 show that it is outperformed by BERT models.
-Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold and tesseract-gold.
+%We fine-tune CmBERT and CmBERT+ptrn on reference-gold and pero-gold to create two versions of each model: the former trained on manually corrected entries and the latter on OCR entries.
+%SpaCy NER is left aside as results from experiment 1 show that it is outperformed by BERT models.
+%Performance metrics are computed for each of the four resulting NER models against the test sets created from reference-gold, pero-gold and tesseract-gold.
 
 \subsection{Results and discussion}
-
-CmBERT and CmBERT+ptrn are fine-tuned on the training sets from reference-gold (manually corrected entries) and pero-gold to produce 4 different models.
-Their F1 score measured against the test tests from reference-gold, pero-gold and tesseract-gold are given in \cref{fig:exp_2_eval_ner}.
-The results clearly show that models perform best when both the pre-training and the NER fine-tuning share the same characteristics (here, OCR noise) as the texts to be processed.
+The measured F1-score are given in \cref{fig:exp_2_eval_ner}.
+Results clearly show that models perform best when both the pre-training and the NER fine-tuning share the same characteristics (here, OCR noise) as the texts to be processed.
 
@@
 We did not pre-train or fine-tune with texts extracted with Tesseract.
-However, despite a loss of performance the model pre-trained and fine-tuned with pero-gold still gives the best results.
+However, despite a loss of performance, the model pre-trained and fine-tuned on NER-pero still gives the best results.
 This is probably due to the fact that the texts produced by Pero-OCR feature characteristics intermediary between human transcriptions and Tesseract.
+This OCR tool removes the characted recognized with a low confidence, which is probably a great help to the NER.
 

### 2022-01-20 src-latex/parts/20-related-work.tex

--- 2022-01-20 src-latex/parts/20-related-work.tex
@@
 They can also be pretrained in an unsupervised way on historical texts for domain adaptation.
-As the directories are written in French, we chose the language model CamemBERT \cite{martin-etal-2020-camembert}, a Transformer model trained on a French corpus.
-\joseph{plutôt tourner ça pour dire qu'il existe des modèles FR / qu'on est limités par la dispo de modèles FR pré-entraînés ?}
+CamemBERT \cite{martin-etal-2020-camembert}, is a weel-known French language model derived from BERT, of which there are already trained versions for the NER task on the Hugging Face library.
 
-\Cref{fig.protocol} depicts the evaluation protocol used to assess the OCR and NER systems.
+The Figure \ref{fig.pipeline} depicts the evaluation protocol used to assess the combined performance of these OCR and NER systems.

### 2022-01-20 src-latex/parts/20-related-work.tex
Reduced section 2

@@
 Most of the time, they have evaluated NER approaches based on deep learning architectures as they seem to adapt more easily to OCR errors than rule-based or more classical supervised approaches.
-\cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by Spacy\footnote{\url{https://spacy.io/}} library to perform NER on \textit{Person}, \textit{GPE}\footnote{Geopolitical Entity} and \textit{Date} on a corpus made of many journal articles with different levels of OCR errors.
+\cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by SpaCy\footnote{\url{https://spacy.io/}} library to perform NER on a corpus made of many journal articles with different levels of OCR errors.
 For each OCRed article, a ground truth text is available so that the Word Error Rate (WER) can be computed.

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Section 6

@@
 This is probably due to the fact that the texts produced by Pero-OCR feature characteristics intermediary between human transcriptions and Tesseract.
-This OCR tool removes the characted recognized with a low confidence, which is probably a great help to the NER.
+This OCR tool removes the characters recognised with a low confidence, which is probably a great help to the NER.
 
@@
     \caption{F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing examples, either manually corrected or raw from Pero-OCR and Tesseract.
-The type of examples used to train the NER task is noted in indice after the model name (e.g.
-Bert$_{reference}$).
-Results show that best performances on OCRed entries are obtained when the BERT model has been pretrained and fine-tuned for NER on examples affected with similar OCR errors.}
+The dataset used to train the NER task is noted in indice after the model name (e.g.
+CmBERT$_{reference}$).
+Results show that best performances on noisy entries are obtained when the model has been pre-trained and fine-tuned for NER on examples affected with a similar OCR noise.}
     \label{fig:exp_2_eval_ner}

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
fig 7 caption

@@
     \includegraphics[width=0.75\textwidth]{images/experiment_2_f1_with_noise_graph.pdf}
-    \caption{F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing examples, either manually corrected or raw from Pero-OCR and Tesseract.
+    \caption{\label{fig:exp_2_eval_ner}F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing examples, grouped by test set.
 The dataset used to train the NER task is noted in indice after the model name (e.g.
@@
 Results show that best performances on noisy entries are obtained when the model has been pre-trained and fine-tuned for NER on examples affected with a similar OCR noise.}
-    \label{fig:exp_2_eval_ner}
 \end{figure}

### 2022-01-20 src-latex/parts/10-introduction.tex
British-isation except for section 3

--- 2022-01-20 src-latex/parts/10-introduction.tex
@@
     \emph{How noisy are modern, out-of-the-box OCR systems?}
-    \emph{What is the behavior of NER when OCR is noisy?}
+    \emph{What is the behaviour of NER when OCR is noisy?}
     \emph{Can NER be made more robust to OCR noise?}
@@
 OCRed texts are generally not sufficient to build a high level, semantic view of a collection of historical documents.
-A subsequent stage is often needed to extract the pieces of information most likely to be searched for by users, such as named entities: people, organizations, dates, places, etc.
+A subsequent stage is often needed to extract the pieces of information most likely to be searched for by users, such as named entities: people, organisations, dates, places, etc.
 Indeed, being able to properly tag text tokens unlocks the ability to relate entities, and provide colleagues from other fields with databases ready for exploitation.
@@
 Being active research topics, OCR and named entity recognition (NER) are still difficult tasks when applied to historical text documents.
-OCR approaches used for modern documents are likely to struggle even on printed historical documents due multiple causes related to text readability (low resolution scans, inconsistent printing rules, artifacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
+OCR approaches used for modern documents are likely to struggle even on printed historical documents due to multiple causes related to text readability (low resolution scans, inconsistent printing rules, artefacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
 On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts, and also from the errors introduced into the text by the OCR step.
@@
 They contain hundred-pages long lists of people, along with their activity and their address, and provide valuable fine-grained knowledge to study the social dynamics of the city over time.
-Despite containing roughly the same information, the diversity in layouts, information organization and digitizing quality is challenging both for OCR and NER.
+Despite containing roughly the same information, the diversity in layouts, information organisation and digitising quality is challenging both for OCR and NER.
 
 In a recent work to locate potentially polluted urban soils \cite{bell2020automated}, trade directories were leveraged to identify and locate all the gas stations in the city of Providence during the XX\textsuperscript{th} century.
-The proposed information extraction pipeline extracts texts with OCR, identifies directories entries and recognizes business type and address within each entry.
+The proposed information extraction pipeline extracts texts with OCR, identifies directories entries and recognises business type and address within each entry.
 In the same way, we aim at producing structured spatio-temporal data from Parisian trade directories.
-These directories originate from several publishers, their content is organized following different indexing methods (by name, by activity or by address), printed in various layouts, use different fonts.
+Because they originate from several publishers, their content is organised following different indexing methods (by name, by activity or by address), is printed in various layouts, and uses different fonts.
 We therefore investigate several state-of-the-art OCR and NER approaches to assess their usability on the corpus.

### 2022-01-20 src-latex/parts/20-related-work.tex
Section 6
British-isation except for section 3

@@
 
-Many approaches have been designed to recognize named entities, ranging from handcrafted rules to supervised approaches \cite{nadeau2007}.
+Many approaches have been designed to recognise named entities, ranging from handcrafted rules to supervised approaches \cite{nadeau2007}.
 Rule based approaches look for portions of the text that match patterns like in \cite{bell2020automated,nouvel2011} or dictionary (gazetteers, author lists, etc.) entries like in \cite{mansouri2008,maurel2011}.
-Such kind of approaches achieve very good results when applied to a specialized domain corpus and when an exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
+Such kind of approaches achieve very good results when applied to a specialised domain corpus and when an exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
 
@@
 In recent years, the latter have grown dramatically, yielding state-of-the-art performances as shown in the recent survey proposed by \cite{li2020}.
-This survey concludes that fine-tuning general-purpose contextualized language models with domain-specific data is very likely to give good performance for use cases with domain-specific texts and few training data.
+This survey concludes that fine-tuning general-purpose contextualised language models with domain-specific data is very likely to give good performance for use cases with domain-specific texts and few training data.
 This strategy has been adopted by \cite{Labusch2020NamedED} to extract named entities in OCRed historical texts in German, French, and English.
@@
 However, the F1-score starts to decrease significantly when OCR errors exceed these thresholds.
-The latter work focuses on adapting the training data to facilitate the generalization of an off-the-shelf NER model from modern texts to historical texts.
+The latter work focuses on adapting the training data to facilitate the generalisation of an off-the-shelf NER model from modern texts to historical texts.
 Finally, reusing a model trained on clean modern data, including embeddings computed on a historical corpus, and fine-tuned on a noisy historical ground truth has proven to be the most effective strategy.
@@
 We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
+
+\nathalie{A compléter ici?}
 
@@
 \textbf{SpaCy} is a software library that offers NLP components assembled in modular pipelines specialised by language.
-Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (jan. 2022).
+Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (as of january 2022).
 Hence, we rely on SpaCy's ad hoc pipeline trained on French corpora and capable of named entity recognition.
@@
 As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics.
-The classification layer relies on a finite-state machine whose transition probabilities are learned using a multilayer perceptron.
+The classification layer relies on a finite-state machine whose transition probabilities are learned using a multi layer perceptron.
 % \bertrand{Drop the next sentences if we need space}
@@
 The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer.
-They can also be pretrained in an unsupervised way on historical texts for domain adaptation.
-CamemBERT \cite{martin-etal-2020-camembert}, is a weel-known French language model derived from BERT, of which there are already trained versions for the NER task on the Hugging Face library.
+They can also be pre-trained in an unsupervised way on historical texts for domain adaptation.
+CamemBERT \cite{martin-etal-2020-camembert}, is a well-known French language model derived from BERT, of which there are already trained versions for the NER task on the Hugging Face library.
 

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Updates from Overleaf
British-isation except for section 3

@@
 \subsection{Results and discussion}
-Table \ref{tab:experiment-1-models-performances} and Figure \ref{fig:f1-vs-trainsize} display the averaged precision, recall and F1-score for all models on the 8 subsets of ground-truth examples.
+Table \ref{tab:experiment-1-models-performances} and Figure \ref{fig:f1-vs-trainsize} display the averaged precision, recall and F1-score for all models on the 8 subsets created from the groundtruth.
 CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.
@@
 
-It appears that pretraining the CamemBERT model on OCR texts seems worth it only when a the training set used to fine-tune the NER layer is small.
+It appears that pre-training the CamemBERT model on OCR texts seems worth it only when a the training set used to fine-tune the NER layer is small.
 This effect might be due to the differences in nature between the training subsets, whose texts are manually corrected, and the noisy OCR texts used to pretrain CamemBERT.
-Indeed, the learned embeddings from the pretraining are specialized on noisy texts and therefore less adapted to clean texts.
-The pretraining aims at adapting the model to the vocabulary of the domain and to the errors caused by the OCR, which reveals not helpful and even counter-productive when the texts do not contain these types of errors.
+Indeed, the learned embeddings from the pret-raining are specialised on noisy texts and therefore less adapted to clean texts.
+The pre-training aims at adapting the model to the vocabulary of the domain and to the errors caused by the OCR, which reveals not helpful and even counter-productive when the texts do not contain these types of errors.
 
-\begin{table}[h!]
+\begin{table}[ht!]
 \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
 	   \center{\includegraphics[width=\textwidth]

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
British-isation except for section 3

@@
 
-We leverage the labeled datasets NER-reference, NER-pero and NER-tesseract created as explained in \cref{sec:dataset}.
+We leverage the labeled sets of entries NER-\emph{reference}, NER-\emph{pero} and NER-\emph{tesseract} created as explained in \cref{sec:dataset}.
 Kraken is left aside as it produces poor results that would require removing 500 entries from all datasets to keep the same set of valid entries.
@@
 
-CmBERT and CmBERT+ptrn are fine-tuned on the training sets built from NER-reference and NER-pero.
+CmBERT and CmBERT+ptrn are fine-tuned on the training sets built from NER-\emph{reference} and NER-\emph{pero}.
 Finally the metrics are measured against the three tests sets.
@@
 The dataset used to train the NER task is noted in indice after the model name (e.g.
-CmBERT$_{reference}$).
+CmBERT$_{pero}$ means CmBERT fine-tuned on NER-\emph{pero}).
 Results show that best performances on noisy entries are obtained when the model has been pre-trained and fine-tuned for NER on examples affected with a similar OCR noise.}

### 2022-01-20 src-latex/parts/20-related-work.tex
Updates from Overleaf

@@
 Most of the time, they have evaluated NER approaches based on deep learning architectures as they seem to adapt more easily to OCR errors than rule-based or more classical supervised approaches.
-\cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by SpaCy\footnote{\url{https://spacy.io/}} library to perform NER on a corpus made of many journal articles with different levels of OCR errors.
+\cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by SpaCy \cite{spacy} library to perform NER on a corpus made of many journal articles with different levels of OCR errors.
 For each OCRed article, a ground truth text is available so that the Word Error Rate (WER) can be computed.
@@
 
-\nathalie{A compléter ici?}
+\textbf{Tesseract} is a long-living project, born as a closed-sourced OCR at Hewlett-Packard in the eighties, it was progressively modernized, then open-sourced in 2005.
+From 2006 until November 2018 it was developed by Google, and is still very active.
+We used in our tests the version 4.1.1, released Dec.
+26, 2019.
+The version 5, released on Nov.
+30, 2021, has not been integrated in our tests yet.
 
-\textbf{Tesseract} 4.1.1 (newest version: 5, released Nov. 2021)
+\textbf{Kraken} is a project created by Benjamin Kiessling several years ago (development can be traced back to 2015), and is actively used in the open-source eScriptorium project \cite{kiessling_escriptorium_2019}.
+As no pre-trained model for modern French was easily available, we used the default English text recognition model trained on modern printed English by Benjamin Kiessling on 2019.
+Models can be easily found and downloaded thanks to their hosting on Zenodo.
 
-\textbf{Pero OCR} version from git repo, master branch updated on Sep 15, 2021
-
-\textbf{Kraken}
+\textbf{\peroocr} is a very recent project (started in 2020) from the Brno University of Technology, in Czech Republic.
+Their authors used many state-of-the-art techniques to train it very efficiently.
+We used the version from the master branch of their GitHub repository, updated on Sep 15, 2021.
+We used the pre-trained weights provided by the authors on the same repository, created on Oct. 9, 2020 from European texts with Latin, Greek and Cyrillic scripts.
+% https://github.com/DCGM/pero-ocr
+% https://www.fit.vut.cz/~ihradis/pero/pero_eu_cz_print_newspapers_2020-10-09.tar.gz
 

### 2022-01-20 src-latex/parts/80-conclusion.tex
New paper structure with 1 section for each experiment
Updates from Overleaf (#11)

@@
 \section{Conclusion}
+
+In this article, we assessed the performance of three modern OCR systems on real historical data.
+Although Pero-OCR clearly outperforms its competitors, the qualitative analysis of its errors shows that it fails on different entries.
+This calls for leveraging both Pero-OCR and Tesseract in a complementary way to get the best out of both systems.
+Moreover, we evaluated the needs of two deep-learning-based NER approaches in terms of training data amount and text quality for pretraining.
+We showed that BERT-based NER can benefit from pretraining and fine-tuning on a noisy OCRed corpus to improve its performance on a noisy OCRed test dataset.
+To perform a NER task on OCRised texts, it seems therefore worthwhile to train the NER model on a corpus OCRised with the same system and not corrected.
+However, the results are not as good as for clean text: it is therefore interesting to investigate how to improve the OCR results or to correct the text in a postprocessing step.
+Furthermore, it seems that the models quickly achieve good performance with very few training examples.
+Leveraging a model fine-tuned on a few examples to automatically produce much larger training sets could reveal a solution to lower the annotation burden.
+Besides, as directory entries always have the same structure - at least within a given list - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout.
+
+Interesting points to discuss:
+\begin{itemize}
+    \item can we train on noisy data? (without manual OCR correction?) => future work? cf Pero OCR training procedure?
+    \item do we need better OCR systems or better post-correction techniques (if NER is reliable enough)?
+    \item Construction of the lexicon and associated cost \nathalie{là je ne comprends pas}
+\end{itemize}
 
@@
     \item Evaluate this: fine-tune Spacy NER on a few samples to generated a bootstrap dataset, feed it to BERT (cleaned or raw) to create an good model with minimal efforts ?
+\nathalie{les résultats de la table 2 ne plaident pas vraiment pour ça: même sur les très petits corpus, spacy est tours derrière...}
     \item Use ML to infer regexes ?

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Updates from Overleaf
Commented out table 2, merged it with figure 6

@@
 The three models are fine-tuned on the NER task 5 times using each of the 8 training subsets, with an early stopping criterion based on the number of training steps without improvement of the F1-score.
-This patience threshold is set to 1600 steps for SpaCY NER and 3 evaluations (1 evaluation every 100 steps) for CmBERT and CmBERT+ptrn.
+This patience threshold is set to 1600 steps for SpaCy NER and 3 evaluations (1 evaluation every 100 steps) for CmBERT and CmBERT+ptrn.
 The metrics are measured for the 24 resulting NER models on the common test set and averaged over the runs.
@@
 \subsection{Results and discussion}
-Table \ref{tab:experiment-1-models-performances} and Figure \ref{fig:f1-vs-trainsize} display the averaged precision, recall and F1-score for all models on the 8 subsets created from the groundtruth.
+Figure \ref{fig:f1-vs-trainsize} displays the averaged precision, recall and F1-score for all models on the 8 subsets created from the groundtruth.
 CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.
@@
 
-\begin{table}[ht!]
-\centering
-\caption{\label{tab:experiment-1-models-performances} F1 score, precision and recall measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER in experiment 1 for 8 training sets of increasing sizes.}
-\begin{tabular}{llrrrrrrrr}
-       & Training examples &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
-       & \% & 0.8   & 1.6   & 3.1   & 6.2   & 12.5  & 25.0  & 50.0  & 100.0 \\
-\midrule\bottomrule
-\multirow{3}{*}{\rotatebox{90}{F1 score}} & CmBERT &  89.5 &  90.5 &  92.7 &  93.3 &  \textbf{94.1} &  \textbf{94.9} &  \textbf{94.6} &  \textbf{95.1} \\
-       & CmBERT-ptrn &  \textbf{92.2} &  \textbf{92.9} &  \textbf{93.6} &  \textbf{93.8} &  93.8 &  94.1 &  \textbf{94.6} &  94.4 \\
-       & SpaCy NER &  87.0 &  89.0 &  90.3 &  91.9 &  92.1 &  92.8 &  93.2 &  93.5 \\
-\cline{1-10}
-\multirow{3}{*}{\rotatebox{90}{Precision}} & CmBERT &  87.4 &  88.7 &  91.5 &  92.7 &  93.3 &  94.9 &  93.9 &  95.1 \\
-       & CmBERT-ptrn &  90.8 &  91.8 &  92.9 &  93.0 &  93.0 &  93.4 &  94.1 &  93.9 \\
-       & SpaCy NER &  85.6 &  87.7 &  90.0 &  92.0 &  92.4 &  92.8 &  93.1 &  93.7 \\
-\cline{1-10}
-\multirow{3}{*}{\rotatebox{90}{Recall}} & CmBERT &  91.6 &  92.5 &  93.9 &  93.9 &  94.9 &  94.9 &  95.4 &  95.1 \\
-       & CmBERT-ptrn &  93.6 &  94.0 &  94.4 &  94.6 &  94.6 &  94.8 &  95.0 &  94.9 \\
-       & SpaCy NER &  88.6 &  90.4 &  90.7 &  91.7 &  91.9 &  92.8 &  93.3 &  93.4 \\
-\end{tabular}
-\end{table}
+% MOVE THIS TABLE TO THE APPENDIX
+%\begin{table}[ht!]
+%\centering
+%\caption{\label{tab:experiment-1-models-performances} F1 score, precision and recall measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER in experiment 1 for 8 training sets of increasing sizes.}
+%\begin{tabular}{llrrrrrrrr}
+%       & Training examples &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
+%       & \% & 0.8   & 1.6   & 3.1   & 6.2   & 12.5  & 25.0  & 50.0  & 100.0 \\
+%\midrule\bottomrule
+%\multirow{3}{*}{\rotatebox{90}{F1 score}} & CmBERT &  89.5 &  90.5 &  92.7 &  93.3 &  \textbf{94.1} &  \textbf{94.9} &  \textbf{94.6} &  \textbf{95.1} \\
+%       & CmBERT-ptrn &  \textbf{92.2} &  \textbf{92.9} &  \textbf{93.6} &  \textbf{93.8} &  93.8 &  94.1 &  \textbf{94.6} &  94.4 \\
+%       & SpaCy NER &  87.0 &  89.0 &  90.3 &  91.9 &  92.1 &  92.8 &  93.2 &  93.5 \\
+%\cline{1-10}
+%\multirow{3}{*}{\rotatebox{90}{Precision}} & CmBERT &  87.4 &  88.7 &  91.5 &  92.7 &  93.3 &  94.9 &  93.9 &  95.1 \\
+%       & CmBERT-ptrn &  90.8 &  91.8 &  92.9 &  93.0 &  93.0 &  93.4 &  94.1 &  93.9 \\
+%       & SpaCy NER &  85.6 &  87.7 &  90.0 &  92.0 &  92.4 &  92.8 &  93.1 &  93.7 \\
+%\cline{1-10}
+%\multirow{3}{*}{\rotatebox{90}{Recall}} & CmBERT &  91.6 &  92.5 &  93.9 &  93.9 &  94.9 &  94.9 &  95.4 &  95.1 \\
+%       & CmBERT-ptrn &  93.6 &  94.0 &  94.4 &  94.6 &  94.6 &  94.8 &  95.0 &  94.9 \\
+%       & SpaCy NER &  88.6 &  90.4 &  90.7 &  91.7 &  91.9 &  92.8 &  93.3 &  93.4 \\
+%\end{tabular}
+%\end{table}
 
@@
 	   \center{\includegraphics[width=\textwidth]
-	       {images/experiment-1-models-performances.pdf}}
-	  \caption{\label{fig:f1-vs-trainsize} Graph view of the metrics displayed in~\cref{tab:experiment-1-models-performances}.
-As SpaCy NER is outperformed by both BERT models from the start, CmBERT slightly surpasses the pre-trained version when the number of training examples becomes large.}.
+	       {figs/eval-ner-exp1.pdf}}
+	  \caption{\label{fig:f1-vs-trainsize} Metrics measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER in experiment 1 for 8 training sets of increasing sizes.
+As SpaCy NER is outperformed by both BERT models from the start, CmBERT slightly surpasses the pre-trained version when the number of training examples becomes large.
+Values for precision and recall are available in the appendix.\bertrand{TODO!}}
 \end{figure}

### 2022-01-20 src-latex/parts/80-conclusion.tex
smaller conclusion

--- 2021-10-28 src-latex/parts/50-xp-setup.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Experimental Setup}
+\section{Conclusion}
 
-\subsection{Implementation details}
-we use spacy \mcite{spacy}
+In this article, we assessed the performance of three modern OCR systems on real historical data.
+Although Pero-OCR clearly outperforms its competitors, the qualitative analysis of its errors shows that it fails on different entries.
+This calls for leveraging both Pero-OCR and Tesseract in a complementary way to get the best out of both systems.
+Moreover, we evaluated the needs of two deep-learning-based NER approaches in terms of training data amount and text quality for pre-training.
+We showed that BERT-based NER can benefit from pre-training and fine-tuning on a noisy OCR corpus to improve its performance on a noisy OCRed test dataset.
+To perform a NER task on OCR texts, it seems therefore worthwhile to train the NER model on a corpus OCRed with the same system and not corrected.
+However, the results are not as good as for clean text: it is therefore interesting to investigate how to improve the OCR results or to correct the text in a post-processing step.
+Furthermore, it seems that the models quickly achieve good performance with very few training examples.
+Leveraging a model fine-tuned on a few examples to automatically produce much larger training sets could reveal a solution to lower the annotation burden.
+Besides, as directory entries always have the same structure - at least within a given list - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout.
+%Interesting points to discuss:
+%\begin{itemize}
+%    \item can we train on noisy data? (without manual OCR correction?) => future work? cf Pero OCR training procedure?
+%    \item do we need better OCR systems or better post-correction techniques (if NER is reliable enough)?
+%    \item Construction of the lexicon and associated cost \nathalie{là je ne comprends pas}
+%\end{itemize}
+%Ideas for future work:
+%\begin{itemize}
+%    \item Evaluate this: fine-tune Spacy NER on a few samples to generated a bootstrap dataset, feed it to BERT (cleaned or raw) to create an good model with minimal efforts ?
+\nathalie{les résultats de la table 2 ne plaident pas vraiment pour ça: même sur les très petits corpus, spacy est tours derrière...}
+%    \item Use ML to infer regexes ?
+%    \item As directories entries are always structured the same way (in a given list at least): use NER to identify entries within pages to save the burden of having to rely on the page layout ?
+%\end{itemize}

### 2022-01-20 src-latex/parts/30-dataset.tex
shorter captions for figs 6 and 7

@@
 Stephen V. Rice's thesis~\cite{santos.2019.wcmel,neudecker.2021.whdip}. The Character Error Rate (CER) is computed at
-the entry level and at the global level defined as the ratio between the number of errors
+the entry level and at the global level, defined as the ratio between the number of errors
 (insertions/deletions/substitutions) over the reference text length.

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
shorter captions for figs 6 and 7

@@
 \label{sec:ner-xp1}
-The constitution of annotated example sets to train a NER model on a target domain is a critical preliminary step.
-Often done manually, possibly from bootstrapped annotations, this task is tedious, time-consuming and error-prone.
-The ability of a model to perform well even with few training examples is a practical criteria to consider.
+The constitution of annotated datasets to train a NER model is a critical preliminary step.
+Often done manually, possibly with bootstrapped annotations, this task is tedious, time-consuming, and error-prone.
+The ability of a model to perform well even with a few training examples is a practical criterion to consider.
 In this first experiment, we investigate the NER performance of SpaCy and CamemBERT when fine-tuned with an increasing number of training examples.
@@
 %The training set is then gradually reduced in size while maintaining the relative frequency of directories within it.
-As the organisation and structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features. 
+As the structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features. 
 To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, $\approx 19\%$) from the ground truth as a test set containing entries from unseen directories.
@@
 To generate smaller training sets, we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as for the train/dev split to maintain the relative frequency of directories.
-We stop if a directory has only one entry left, or if the current training subset contains less than 30 entries maintaining the relative frequency of directories within it
-Applying this procedure to the initial training set produced 8 training subsets containing 49, 99, 199, 398, 796, 1593, 3186, and 6373 entries.
+We stop if a directory has only one entry left, or if the current training subset contains less than 30 entries, maintaining the relative frequency of directories within it
+Applying this procedure to the initial training set produces 8 training subsets containing 49, 99, 199, 398, 796, 1593, 3186, and 6373 entries.
 
@@
 	  \caption{\label{fig:f1-vs-trainsize} Metrics measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER in experiment 1 for 8 training sets of increasing sizes.
-As SpaCy NER is outperformed by both BERT models from the start, CmBERT slightly surpasses the pre-trained version when the number of training examples becomes large.
-Values for precision and recall are available in the appendix.\bertrand{TODO!}}
+Precision and recall values are available in the appendix.}
 \end{figure}

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Updates from Overleaf
shorter captions for figs 6 and 7

@@
 CmBERT and CmBERT+ptrn are fine-tuned on the training sets built from NER-\emph{reference} and NER-\emph{pero}.
+The training parameters are mostly the same as in \cref{sec:ner-xp1-protocol}, only this time the patience threshold is set to 5 evaluations.
 Finally the metrics are measured against the three tests sets.
@@
     \includegraphics[width=0.75\textwidth]{images/experiment_2_f1_with_noise_graph.pdf}
-    \caption{\label{fig:exp_2_eval_ner}F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing examples, grouped by test set.
+    \caption{\label{fig:exp_2_eval_ner}F1 scores of NER predictions in presence of OCR noise in the training and testing examples, grouped by test set.
 The dataset used to train the NER task is noted in indice after the model name (e.g.
-CmBERT$_{pero}$ means CmBERT fine-tuned on NER-\emph{pero}).
-Results show that best performances on noisy entries are obtained when the model has been pre-trained and fine-tuned for NER on examples affected with a similar OCR noise.}
+CmBERT$_{pero}$ means CmBERT fine-tuned on NER-\emph{pero}).}
 \end{figure}

### 2022-01-20 src-latex/parts/40-ocr-bench.tex
Move stuff
shorter captions for figs 6 and 7
OCR XP protocol

@@
 
-\subsection{Protocol}
+This section focuses on the evaluation of the performance of the three open-source OCR systems we selected, as described in \cref{sec:pipeline-summary}: Tesseract v4, Kraken and \peroocr.
+The dataset used to perform this evaluation is composed of the $8,765$ entries (containing $424,764$ characters) from the dataset we previously introduced.
+The single-column, cropped image of entries are used as input of each OCR system after a simple background whitening.
+As pages were previously deskewed, text is mostly horizontal except for a few cases.
+The expected output is the human transcription of these images provided in the dataset.
+Before computing the Character Error Rate (CER) for each entry, each text prediction is normalized with the same basic rules as the ones used to post-process human transcriptions: dashes, quotes, and character code for some glyphs like stars or hands are normalized.
+% Regarding the charset of each OCR, Tesseract predicts characters covering a large portion of Latin script, \peroocr generates characters with a similar range, sometimes predicting some Cyrillic characters, and Kraken is constrained by its English model to predict only ASCII characters.
 
-Dataset:
-- input: image crops for each entry
-- expected output: human transcription
-- 8765 entries used, \joseph{FIXME number of chars}
-
-Systems
-In this benchmark, we consider 3 OCR systems well known for historical document analysis: Pero OCR, Tesseract, and Kraken.
-
-\subsection{Results and discussion}
+%\subsection{Results and discussion}
 
@@
 \toprule
- & Pero-OCR  & Tesseract & Kraken \\    
+ & \peroocr  & Tesseract & Kraken \\    
 \midrule
@@
 }
-\caption{Character error rates at entry-level for Pero-OCR, Kraken and Tesseract. (a) Global CER and distribution of the CER per entry. (b)
-joint plot of the per-entry error rate showing that Pero-OCR and Tesseract do not fail on the same entries.}  
+\caption{CER at entry-level for \peroocr, Kraken and Tesseract. (a) Global CER and distribution of the CER per entry. (b)
+joint plot of the per-entry error rate showing that \peroocr and Tesseract do not fail on the same entries.}  
 \label{fig.ocr-results}
 \end{figure}
-
+%
 \Cref{fig.ocr-results} compares the performance of the OCR systems on our dataset. We can see Kraken's performance are
@@
 and so it misses French specific symbols. On the over hand, even when using a French model trained on French 19th
-documents, performance does not increase (and relaxing the character match rules does not help either). Tesseract and
-Pero-OCR are performing better on this dataset ``out-of-the-box''. With no fine-tuning, Pero-OCR gets the best accuracy
+documents, performance does not increase (and relaxing the character matching rules does not help either). Tesseract and
+\peroocr are performing better on this dataset ``out-of-the-box''. With no fine-tuning, \peroocr gets the best accuracy
 with less than 4\% character errors. Many of them are even due to a bad line detection in case of multi-lines entries
@@
 committed on the same entries (if so, all points would be on the diagonal line) and that combining the outputs of
-Pero-OCR and Tesseract would improve the overall recognition quality. Also, we will not consider Kraken in the following
-NER experiments because of its recognition rate is already low and cannot be artificially degraded in a controlled setup.
+\peroocr and Tesseract could possibly improve the overall recognition quality. 
+Also, we will not consider Kraken in the following NER experiments because its recognition rate is already too low.

### 2022-01-20 src-latex/parts/10-introduction.tex
Updates from Overleaf

@@
 % General context + scope limitation
-OCRed texts are generally not sufficient to build a high level, semantic view of a collection of historical documents.
-A subsequent stage is often needed to extract the pieces of information most likely to be searched for by users, such as named entities: people, organisations, dates, places, etc.
-Indeed, being able to properly tag text tokens unlocks the ability to relate entities, and provide colleagues from other fields with databases ready for exploitation.
+OCRed texts are generally not sufficient to build a high level semantic view of a collection of historical documents.
+A subsequent stage is often needed to extract the pieces of information most likely to be searched for by users, such as named entities: persons, organisations, dates, places, etc.
+Indeed, being able to properly tag text tokens unlocks the ability to relate entities and provide colleagues from other fields with databases ready for exploitation.
 
 Being active research topics, OCR and named entity recognition (NER) are still difficult tasks when applied to historical text documents.
-OCR approaches used for modern documents are likely to struggle even on printed historical documents due to multiple causes related to text readability (low resolution scans, inconsistent printing rules, artefacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the huge diversity of historical sources.
-On the other hand, NER approaches developed for modern texts suffer from the time gap between the contemporary entities and the entities mentioned in older texts, and also from the errors introduced into the text by the OCR step.
+OCR approaches used for modern documents are likely to struggle even on printed historical documents due to multiple causes related to text readability (low resolution scans, inconsistent printing rules, artefacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the great diversity of historical sources.
+On the other hand, the semantics of entities in NER approaches developed for modern texts may be different from those in ancient texts.
 
-In this article, we focus on a peculiar category of historical printed books: city trade directories of Paris from the XIX\textsuperscript{th} century (1801 - 1854).
-They contain hundred-pages long lists of people, along with their activity and their address, and provide valuable fine-grained knowledge to study the social dynamics of the city over time.
-Despite containing roughly the same information, the diversity in layouts, information organisation and digitising quality is challenging both for OCR and NER.
+In this article, we focus on a corpus of printed trade directories of Paris from the XIX\textsuperscript{th} century, containing hundreds of pages long lists of people with their activity and address.
+They provide fine-grained knowledge to study the social dynamics of the city over time.
+As they originate from different publishers, they show a diversity in layout, information organisation and printing quality, which adds to the poor digitising quality to make OCR and NER challenging tasks.
 
-In a recent work to locate potentially polluted urban soils \cite{bell2020automated}, trade directories were leveraged to identify and locate all the gas stations in the city of Providence during the XX\textsuperscript{th} century.
-The proposed information extraction pipeline extracts texts with OCR, identifies directories entries and recognises business type and address within each entry.
-In the same way, we aim at producing structured spatio-temporal data from Parisian trade directories.
-Because they originate from several publishers, their content is organised following different indexing methods (by name, by activity or by address), is printed in various layouts, and uses different fonts.
-We therefore investigate several state-of-the-art OCR and NER approaches to assess their usability on the corpus.
+Trade directories have been leveraged in a recent work to identify polluted urban soils \cite{bell2020automated} and locate all gas stations in the city of Providence over the last century.
+In an ongoing research project we aim at producing structured spatio-temporal data from the entries in the Paris trade directories to study the transformations of the fraction of XIX\textsuperscript{th} century Parisian society reachable through these sources.
+%Because they originate from several publishers, their content is organised following different indexing methods (by name, by activity or by address), is printed in various layouts, and uses different fonts.
+Therefor, we investigate several state-of-the-art OCR and NER approaches to assess their usability to process the corpus.
 

### 2022-01-20 src-latex/parts/30-dataset.tex
Updates from Overleaf

@@
 may vary a lot from one directory to another (see Figure \ref{fig:directories}).
+For training and testing purposes, our groundtruth dataset must be as representative as possible of the diversity of the available directories.
 
@@
 
-For training as testing purposes, our groundtruth dataset must be as representative as possible of the diversity of the available directories.
-
 For example, the Didot directory of 1854 is organised according to three indexes: by name, by activity, and by street name.
 The structure of the following entry is characteristic of its index by name: "Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47." This entry begins with a person's name, followed by the initials of their first name in parentheses.The symbol indicates that this person was awarded the Légion d'Honneur.
-The following part of this entry represents the person's activity,i.e.
-their profession, or social status, here "professor at the Conservatory of music and declamation".
-The street name and number where the person lives or carries out their activity are written at the end of the entry.
+The following part of this entry represents the person's activity,i.e., their profession, or social status, here "professor at the Conservatory of music and declamation".
+The street name and number where the person lives or carries out their activities are written at the end of the entry.
 
-These pieces of information constituting the structure of the entries are the basic data that we want to extract, deduplicate and structure in order to build a spatio-temporal database of XIX\textsuperscript{th} century Parisian inhabitants and businesses.
-With the exception of some potentially wordy activity descriptions, they correspond to named entities.
-However, as Table \ref{tab:directories} shows, while most entries contain the same types of named entities, their order and the way they are written vary from one directory or index to another.
-To provide examples of each entry structure, pages from each type have thus been annotated\footnote{In this work, the street name index entries have not been included as they are very different from the others.
+These pieces of information constituting the structure of the entries are the basic data that we want to extract, deduplicate, and structure to build a spatio-temporal database of XIX\textsuperscript{th} century Parisian inhabitants and businesses.
+Except for some potentially wordy activity descriptions, they correspond to named entities.
+However, while most entries contain the same types of named entities, their order and the way they are written vary from one directory or index to another.
+To provide examples of each entry structure, pages from each type have thus been annotated.
+%\footnote{In this work, the street name index entries have not been included as they are very different from the others.
 The strategy to be adopted to process them without loss of quality for the NER model will be the subject of future work.}.
@@
     \includegraphics[width=\linewidth]{figs/protocol.pdf}
-    \caption{Scheme of the data extraction pipeline with two quality control checkpoints. \joseph{Missing "Ground truth" for NER stage?}}
+    \caption{Data extraction pipeline with two quality control checkpoints. }
     \label{fig.pipeline}
@@
 The simplified data extraction pipeline depicted in~\cref{fig.pipeline} processes the documents presented
-in~\cref{sec:dataset}. First, the layout extraction and entry segmentation of the page are performed with a
+in~\cref{sec:dataset}. First, the page layout extraction and entry segmentation are performed with a
 semi-automated system and checked by a human. The resulting images, representing 8765 entries, are the inputs of a customizable two-stage pipeline.

### 2022-01-20 src-latex/parts/80-conclusion.tex

@@
 This calls for leveraging both Pero-OCR and Tesseract in a complementary way to get the best out of both systems.
-Moreover, we evaluated the needs of two deep-learning-based NER approaches in terms of training data amount and text quality for pre-training.
-We showed that BERT-based NER can benefit from pre-training and fine-tuning on a noisy OCR corpus to improve its performance on a noisy OCRed test dataset.
+Moreover, we evaluated the needs of two deep-learning-based NER approaches in terms of training data amount and text quality for pretraining.
+We showed that BERT-based NER can benefit from pretraining and fine-tuning on a noisy OCR corpus to improve its performance on a noisy OCRed test dataset.
 To perform a NER task on OCR texts, it seems therefore worthwhile to train the NER model on a corpus OCRed with the same system and not corrected.
-However, the results are not as good as for clean text: it is therefore interesting to investigate how to improve the OCR results or to correct the text in a post-processing step.
+However, the results are not as good as for clean text: it is therefore interesting to investigate how to improve the OCR results or to correct the text in a postprocessing step.
 Furthermore, it seems that the models quickly achieve good performance with very few training examples.
 Leveraging a model fine-tuned on a few examples to automatically produce much larger training sets could reveal a solution to lower the annotation burden.
-Besides, as directory entries always have the same structure - at least within a given list - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout.
+Besides, as directory entries always have the same structure - at least within a given index - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout.
 %Interesting points to discuss:

### 2022-01-20 src-latex/parts/30-dataset.tex
Remove prec/recall equations

@@
 precision, recall, and f-measure (or f-score) are computed considering only the exact matches between entities of the
-\emph{NER-target} and those from the \emph{NER-prediction}. Precision and recall are defined by \cref{eqn.prec-recall};
+\emph{NER-target} and those from the \emph{NER-prediction}, i.e. pairs of entries for which the type, start and end positions are exactly the same.
+Precision is the ratio of exactly matches divided by the number of predicted entries,
+and recall is defined as the ratio of exact matches divided by the number of expected entries;
 the f-measure is their harmonic mean.
 
-\begin{align}
-    \mathrm{precision} &= \frac{\text{\#exact matches}}{\text{\#entries in prediction}} & \mathrm{recall} &= \frac{\text{\#exact matches}}{\text{\#entries in reference}}
-    \label{eqn.prec-recall}
-\end{align}
-
-The evaluation metrics are illustrated on~\cref{fig.eval-ocr-ner}. The OCR text and the groundtruth are first aligned to
+The evaluation process illustrated on~\cref{fig.eval-ocr-ner}. The OCR text and the groundtruth are first aligned to
 evaluate the OCR accuracy. As there are 11 mismatches over 56 aligned characters, the CER is thus 24\%. This alignment
@@
 while only 2 over 4 entities are matched in the reference (recall). It follows an overall f-score of 0.4.
+
+\joseph{FIXME say we need to discard some entries}

### 2022-01-20 src-latex/parts/30-dataset.tex
Add mention about degenerated tags in NER metric

@@
 \begin{table}
+\caption{Entities to recognize in the dataset.}
+\label{tab.entities}
 \small
@@
 \end{tabular}
-\caption{Entities to recognize in the dataset.}
-\label{tab.entities}
 \end{table}
@@
 while only 2 over 4 entities are matched in the reference (recall). It follows an overall f-score of 0.4.
-
-\joseph{FIXME say we need to discard some entries}
+%
+A side-effect of tag projections is that is can produce degenerated tags, i.e. empty or containing only spaces.
+In such cases, the corresponding entries must be removed from the dataset for NER evaluation,
+and only the intersection of the sets of valid entries must be used to compare NER systems on multiple OCR predictions or references.

### 2022-01-20 src-latex/parts/20-related-work.tex
OCR XP protocol
Updates from Overleaf

@@
 Most of the current state-of-the-art open-source OCR systems, like Tesseract \cite{smith2007overview}, OCRopus \cite{breuel2008ocropus}, and the recent Kraken~\cite{kraken}, Calamari~\cite{wick_calamari_2020} and \peroocr \cite{kohut2021ts} are based on a pipeline of convolutional neural networks (CNNs) and long short-term memory networks (LSTM).
-Although this kind of model produces good results with modern texts, it faces several challenges with ancient texts, such as the lack of annotated data for learning, or different transcription styles in training data.
+Although this model produces good results with modern texts, it still faces challenges with ancient texts, such as the lack of annotated data for learning, or different transcription styles for training data.
 %
@@
 \subsection{Pipeline summary}
+\label{sec:pipeline-summary}
 

### 2022-01-20 src-latex/parts/30-dataset.tex
Updates from Overleaf

--- 2022-01-20 src-latex/parts/30-dataset.tex
@@
 
-These pieces of information constituting the structure of the entries are the basic data that we want to extract, deduplicate, and structure to build a spatio-temporal database of XIX\textsuperscript{th} century Parisian inhabitants and businesses.
+These pieces of information stored in the entries are the basic data that we want to extract, deduplicate, and structure to build a spatio-temporal database of XIX\textsuperscript{th} century Parisian inhabitants and businesses.
 Except for some potentially wordy activity descriptions, they correspond to named entities.
-However, while most entries contain the same types of named entities, their order and the way they are written vary from one directory or index to another.
+However, while most entries contain the same types of named entities, their order and the way they are written vary from one directory/index to another.
 To provide examples of each entry structure, pages from each type have thus been annotated.
@@
 OCR text; its output is compared to the NER groundtruth. There is only 2 over 3 tags matching in the prediction (precision),
-while only 2 over 4 entities are matched in the reference (recall). It follows an overall f-score of 0.4.
+while only 2 over 4 entities are matched in the reference (recall). It follows an overall F1-score of 0.4.

### 2022-01-20 src-latex/parts/80-conclusion.tex
Updates from Overleaf

@@
 \section{Conclusion}
-
-In this article, we assessed the performance of three modern OCR systems on real historical data.
-Although Pero-OCR clearly outperforms its competitors, the qualitative analysis of its errors shows that it fails on different entries.
-This calls for leveraging both Pero-OCR and Tesseract in a complementary way to get the best out of both systems.
-Moreover, we evaluated the needs of two deep-learning-based NER approaches in terms of training data amount and text quality for pretraining.
-We showed that BERT-based NER can benefit from pretraining and fine-tuning on a noisy OCR corpus to improve its performance on a noisy OCRed test dataset.
-To perform a NER task on OCR texts, it seems therefore worthwhile to train the NER model on a corpus OCRed with the same system and not corrected.
-However, the results are not as good as for clean text: it is therefore interesting to investigate how to improve the OCR results or to correct the text in a postprocessing step.
-Furthermore, it seems that the models quickly achieve good performance with very few training examples.
-Leveraging a model fine-tuned on a few examples to automatically produce much larger training sets could reveal a solution to lower the annotation burden.
+We assessed the performance of three modern OCR systems on real historical data.
+Although Pero-OCR clearly outperforms its competitors, the qualitative analysis of OCR errors shows that the failure cases are not the same for Pero and Tesseract.
+This calls for leveraging both OCR systems in a complementary way to get the best of the two worlds.
+%Moreover, we evaluate two deep-learning-based NER approaches in terms of their sensibility to the amount of training data and to noisy OCR texts.
+The evaluation of SpaCy NER and CamemBERT (with and without pre-training) showed that BERT-based NER can benefit from pre-training and fine-tuning on a corpus produced with the same process than the texts to annotate.
+%However, the results are not as good as for clean text: it is therefore interesting to investigate how to improve the OCR results or to correct the text in a postprocessing step.
+Furthermore, it seems that all three models achieve good performance with relatively few training examples.
+With a F1-score of 92\% with only 49 training examples, the pre-trained CamemBERT model seems to be a good choice to serve as a bootstrapping model in order to quickly produce large training sets and therefore lower the burden of creating a ground truth from scratch.
 Besides, as directory entries always have the same structure - at least within a given index - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout.

### 2022-01-20 src-latex/parts/10-introduction.tex
Typo in intro
Updates from Overleaf (#12)

@@
 
-Trade directories have been leveraged in a recent work to identify polluted urban soils \cite{bell2020automated} and locate all gas stations in the city of Providence over the last century.
-In an ongoing research project we aim at producing structured spatio-temporal data from the entries in the Paris trade directories to study the transformations of the fraction of XIX\textsuperscript{th} century Parisian society reachable through these sources.
+Trade directories have been leveraged in recent work to identify polluted urban soils \cite{bell2020automated} and locate all gas stations in the city of Providence over the last century.
+In an ongoing research project, we aim at producing structured spatio-temporal data from the entries of the Paris trade directories to study the dynamics of the fraction of the the XIX\textsuperscript{th} century Parisian society reachable through these sources.
 %Because they originate from several publishers, their content is organised following different indexing methods (by name, by activity or by address), is printed in various layouts, and uses different fonts.
-Therefor, we investigate several state-of-the-art OCR and NER approaches to assess their usability to process the corpus.
+Therefore, we investigate several state-of-the-art OCR and NER approaches to assess their usability to process the corpus.
 

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Updates from Overleaf
Overleaf 2022 01 20 2159 (#13)
* Updates from Overleaf

* Shorted conclusion

@@
 This is probably due to the fact that the texts produced by Pero-OCR feature characteristics intermediary between human transcriptions and Tesseract.
-This OCR tool removes the characters recognised with a low confidence, which is probably a great help to the NER.
+This OCR tool removes the characters recognised with low confidence, which is probably a great help to the NER.
 
@@
     \includegraphics[width=0.75\textwidth]{images/experiment_2_f1_with_noise_graph.pdf}
-    \caption{\label{fig:exp_2_eval_ner}F1 scores of NER predictions in presence of OCR noise in the training and testing examples, grouped by test set.
+    \caption{\label{fig:exp_2_eval_ner}F1-scores of NER predictions in presence of OCR noise in the training and testing examples, grouped by test set.
 The dataset used to train the NER task is noted in indice after the model name (e.g.

### 2022-01-20 src-latex/parts/20-related-work.tex
Avoid isolated lines
Comment redundant paragraph

@@
 
-In this study, we consider Tesseract version 4, Kraken and \peroocr.
-They were selected because of their ease of use (in the case of Tesseract and Kraken)
-or because of their performance (in the case of \peroocr),
-but further work should include more systems from our initial selection.
+% In this study, we consider Tesseract version 4, Kraken and \peroocr.
+% They were selected because of their ease of use (in the case of Tesseract and Kraken)
+% or because of their performance (in the case of \peroocr),
+% but further work should include more systems from our initial selection.
 
@@
 We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
+\Cref{fig.pipeline} depicts the evaluation protocol used to assess the combined performance of these OCR and NER systems.
 
@@
 CamemBERT \cite{martin-etal-2020-camembert}, is a well-known French language model derived from BERT, of which there are already trained versions for the NER task on the Hugging Face library.
-
-The Figure \ref{fig.pipeline} depicts the evaluation protocol used to assess the combined performance of these OCR and NER systems.

### 2022-01-20 src-latex/figs/protocol.tex
Reduce Fig 3 height

@@
   };
-  \matrix[matrix of nodes, nodes={draw, align=center, text width=3cm}, column sep=2cm, row sep=.5cm, row 1/.style=title] (A)
+  \matrix[matrix of nodes, nodes={draw, align=center, text width=3cm, anchor=center}, column sep=2cm, row sep=.5cm, row 1/.style=title] (A)
   {
@@
     & |(tess)| Tesseract          &   & |(cam)| CamemBERT &   \\
-    |(a)| Layout \& Entry extraction &           & |[qa] (b)|  & & |[qa] (c)|  \\
-    & |(kraken)| Kraken   &    & |(spacy)| Spacy       &    \\
-    & |(gt)| \emph{GroundTruth}   &    &        &    \\
+    |(a)| Layout \& Entry extraction & |(kraken)| Kraken           & |[qa] (b)|  & & |[qa] (c)|  \\
+    & |(gt)| \emph{GroundTruth}   &    & |(spacy)| Spacy       &    \\
   };

### 2022-01-20 src-latex/main-paper.tex
Shrink abstract and put fig in 1st page
Shrink intro
New paper structure with 1 section for each experiment
Add pero macro
Updates from Overleaf (#11)
small hack to reduce acknowledgements

@@
 % \usepackage{nth}
-% \usepackage{subcaption}
 \usepackage{booktabs}
@@
 \usepackage{tabularx} % Addition
-\usepackage{booktabs} % Addition
 \usepackage{siunitx} % Addition
 \usepackage{scalerel}
-\usepackage{cleveref}
+\usepackage[inline,shortlabels]{enumitem}
+\usepackage{xspace}
 
@@
 % \renewcommand{\baselinestretch}{0.97}
+\newcommand{\peroocr}{PERO OCR\xspace}
 
@@
 \setlength{\tabcolsep}{5pt}
-\renewcommand{\arraystretch}{1.4}
+\renewcommand{\arraystretch}{1.3}
 
@@
 % \url{http://crh.ehess.fr/index.php?5206}\\
-\joseph{Je propose de supprimer les URLs.}%
 }
@@
 \input{parts/30-dataset.tex}
-\input{parts/40-ner-approaches.tex}
-\input{parts/50-xp-setup.tex}
-\input{parts/60-results.tex}
+\input{parts/40-ocr-bench.tex}
+\input{parts/50-ner-xp1-training-set-size-and-pretraining.tex}
+\input{parts/60-ner-xp2-pretraining-and-finetuning.tex}
 \input{parts/80-conclusion.tex}
 
-\section*{Acknowledgments}
+\bigskip
+\noindent\textbf{Acknowledgments}\\
+{
+\footnotesize
 This work is supported by the French National Research Agency (ANR), as part of the SODUCO project, under Grant ANR-18-CE38-0013.
@@
 as well as G. Thomas, P. Abi Saad, R. Lelièvre, D. Mignon, T. Cavaciuti and P. Sadki for contributing to the annotation platform.
+}
 %

### 2022-01-20 src-latex/parts/30-dataset.tex
Overleaf 2022 01 20 2159 (#13)
* Updates from Overleaf

* Shorted conclusion
Fix typo

@@
 \textbf{NER Stage.} The second stage of the pipeline aims an extracting the named entities from a text with a NER
-system. This text originates from the OCR outputs in a real-word scenario but might be groundtruth text in order to
-evaluate the NER performance standalone. There are 5 type of entity to detect~(see.~\cref{tab.entities}).
+system. This text originates from the OCR outputs in a real-word scenario but might be the groundtruth text in order to
+evaluate the NER performance independently. There are 5 type of entity to detect~(see.~\cref{tab.entities}).
 The NER system has to classify non-overlapping parts of the text into one of these entity (or none of them). A second
@@
 
-Note that the dataset used to assess the NER stage is subset of the entries. Indeed, we need to ensure that the datasets
+Note that the dataset used to assess the NER stage is a subset of the entries. Indeed, we need to ensure that the datasets
 contain the same entries whichever the OCR used in the previous stage. Entries where the OCR produced an empty string

### 2022-01-20 src-latex/parts/30-dataset.tex
Updates from Overleaf

--- 2022-01-20 src-latex/parts/30-dataset.tex
@@
 
-The directories to be processed are stored in different libraries in Paris.
-They have been scanned independently and with various levels of quality.
-Moreover, they cover a wide period and have been produced by different publishers and printers.
-Therefore, their contents, indexes, layouts, methods of printing, etc.
-may vary a lot from one directory to another (see Figure \ref{fig:directories}).
-For training and testing purposes, our groundtruth dataset must be as representative as possible of the diversity of the available directories.
+The directories are available from different libraries in Paris and have been digitised independently in various levels of quality. 
+They cover a large period of time and originate from several publishers.
+Therefore, their contents, indexes, layouts, methods of printing, etc. may vary a lot from one directory to another (see Figure \ref{fig:directories}).
+We want our groundtruth dataset to be representative of the diversity of directories available on the period.
 
@@
 
-For example, the Didot directory of 1854 is organised according to three indexes: by name, by activity, and by street name.
-The structure of the following entry is characteristic of its index by name: "Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47." This entry begins with a person's name, followed by the initials of their first name in parentheses.The symbol indicates that this person was awarded the Légion d'Honneur.
-The following part of this entry represents the person's activity,i.e., their profession, or social status, here "professor at the Conservatory of music and declamation".
-The street name and number where the person lives or carries out their activities are written at the end of the entry.
+Directories contain lists of people with various information attached.
+For instance the directory published by \emph{Didot} in 1854 contains three redundant lists or people sorted by name, by activity and by street name.
+An typical example entry from this directory is "Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47.":  it begins with the person's name and surname (here in parentheses).
+The glyph denotes that this person was awarded the Légion d'Honneur.
+Then comes a descriptions of the person's activity, here his profession (professor at the Conservatory of music and declamation), but it can also be a social status.
+Such descriptions range from a single word to paragraphs describing the occupation in full detail.
+The street name and number where the person lives or carries out their activities ends the entry.
 
-These pieces of information stored in the entries are the basic data that we want to extract, deduplicate, and structure to build a spatio-temporal database of XIX\textsuperscript{th} century Parisian inhabitants and businesses.
+These are the structured data that we want to extract, deduplicate  and structure to build a spatio-temporal database.
 Except for some potentially wordy activity descriptions, they correspond to named entities.
@@
 \emph{NER-target} and those from the \emph{NER-prediction}, i.e. pairs of entries for which the type, start and end positions are exactly the same.
-Precision is the ratio of exactly matches divided by the number of predicted entries,
+Precision is the ratio of exact matches divided by the number of predicted entries,
 and recall is defined as the ratio of exact matches divided by the number of expected entries;
@@
 
-The evaluation process illustrated on~\cref{fig.eval-ocr-ner}. The OCR text and the groundtruth are first aligned to
+The evaluation process is illustrated on~\cref{fig.eval-ocr-ner}. The OCR and the groundtruth texts are first aligned to
 evaluate the OCR accuracy. As there are 11 mismatches over 56 aligned characters, the CER is thus 24\%. This alignment
@@
 %
-A side-effect of tag projections is that is can produce degenerated tags, i.e. empty or containing only spaces.
-In such cases, the corresponding entries must be removed from the dataset for NER evaluation,
-and only the intersection of the sets of valid entries must be used to compare NER systems on multiple OCR predictions or references.
+A side effect of tag projection is that it can produce degenerated tags, i.e. empty or containing only spaces.
+In such cases, the corresponding entries must be removed from the dataset,
+and only the intersection of the sets of valid entries must be used to compare NER systems with multiple OCR predictions or references.

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Updates from Overleaf

@@
 \label{sec:ner-xp1-protocol}
-The following three models form our baseline for both NER experiments.
+The following models form our baseline for both NER experiments.
 Their short names written in square brackets will be used to reference them from now on.
@@
     \item \textbf{Huggingface CamemBERT~[CmBERT]}: We rely on the implementation of BERT models provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
-We chose to reuse a CamemBERT model published on the Hugging Face repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
+We chose to reuse a CamemBERT model published on the Huggingface repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
 \item \textbf{CamemBERT pretrained on Paris directories~[CmBERT+ptrn]}: 

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex

@@
 \label{sec:ner-xp2}
-Noise introduced by OCR is known to have a negative impact on named entity recognition, because it alters the structure and lexicon of the input texts, moving them away from the language model known to the NER process.
-In real-life situations the models used are often trained on texts without such noise, even though the texts to be annotated are extracted with OCR.
-This raises the question of the most appropriate strategy to build a NER model tolerant to OCR noise.
-In this second experiment we address this question based on our benchmark.
+Noise introduced by OCR is known to have a negative impact on NER, because it alters the structure and lexicon of the input texts, moving them away from the language model known to the NER process.
+In real-life situations, the models are often trained on texts without such noise, even though the texts to be annotated are extracted with OCR.
+In this second experiment, we aim at assessing the most appropriate strategy to build a NER model tolerant to OCR noise.
 
@@
 
-Each dataset is split into a training, a development and a test set following the exact same protocol as described in \cref{sec:ner-xp1-protocol}, except this time we do not need to create smaller training sets.
-As the NER sets contain 8341 entries, the producing train sets (resp. development and test) contain 6004 entries - 72\% of the total (resp. 668 - 8\% and 1669 - 20\%).
+Each dataset is split into training development, and test sets following the same protocol as described in \cref{sec:ner-xp1-protocol}, except this time we do not need to create smaller training sets.
+As the NER sets contain 8341 entries, the produced train sets (resp. development and test) contain 6004 entries - 72\% of the total (resp. 668 - 8\% and 1669 - 20\%).
 
@@
     \includegraphics[width=0.75\textwidth]{images/experiment_2_f1_with_noise_graph.pdf}
-    \caption{\label{fig:exp_2_eval_ner}F1-scores of NER predictions in presence of OCR noise in the training and testing examples, grouped by test set.
+    \caption{\label{fig:exp_2_eval_ner}NER F1-scores in presence of OCR noise in the training and testing examples, grouped by test set.
 The dataset used to train the NER task is noted in indice after the model name (e.g.

### 2022-01-20 src-latex/parts/30-dataset.tex
French goes in italic

@@
 Directories contain lists of people with various information attached.
-For instance the directory published by \emph{Didot} in 1854 contains three redundant lists or people sorted by name, by activity and by street name.
-An typical example entry from this directory is "Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47.":  it begins with the person's name and surname (here in parentheses).
-The glyph denotes that this person was awarded the Légion d'Honneur.
+For instance the directory published by \emph{Didot} in 1854 contains three redundant lists of people sorted by name, by activity and by street name.
+An typical example entry from this directory is "\textit{Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47.}".
+It begins with the person's name and surname, here inside parentheses.
+The glyph denotes that this person was awarded the \textit{Légion d'Honneur}.
 Then comes a descriptions of the person's activity, here his profession (professor at the Conservatory of music and declamation), but it can also be a social status.

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Part 3 shorter (i guess ?)
French goes in italic

@@
     \includegraphics[width=0.75\textwidth]{images/experiment_2_f1_with_noise_graph.pdf}
-    \caption{\label{fig:exp_2_eval_ner}NER F1-scores in presence of OCR noise in the training and testing examples, grouped by test set.
-The dataset used to train the NER task is noted in indice after the model name (e.g.
+    \caption{\label{fig:exp_2_eval_ner}NER F1-scores in presence of OCR noise in the training and testing sets, grouped by test set.
+The dataset used for training is noted in indice after the model name (e.g.
 CmBERT$_{pero}$ means CmBERT fine-tuned on NER-\emph{pero}).}

### 2022-01-20 src-latex/parts/20-related-work.tex
Conclusion

@@
 
-Based on those works, we chose to test three OCR systems, namely, Tesseract, Pero OCR, and Kraken.
+Based on those works, we chose to test three OCR systems, namely, Tesseract, PERO OCR, and Kraken.
 We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
@@
 They can also be pre-trained in an unsupervised way on historical texts for domain adaptation.
-CamemBERT \cite{martin-etal-2020-camembert}, is a well-known French language model derived from BERT, of which there are already trained versions for the NER task on the Hugging Face library.
+CamemBERT \cite{martin-etal-2020-camembert}, is a well-known French language model derived from BERT, of which there are already trained versions for the NER task on the Huggingface library.

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Conclusion

@@
 % !TeX root = ../main-paper.tex
-\section{Experiment 1: NER sensibility to the number of training examples}
+\section{NER sensibility to the number of training examples}
 \label{sec:ner-xp1}
@@
 	       {figs/eval-ner-exp1.pdf}}
-	  \caption{\label{fig:f1-vs-trainsize} Metrics measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER in experiment 1 for 8 training sets of increasing sizes.
+	  \caption{\label{fig:f1-vs-trainsize} Metrics measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER for 8 training sets of increasing sizes.
 Precision and recall values are available in the appendix.}

### 2022-01-20 src-latex/parts/30-dataset.tex
French gos in italic + british-isation

@@
 \emph{block} mode for \emph{Tesseract}). The glyphs used in this dataset might be unknown to the OCR. Some of them, like
-\scalerel*{\includegraphics{./images/LH.png}}{H}, do not even have a unicode codepoint. Text normalization is thus required
+\scalerel*{\includegraphics{./images/LH.png}}{H}, do not even have a unicode codepoint. Text normalisation is thus required
 on the OCR outputs and consists in projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes
-sense), homogenize extra symbols (star and hand symbols) and alleviate some ambiguities (e.g. \emph{no} and \emph{n°}).
+sense), homogenise extra symbols (star and hand symbols) and alleviate some ambiguities (e.g. \emph{no} and \emph{n°}).
 
-At the end of this first stage, an OCR Quality Assessment (Q.A.) between the \emph{normalized} groundtruth text and the
+At the end of this first stage, an OCR Quality Assessment (Q.A.) between the \emph{normalised} groundtruth text and the
 OCR output is performed.
@@
 \begin{table}
-\caption{Entities to recognize in the dataset.}
+\caption{Entities to recognise in the dataset.}
 \label{tab.entities}
@@
 First names, initials, or civility mentions are included.
-E.g.: "Alibert (Prosper)", "Allamand frères et Hersent", "Heurtemotte (Vve)", etc.
+E.g.: \textit{Alibert (Prosper)}, \textit{Allamand frères et Hersent},\textit{Heurtemotte (Vve)}.
 \\
 TITLE & 483 & an honorary title, either text, glyphs or a combination of the two.
-E.g: "O.
-\scalerel*{\includegraphics{./images/LH.png}} {H}" for "Officier de la Légion d'Honneur" or "\scalerel*{\includegraphics{./images/GMLondon.png}}{H}" to represent the Great Medal at the London exhibition.\\
-ACT & 6472 & the profession or social status of a person summarized in the single concept of activity. E.g.: "horlogerie", "export en tous genres", "Conseiller d'Etat", "propriétaire".\\
-LOC & 9709 & mostly street names ("r. de la Jussienne"), but may also be neighborhoods ("Marais du Temple") or any indirect spatial references ("au Palais Royal").\\
-CARDINAL & 8747 & a street number, as part of an address ("16", "5 bis"), or a range of numbers (e.g. "23-25" or "5 à 9").\\
-FT & 43 & a geographic entity type, used to give more details on a location, e.g. "boutique", "atelier", "fab." or "dépôt". \\
+E.g: O.
+\scalerel*{\includegraphics{./images/LH.png}} {H} for \textit{Officier de la Légion d'Honneur} or \scalerel*{\includegraphics{./images/GMLondon.png}}{H} for the Great Medal at the London exhibition.\\
+ACT & 6472 & the profession or social status of a person summarised in the single concept of activity.
+E.g.: \textit{horlogerie}, \textit{export en tous genres}, \textit{Conseiller d'État}, \textit{propriétaire}.\\
+LOC & 9709 & mostly street names (\textit{r. de la Jussienne}), but may also be neighbourhoods (\textit{Marais du Temple}) or any indirect spatial references (\textit{au Palais Royal}).\\
+CARDINAL & 8747 & a street number, as part of an address (16, 5 bis), or a range of numbers (e.g. 23-25 or 5 \textit{à} 9).\\
+FT & 43 & a geographic entity type, used to give more details on a location, e.g. \textit{boutique}, \textit{atelier}, \textit{fab.} or \textit{dépôt}. \\
 \end{tabular}

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
French gos in italic + british-isation

@@
 \item \textbf{CamemBERT pretrained on Paris directories~[CmBERT+ptrn]}: 
-In order to evaluate whether adapting CamemBERT to the domain increases its performances, we do an unsupervised pre-training of CmBERT for next sentence prediction and masked language modeling, using \num{845000} entries randomly sampled and OCRed with Pero.
+To evaluate whether adapting CamemBERT to the domain increases its performance, we do an unsupervised pre-training of CmBERT for next sentence prediction and masked language modeling, using \num{845000} entries randomly sampled and OCRed with PERO.
 The model is trained for 3 epochs and is available online\footnote{\url{https://huggingface.co/HueyNemud/das22-10-camembert_pretrained}}.

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Conclusion
French gos in italic + british-isation

@@
 % !TeX root = ../main-paper.tex
-\section{Experiment 2: Impact of OCR noise on named entity recognition}
+\section{Impact of OCR noise on named entity recognition}
 \label{sec:ner-xp2}
@@
 However, despite a loss of performance, the model pre-trained and fine-tuned on NER-pero still gives the best results.
-This is probably due to the fact that the texts produced by Pero-OCR feature characteristics intermediary between human transcriptions and Tesseract.
+This is probably due to the fact that the texts produced by PERO-OCR feature characteristics intermediary between human transcriptions and Tesseract.
 This OCR tool removes the characters recognised with low confidence, which is probably a great help to the NER.
@@
 The dataset used for training is noted in indice after the model name (e.g.
-CmBERT$_{pero}$ means CmBERT fine-tuned on NER-\emph{pero}).}
+CmBERT$_{pero}$ for CmBERT fine-tuned on NER-\emph{pero}).}
 \end{figure}

### 2022-01-20 src-latex/parts/80-conclusion.tex
Conclusion
French gos in italic + british-isation

@@
 \section{Conclusion}
-We assessed the performance of three modern OCR systems on real historical data.
-Although Pero-OCR clearly outperforms its competitors, the qualitative analysis of OCR errors shows that the failure cases are not the same for Pero and Tesseract.
+We assessed the performances of three modern OCR systems on a set of historical sources of great interest in social history.
+Although PERO-OCR clearly outperforms its competitors, the qualitative analysis of OCR errors shows that the failure cases are not the same for PERO and Tesseract.
 This calls for leveraging both OCR systems in a complementary way to get the best of the two worlds.

### 2022-01-20 src-latex/parts/20-related-work.tex
Replace forward reference to fig with fwd ref to section.

@@
 The directory processing pipeline presented in \cite{bell2020automated} includes an OCR step, done with Tesseract, and a NER step to identify company names and addresses, performed using regular expressions.
-This section reviews existing OCR and NER approaches with historical texts and presents some works assessing the effects of OCR quality on the NER performance and the proposed solutions.
+This section reviews existing OCR and NER approaches on historical texts and presents some works assessing the effects of OCR quality on the NER performance and the proposed solutions.
 
 \subsection{Optical Character Recognition of historical texts}
-Among the large amount of OCR solutions, being either open, free or paid software, available as libraries, Python packages, binaries or cloud API, not all options seem suitable for historical document processing.
+Among the large number of OCR solutions, being either open, free, or paid software, available as libraries, Python packages, binaries, or cloud API, not all options seem suitable for historical document processing.
 We chose to avoid in our study paid and closed-source solutions.
@@
 
-Based on those works, we chose to test three OCR systems, namely, Tesseract, PERO OCR, and Kraken.
+Based on those works, we chose to test three OCR systems, namely, Tesseract, \peroocr, and Kraken.
 We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
-\Cref{fig.pipeline} depicts the evaluation protocol used to assess the combined performance of these OCR and NER systems.
+In \cref{sec:dataset}, we will explain the evaluation protocol used to assess the combined performance of these OCR and NER systems.
 
 \textbf{Tesseract} is a long-living project, born as a closed-sourced OCR at Hewlett-Packard in the eighties, it was progressively modernized, then open-sourced in 2005.
-From 2006 until November 2018 it was developed by Google, and is still very active.
-We used in our tests the version 4.1.1, released Dec.
+From 2006 until November 2018, it was developed by Google and is still very active.
+We used in our tests version 4.1.1, released Dec.
 26, 2019.
-The version 5, released on Nov.
+Version 5, released on Nov.
 30, 2021, has not been integrated in our tests yet.
@@
 
-\textbf{\peroocr} is a very recent project (started in 2020) from the Brno University of Technology, in Czech Republic.
+\textbf{\peroocr} is a very recent project (started in 2020) from the Brno University of Technology in Czech Republic.
 Their authors used many state-of-the-art techniques to train it very efficiently.
 We used the version from the master branch of their GitHub repository, updated on Sep 15, 2021.
-We used the pre-trained weights provided by the authors on the same repository, created on Oct. 9, 2020 from European texts with Latin, Greek and Cyrillic scripts.
+We used the pre-trained weights provided by the authors on the same repository, created on Oct. 9, 2020 from European texts with Latin, Greek, and Cyrillic scripts.
 % https://github.com/DCGM/pero-ocr
@@
 \textbf{SpaCy} is a software library that offers NLP components assembled in modular pipelines specialised by language.
-Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (as of january 2022).
-Hence, we rely on SpaCy's ad hoc pipeline trained on French corpora and capable of named entity recognition.
+Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (as of January 2022).
+Hence, we rely on SpaCy's ad hoc pipeline trained on French corpora and capable of Named Entity Recognition.
 % deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.
@@
 As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics.
-The classification layer relies on a finite-state machine whose transition probabilities are learned using a multi layer perceptron.
+The classification layer relies on a finite-state machine whose transition probabilities are learned using a multi-layer perceptron.
 % \bertrand{Drop the next sentences if we need space}
@@
 
-\textbf{CamemBERT}
-Language model based on \textit{Transformers} \cite{vaswani2017attention}, like BERT \cite{devlin2018bert}, have become a new paradigm for NER\cite{li2020}. 
+\textbf{CamemBERT} is an adaptation of the BERT transformer-based model for the French language\cite{vaswani2017attention,devlin2018bert}.
+Such language models have become a new paradigm for NER\cite{li2020}.
 The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer.

### 2022-01-20 src-latex/parts/30-dataset.tex
Fix bad section reference in sec. 3.

@@
 
+This section presents the historical sources that we selected and their contents.
+It also details the construction of the groundtruth dataset leveraged in our experiments and the metrics used to evaluate OCR and NER results.
+
 \subsection{A selection of Paris trade directories from 1798 to 1854}
+\label{sec:corpus}
 
@@
 Therefore, their contents, indexes, layouts, methods of printing, etc. may vary a lot from one directory to another (see Figure \ref{fig:directories}).
-We want our groundtruth dataset to be representative of the diversity of directories available on the period.
+We want our groundtruth dataset to be representative of the diversity of directories available in the period.
 
@@
 Directories contain lists of people with various information attached.
-For instance the directory published by \emph{Didot} in 1854 contains three redundant lists of people sorted by name, by activity and by street name.
+For instance, the directory published by \emph{Didot} in 1854 contains three redundant lists of people sorted by name, by activity, and by street name.
 An typical example entry from this directory is "\textit{Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47.}".
@@
 The glyph denotes that this person was awarded the \textit{Légion d'Honneur}.
-Then comes a descriptions of the person's activity, here his profession (professor at the Conservatory of music and declamation), but it can also be a social status.
+Then comes a description of the person's activity, here his profession (professor at the Conservatory of music and declamation), but it can also be a social status.
 Such descriptions range from a single word to paragraphs describing the occupation in full detail.
@@
 
-These are the structured data that we want to extract, deduplicate  and structure to build a spatio-temporal database.
+These are the pieces of information that we want to extract, deduplicate  and structure to build a spatio-temporal database.
 Except for some potentially wordy activity descriptions, they correspond to named entities.
@@
 The simplified data extraction pipeline depicted in~\cref{fig.pipeline} processes the documents presented
-in~\cref{sec:dataset}. First, the page layout extraction and entry segmentation are performed with a
+in~\cref{sec:corpus}. First, the page layout extraction and entry segmentation are performed with a
 semi-automated system and checked by a human. The resulting images, representing 8765 entries, are the inputs of a customizable two-stage pipeline.
@@
 \emph{block} mode for \emph{Tesseract}). The glyphs used in this dataset might be unknown to the OCR. Some of them, like
-\scalerel*{\includegraphics{./images/LH.png}}{H}, do not even have a unicode codepoint. Text normalisation is thus required
+\scalerel*{\includegraphics{./images/LH.png}}{H}, do not even have a Unicode codepoint. Text normalisation is thus required
 on the OCR outputs and consists in projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes
-sense), homogenise extra symbols (star and hand symbols) and alleviate some ambiguities (e.g. \emph{no} and \emph{n°}).
+sense), homogenise extra symbols (star and hand symbols), and alleviate some ambiguities (e.g. \emph{no} and \emph{n°}).
 
@@
 %2. Entrée/Sortie
-\textbf{NER Stage.} The second stage of the pipeline aims an extracting the named entities from a text with a NER
+\textbf{NER Stage.} The second stage of the pipeline aims at extracting the named entities from a text with a NER
 system. This text originates from the OCR outputs in a real-word scenario but might be the groundtruth text in order to
-evaluate the NER performance independently. There are 5 type of entity to detect~(see.~\cref{tab.entities}).
-The NER system has to classify non-overlapping parts of the text into one of these entity (or none of them). A second
-Q.A. (namely NER Q.A.) is performed at the end of this stage between the groundtruth and the NER output.
+evaluate the NER performance independently. There are 5 types of entity to detect~(see.~\cref{tab.entities}).
+The NER system has to classify non-overlapping parts of the text into one of these entities (or none of them). A second
+Q.A. (namely, NER Q.A.) is performed at the end of this stage between the groundtruth and the NER output.
 
@@
 and those for which no entity could be projected from the groundtruth have to be ignored. We filter the set of entries by
-keeping only the entries that are always valid at the end of the stage 1. At the cost of losing 5\% of the references
-entries, we thus have 8341 valid entries at the input of the stage 2.
+keeping only the entries that are always valid at the end of stage 1. At the cost of losing 5\% of the references
+entries, we thus have 8341 valid entries at the input of stage 2.
 
@@
 %
-A side effect of tag projection is that it can produce degenerated tags, i.e. empty or containing only spaces.
+A side effect of tag projection is that it can produce degenerated tags, i.e., empty or containing only spaces.
 In such cases, the corresponding entries must be removed from the dataset,

### 2022-01-20 src-latex/parts/30-dataset.tex
Typos

@@
 Directories contain lists of people with various information attached.
-For instance, the directory published by \emph{Didot} in 1854 contains three redundant lists of people sorted by name, by activity, and by street name.
-An typical example entry from this directory is "\textit{Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47.}".
+For instance, the directory published by \emph{Didot} in 1854 contains three redundant lists of people sorted by name, by activity and by street name.
+A typical example entry from this directory is "\textit{Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47.}".
 It begins with the person's name and surname, here inside parentheses.
@@
     \includegraphics[width=\linewidth]{figs/protocol.pdf}
-    \caption{Data extraction pipeline with two quality control checkpoints. }
+    \caption{Data extraction pipeline with two quality control checkpoints. The
+    NER Q.A. checkpoint may either assess the NER system lonely (the dashed
+    \emph{groundtruth} path) or may evaluate the joint work of a NER
+    system with an OCR system.}
     \label{fig.pipeline}

### 2022-01-20 src-latex/main-supplementary-material.tex
Move some ref to supp. material
Minor comment

@@
 \subsection{Content}
+
+The Table \ref{tab:directories} lists the directories selected, their layout, the structure of their entries, and the number of annotated entries according to the considered index.
 \begin{table}[h!]
@@
     \hline
-    \multicolumn{1}{|c|}{\textbf{Directory publisher or name}} & \multicolumn{1}{c|}{\textbf{Year}} & \multicolumn{1}{c|}{\textbf{\# Col}} & \multicolumn{1}{c|}{\textbf{Index}} & \multicolumn{1}{c|}{\textbf{Entry structure}} & \multicolumn{1}{c|}{\textbf{\# Ann}}
+    \multicolumn{1}{|c|}{\textbf{Directory publisher or name}} & \multicolumn{1}{c|}{\textbf{Year}} & \multicolumn{1}{c|}{\textbf{\# Col}} & \multicolumn{1}{c|}{\textbf{Index}} & \multicolumn{1}{c|}{\textbf{General entry structure}} & \multicolumn{1}{c|}{\textbf{\# Ann}}
     \\ \hline

### 2022-01-20 src-latex/parts/10-introduction.tex
small fix in 2.3

@@
 
-The contributions of this article are the following.
+The contributions of this article are as follows:
 % Contributions
 \begin{enumerate*}[(i)]
-    \item We review state of the art OCR and NER systems for historical documents (\cref{sec:related-work}).
+    \item We review state-of-the-art OCR and NER systems for historical documents (\cref{sec:related-work}).
     \item We introduce a new dataset suitable for OCR and NER evaluation (\cref{sec:dataset}).

### 2022-01-20 src-latex/parts/20-related-work.tex
Fix US spelling.
small fix in 2.3

@@
 
-Based on those works, we chose to test three OCR systems, namely, Tesseract, \peroocr, and Kraken.
+Based on those works, we chose to test three OCR systems, namely, Tesseract, PERO OCR, and Kraken.
 We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
-In \cref{sec:dataset}, we will explain the evaluation protocol used to assess the combined performance of these OCR and NER systems.
+\Cref{fig.pipeline} depicts the evaluation protocol used to assess the combined performance of these OCR and NER systems.
 

### 2022-01-20 src-latex/parts/30-dataset.tex
small fix in 2.3

--- 2022-01-20 src-latex/parts/30-dataset.tex
@@
 
-This section presents the historical sources that we selected and their contents.
-It also details the construction of the groundtruth dataset leveraged in our experiments and the metrics used to evaluate OCR and NER results.
+This section presents the historical sources that we selected and their contents. It also details the construction of the groundtruth dataset leveraged in our experiments and the metrics used.
 
 \subsection{A selection of Paris trade directories from 1798 to 1854}
-\label{sec:corpus}
 
@@
 The simplified data extraction pipeline depicted in~\cref{fig.pipeline} processes the documents presented
-in~\cref{sec:corpus}. First, the page layout extraction and entry segmentation are performed with a
+in~\cref{sec:dataset}. First, the page layout extraction and entry segmentation are performed with a
 semi-automated system and checked by a human. The resulting images, representing 8765 entries, are the inputs of a customizable two-stage pipeline.

### 2022-01-20 src-latex/parts/40-ocr-bench.tex
Small corrections everywhere

@@
 The dataset used to perform this evaluation is composed of the $8,765$ entries (containing $424,764$ characters) from the dataset we previously introduced.
-The single-column, cropped image of entries are used as input of each OCR system after a simple background whitening.
-As pages were previously deskewed, text is mostly horizontal except for a few cases.
+The single-column, cropped images of entries are used as input of each OCR system after a simple background whitening.
+As the pages were previously deskewed, the text is mostly horizontal except for a few cases.
 The expected output is the human transcription of these images provided in the dataset.
-Before computing the Character Error Rate (CER) for each entry, each text prediction is normalized with the same basic rules as the ones used to post-process human transcriptions: dashes, quotes, and character code for some glyphs like stars or hands are normalized.
+Before computing the Character Error Rate (CER) for each entry, each text prediction is normalized with the same basic rules as the ones used to post-process human transcription: dashes, quotes, and character codes for some glyphs like stars or hands are normalized.
 % Regarding the charset of each OCR, Tesseract predicts characters covering a large portion of Latin script, \peroocr generates characters with a similar range, sometimes predicting some Cyrillic characters, and Kraken is constrained by its English model to predict only ASCII characters.
@@
 not as good as the two first OCR. This is partially due to the fact that the closest available model is for English text
-and so it misses French specific symbols. On the over hand, even when using a French model trained on French 19th
-documents, performance does not increase (and relaxing the character matching rules does not help either). Tesseract and
+and so it misses French specific symbols. On the other hand, even when using a French model trained on French 19th
+documents, the performance does not increase (and relaxing the character matching rules does not help either). Tesseract and
 \peroocr are performing better on this dataset ``out-of-the-box''. With no fine-tuning, \peroocr gets the best accuracy
@@
 and are not related to the OCR system itself. \Cref{fig.ocr-results} (b) shows that errors from the two best OCR are not
-committed on the same entries (if so, all points would be on the diagonal line) and that combining the outputs of
-\peroocr and Tesseract could possibly improve the overall recognition quality. 
-Also, we will not consider Kraken in the following NER experiments because its recognition rate is already too low.
+committed to the same entries (if so, all points would be on the diagonal line) and that combining the outputs of
+\peroocr and Tesseract could improve the overall recognition quality. 
+In addition, we will not consider Kraken in the following NER experiments because its recognition rate is already too low.

### 2022-01-20 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Small corrections everywhere

@@
 \begin{itemize}
-    \item \textbf{SpaCy NER pipeline for French~[SpaCy NER]}: We use the pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the french corpora deep-sequoia and wikiner-fr.
+    \item \textbf{SpaCy NER pipeline for French~[SpaCy NER]}: We use the pipeline \textit{fr\_core\_news\_lg} provided by SpaCy v.3.2.1\mcite{spacy}, already trained for NER on the French corpora deep-sequoia and wikiner-fr.
     We stress again that we use the CNN version of this pipeline, not the transformer-based available in SpaCy v3.
@@
 \subsection{Results and discussion}
-Figure \ref{fig:f1-vs-trainsize} displays the averaged precision, recall and F1-score for all models on the 8 subsets created from the groundtruth.
-CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.
+Figure \ref{fig:f1-vs-trainsize} displays the averaged precision, recall, and F1-score for all models on the 8 subsets created from the groundtruth.
+CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: the performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.
 The F1 score increases by 4.6 points between 49 and 796 examples for CmBERT (resp. 1.6 for CmBERT+ptrn and 5.1 for SpaCy NER) but only by 1 point between 796 and 6373 examples (resp. 0.6 and 1.4).
-The models derived from CamemBERT always outperform the Spacy model.
+The models derived from CamemBERT always outperform the SpaCy model.
 
-It appears that pre-training the CamemBERT model on OCR texts seems worth it only when a the training set used to fine-tune the NER layer is small.
+It appears that pre-training the CamemBERT model on OCR text seems worth it only when a the training set used to fine-tune the NER layer is small.
 This effect might be due to the differences in nature between the training subsets, whose texts are manually corrected, and the noisy OCR texts used to pretrain CamemBERT.
-Indeed, the learned embeddings from the pret-raining are specialised on noisy texts and therefore less adapted to clean texts.
-The pre-training aims at adapting the model to the vocabulary of the domain and to the errors caused by the OCR, which reveals not helpful and even counter-productive when the texts do not contain these types of errors.
+Indeed, the learned embeddings from pre-training are specialised to noisy texts and therefore less adapted to clean text.
+The pre-training aims at adapting the model to the vocabulary of the domain and to the errors caused by the OCR, which reveals not helpful and even counterproductive when the texts do not contain these types of errors.
 

### 2022-01-20 src-latex/parts/80-conclusion.tex
Small corrections everywhere

@@
 % !TeX root = ../main-paper.tex
-\section{Conclusion}
-We assessed the performances of three modern OCR systems on a set of historical sources of great interest in social history.
+\section{Conclusion and future works}
+We assessed the performance of three modern OCR systems on a set of historical sources of great interest in social history.
 Although PERO-OCR clearly outperforms its competitors, the qualitative analysis of OCR errors shows that the failure cases are not the same for PERO and Tesseract.
@@
 Furthermore, it seems that all three models achieve good performance with relatively few training examples.
-With a F1-score of 92\% with only 49 training examples, the pre-trained CamemBERT model seems to be a good choice to serve as a bootstrapping model in order to quickly produce large training sets and therefore lower the burden of creating a ground truth from scratch.
+With a F1-score of 92\% with only 49 training examples, the pre-trained CamemBERT model seems to be a good choice to serve as a bootstrapping model to quickly produce large training sets and therefore lower the burden of creating a ground truth from scratch.
 Besides, as directory entries always have the same structure - at least within a given index - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout.

### 2022-01-20 src-latex/figs/protocol.tex
Fix Fig 3 entry block + extend its caption.

--- 2022-01-17 src-latex/figs/protocol.tex
@@
   };
-  \matrix[matrix of nodes, nodes={draw, align=center, text width=3cm}, column sep=2cm, row sep=.5cm, row 1/.style=title] (A)
+  \matrix[matrix of nodes, nodes={draw, align=center, text width=3cm, anchor=center}, column sep=2cm, row sep=.3cm, row 1/.style=title] (A)
   {
@@
     & |(tess)| Tesseract          &   & |(cam)| CamemBERT &   \\
-    |(a)| Layout \& Entry extraction &           & |[qa] (b)|  & & |[qa] (c)|  \\
-    & |(kraken)| Kraken   &    & |(spacy)| Spacy       &    \\
-    & |(gt)| \emph{GroundTruth}   &    &        &    \\
+    |(a)| Single-block Entry Image & |(kraken)| Kraken           & |[qa] (b)|  & & |[qa] (c)|  \\
+    & |(gt)| \emph{GroundTruth}   &    & |(spacy)| Spacy       &    \\
   };

### 2022-01-20 src-latex/parts/30-dataset.tex
Small corrections everywhere
Fix Fig 3 entry block + extend its caption.
Fix wrong commit

--- 2022-01-20 src-latex/parts/30-dataset.tex
@@
 For instance, the directory published by \emph{Didot} in 1854 contains three redundant lists of people sorted by name, by activity and by street name.
-A typical example entry from this directory is "\textit{Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47.}".
+A typical example entry from this directory is ``\textit{Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47.}''.
 It begins with the person's name and surname, here inside parentheses.
@@
 
-%\joseph{TODO: separate entry detection, OCR and NER stages explicitely}
-
 The simplified data extraction pipeline depicted in~\cref{fig.pipeline} processes the documents presented
@@
 always a single block. Thus, the most adapted mode is chosen when the OCR system allows for the detection mode (e.g. the
-\emph{block} mode for \emph{Tesseract}). The glyphs used in this dataset might be unknown to the OCR. Some of them, like
-\scalerel*{\includegraphics{./images/LH.png}}{H}, do not even have a Unicode codepoint. Text normalisation is thus required
-on the OCR outputs and consists in projecting Unicode characters onto the latin-1 \emph{charset} (whenever it makes
-sense), homogenise extra symbols (star and hand symbols), and alleviate some ambiguities (e.g. \emph{no} and \emph{n°}).
-
+\emph{block} mode for \emph{Tesseract}). 
+Some glyphs used in this dataset might be unknown to the OCR, and some like
+\scalerel*{\includegraphics{./images/LH.png}}{H} do not even have a Unicode codepoint, and were annotated using Unicode Private User Area 1.
+Furthermore, as some annotations guidelines where unclear to human annotators, some projections rules were applied: whitespace, dash, dots and a couple of commonly confused characters where projected to well-defined codepoints.
+The same normalization was applied to OCR predictions.
 At the end of this first stage, an OCR Quality Assessment (Q.A.) between the \emph{normalised} groundtruth text and the
-OCR output is performed.
-
-%\joseph{[DONE] a word about the charset, and existence of glyphs unknown to the OCR}
-%\joseph{[DONE] parler de l'étage d'extraction d'entrées qui est validé manuellement, et des quelques règles de simplification du charset pour homogénéiser entre les annotateurs / pallier des ambiguïtés dans les règles d'annotation (qu'on pourrait encore améliorer)}
+OCR output is performed on the basis of the $8,765$ entries which were manually controlled, totalling $424,764$ characters (including $54,387$ spacing characters).
+Entries are $49.0$-characters long on average.
 
@@
 contain the same entries whichever the OCR used in the previous stage. Entries where the OCR produced an empty string
-and those for which no entity could be projected from the groundtruth have to be ignored. We filter the set of entries by
-keeping only the entries that are always valid at the end of stage 1. At the cost of losing 5\% of the references
-entries, we thus have 8341 valid entries at the input of stage 2.
+and those for which no entity could be projected from the groundtruth have to be ignored. 
+We filter the set of entries by keeping only the entries that are always valid at the end of the stage 1.
+%At the cost of losing 5\% of the references entries, we thus have 8341 valid entries at the input of the stage 2.
+% /!\ True only for \peroocr and Tesseract predictions, not considering Kraken (we loose 500 more entries)
+Therefore, the $8,765$ reference entries were manually annotated with $34,242$ entities; entries contain $3.9$ entities on average.
+Projecting reference tagged entities on OCR predictions resulted in a variable loss of entries.
+For \peroocr, $8,392$ valid entries were generated, for Tesseract $8,700$ and for Kraken $7,990$.
+The resulting intersection of the sets of valid entries contained $7,725$ entries for the tree OCR systems (and the reference),
+or $8,341$ entries if we consider \peroocr and Tesseract only.
 
@@
 
-% It can therefore be used to train and improve the entry detection, OCR and NER tools.
-% book, page, box, text_ocr, ner_xml (+uid within a page)
-
-%\joseph{TODO intégrer dans les sections avant ?}
 %
-%
-% valid ner (ref \& pero \& tess): 8341
-%
-%TODO average entry len (in char)
 %TODO average number of tags/entities per entry
@@
 %\end{itemize*}
-
-%\joseph{TODO expliquer qu'il faut faire la fusion des sous-datasets NER}
-
-%\subsection{Focus on how NER GT was aligned on noisy OCR}
 
@@
 (insertions/deletions/substitutions) over the reference text length.
+Word Error Rate is hard to define for our tokens and was not considered.
 
@@
 %
-A side effect of tag projection is that it can produce degenerated tags, i.e., empty or containing only spaces.
-In such cases, the corresponding entries must be removed from the dataset,
-and only the intersection of the sets of valid entries must be used to compare NER systems with multiple OCR predictions or references.
+% A side effect of tag projection is that it can produce degenerated tags, i.e. empty or containing only spaces.
+% In such cases, the corresponding entries must be removed from the dataset,
+% and only the intersection of the sets of valid entries must be used to compare NER systems with multiple OCR predictions or references.

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Add small precision about valid entries in XP2
Updates from Overleaf

@@
 We leverage the labeled sets of entries NER-\emph{reference}, NER-\emph{pero} and NER-\emph{tesseract} created as explained in \cref{sec:dataset}.
-Kraken is left aside as it produces poor results that would require removing 500 entries from all datasets to keep the same set of valid entries.
-
 Each dataset is split into training development, and test sets following the same protocol as described in \cref{sec:ner-xp1-protocol}, except this time we do not need to create smaller training sets.
-As the NER sets contain 8341 entries, the produced train sets (resp. development and test) contain 6004 entries - 72\% of the total (resp. 668 - 8\% and 1669 - 20\%).
+As the NER sets contain 8341 entries (see \cref{sec:dataset-for-eval}), the produced train sets (resp. development and test) contain 6004 entries - 72\% of the total (resp. 668 - 8\% and 1669 - 20\%).
 

### 2022-01-20 src-latex/parts/40-ocr-bench.tex
Fix Sec 4 - OCR + enlarge font in figure 5.

@@
 The dataset used to perform this evaluation is composed of the $8,765$ entries (containing $424,764$ characters) from the dataset we previously introduced.
-The single-column, cropped images of entries are used as input of each OCR system after a simple background whitening.
+The single-column, cropped images of entries are used as input of each OCR system.
 As the pages were previously deskewed, the text is mostly horizontal except for a few cases.
@@
 and are not related to the OCR system itself. \Cref{fig.ocr-results} (b) shows that errors from the two best OCR are not
-committed to the same entries (if so, all points would be on the diagonal line) and that combining the outputs of
+committed on the same entries (if so, all points would be on the diagonal line) and that combining the outputs of
 \peroocr and Tesseract could improve the overall recognition quality. 

### 2022-01-20 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Remove note on Kraken in 6.1 to avoid repeating section 4

@@
 \subsection{Training and evaluation protocol}
-Only CmBERT and CmBERT+ptrn are considered since experiment 1 shows that SpaCy NER is systematically outperformed by these two models.
+Only CmBERT and CmBERT+ptrn are considered since the first experiment shows that SpaCy NER is outperformed by these two models for all sizes of training sets.
 

### 2022-01-20 src-latex/parts/20-related-work.tex
Updates from Overleaf

--- 2022-01-20 src-latex/parts/20-related-work.tex
@@
 As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics.
-The classification layer relies on a finite-state machine whose transition probabilities are learned using a multi-layer perceptron.
+The classification layer relies on a finite-state machine whose transition probabilities are learnt using a multi-layer perceptron.
 % \bertrand{Drop the next sentences if we need space}
@@
 
-\textbf{CamemBERT} is an adaptation of the BERT transformer-based model for the French language\cite{vaswani2017attention,devlin2018bert}.
+\textbf{CamemBERT} is well-known adaptation of the BERT transformer-based model for the French language\cite{vaswani2017attention,devlin2018bert,martin-etal-2020-camembert}.
 Such language models have become a new paradigm for NER\cite{li2020}.
-The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer.
-They can also be pre-trained in an unsupervised way on historical texts for domain adaptation.
-CamemBERT \cite{martin-etal-2020-camembert}, is a well-known French language model derived from BERT, of which there are already trained versions for the NER task on the Huggingface library.
+The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually refered as a "head".
+They can also be pre-trained in an unsupervised way on large amount of unlabeled texts for domain adaptation.

### 2022-01-21 src-latex/parts/80-conclusion.tex
conclusion minor change

@@
 We assessed the performance of three modern OCR systems on a set of historical sources of great interest in social history.
-Although PERO-OCR clearly outperforms its competitors, the qualitative analysis of OCR errors shows that the failure cases are not the same for PERO and Tesseract.
+Although \peroocr clearly outperforms its competitors, the qualitative analysis of OCR errors shows that its failure cases are not the same as Tesseract.
 This calls for leveraging both OCR systems in a complementary way to get the best of the two worlds.
@@
 With a F1-score of 92\% with only 49 training examples, the pre-trained CamemBERT model seems to be a good choice to serve as a bootstrapping model to quickly produce large training sets and therefore lower the burden of creating a ground truth from scratch.
-Besides, as directory entries always have the same structure - at least within a given index - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout.
+Besides, as directory entries always have the same structure - at least within a given index - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout only.
+%, or, leveraging the low amount of training data required, interactively generate per-index NER models.
 %Interesting points to discuss:

### 2022-01-21 src-latex/parts/30-dataset.tex
Add small precision about valid entries in XP2

@@
 
-These are the pieces of information that we want to extract, deduplicate  and structure to build a spatio-temporal database.
+These are the pieces of information we want to extract, deduplicate and structure to build a spatio-temporal database.
 Except for some potentially wordy activity descriptions, they correspond to named entities.
@@
 \subsection{A dataset for OCR and NER evaluation}
+\label{sec:dataset-for-eval}
 
@@
 Furthermore, as some annotations guidelines where unclear to human annotators, some projections rules were applied: whitespace, dash, dots and a couple of commonly confused characters where projected to well-defined codepoints.
-The same normalization was applied to OCR predictions.
+The same normalisation was applied to OCR predictions.
 At the end of this first stage, an OCR Quality Assessment (Q.A.) between the \emph{normalised} groundtruth text and the

### 2022-01-21 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex

@@
 We chose to reuse a CamemBERT model published on the Huggingface repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
-\item \textbf{CamemBERT pretrained on Paris directories~[CmBERT+ptrn]}: 
-To evaluate whether adapting CamemBERT to the domain increases its performance, we do an unsupervised pre-training of CmBERT for next sentence prediction and masked language modeling, using \num{845000} entries randomly sampled and OCRed with PERO.
+\item \textbf{CamemBERT pre-trained on Paris directories~[CmBERT+ptrn]}: 
+To evaluate whether adapting CamemBERT to the domain increases its performance, we do an unsupervised pre-training of CmBERT for next sentence prediction and masked language modeling, using approx.
+$845,000$ entries randomly sampled and OCRed with PERO.
 The model is trained for 3 epochs and is available online\footnote{\url{https://huggingface.co/HueyNemud/das22-10-camembert_pretrained}}.
@@
 As the structure of entries varies across directories, the models may learn to overfit on a subset of directories with specific features. 
-To reduce the evaluation bias, we start by leaving out 3 directories (1690 entries, $\approx 19\%$) from the ground truth as a test set containing entries from unseen directories.
+To reduce the evaluation bias, we start by leaving out 3 directories ($1,690$ entries, $\approx 19\%$) from the ground truth as a test set containing entries from unseen directories.
 
-Then, a stratified sampling based on the source directory of each entry is run on the remainder to create a training set (6373 entries, $\approx 73\%$ of the gold reference) and a development set (709 entries, $\approx 8\%$).
+Then, a stratified sampling based on the source directory of each entry is run on the remainder to create a training set ($6,373$ entries, $\approx 73\%$ of the gold reference) and a development set (709 entries, $\approx 8\%$).
 The development set is used to evaluate the model during the training phase.
@@
 The three models are fine-tuned on the NER task 5 times using each of the 8 training subsets, with an early stopping criterion based on the number of training steps without improvement of the F1-score.
-This patience threshold is set to 1600 steps for SpaCy NER and 3 evaluations (1 evaluation every 100 steps) for CmBERT and CmBERT+ptrn.
+This patience threshold is set to $1,600$ steps for SpaCy NER and 3 evaluations (1 evaluation every 100 steps) for CmBERT and CmBERT+ptrn.
 The metrics are measured for the 24 resulting NER models on the common test set and averaged over the runs.

### 2022-01-21 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex

@@
 Each dataset is split into training development, and test sets following the same protocol as described in \cref{sec:ner-xp1-protocol}, except this time we do not need to create smaller training sets.
-As the NER sets contain 8341 entries (see \cref{sec:dataset-for-eval}), the produced train sets (resp. development and test) contain 6004 entries - 72\% of the total (resp. 668 - 8\% and 1669 - 20\%).
+As the NER sets contain $8,341$ entries (see \cref{sec:dataset-for-eval}), the produced train sets (resp.
+development and test) hold $6,004$ entries - 72\% of the total (resp.
+668 - 8\% and $1,669$ - 20\%).
 

### 2022-01-21 src-latex/parts/40-ocr-bench.tex
YES !

@@
 The expected output is the human transcription of these images provided in the dataset.
-Before computing the Character Error Rate (CER) for each entry, each text prediction is normalized with the same basic rules as the ones used to post-process human transcription: dashes, quotes, and character codes for some glyphs like stars or hands are normalized.
+Before computing the Character Error Rate (CER) for each entry, each text prediction is normalised with the same basic rules as the ones used to post-process human transcription: dashes, quotes and character codes for glyphs like stars or hands are normalised.
 % Regarding the charset of each OCR, Tesseract predicts characters covering a large portion of Latin script, \peroocr generates characters with a similar range, sometimes predicting some Cyrillic characters, and Kraken is constrained by its English model to predict only ASCII characters.

### 2022-01-21 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
YES !

@@
     \item \textbf{Huggingface CamemBERT~[CmBERT]}: We rely on the implementation of BERT models provided by the software library Huggingface (transformers v.4.15.0, datasets v.1.17.0).
-We chose to reuse a CamemBERT model published on the Huggingface repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and already trained for NER on wikiner-fr.
+We chose to reuse a CamemBERT model published on the Huggingface repository \footnote{\url{https://huggingface.co/Jean-Baptiste/camembert-ner}} and trained for NER on wikiner-fr.
 \item \textbf{CamemBERT pre-trained on Paris directories~[CmBERT+ptrn]}: 
@@
 	       {figs/eval-ner-exp1.pdf}}
-	  \caption{\label{fig:f1-vs-trainsize} Metrics measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER for 8 training sets of increasing sizes.
-Precision and recall values are available in the appendix.}
+	  \caption{\label{fig:f1-vs-trainsize} Metrics measured on the fine-tuned models CmBERT, CmBERT+ptrn and SpaCy NER for 8 training sets of increasing sizes.}
 \end{figure}

### 2022-01-21 src-latex/parts/80-conclusion.tex
YES !

@@
 Furthermore, it seems that all three models achieve good performance with relatively few training examples.
-With a F1-score of 92\% with only 49 training examples, the pre-trained CamemBERT model seems to be a good choice to serve as a bootstrapping model to quickly produce large training sets and therefore lower the burden of creating a ground truth from scratch.
-Besides, as directory entries always have the same structure - at least within a given index - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout only.
-%, or, leveraging the low amount of training data required, interactively generate per-index NER models.
+With a F1-score of 92\% with only 49 training examples, the pre-trained CamemBERT model is a good choice to serve as a bootstrapping model to quickly produce large training sets and therefore lower the burden of creating a ground truth from scratch.
+Besides, as directory entries always have the same structure - at least within a given index - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout only, or even interactively generate per-index NER models to take advantage of the low amount of training examples required.
 %Interesting points to discuss:

### 2022-01-21 src-latex/parts/80-conclusion.tex
example > sample

@@
 With a F1-score of 92\% with only 49 training examples, the pre-trained CamemBERT model is a good choice to serve as a bootstrapping model to quickly produce large training sets and therefore lower the burden of creating a ground truth from scratch.
-Besides, as directory entries always have the same structure - at least within a given index - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout only, or even interactively generate per-index NER models to take advantage of the low amount of training examples required.
+Besides, as directory entries always have the same structure - at least within a given index - we could take advantage of NER results and some simple rules to identify entries within pages instead of relying on the page layout only, or even interactively generate per-index NER models to take advantage of the low amount of training samples required.
 %Interesting points to discuss:

### 2022-01-21 src-latex/parts/20-related-work.tex
Make Language Tools happy

@@
 Such language models have become a new paradigm for NER\cite{li2020}.
-The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually refered as a "head".
+The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually referred as a "head".
 They can also be pre-trained in an unsupervised way on large amount of unlabeled texts for domain adaptation.

### 2022-01-21 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
YES !
Make Language Tools happy

@@
 Only CmBERT and CmBERT+ptrn are considered since the first experiment shows that SpaCy NER is outperformed by these two models for all sizes of training sets.
-
-We leverage the labeled sets of entries NER-\emph{reference}, NER-\emph{pero} and NER-\emph{tesseract} created as explained in \cref{sec:dataset}.
+We leverage the labelled sets of entries NER-\emph{reference}, NER-\emph{pero} and NER-\emph{tesseract} created as explained in \cref{sec:dataset}.
 Each dataset is split into training development, and test sets following the same protocol as described in \cref{sec:ner-xp1-protocol}, except this time we do not need to create smaller training sets.
 As the NER sets contain $8,341$ entries (see \cref{sec:dataset-for-eval}), the produced train sets (resp.
-development and test) hold $6,004$ entries - 72\% of the total (resp.
+development and test) count $6,004$ entries - 72\% of the total (resp.
 668 - 8\% and $1,669$ - 20\%).
-
 CmBERT and CmBERT+ptrn are fine-tuned on the training sets built from NER-\emph{reference} and NER-\emph{pero}.
@@
 However, despite a loss of performance, the model pre-trained and fine-tuned on NER-pero still gives the best results.
-This is probably due to the fact that the texts produced by PERO-OCR feature characteristics intermediary between human transcriptions and Tesseract.
+This is probably due to the fact that the texts produced by \peroocr feature characteristics intermediary between human transcriptions and Tesseract.
 This OCR tool removes the characters recognised with low confidence, which is probably a great help to the NER.

### 2022-01-21 src-latex/parts/80-conclusion.tex
Make Language Tools happy

@@
 %Moreover, we evaluate two deep-learning-based NER approaches in terms of their sensibility to the amount of training data and to noisy OCR texts.
-The evaluation of SpaCy NER and CamemBERT (with and without pre-training) showed that BERT-based NER can benefit from pre-training and fine-tuning on a corpus produced with the same process than the texts to annotate.
+The evaluation of SpaCy NER and CamemBERT (with and without pre-training) showed that BERT-based NER can benefit from pre-training and fine-tuning on a corpus produced with the same process as the texts to annotate.
 %However, the results are not as good as for clean text: it is therefore interesting to investigate how to improve the OCR results or to correct the text in a postprocessing step.

### 2022-01-21 src/ocr/OCRMetrics.py
OCR pipeline (#4)
* Add TODO, enable PR with non-empty branch

* Add isri binding tools

* WIP add stub for ner gt sync tool

* Add TAG adding.

* Add comments

* Unicode tests and notes

* More tests

* Tools for NER GT tags alignment almost ready

* Update ner_sync_ref.py

Fix.

* doc

* more tests

* Add accuracy tools

* Update demo

* wip new code

* Add accuracy summary report.

* wip rewrite of text preparation tools

* new normalization and alignment functions

* generate pretraining set

* add dataset!

* generate pretraining

* more charset adjustments

* Fix some dataset entries (ch

@@ -1,0 +1,99 @@
+'''
+Module for Text Alignment and OCR Metrics
+'''
+
+
+
+import isri_tools
+from typing import List, Union
+from dataclasses import dataclass
+
+@dataclass
+class AlignmentStats:
+    """Provide statistics about an aligment
+    """
+    errors: int = 0
+    matches: int = 0
+    reference_length: int = 0
+    x_length: int = 0
+
+    @property
+    def length(self):
+        return self.errors + self.matches
+
+    @property
+    def precision(self) -> float:
+        """Get the alignment recall (ratio match/reference length)
+
+        Returns:
+            float: Precision score
+        """
+        return self.matches / self.x_length
+    
+    @property
+    def recall(self) -> float:
+        '''
+        Get the alignment recall (ratio match/reference length)
+
+        Returns:
+            float: Recall score
+        '''
+        return self.matches / self.reference_length
+
+    @property
+    def accuracy(self) -> float:
+        '''
+        Get the alignment accuracy (number of match over the alignment length)
+        '''
+        return self.matches / self.length
+
+    
+    @property
+    def CER(self) -> float:
+        '''
+        Get the alignment error rare (number of errors over the reference length)
+        https://towardsdatascience.com/evaluating-ocr-output-quality-with-character-error-rate-cer-and-word-error-rate-wer-853175297510#5aec
+        '''
+        return self.errors / self.reference_length
+
+    @property
+    def CERnorm(self) -> float:
+        '''
+        Get the alignment error rare (number of errors over the reference length)
+        https://towardsdatascience.com/evaluating-ocr-output-quality-with-character-error-rate-cer-and-word-error-rate-wer-853175297510#5aec
+        https://sites.google.com/site/textdigitisation/qualitymeasures/computingerrorrates
+        '''
+        return self.errors / self.length
+
+
+@dataclass
+class AlignmentReport:
+    '''
+    Provide an alignment report between a reference string and another string
+    '''
+    reference: str
+    x: str
+    aligned_ref: str
+    aligned_x: str
+
+    def __init__(self, reference: str, x: str) -> None:
+        self.reference = reference
+        self.x = x
+        self.aligned_ref, self.aligned_x = isri_tools.align(reference, x, "⌴")
+
+        match = 0
+        for a,b in zip(self.aligned_ref, self.aligned_x):
+            match += int(a == b)
+
+        n = len(self.aligned_ref)
+        self._stats = AlignmentStats(matches=match,
[…]

### 2022-01-21 src/ocr/isri_tools_src/Modules/accrpt.h

@@ -1,0 +1,104 @@
+/**********************************************************************
+ *
+ *  accrpt.h
+ *
+ *  This module provides support for reading and writing character
+ *  accuracy reports.  The contents of one of these reports is
+ *  represented by an "Accdata" structure.
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#ifndef _ACCRPT_
+#define _ACCRPT_
+
+#include "charclass.h"
+#include "table.h"
+
+typedef
+struct
+{
+    long ins;		/* number of insertions */
+    long subst;		/* number of substitutions */
+    long del;		/* number of deletions */
+    long errors;	/* number of errors = ins + subst + del */
+} Accops;
+
+typedef
+struct
+{
+    long count;		/* number of ground-truth characters in this class */
+    long missed;	/* number of these that were misrecognized */
+} Accclass;
+
+BEGIN_ENTRY(Conf)
+    long errors;	/* number of errors caused by this confusion */
+    long marked;	/* number of these that were marked */
+END_ENTRY(Conf);
+
+BEGIN_TABLE_OF(Conf, Conflist)
+END_TABLE(Conftable);	/* table of confusions */
+
+typedef
+struct
+{
+    long characters;	/* number of ground-truth characters */
+    long errors;	/* number of errors made */
+    long reject_characters;
+			/* number of reject characters generated */
+    long suspect_markers;
+			/* number of characters marked as suspect */
+    long false_marks;	/* number of false marks */
+    Accops marked_ops;	/* edit operations for marked errors */
+    Accops unmarked_ops;/* edit operations for unmarked errors */
+    Accops total_ops;	/* edit operations for all errors */
[…]

### 2022-01-21 src/ocr/isri_tools_src/Modules/charclass.c

@@ -1,0 +1,169 @@
+/**********************************************************************
+ *
+ *  charclass.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include "charclass.h"
+
+static short num_classes;
+static char *class_name[MAX_CHARCLASSES];
+static Charclass class[NUM_CHARVALUES];
+
+static struct range
+{
+    Charvalue start, stop;
+    char *name;
+} range[] =
+{
+    {0x0000, 0x0000, "Unassigned"},
+    {0x0000, 0x0008, "ASCII Control Codes"},
+    {0x0009, 0x000D, "ASCII Spacing Characters"},
+    {0x000E, 0x001F, "ASCII Control Codes"},
+    {0x0020, 0x0020, "ASCII Spacing Characters"},
+    {0x0021, 0x002F, "ASCII Special Symbols"},
+    {0x0030, 0x0039, "ASCII Digits"},
+    {0x003A, 0x0040, "ASCII Special Symbols"},
+    {0x0041, 0x005A, "ASCII Uppercase Letters"},
+    {0x005B, 0x0060, "ASCII Special Symbols"},
+    {0x0061, 0x007A, "ASCII Lowercase Letters"},
+    {0x007B, 0x007E, "ASCII Special Symbols"},
[…]

### 2022-01-21 src/ocr/isri_tools_src/Modules/list.c

@@ -1,0 +1,149 @@
+/**********************************************************************
+ *
+ *  list.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include "list.h"
+#include "util.h"
+
+BEGIN_ITEM(Item)
+END_ITEM(Item);
+
+BEGIN_LIST_OF(Item)
+END_LIST(List);
+
+/**********************************************************************/
+
+void list_initialize(void *list_)
+{
+  List* list = list_;
+  list->first = list->last = NULL;
+  list->array = NULL;
+  list->count = 0;
+}
+/**********************************************************************/
+
+static void free_array(list)
+List *list;
+{
+    if (list->array)
+    {
+	free(list->array);
+	list->array = NULL;
+    }
+}
+/**********************************************************************/
+
+static void list_insert(list, ref1, item, ref2)
+List *list;
+Item *ref1, *item, *ref2;
+{
+    item->prev = ref1;
+    item->next = ref2;
+    if (ref1)
+	ref1->next = item;
+    else
+	list->first = item;
+    if (ref2)
+	ref2->prev = item;
+    else
+	list->last = item;
+    free_array(list);
+    list->count++;
+}
+/**********************************************************************/
+
+void list_insert_first(list, item)
+List *list;
+Item *item;
+{
+    list_insert(list, NULL, item, list->first);
+}
+/**********************************************************************/
+
+void list_insert_last(list, item)
+List *list;
+Item *item;
+{
+    list_insert(list, list->last, item, NULL);
+}
+/**********************************************************************/
+
+void list_insert_before(list, item, ref)
+List *list;
+Item *item, *ref;
+{
+    list_insert(list, ref->prev, item, ref);
[…]

### 2022-01-21 src/ocr/isri_tools_src/Modules/sync.h

@@ -1,0 +1,78 @@
+/**********************************************************************
+ *
+ *  sync.h
+ *
+ *  This module provides support for string synchronization.  There are
+ *  two algorithms based on the detection of long common substrings:
+ *  "synchronize" can match two or more text streams but does not find
+ *  transposed matches, while "transpose_sync" locates transposed
+ *  matches but can be applied to only two text streams.  A third
+ *  algorithm, "fastukk_sync", is based on an algorithm by Ukkonen, and
+ *  finds an optimal match of two text streams using cost function
+ *  (1,1,1).
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#ifndef _SYNC_
+#define _SYNC_
+
+#include "text.h"
+
+typedef
+struct
+{
+    long start;		/* index of the first character of the substring */
+    long stop;		/* index of the last character of the substring */
+    long length;	/* length of the substring in characters */
+} Substr;		/* describes a substring of a text value by specifying
+			   its indices within the "array" representation of the
+			   text value */
+
+BEGIN_ITEM(Sync)
+    Substr *substr;	/* array describing one substring for each text value */
+    long *match;	/* if non-NULL, then the substrings have been matched
+			   and this points to the match number; otherwise, the
+			   substrings are unmatched */
+END_ITEM(Sync);
+
+BEGIN_LIST_OF(Sync)
+END_LIST(Synclist);	/* list of matched and unmatched substrings */
+
+void synchronize(/* Synclist *synclist, short num_text, Text *text */);
+			/* given "num_text" streams of text, synchronizes the
+			   streams and stores the results in "synclist"; each
+			   item in the list points to an array of "num_text"
+			   substrings */
+
[…]

### 2022-01-21 src/ocr/isri_tools_src/Modules/table.c

@@ -1,0 +1,122 @@
+/**********************************************************************
+ *
+ *  table.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include "table.h"
+#include "util.h"
+
+BEGIN_ENTRY(Entry)
+END_ENTRY(Entry);
+
+BEGIN_TABLE_OF(Entry, Entrylist)
+END_TABLE(Table);
+
+/**********************************************************************/
+
+void table_initialize(table)
+Table *table;
+{
+    short i;
+    for (i = 0; i < TABLE_SIZE; i++)
+	list_initialize(&table->list[i]);
+    table->array = NULL;
+    table->count = 0;
+}
+/**********************************************************************/
+
+static short table_index(key)
+unsigned char *key;
+{
+    long i, sum = 0;
+    for (i = 0; key[i]; i++)
+	sum += key[i];
+    return(sum % TABLE_SIZE);
+}
+/**********************************************************************/
+
+void *table_lookup(table, key)
+Table *table;
+char *key;
+{
+    Entry *entry;
+    for (entry = table->list[table_index(key)].first; entry &&
+    strcmp(key, entry->key) != 0; entry = entry->next);
+    return(entry);
+}
+/**********************************************************************/
+
+static void free_array(table)
+Table *table;
+{
+    if (table->array)
+    {
+	free(table->array);
+	table->array = NULL;
+    }
+}
+/**********************************************************************/
+
+void table_insert(table, entry)
+Table *table;
+Entry *entry;
+{
+    list_insert_first(&table->list[table_index(entry->key)], entry);
+    free_array(table);
+    table->count++;
+}
[…]

### 2022-01-21 src/ocr/isri_tools_src/Modules/text.h

@@ -1,0 +1,108 @@
+/**********************************************************************
+ *
+ *  text.h
+ *
+ *  This module provides definitions and routines to support the reading
+ *  and writing of OCR-generated text.  A "Text" structure is defined to
+ *  be a linked list of "Char" structures, where each "Char" structure
+ *  gives the value of one 16-bit character and indicates whether it
+ *  is suspect.
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#ifndef _TEXT_
+#define _TEXT_
+
+#include "list.h"
+#include "util.h"
+
+#define BLANK             ' '
+#define NEWLINE           '\n'
+#define REJECT_CHARACTER  '~'
+#define SUSPECT_MARKER    '^'
+
+typedef unsigned short Charvalue;
+#define NUM_CHARVALUES  65536
+
+BEGIN_ITEM(Char)
+    Boolean suspect;
+    Charvalue value;
+END_ITEM(Char);
+
+BEGIN_LIST_OF(Char)
+END_LIST(Text);
+
+void append_char(Text *text, Boolean suspect, Charvalue value);
+			/* appends the given character to "text" */
+
+typedef
+struct
+{
+    Boolean find_header;/* if True, a header will be looked for and skipped if
+			   present; if False and a header is present, it will be
+			   stored as text */
+    Boolean find_markers;
+			/* if True, any occurrence of the "suspect_marker"
+			   character will be interpreted as marking the
+			   following character as suspect */
+    int suspect_marker; /* applicable when "find_markers" is True; if zero,
+			   SUSPECT_MARKER will be used */
+    Boolean find_hex_values;
+			/* if True, any occurrence of a hex value of the form
+			   <FF> or <FFFF> will be converted to the corresponding
[…]

### 2022-01-21 src/ocr/isri_tools_src/Modules/util.c

@@ -1,0 +1,272 @@
+/**********************************************************************
+ *
+ *  util.c
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#include <stdlib.h>
+#include <signal.h>
+
+#ifdef unix
+#include <sys/types.h>
+#include <sys/stat.h>
+#else
+#include <sys\types.h>
+#include <sys\stat.h>
+#endif
+
+#include "util.h"
+
+char *exec_name;
+
+Boolean usage_when_no_args = True;
+void (*usage_routine)();
+
+void (*cleanup_routine)();
+
+int errstatus = 1;
+
+/**********************************************************************/
+
+void *allocate(number, size)
+size_t number, size;
+{
+    void *p;
+    p = calloc(number, size);
+    if (!p)
+	error("unable to allocate memory", Exit);
+    return(p);
+}
+/**********************************************************************/
+
+int ustrcmp(s1, s2)
+unsigned char *s1, *s2;
+{
+    long i;
+    for (i = 0; s1[i] && s1[i] == s2[i]; i++);
+    if (s1[i] || s2[i])
+	return(s1[i] < s2[i] ? -1 : 1);
+    else
+	return(0);
+}
+/**********************************************************************/
+
+FILE *open_file(filename, mode)
+char *filename, *mode;
+{
+    FILE *f;
+    if (!filename)
+	return(mode[0] == 'r' ? stdin : stdout);
+    f = fopen(filename, mode);
+    if (f)
+	return(f);
+    if (mode[0] == 'w')
+	error_string("unable to create file", filename, Exit);
+    else
+	error_string("unable to open file", filename, Exit);
+    return NULL;
+}
+/**********************************************************************/
+
[…]

### 2022-01-21 src/ocr/isri_tools_src/Modules/util.h

@@ -1,0 +1,139 @@
+/**********************************************************************
+ *
+ *  util.h
+ *
+ *  This module contains basic definitions and utility routines that
+ *  are needed by almost every module/program in the OCR Experimental
+ *  Environment.
+ *
+ *  Author: Stephen V. Rice
+ *  
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+#ifndef _UTIL_
+#define _UTIL_
+
+#include <malloc.h>
+#include <stdio.h>
+#include <string.h>
+
+#ifndef True
+typedef char Boolean;
+#define True   1
+#define False  0
+#endif
+
+#ifndef max
+#define max(a, b)  ((a) > (b) ? (a) : (b))
+#define min(a, b)  ((a) < (b) ? (a) : (b))
+#endif
+
+#define NEW(type)  ((type *) allocate((size_t) 1, sizeof(type)))
+			/* allocate one instance of "type" */
+#define NEW_ARRAY(number, type) \
+                   ((type *) allocate((size_t) (number), sizeof(type)))
+			/* allocate an array of "type" */
+void *allocate(/* size_t number, size_t size */);
+
+int ustrcmp(/* unsigned char *s1, unsigned char *s2 */);
+			/* compares strings like "strcmp" but treats characters
+			   as unsigned */
+
+FILE *open_file(/* char *filename, char *mode */);
+			/* opens the named file in the specified mode; reports
+			   an error and quits if unable to open the file; if
+			   "filename" is NULL, returns stdin or stdout,
+			   depending on the mode */
+
+void close_file(/* FILE *f */);
+			/* closes the specified file */
+
+Boolean file_exists(/* char *filename */);
+			/* returns True if the named file exists */
+
[…]

### 2022-01-21 src/ocr/isri_tools_src/accuracy.cpp

@@ -1,0 +1,286 @@
+
+/**********************************************************************
+ *
+ *  accuracy.c
+ *
+ *  Author: Stephen V. Rice
+ *
+ * Copyright 1996 The Board of Regents of the Nevada System of Higher
+ * Education, on behalf, of the University of Nevada, Las Vegas,
+ * Information Science Research Institute
+ *
+ * Licensed under the Apache License, Version 2.0 (the "License"); you
+ * may not use this file except in compliance with the License.  You
+ * may obtain a copy of the License at
+ *
+ *    http://www.apache.org/licenses/LICENSE-2.0
+ *
+ * Unless required by applicable law or agreed to in writing, software
+ * distributed under the License is distributed on an "AS IS" BASIS,
+ * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or
+ * implied. See the License for the specific language governing
+ * permissions and limitations under the License.
+ *
+ **********************************************************************/
+
+extern "C"
+{
+#include "Modules/accrpt.h"
+#include "Modules/sync.h"
+}
+
+#include <pybind11/iostream.h>
+#include <pybind11/pybind11.h>
+#include <pybind11/stl.h>
+#include <string>
+#include <fstream>
+
+namespace py = pybind11;
+
+namespace
+{
+constexpr int MAX_DISPLAY = 24;
+
+void
+make_key (Text *texts, char *key, Sync *sync)
+{
+  long i, j;
+  char buffer[2][MAX_DISPLAY + 4], string[STRING_SIZE];
+  for (i = 0; i < 2; i++)
+    {
+      buffer[i][0] = '\0';
+      for (j = sync->substr[i].start; j <= sync->substr[i].stop; j++)
+        {
+          char_to_string (False, texts[i].array[j]->value, string, True);
+          if (strlen (buffer[i]) + strlen (string) > MAX_DISPLAY)
+            {
+              strcat (buffer[i], "...");
+              break;
+            }
+          strcat (buffer[i], string);
+        }
+    }
+  sprintf (key, "{%s}-{%s}\n", buffer[0], buffer[1]);
+}
+/**********************************************************************/
+
+void
+add_ops (Accops *sum_ops, Accops *ops)
+{
+  sum_ops->ins += ops->ins;
+  sum_ops->subst += ops->subst;
+  sum_ops->del += ops->del;
+  sum_ops->errors += ops->errors;
+}
+
[…]

### 2022-01-21 src/ocr/isri_tools_src/align.cpp

@@ -1,0 +1,109 @@
+extern "C"
+{
+  #include "Modules/text.h"
+  #include "Modules/sync.h"
+}
+
+#include <string>
+#include <vector>
+#include <pybind11/pybind11.h>
+#include <pybind11/stl.h>
+
+
+namespace
+{
+
+  void text_from_string(Text* text, std::wstring a)
+  {
+    for (auto c: a)
+      append_char(text, /* suspect = */false, c);
+  }
+
+  std::wstring text_to_string(Text* text)
+  {
+    std::wstring w;
+    w.reserve(text->count);
+    for (auto* item = text->first; item; item = item->next)
+      w.push_back(item->value);
+    return w;
+  }
+
+  int compute_mapping(Synclist* synclist, int* A, int* B)
+  {
+
+    int i = 0;
+    for (Sync* sync = synclist->first; sync; sync = sync->next)
+    {
+      int ma = sync->substr[0].length;
+      int mb = sync->substr[1].length;
+      int m = std::max(ma, mb);
+      for (int k = 0; k < ma; ++k)
+        A[sync->substr[0].start + k] = i + k;
+      for (int k = 0; k < mb; ++k)
+        B[sync->substr[1].start + k] = i + k;
+      i += m;
+    }
+    return i;
+  }
+
+int compute_align_map(std::wstring a, std::wstring b, std::vector<int>& map_A, std::vector<int>& map_B)
+{
+  Text A;
+  Text B;
+  list_initialize(&A);
+  list_initialize(&B);
+  text_from_string(&A, a);  // 16 bit fixed encoding only for now
+  text_from_string(&B, b);  // 16 bit fixed encoding only for now
+
+  map_A.resize(a.size());
+  map_B.resize(b.size());
+
+  int alignment_size;
+  {
+    Text texts[2] = {A, B};
+
+    Synclist synclist;
+    fastukk_sync(&synclist, texts); // Actual call to original UNLV/ISRI tools
+    alignment_size = compute_mapping(&synclist, map_A.data(), map_B.data());
+  }
+
+  list_empty(&A, free);
+  list_empty(&B, free);
+
+  return alignment_size;
+}
+}
+
+std::pair<std::vector<int>, std::vector<int>> get_align_map(std::wstring a, std::wstring b)
+{
+  std::vector<int> map_A, map_B;
+  compute_align_map(a, b, map_A, map_B);
+
[…]

### 2022-01-21 src/ocr/isri_tools_src/main.cpp

@@ -1,0 +1,10 @@
+#include <pybind11/pybind11.h>
+
+void init_align(pybind11::module* m);
+void init_accuracy(pybind11::module& m);
+
+PYBIND11_MODULE(isri_tools, m) {
+    m.doc() = "ISRI Analytic tools"; // optional module docstring
+    init_align(&m);
+    init_accuracy(m);
+}

### 2022-01-21 src/ocr/setup.py

@@ -1,0 +1,33 @@
+from setuptools import Extension, setup
+from pybind11.setup_helpers import Pybind11Extension
+
+sourcefiles = [
+    "isri_tools_src/align.cpp",
+    "isri_tools_src/accuracy.cpp",
+    "isri_tools_src/main.cpp",
+    "isri_tools_src/Modules/accrpt.c",
+    "isri_tools_src/Modules/charclass.c",
+    "isri_tools_src/Modules/ci.c",
+    "isri_tools_src/Modules/dist.c",
+    "isri_tools_src/Modules/edorpt.c",
+    "isri_tools_src/Modules/list.c",
+    "isri_tools_src/Modules/sort.c",
+    "isri_tools_src/Modules/stopword.c",
+    "isri_tools_src/Modules/sync.c",
+    "isri_tools_src/Modules/table.c",
+    "isri_tools_src/Modules/text.c",
+    "isri_tools_src/Modules/unicode.c",
+    "isri_tools_src/Modules/util.c",
+    "isri_tools_src/Modules/wacrpt.c",
+    "isri_tools_src/Modules/word.c",
+    ]
+
+
+
+ext = Pybind11Extension("isri_tools", sourcefiles)
+
+setup(
+    name="isri_tools",
+    version="20210115_1751",
+    ext_modules = [ext]
+)

### 2022-01-21 src/ocr/test_text_utils.py

@@ -1,0 +1,214 @@
+
+from text_utils import (
+    UNICODE_SIMPLIFICATIONS_SINGLE_CHAR,
+    charset_stats,
+    ocr_simplifications_for_evaluation, 
+    simplify_unicode_charset,
+    check_alignment_charset,
+    add_tags_prediction,
+    xml_contains_empty_tags,
+    xml_remove_empty_tags,
+    fix_manual_ner_xml)
+
+test_str = ("\U0001F449,AB\u000D\u000Aab "
+     + "".join(chr(x) for x in UNICODE_SIMPLIFICATIONS_SINGLE_CHAR.keys()) 
+     + "\u0001\u0000-\u0009\u000B -\u001F\u007F \u0080-\u009F \uFE00-\uFE0F "
+     + " \uD800-\uDFFF \uFDD0-\uFDEF\uFFFE\uFFFF 123")
+    
+def test_check_alignment_charset():
+    assert(not check_alignment_charset(test_str))
+
+def test_simplify_unicode_charset():
+    assert(check_alignment_charset(simplify_unicode_charset(test_str)))
+
+def test_str_rstrip_removes_all_spaces():
+    s = "abc\u0009 \u00A0 \u2000 \u2001 \u2002 \u2003 \u2004 \u2005 \u2006 \u2007 \u2008 \u2009 \u200A \u202F \u205F"
+    assert(s.rstrip() == "abc")
+
+def test_str_rstrip_removes_all_newlines():
+    s = "abc \u000A \u000B \u000C \u000D \u0085 \u2028 \u2029"
+    assert(s.rstrip() == "abc")
+
+def test_add_tags_prediction():
+    A = "<PER>Anthony</PER> \uE001, <FT>fab.</FT> du <ACT>pêche</ACT>, <LOC>Châtelet</LOC>xx"
+    B = "\u261E Mme Antoine, \uE00A fab la p&cheur et \ndu ch>telet et du Faub. St Antoine"
+    R = add_tags_prediction(A, B)
+    assert(R == "\u261E Mme <PER>Antoine</PER>, \uE00A <FT>fab</FT> la <ACT>p&amp;che</ACT>ur et \u2029du <LOC>ch&gt;telet</LOC> et du Faub. St Antoine")
+
+    A = "<PER>Anthony</PER> \uE001, <FT>fab</FT><FT>.</FT> du <ACT>pêche</ACT>, <LOC>Châtelet</LOC>xx"
[…]

### 2022-01-21 src/ocr/text_utils.py

@@ -1,0 +1,921 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+
+# ==============================================================================
+# Imports
+from collections import Counter
+import io
+import logging
+from typing import List, Tuple, Iterable
+import unicodedata
+from warnings import warn
+
+import isri_tools
+import numpy as np
+import pandas as pd
+import regex
+
+# ==============================================================================
+# Logging
+logger = logging.getLogger(__name__)
+
+
+
+# ==============================================================================
+# Some constants
+TAG_LIST = ["PER", "LOC", "ACT", "CARDINAL", "FT", "TITRE"]
+
+_newline_chars = "\u000A\u000B\u000C\u000D\u0085\u2028\u2029"
+_space_chars = "\u0020\u0009\u00A0\u2000\u2001\u2002\u2003\u2004\u2005\u2006\u2007\u2008\u2009\u200A\u202F\u205F"
+# _empty_chars = ""  # Not managed here
+
+REGEX_EMPTY_TAGS = (
+    r"<(\L<tag>)>(["
+         + _newline_chars 
+         + _space_chars 
+         + r"]*)</(?1)>")
+
+WHITESPACE_ONLY_STR_RE = "^[" + _newline_chars + _space_chars + "]*$"
+
+# ==============================================================================
+# Annotation codes projection to Unicode chars
+# Expected input: raw OCR human transcription and annotation (special glyphs w/o public codepoints)
+# Output: OCR reference text with projected annotations (before normalization)
+# Transform type: (multiple chars) -> (1+ chars), can change string length
+
+# From https://hackmd.io/ocGrovCAQZuKAoGs1zTYIg#Liste-des-glyphes-et-de-leur-repr%C3%A9sentation-en-caract%C3%A8res
+ANNOTATION_CODES = [
+    # We use unicode private user area when needed (no dedicated unicode char)
+    ("::LH::", "\uE001"),  # PUA #1 
[…]

### 2022-01-27 src-latex/parts/20-related-work.tex

--- 2022-01-20 src-latex/parts/20-related-work.tex
@@
 % !TeX root = ../main-paper.tex
-\section{OCR and NER on historical texts}
-\label{sec:related-work}
+\section{Named Entity Recognition in Historical Texts}
 
-The directory processing pipeline presented in \cite{bell2020automated} includes an OCR step, done with Tesseract, and a NER step to identify company names and addresses, performed using regular expressions.
-This section reviews existing OCR and NER approaches on historical texts and presents some works assessing the effects of OCR quality on the NER performance and the proposed solutions.
+The directory processing pipeline presented in \cite{bell2020automated} to locate old gas stations using mid-20th century directories includes an OCR step, done with Tesseract, and a NER step to identify company names and addresses within the extracted text, performed using rules built as regular expressions.
+%They look for lexical and layout patterns to retrieve portions of text that refer to street names, numbers, and company names.
 
-\subsection{Optical Character Recognition of historical texts}
-Among the large number of OCR solutions, being either open, free, or paid software, available as libraries, Python packages, binaries, or cloud API, not all options seem suitable for historical document processing.
-We chose to avoid in our study paid and closed-source solutions.
-This notably discards Transkribus~\cite{transkribus}, which relies on the commercial system ABBYY's Finereader as well as on two handwritten transcription engines, to process text.
-% for Transkribus HTR engines, there is one closed source and one apparently open source
-% we should test the open source one: https://github.com/jpuigcerver/pylaia
+Many approaches have been designed to recognize named entities, ranging from handcrafted rules to supervised approaches \cite{nadeau2007}.
+Rule based approaches look for portions of text that match patterns based on word characteristics such as their case, whether they are digits or not, their part-of-speech tags, etc.
+These patterns can be fully handcrafted like in \cite{bell2020automated} or generated by data mining approaches like in \cite{nouvel2011}.
+Rule based approaches can also combine a set of patterns with a set of dictionaries (gazetteers, author lists, etc.) that help recognizing named entities when an exhaustive lexicon of searched entities is available \cite{mansouri2008,maurel2011}.
+For example, the CasEN\footnote{\url{https://tln.lifat.univ-tours.fr/version-francaise/ressources/casen}} transducer cascade leverages syntatic-lexical patterns and dictionnaries to recognize Named Entities in French texts.
+Such kind of approaches achieve very good results when applied to specialized domain corpus - like directories - and when exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
 
-Most of the current state-of-the-art open-source OCR systems, like Tesseract \cite{smith2007overview}, OCRopus \cite{breuel2008ocropus}, and the recent Kraken~\cite{kraken}, Calamari~\cite{wick_calamari_2020} and \peroocr \cite{kohut2021ts} are based on a pipeline of convolutional neural networks (CNNs) and long short-term memory networks (LSTM).
-Although this model produces good results with modern texts, it still faces challenges with ancient texts, such as the lack of annotated data for learning, or different transcription styles for training data.
-%
-% \cite{martinek2019hybrid} propose an approach to generate synthetic annotated text for historical OCR training, based on manually collected characters from historical text images.
-% This work proposes to train an OCR system based on a CNN-LSTM network with synthetic data and then to fine-tune the model with some pages of real historical annotated text.
-% The results show that this approach gives state-of-the-art results.
+Supervised approaches include both approaches implementing supervised learning algorithms with careful text feature engineering, and deep learning based approaches which automatically build their own features to classify words into one category of named entities or another.
+In recent years, the latter have grown significantly, surpassing the performance of rule-based systems and feature-based supervised approaches and establishing new baselines \cite{li2020}.
 
-To overcome the limitations due to different transcription styles in training data, \peroocr adds a Transcription Style Block layer to a classical model based on a CNN and a Recurrent Neural Network components \cite{kohut2021ts}.
-This block takes the image of the text and a Transcription Style Identifier as inputs and helps the network decide what kind of transcription style to use as output.
-
-% In this study, we consider Tesseract version 4, Kraken and \peroocr.
-% They were selected because of their ease of use (in the case of Tesseract and Kraken)
-% or because of their performance (in the case of \peroocr),
-% but further work should include more systems from our initial selection.
-
-% Finally don't mention OCR-D as it relies on either Tesseract, Kraken, Ocropus or Calamari
-
-% PERO refs
-% O Kodym, M Hradiš: Page Layout Analysis System for Unconstrained Historic Documents. ICDAR, 2021.
-% M Kišš, K Beneš, M Hradiš: AT-ST: Self-Training Adaptation Strategy for OCR in Domains with Limited Transcriptions. ICDAR, 2021.
-% J Kohút, M Hradiš: TS-Net: OCR Trained to Switch Between Text Transcription Styles. ICDAR, 2021.
-
-\subsection{Named Entity Recognition in historical texts}
-\label{subsection:stoa-ner-on-historical-texts}
-
-Many approaches have been designed to recognise named entities, ranging from handcrafted rules to supervised approaches \cite{nadeau2007}.
-Rule based approaches look for portions of the text that match patterns like in \cite{bell2020automated,nouvel2011} or dictionary (gazetteers, author lists, etc.) entries like in \cite{mansouri2008,maurel2011}.
-Such kind of approaches achieve very good results when applied to a specialised domain corpus and when an exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
-
-Supervised approaches include both approaches implementing supervised learning algorithms with careful text feature engineering, and deep learning based approaches which automatically build their own features to classify tokens into named entity categories.
-In recent years, the latter have grown dramatically, yielding state-of-the-art performances as shown in the recent survey proposed by \cite{li2020}.
-This survey concludes that fine-tuning general-purpose contextualised language models with domain-specific data is very likely to give good performance for use cases with domain-specific texts and few training data.
-This strategy has been adopted by \cite{Labusch2020NamedED} to extract named entities in OCRed historical texts in German, French, and English.
-However, the NER performance drops significantly as the quality of the OCR decreases and is correlated with its decrease.
-
-Several recent studies have focused on the impact of OCR quality on NER results.
-Most of the time, they have evaluated NER approaches based on deep learning architectures as they seem to adapt more easily to OCR errors than rule-based or more classical supervised approaches.
-\cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by SpaCy \cite{spacy} library to perform NER on a corpus made of many journal articles with different levels of OCR errors.
-For each OCRed article, a ground truth text is available so that the Word Error Rate (WER) can be computed.
-The performance of the NER model with respect to OCR quality is eventually assessed by computing the F1-score for each NER class and each article, i.e., each WER value.
-\cite{hamdi2020assessing} performed a similar but more extensive evaluation on four supervised NER models: CoreNLP using Conditional Random Fields and three deep neural models, BLSTM-CNN, BLSTM-CRF, and BLSTM-CNN-CRF.
-They tested them on CoNLL-02 and CoNLL-03 NER benchmark corpora, degraded by applying four different types of OCR noise.
-Overall, NER F-measure drops from 90\% to 50\% when the Word Error Rate increases from 8\% to 50\%.
-However, models based on deep neural networks seem less sensitive to OCR errors.
-Two approaches have been proposed by \cite{huynh2020use} and \cite{marz2021data}  to reduce the negative impact of OCR errors on NER performance on historical texts.
-The former applies a spelling correction tool to several corpora with variable OCR error rates. 
-As long as OCR errors remain low ($CER<2\%$ and $WER<10\%$), this strategy makes it possible to maintain good NER results.
-However, the F1-score starts to decrease significantly when OCR errors exceed these thresholds.
-The latter work focuses on adapting the training data to facilitate the generalisation of an off-the-shelf NER model from modern texts to historical texts.
-Finally, reusing a model trained on clean modern data, including embeddings computed on a historical corpus, and fine-tuned on a noisy historical ground truth has proven to be the most effective strategy.
-
-\subsection{Pipeline summary}
-\label{sec:pipeline-summary}
-
-Based on those works, we chose to test three OCR systems, namely, Tesseract, \peroocr, and Kraken.
-We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
-In \cref{sec:dataset}, we will explain the evaluation protocol used to assess the combined performance of these OCR and NER systems.
-
-\textbf{Tesseract} is a long-living project, born as a closed-sourced OCR at Hewlett-Packard in the eighties, it was progressively modernized, then open-sourced in 2005.
-From 2006 until November 2018, it was developed by Google and is still very active.
-We used in our tests version 4.1.1, released Dec.
-26, 2019.
-Version 5, released on Nov.
-30, 2021, has not been integrated in our tests yet.
-
-\textbf{Kraken} is a project created by Benjamin Kiessling several years ago (development can be traced back to 2015), and is actively used in the open-source eScriptorium project \cite{kiessling_escriptorium_2019}.
-As no pre-trained model for modern French was easily available, we used the default English text recognition model trained on modern printed English by Benjamin Kiessling on 2019.
-Models can be easily found and downloaded thanks to their hosting on Zenodo.
-
-\textbf{\peroocr} is a very recent project (started in 2020) from the Brno University of Technology in Czech Republic.
-Their authors used many state-of-the-art techniques to train it very efficiently.
-We used the version from the master branch of their GitHub repository, updated on Sep 15, 2021.
-We used the pre-trained weights provided by the authors on the same repository, created on Oct. 9, 2020 from European texts with Latin, Greek, and Cyrillic scripts.
-% https://github.com/DCGM/pero-ocr
-% https://www.fit.vut.cz/~ihradis/pero/pero_eu_cz_print_newspapers_2020-10-09.tar.gz
-
-\textbf{SpaCy} is a software library that offers NLP components assembled in modular pipelines specialised by language.
-Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (as of January 2022).
-Hence, we rely on SpaCy's ad hoc pipeline trained on French corpora and capable of Named Entity Recognition.
-% deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.
-The global architecture of these pipelines have not been yet published but are explained by the developers on their website.
-Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
-The decision layer is an adaptation of the transition-based model presented in~\cite{lample2016}.
-As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics.
-The classification layer relies on a finite-state machine whose transition probabilities are learnt using a multi-layer perceptron.
-% \bertrand{Drop the next sentences if we need space}
-% In 2018~\cite{won2018} evaluated SpaCy's NER ability to detect place names in five corpora of ancient letters written in English.
-% They measured an average F1 score of 0.57.
-% The SpaCy developers claim an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
-%For our experiments, the French model \textit{fr\_core\_news\_lg} is fine-tuned using our ground truth corpus.
-
-\textbf{CamemBERT} is well-known adaptation of the BERT transformer-based model for the French language\cite{vaswani2017attention,devlin2018bert,martin-etal-2020-camembert}.
-Such language models have become a new paradigm for NER\cite{li2020}.
-The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually refered as a "head".
-They can also be pre-trained in an unsupervised way on large amount of unlabeled texts for domain adaptation.
+NER for historical documents given our constraints

### 2022-01-27 src-latex/parts/40-ner-approaches.tex

@@
 \section{NER Approaches}
-We select two deep-learning-based NER models available in packaged software libraries: SpaCy NLP pipelines and CamemBERT.
-We also create an additional CamemBERT model pretrained on a collection of unannotated directory entries extracted with Pero-OCR.
-The NER layer of each three models is then fine-tuned on historical directories as detailed in\cref{subsection:experiment-1-setup} and\cref{subsection:experiment-2-setup}.
 
-\paragraph{Naming conventions} The short name of each model, indicated in square brackets, is used to refer to it from now on.
-
-\subsubsection{SpaCy NER pipeline~[SpaCy NER]}
-Spacy is a software library that offers NLP components assembled in modular pipelines specialised by language.
-Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (jan. 2022).
-Hence we rely on SpaCy's ad hoc pipelines trained on French corpora and capable of named entity recognition.
-% deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.
-The global architecture of these pipelines have not been yet published but are explained by the developers on their website.
-Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
-The decision layer is an adaptation of the transition-based model presented in~\cite{lample2016}.
-As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics.
-The classification layer relies on a finite-state machine whose transition probabilities are learned using a multilayer perceptron.
-\bertrand{Drop the next sentences if we need space}
-In 2018~\cite{won2018} evaluated SpaCy's NER ability to detect place names in five corpora of ancient letters written in English.
-They measured an average F1 score of 0.57.
-The SpaCy developers claim an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
-%For our experiments, the French model \textit{fr\_core\_news\_lg} is fine-tuned using our ground truth corpus.
-
-\subsubsection{CamemBERT~[CmBERT]}
-Language model based on Transformers, like BERT, have become a new paradigm for NER\cite{li2020}. 
-The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer.
-They can also be pretrained in an unsupervised way on historical texts for domain adaptation.
-As the directories are written in French, we chose the language model CamemBERT \cite{martin-etal-2020-camembert}, a Transformer model trained on a French corpus.
-\bertrand{WIP}
-\subsubsection{CamemBERT pretrained on directories~[CmBERT+ptrn]}
-In order to evaluate the benefits of pre-training the language model on OCR texts from domain-related documents, we adapt the CmBERT embeddings using \num{845.000} entries randomly selected wihtin the collection of directories and extracted with Pero-OCR.
-We achieve this by training the CmBERT model for 3 epochs on two unsupervised tasks: next sentence prediction and masked language model.
+\begin{itemize}
+    \item regexp
+    \item Spacy CNN  : tok2vec tokenizer + transition based NER
+    \item CamemBERT finetuned with wikiner-fr + gold dataset (clean texts)
+    \item CamemBERT finetuned with wikiner-fr + gold dataset (clean texts + pretrained on raw texts (Pero ? Kraken ? Both ?)
+\end{itemize}
 
 % Other possibilities:
-%1. Traditional ML based:
-%    Conditional Random Fields (CRF) - https://pypi.org/project/sklearn-crfsuite/
-%    Maximum-entropy Markov model
+1. Traditional ML based:
+    Conditional Random Fields (CRF) - https://pypi.org/project/sklearn-crfsuite/
+    Maximum-entropy Markov model
 
-%2. Neural Networks based:
-%    LSTMs, bi-LSTM - https://github.com/flairNLP/flair
-%    CNNs (SpaCy uses CNN based architecture)
-%    Transformers (Spacy has recently launched it) - %https://spacy.io/universe/project/spacy-transformers
+2. Neural Networks based:
+    LSTMs, bi-LSTM - https://github.com/flairNLP/flair
+    CNNs (SpaCy uses CNN based architecture)
+    Transformers (Spacy has recently launched it) - https://spacy.io/universe/project/spacy-transformers

### 2022-01-27 src-latex/parts/60-results.tex

--- 2022-01-18 src-latex/parts/60-results.tex
@@
 
-\subsection{OCR evaluation}
-
-\begin{figure}
-
-\subcaptionbox{}[.5\linewidth]{
-\begin{tabular}{rll}
-          & CER & CER$_{norm.}$ \\
-\midrule
-Pero      & 3.78\% & 3.76\% \\   
-Tesseract & 6.56\% & 6.45\% \\
-\bottomrule
-\end{tabular}
-
-\bigskip
-
-\includegraphics[width=\linewidth]{images/ocr-eval-2.pdf}
-}
-\subcaptionbox{}[.5\linewidth]{
-\includegraphics[width=\linewidth]{images/ocr-eval-1.pdf}
-}
-\caption{Character error rates at entry-level for Pero OCR and Tesseract. (a) Global CER and distribution of the CER per entry. (b):
-joint plot of the dataset showing that the two systems do not fail on the same entries.}  
-\end{figure}
-
-\subsection{Experiment 1: NER sensibility to the number of training samples}
+\subsection{Experiment 1: Accuracy vs NER approach vs training set size}
 
@@
 
-\begin{table}[h!]
+\begin{table}[!h]
+
+\caption{Models F1 score on unseen data vs trainset size}
 \centering
-\caption{NER performances at entry-level for SpaCy NER, CamemBERT and CamemBERT-pretrained when fine-tuned with an increasing number of examples.}
-\begin{tabular}{llrrrrrrrr}
-       & Training examples &  49   &  99   &  199  &  398  &  796  &  1593 &  3186 &  6373 \\
-       & \% & 0.8   & 1.6   & 3.1   & 6.2   & 12.5  & 25.0  & 50.0  & 100.0 \\
-\midrule\bottomrule
-\multirow{3}{*}{\rotatebox{90}{F1 score}} & Camembert &  89.5 &  90.5 &  92.7 &  93.3 &  \textbf{94.1} &  \textbf{94.9} &  \textbf{94.6} &  \textbf{95.1} \\
-       & Camembert-pretrained &  \textbf{92.2} &  \textbf{92.9} &  \textbf{93.6} &  \textbf{93.8} &  93.8 &  94.1 &  \textbf{94.6} &  94.4 \\
-       & SpaCy NER &  87.0 &  89.0 &  90.3 &  91.9 &  92.1 &  92.8 &  93.2 &  93.5 \\
-\cline{1-10}
-\multirow{3}{*}{\rotatebox{90}{Precision}} & Camembert &  87.4 &  88.7 &  91.5 &  92.7 &  93.3 &  94.9 &  93.9 &  95.1 \\
-       & Camembert-pretrained &  90.8 &  91.8 &  92.9 &  93.0 &  93.0 &  93.4 &  94.1 &  93.9 \\
-       & SpaCy NER &  85.6 &  87.7 &  90.0 &  92.0 &  92.4 &  92.8 &  93.1 &  93.7 \\
-\cline{1-10}
-\multirow{3}{*}{\rotatebox{90}{Recall}} & Camembert &  91.6 &  92.5 &  93.9 &  93.9 &  94.9 &  94.9 &  95.4 &  95.1 \\
-       & Camembert-pretrained &  93.6 &  94.0 &  94.4 &  94.6 &  94.6 &  94.8 &  95.0 &  94.9 \\
-       & SpaCy NER &  88.6 &  90.4 &  90.7 &  91.7 &  91.9 &  92.8 &  93.3 &  93.4 \\
+\begin{tabular}[t]{lcccccccc}
+\toprule
+Trainset size (\% of the full trainset) & 0.8 & 1.5 & 3.1 & 6.2 & 12.5 & 25 & 50 & 100\\
+\midrule
+Cnn & \textbf{0.855} & \textbf{0.885} & \textbf{0.910} & 0.911 & 0.919 & 0.925 & 0.926 & 0.929\\
+Camembert & 0.236 & 0.626 & 0.856 & \textbf{0.916} & \textbf{0.930} & 0.933 & \textbf{0.948} & 0.953\\
+Camembert+pretraining & 0.190 & 0.685 & 0.889 & 0.913 & \textbf{0.930} & \textbf{0.946} & \textbf{0.948} & \textbf{0.960}\\
+\bottomrule
 \end{tabular}
-\label{tab:experiment-1-models-performances}
 \end{table}
 
-\begin{figure}[h!]
+\begin{figure}[htb!]
 	   \center{\includegraphics[width=\textwidth]
-	       {images/experiment-1-models-performances.pdf}}
-	  \caption{\label{fig:f1-vs-trainsize} Precision, recall and f1 score from table~\ref{tab:experiment-1-models-performances}}.
+	       {../material/experiment_1/f1_vs_trainsize.png}}
+	  \vspace{3in}
+	  \caption{\label{fig:f1-vs-trainsize} Models F1 score on unseen data vs trainset size}
 \end{figure}
 
-\subsection{Experiment 2: NER in the presence of noisy OCR texts}
+\subsection{Experiment 2: sensibility to training on noisy data}
+Idea : Measure the impact of training with noisy data when predicting NER tags on both clean and noisy data.
+
+What's interesting : do models perform better on noisy data when trained on noisy data ? Does it impact predictions on clean data ?
+
+Table : NER NN models VS train on {noisy, clean} datasets VS evaluate on {noisy, clean} datasets
+
+\subsection{Experiment 3: Accuracy vs NER approach vs OCR noise}
 
@@
 
-Table : NER NN models VS train on {noisy, clean} datasets VS evaluate on {noisy, clean} datasets
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{src-latex/figs/eval-ner-exp2.pdf}
-    \caption{F1 scores (in \%) of NER predictions in presence of OCR noise in the training and testing examples, either manually corrected or raw from Pero-OCR and Tesseract.
-The type of examples used to train the NER task is noted in indice after the model name (e.g.
-Bert$_{reference}$).
-Results show that best performances on OCRed entries are obtained when the BERT model has been pretrained and fine-tuned for NER on examples affected with similar OCR errors.}
-    \label{fig:exp_2_eval_ner}
-\end{figure}
 Opt. use synthetic text perturbation as well? Maybe not interesting and too artificial if we have access to 2+ OCR systems.

### 2022-02-08 src-latex/main-paper.tex
Update ack

@@
 %
-The authors want to thank S. Bacciochi and P. Cristofoli for helping to create the reference dataset, 
-L. Morice for annotating data,
-as well as G. Thomas, P. Abi Saad, R. Lelièvre, D. Mignon, T. Cavaciuti and P. Sadki for contributing to the annotation platform.
+The authors want to thank S.~Bacciochi, P.~Cristofoli and J.~Perret for helping to create the reference dataset, 
+L.~Morice for annotating data,
+as well as G.~Thomas, P.~Abi Saad, R.~Lelièvre, D.~Mignon, T.~Cavaciuti and P.~Sadki for contributing to the annotation platform.
 }

### 2022-02-25 rebuttal/rebuttal_response_max_500_words.txt
Add reviews and tentative response

We thank the reviewers for their comments. We summarize and respond to them in what follows.

[Reviewer 1] ICDAR Contest in 2011 related to that topic
Thanks for pointing this out. They are essentially using a full-cross evaluation of all methods to perform end-to-end evaluation, just like us. Also, they do not report NER evaluation, unlike us.
[TODO add ref to https://hal.archives-ouvertes.fr/inria-00608371 + small comment]  We updated the paper with these precisions.

[Reviewer 1] Qualitative analysis of influence of OCR quality on NER
[Reviewer 1] Metrics uses, potential biases

[Bertrand] Ajouter un paragraphe d'évaluation qualitative des résultats de l'ocr+ner comme on avait prévu initialement, et un commentaire la projection des labels dans les chaînes bruitée. 
[Joseph] say we used metrics as reported on previous work, and also that one of our goals is to assess whether NER could help correcting OCR errors?
Here OCR is either pretty good or very wrong, so alignments usually make sense or can be detected as degenerated cases. However, it is very true that some borderline cases may cause trouble. This requires more study and [TODO] we updated the paper to be more explicit about these limitations.

[Reviewer 2] comparing two runs of Pero OCR together, or two runs of Tesseract together
Very good point, but we could not compete with the large, private training datasets used to produce the models we used. We chose to report, as objectively as possible, the performance of actual, off-the-shelf systems.

[Reviewer 2] reason for not reporting Kraken performance on downstream NER
[Bertrand] Ajouter Kraken à l'évaluation NER ? 
(say we chose not to report this because would have reduced a lot the size of the test set? This would require to add one extra set of experiment; not possible because of space constraints)

[Reviewer 2] missing Figure 1 reference, inconsistent reference to figures, Tesseract spelling
Thanks for pointing these issues, [TODO] we fixed them.

[Reviewer 2] Experiment with artificial OCR noise in future work
This is planned, but it seems to be very hard to generate realistic noise.
We decided to first focus on real cases with no further assumptions.

[Reviewer 3] Section 2.3 should be a separate section about the pipeline
[RBertrand] Transformer la sous-section 2.3 "Pipeline summary" en section 3. 
[TODO attention espace supplémentaire] We extracted this content as a new section as suggested.

[Reviewer 3] What are "thumbnails" in page 7
We refer to "thumbnails" as image crops, have the same resolution as the original image.
We updated the text to clarify this.

[Reviewer 3] Description of NER QA is confusing
[TODO] we clarified the relevant section

[Reviewer 3] Figure 4 needs clarification
[Bertrand] Détailler la légende de la figure 4 ? 
[TODO] We improved the description of fig. 4.

### 2022-02-27 rebuttal/rebuttal_response_max_500_words.txt
wip rebuttal

@@
 [Reviewer 1] Qualitative analysis of influence of OCR quality on NER
+[Bertrand] say that effects of OCR noise are indeed non-uniform on NER results.
+The impact on structuring elements of the entries like punctuation symbols appears to be most significant, as they mark the separation between semantically homogeneous blocks (e.g.
+between PERS and ACT).
+We plan to carry out a more in-depth qualitative study in the near future ?
+
 [Reviewer 1] Metrics uses, potential biases
-
 [Bertrand] Ajouter un paragraphe d'évaluation qualitative des résultats de l'ocr+ner comme on avait prévu initialement, et un commentaire la projection des labels dans les chaînes bruitée. 
@@
 [Reviewer 2] reason for not reporting Kraken performance on downstream NER
-[Bertrand] Ajouter Kraken à l'évaluation NER ? 
-(say we chose not to report this because would have reduced a lot the size of the test set? This would require to add one extra set of experiment; not possible because of space constraints)
+We chose not to report Kraken performances as it would have drastically reduced the size of the labelled sets "NER-x" and thus hampered the other evaluations on the reference, pero and tesseract datasets.
+All labelled datasets contains the same subset of directories entries by construction (section 3).
+Kraken performs quite poorly, which hinders the projection of labels from the ground truth and drastically reduces the size of the subset of labelled entries.
 
@@
 Thanks for pointing these issues, [TODO] we fixed them.
+(Refs. to figures fixed)
+(About T/tesseract : is he refering to fig 7 ? If so, "tesseract" is the name of the test set from "NER-tesseract").
 
 [Reviewer 2] Experiment with artificial OCR noise in future work
-This is planned, but it seems to be very hard to generate realistic noise.
-We decided to first focus on real cases with no further assumptions.
+We plan to perform such experiments following e.g. the techniques presented in https://doi.org/10.1007/978-3-030-86331-9_48
+As it seems to be quite hard to generate realistic noise, we decided to first focus on real cases with no further assumptions.
 
 [Reviewer 3] Section 2.3 should be a separate section about the pipeline
-[RBertrand] Transformer la sous-section 2.3 "Pipeline summary" en section 3. 
 [TODO attention espace supplémentaire] We extracted this content as a new section as suggested.
@@
 We refer to "thumbnails" as image crops, have the same resolution as the original image.
-We updated the text to clarify this.
+We updated the text to clarify this. (TODO ?)
 

### 2022-02-28 src-latex/parts/20-related-work.tex
Add section 3

--- 2022-01-20 src-latex/parts/20-related-work.tex
@@
 Finally, reusing a model trained on clean modern data, including embeddings computed on a historical corpus, and fine-tuned on a noisy historical ground truth has proven to be the most effective strategy.
-
-\subsection{Pipeline summary}
-\label{sec:pipeline-summary}
-
-Based on those works, we chose to test three OCR systems, namely, Tesseract, \peroocr, and Kraken.
-We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
-In \cref{sec:dataset}, we will explain the evaluation protocol used to assess the combined performance of these OCR and NER systems.
-
-\textbf{Tesseract} is a long-living project, born as a closed-sourced OCR at Hewlett-Packard in the eighties, it was progressively modernized, then open-sourced in 2005.
-From 2006 until November 2018, it was developed by Google and is still very active.
-We used in our tests version 4.1.1, released Dec.
-26, 2019.
-Version 5, released on Nov.
-30, 2021, has not been integrated in our tests yet.
-
-\textbf{Kraken} is a project created by Benjamin Kiessling several years ago (development can be traced back to 2015), and is actively used in the open-source eScriptorium project \cite{kiessling_escriptorium_2019}.
-As no pre-trained model for modern French was easily available, we used the default English text recognition model trained on modern printed English by Benjamin Kiessling on 2019.
-Models can be easily found and downloaded thanks to their hosting on Zenodo.
-
-\textbf{\peroocr} is a very recent project (started in 2020) from the Brno University of Technology in Czech Republic.
-Their authors used many state-of-the-art techniques to train it very efficiently.
-We used the version from the master branch of their GitHub repository, updated on Sep 15, 2021.
-We used the pre-trained weights provided by the authors on the same repository, created on Oct. 9, 2020 from European texts with Latin, Greek, and Cyrillic scripts.
-% https://github.com/DCGM/pero-ocr
-% https://www.fit.vut.cz/~ihradis/pero/pero_eu_cz_print_newspapers_2020-10-09.tar.gz
-
-\textbf{SpaCy} is a software library that offers NLP components assembled in modular pipelines specialised by language.
-Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (as of January 2022).
-Hence, we rely on SpaCy's ad hoc pipeline trained on French corpora and capable of Named Entity Recognition.
-% deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.
-The global architecture of these pipelines have not been yet published but are explained by the developers on their website.
-Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
-The decision layer is an adaptation of the transition-based model presented in~\cite{lample2016}.
-As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics.
-The classification layer relies on a finite-state machine whose transition probabilities are learnt using a multi-layer perceptron.
-% \bertrand{Drop the next sentences if we need space}
-% In 2018~\cite{won2018} evaluated SpaCy's NER ability to detect place names in five corpora of ancient letters written in English.
-% They measured an average F1 score of 0.57.
-% The SpaCy developers claim an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
-%For our experiments, the French model \textit{fr\_core\_news\_lg} is fine-tuned using our ground truth corpus.
-
-\textbf{CamemBERT} is well-known adaptation of the BERT transformer-based model for the French language\cite{vaswani2017attention,devlin2018bert,martin-etal-2020-camembert}.
-Such language models have become a new paradigm for NER\cite{li2020}.
-The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually refered as a "head".
-They can also be pre-trained in an unsupervised way on large amount of unlabeled texts for domain adaptation.

### 2022-02-28 src-latex/parts/30-dataset.tex
wip rebuttal
Add section 3

--- 2022-01-20 src-latex/parts/30-dataset.tex
@@
 % !TeX root = ../main-paper.tex
-\section{Dataset}
-\label{sec:dataset}
 
-% Conventions nommage
-% Dataset annoté à la main = "ground truth"
-% ground truth aligné = "ner-reference"
-% entrées pero aligné = "ner-pero"
-% entrées tesseract alignées = "ner-tesseract"
+\section{A two-stage pipeline to assess OCR and NER performances on historical texts}
+\label{sec:pipeline-summary}
 
-%In this work, we  aim  at  producing  structured  spatio-temporal  data  from the information contained in the XIX\textsuperscript{th} century Parisian trade directories.
-%To do this, it is therefore necessary to extract the text from the scans of the directories to be processed and to identify the entries in the various indexing lists used.
-%Then, in each of these entries, the named entities they contain have to be extracted to produce structured spatio-temporal data representing the evolution over time of the people and businesses listed in these directories, their descriptions and their locations.
-%Both of these extraction tasks require data to train and evaluate the OCR and NER approaches identified as potentially relevant: Tesseract and PERO OCR for the OCR task and Spacy and CamemBERT - pre-trained or fine-tuned - neural models for the NER task.
-
-This section presents the historical sources that we selected and their contents.
-It also details the construction of the groundtruth dataset leveraged in our experiments and the metrics used to evaluate OCR and NER results.
-
-\subsection{A selection of Paris trade directories from 1798 to 1854}
-\label{sec:corpus}
-
-The directories are available from different libraries in Paris and have been digitised independently in various levels of quality. 
-They cover a large period of time and originate from several publishers.
-Therefore, their contents, indexes, layouts, methods of printing, etc. may vary a lot from one directory to another (see Figure \ref{fig:directories}).
-We want our groundtruth dataset to be representative of the diversity of directories available in the period.
-
-\begin{figure}[htb!]
-	   \center{\includegraphics[width=0.9\textwidth]
-	       {./images/DirectoryExcerpts2.png}}
-	  \caption{\label{fig:directories} Examples of directory layouts and contents: 1) Duverneuil et La Tynna 1806 - index by name; 2) Deflandre 1828 - index by activity ; 3) Bottin 1851 - index by street name}
-\end{figure}
-
-Directories contain lists of people with various information attached.
-For instance, the directory published by \emph{Didot} in 1854 contains three redundant lists of people sorted by name, by activity and by street name.
-A typical example entry from this directory is ``\textit{Batton (D.-A.) \scalerel*{\includegraphics{./images/LH.png}}{H}, professeur au Conservatoire de musique et de déclamation, Saint-Georges, 47.}''.
-It begins with the person's name and surname, here inside parentheses.
-The glyph denotes that this person was awarded the \textit{Légion d'Honneur}.
-Then comes a description of the person's activity, here his profession (professor at the Conservatory of music and declamation), but it can also be a social status.
-Such descriptions range from a single word to paragraphs describing the occupation in full detail.
-The street name and number where the person lives or carries out their activities ends the entry.
-
-These are the pieces of information that we want to extract, deduplicate  and structure to build a spatio-temporal database.
-Except for some potentially wordy activity descriptions, they correspond to named entities.
-However, while most entries contain the same types of named entities, their order and the way they are written vary from one directory/index to another.
-To provide examples of each entry structure, pages from each type have thus been annotated.
-%\footnote{In this work, the street name index entries have not been included as they are very different from the others.
-The strategy to be adopted to process them without loss of quality for the NER model will be the subject of future work.}.
-
-\subsection{A dataset for OCR and NER evaluation}
-
-\begin{figure}[tb]
+\begin{figure}[h!]
     \includegraphics[width=\linewidth]{figs/protocol.pdf}
@@
 
-The simplified data extraction pipeline depicted in~\cref{fig.pipeline} processes the documents presented
-in~\cref{sec:corpus}. First, the page layout extraction and entry segmentation are performed with a
-semi-automated system and checked by a human. The resulting images, representing 8765 entries, are the inputs of a customizable two-stage pipeline.
+The proposed pipeline to evaluate the extraction and annotation of historical directories is depicted in~\cref{fig.pipeline}
+First, the page layout extraction and entry segmentation are performed with a
+semi-automated system and checked by a human.
+The resulting images are the inputs of a customizable two-stage pipeline detailed in this section.
 
-%1. Entrée/Sortie de l'OCR
-\textbf{OCR stage.} The first stage of the pipeline aims at extracting the raw text from the images. An OCR system runs
-on the thumbnails of each segmented entry to extract its text. An entry might span over multiple text lines but is
-always a single block. Thus, the most adapted mode is chosen when the OCR system allows for the detection mode (e.g. the
-\emph{block} mode for \emph{Tesseract}). 
-Some glyphs used in this dataset might be unknown to the OCR, and some like
-\scalerel*{\includegraphics{./images/LH.png}}{H} do not even have a Unicode codepoint, and were annotated using Unicode Private User Area 1.
-Furthermore, as some annotations guidelines where unclear to human annotators, some projections rules were applied: whitespace, dash, dots and a couple of commonly confused characters where projected to well-defined codepoints.
-The same normalization was applied to OCR predictions.
-At the end of this first stage, an OCR Quality Assessment (Q.A.) between the \emph{normalised} groundtruth text and the
-OCR output is performed on the basis of the $8,765$ entries which were manually controlled, totalling $424,764$ characters (including $54,387$ spacing characters).
-Entries are $49.0$-characters long on average.
+\subsection{OCR and NER systems considered}
+\label{subsec:ocr-ner-tools}
+We chose to test three OCR systems: Tesseract, \peroocr, and Kraken. 
+We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
+In \cref{sec:dataset}, we will explain the evaluation protocol used to assess the combined performance of these OCR and NER systems.
 
-%2. Entrée/Sortie
-\textbf{NER Stage.} The second stage of the pipeline aims at extracting the named entities from a text with a NER
-system. This text originates from the OCR outputs in a real-word scenario but might be the groundtruth text in order to
-evaluate the NER performance independently. There are 5 types of entity to detect~(see.~\cref{tab.entities}).
-The NER system has to classify non-overlapping parts of the text into one of these entities (or none of them). A second
-Q.A. (namely, NER Q.A.) is performed at the end of this stage between the groundtruth and the NER output.
+\textbf{Tesseract} is a long-living project, born as a closed-sourced OCR at Hewlett-Packard in the eighties, it was progressively modernized, then open-sourced in 2005.
+From 2006 until November 2018, it was developed by Google and is still very active.
+We used in our tests version 4.1.1, released Dec.
+26, 2019.
+Version 5, released on Nov.
+30, 2021, has not been integrated in our tests yet.
 
-Note that the dataset used to assess the NER stage is a subset of the entries. Indeed, we need to ensure that the datasets
-contain the same entries whichever the OCR used in the previous stage. Entries where the OCR produced an empty string
-and those for which no entity could be projected from the groundtruth have to be ignored. 
-We filter the set of entries by keeping only the entries that are always valid at the end of the stage 1.
-%At the cost of losing 5\% of the references entries, we thus have 8341 valid entries at the input of the stage 2.
-% /!\ True only for \peroocr and Tesseract predictions, not considering Kraken (we loose 500 more entries)
-Therefore, the $8,765$ reference entries were manually annotated with $34,242$ entities; entries contain $3.9$ entities on average.
-Projecting reference tagged entities on OCR predictions resulted in a variable loss of entries.
-For \peroocr, $8,392$ valid entries were generated, for Tesseract $8,700$ and for Kraken $7,990$.
-The resulting intersection of the sets of valid entries contained $7,725$ entries for the tree OCR systems (and the reference),
-or $8,341$ entries if we consider \peroocr and Tesseract only.
+\textbf{Kraken} is a project created by Benjamin Kiessling several years ago (development can be traced back to 2015), and is actively used in the open-source eScriptorium project \cite{kiessling_escriptorium_2019}.
+As no pre-trained model for modern French was easily available, we used the default English text recognition model trained on modern printed English by Benjamin Kiessling on 2019.
+Models can be easily found and downloaded thanks to their hosting on Zenodo.
 
-\begin{table}
-\caption{Entities to recognise in the dataset.}
-\label{tab.entities}
-\small
-\begin{tabular}{llp{8cm}}
-Entity & Count & Description\\    
-PER & 8788 & a person name or a business name, usually referred to as several person names.
-First names, initials, or civility mentions are included.
-E.g.: \textit{Alibert (Prosper)}, \textit{Allamand frères et Hersent},\textit{Heurtemotte (Vve)}.
-\\
-TITLE & 483 & an honorary title, either text, glyphs or a combination of the two.
-E.g: O.
-\scalerel*{\includegraphics{./images/LH.png}} {H} for \textit{Officier de la Légion d'Honneur} or \scalerel*{\includegraphics{./images/GMLondon.png}}{H} for the Great Medal at the London exhibition.\\
-ACT & 6472 & the profession or social status of a person summarised in the single concept of activity.
-E.g.: \textit{horlogerie}, \textit{export en tous genres}, \textit{Conseiller d'État}, \textit{propriétaire}.\\
-LOC & 9709 & mostly street names (\textit{r. de la Jussienne}), but may also be neighbourhoods (\textit{Marais du Temple}) or any indirect spatial references (\textit{au Palais Royal}).\\
-CARDINAL & 8747 & a street number, as part of an address (16, 5 bis), or a range of numbers (e.g. 23-25 or 5 \textit{à} 9).\\
-FT & 43 & a geographic entity type, used to give more details on a location, e.g. \textit{boutique}, \textit{atelier}, \textit{fab.} or \textit{dépôt}. \\
-\end{tabular}
-\end{table}
+\textbf{\peroocr} is a very recent project (started in 2020) from the Brno University of Technology in Czech Republic.
+Their authors used many state-of-the-art techniques to train it very efficiently.
+We used the version from the master branch of their GitHub repository, updated on Sep 15, 2021.
+We used the pre-trained weights provided by the authors on the same repository, created on Oct. 9, 2020 from European texts with Latin, Greek, and Cyrillic scripts.
+% https://github.com/DCGM/pero-ocr
+% https://www.fit.vut.cz/~ihradis/pero/pero_eu_cz_print_newspapers_2020-10-09.tar.gz
 
-%
-%TODO average number of tags/entities per entry
-%TODO lists actual dataset content: set of entries. For each entry:
-%\begin{itemize*}
-%    \item ref to original book, page, region
-%    \item human transcription (OCR GT)
-%    \item human NER (NER GT)
-%    \item [GENERATED] OCR produced by Pero OCR, Tesseract v4, Kraken (precise model used)
-%    \item [GENERATED] NER GT aligned on text predictions from Pero OCR, Tesseract v4, Kraken
-%\end{itemize*}
+\textbf{SpaCy} is a software library that offers NLP components assembled in modular pipelines specialised by language.
+Although BERT is available in the latest version of SpaCy (v3), the pipeline for French does not provide a NER layer at the time of our experiments (as of January 2022).
+Hence, we rely on SpaCy's ad hoc pipeline trained on French corpora and capable of Named Entity Recognition.
+% deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.
+The global architecture of these pipelines have not been yet published but are explained by the developers on their website.
+Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
+The decision layer is an adaptation of the transition-based model presented in~\cite{lample2016}.
+As words are processed sequentially, their vectors are concatenated with those of the last known entities to encode the nearby predicted semantics.
+The classification layer relies on a finite-state machine whose transition probabilities are learnt using a multi-layer perceptron.
+% \bertrand{Drop the next sentences if we need space}
+% In 2018~\cite{won2018} evaluated SpaCy's NER ability to detect place names in five corpora of ancient letters written in English.
+% They measured an average F1 score of 0.57.
+% The SpaCy developers claim an accuracy of 0.85 for the English NER pipeline \textit{en\_core\_web\_lg} on the OntoNotes 5.0 corpus\footnote{https://spacy.io/usage/facts-figures}.
+%For our experiments, the French model \textit{fr\_core\_news\_lg} is fine-tuned using our ground truth corpus.
+
+\textbf{CamemBERT} is well-known adaptation of the BERT transformer-based model for the French language\cite{vaswani2017attention,devlin2018bert,martin-etal-2020-camembert}.
+Such language models have become a new paradigm for NER\cite{li2020}.
+The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually referred as a "head".
+They can also be pre-trained in an unsupervised way on large amount of unlabeled texts for domain adaptation.
 

### 2022-02-28 rebuttal/rebuttal_response_max_500_words.txt
update rebuttal response

@@
 [Reviewer 3] Section 2.3 should be a separate section about the pipeline
-[TODO attention espace supplémentaire] We extracted this content as a new section as suggested.
+We extracted this content as a new section as suggested.
+We have also moved the subsection "Metrics for OCR and NER Quality assessment" from the former section 3 "Dataset" to this new section.
 

### 2022-02-28 src-latex/parts/00-abstract.tex
WIP rebuttal
Ref to fig. 1 + mention to thumbnails

--- 2022-01-12 src-latex/parts/00-abstract.tex
@@
 % !TeX root = ../main-paper.tex
-This paper features a comparison of four (\bertrand{Ensure it's 4!} named entity recognition (NER) techniques which can be reasonably used in the context of historical document analysis.
-Such natural language processing techniques identify which class each text token belong to, like "person name", "location", "number" or "activity".
+Named entity recognition (NER) is a necessary step in many pipelines targeting historical documents.
+Indeed, such natural language processing techniques identify which class each text token belongs to,
+e.g. ``person name'', ``location'', ``number''.
 Introducing a new public dataset built from 19th century French directories,
-we perform a practical assessment of the accuracy of, and the annotation cost (considering various training set sizes) needed by each technique.
-We also study the impact of real OCR noise for several OCR systems on the performance of NER.
-Experiments suggest that FIXME APPROACH NAME is the most effective approach on the dataset we processed.
-FIXME add numerical results
-Results can be reproduced using data and code freely available at \url{https://github.com/soduco/paper-ner-bench-das22}
-and the latest version of is paper can be obtained at \url{TODO arxiv URL}.
+we first assess how noisy modern, off-the-shelf OCR are.
+Then, we compare modern CNN- and Transformer-based NER techniques
+which can be reasonably used in the context of historical document analysis.
+We measure their requirements in terms of training data,
+the effects of OCR noise on their performance,
+and show how Transformer-based NER can benefit from unsupervised pre-training 
+and supervised fine-tuning on noisy data (\cref{fig.pipeline-overview}).
+Results can be reproduced using resources available at \url{https://github.com/soduco/paper-ner-bench-das22}.
+% The latest version of is paper can be obtained at \url{https://arxiv.org/abs/xxxx.xxxxx}.

### 2022-02-28 src-latex/parts/30-dataset.tex
WIP rebuttal
Ref to fig. 1 + mention to thumbnails

--- 2022-01-21 src-latex/parts/30-dataset.tex
@@
 They cover a large period of time and originate from several publishers.
-Therefore, their contents, indexes, layouts, methods of printing, etc. may vary a lot from one directory to another (see Figure \ref{fig:directories}).
+Therefore, their contents, indexes, layouts, methods of printing, etc. may vary a lot from one directory to another (see fig. \ref{fig:directories}).
 We want our groundtruth dataset to be representative of the diversity of directories available in the period.
@@
 %1. Entrée/Sortie de l'OCR
-\textbf{OCR stage.} The first stage of the pipeline aims at extracting the raw text from the images. An OCR system runs
-on the thumbnails of each segmented entry to extract its text. An entry might span over multiple text lines but is
+\textbf{OCR stage.} The first stage of the pipeline aims at extracting the raw text from the images. 
+Individual image fragments are created from each segmented entry and fed to an OCR system for text extraction.
+An entry might span over multiple text lines but is
 always a single block. Thus, the most adapted mode is chosen when the OCR system allows for the detection mode (e.g. the

### 2022-02-28 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Make Language Tools happy

@@
 \subsection{Results and discussion}
-Figure \ref{fig:f1-vs-trainsize} displays the averaged precision, recall, and F1-score for all models on the 8 subsets created from the groundtruth.
+Fig. \ref{fig:f1-vs-trainsize} displays the averaged precision, recall, and F1-score for all models on the 8 subsets created from the groundtruth.
 CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: the performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.
@@
 
-It appears that pre-training the CamemBERT model on OCR text seems worth it only when a the training set used to fine-tune the NER layer is small.
+It appears that pre-training the CamemBERT model on OCR text seems worth it only when the training set used to fine-tune the NER layer is small.
 This effect might be due to the differences in nature between the training subsets, whose texts are manually corrected, and the noisy OCR texts used to pretrain CamemBERT.

### 2022-02-28 rebuttal/rebuttal_response_max_500_words.txt
Update rebuttal response with Edwin's version

--- 2022-02-27 rebuttal/rebuttal_response_max_500_words.txt
@@
 
-[Reviewer 1] ICDAR Contest in 2011 related to that topic
-Thanks for pointing this out. They are essentially using a full-cross evaluation of all methods to perform end-to-end evaluation, just like us. Also, they do not report NER evaluation, unlike us.
-[TODO add ref to https://hal.archives-ouvertes.fr/inria-00608371 + small comment]  We updated the paper with these precisions.
+## Metrics and evaluation protocol
 
-[Reviewer 1] Qualitative analysis of influence of OCR quality on NER
-[Bertrand] say that effects of OCR noise are indeed non-uniform on NER results.
-The impact on structuring elements of the entries like punctuation symbols appears to be most significant, as they mark the separation between semantically homogeneous blocks (e.g.
-between PERS and ACT).
-We plan to carry out a more in-depth qualitative study in the near future ?
+> [Reviewer 1] ICDAR Contest in 2011 related to that topic
 
-[Reviewer 1] Metrics uses, potential biases
-[Bertrand] Ajouter un paragraphe d'évaluation qualitative des résultats de l'ocr+ner comme on avait prévu initialement, et un commentaire la projection des labels dans les chaînes bruitée. 
-[Joseph] say we used metrics as reported on previous work, and also that one of our goals is to assess whether NER could help correcting OCR errors?
-Here OCR is either pretty good or very wrong, so alignments usually make sense or can be detected as degenerated cases.
-However, it is very true that some borderline cases may cause trouble.
-This requires more study and [TODO] we updated the paper to be more explicit about these limitations.
+Thanks for pointing this out. They are essentially using a full-cross evaluation of all methods to perform end-to-end
+evaluation, just like us. Also, they do not report NER evaluation, unlike us. We updated the paper with these
+clarifications.
 
-[Reviewer 2] comparing two runs of Pero OCR together, or two runs of Tesseract together
+> [Reviewer 1] Qualitative analysis of influence of OCR quality on NER
+
+Effects of the OCR noise are indeed non-uniform on NER results. The impact on structuring elements of the entries like
+punctuation symbols appears to be most significant, as they mark the separation between semantically homogeneous blocks
+(e.g. between PERS and ACT). We plan to carry out a more in-depth qualitative study in the near future.
+
+> [Reviewer 1] The authors take great care to have a clean and reproducible way of projecting expected outcomes onto
+> noisy OCR results. The question arises to how useful this is and whether it is not introducing biases in various ways,
+> overestimating or underestimating the effects of the NER labeling
+
+The metrics used are the ones commonly used in previous works. Projecting the ground-truth labels on a noisy text was a
+sensible way to assess whether NERs could help to correct OCR errors. Here OCR is either pretty good or very wrong, so
+alignments usually make sense or can be detected as degenerated cases. However, it is very true that some borderline
+cases may cause trouble. This requires more study, and we updated the paper to be more explicit about these limitations.
+
+> [Reviewer 2] Comparing two runs of PERO OCR together, or two runs of Tesseract together
+
 Very good point, but we could not compete with the large, private training datasets used to produce the models we used.
@@
 
-[Reviewer 2] reason for not reporting Kraken performance on downstream NER
-We chose not to report Kraken performances as it would have drastically reduced the size of the labelled sets "NER-x" and thus hampered the other evaluations on the reference, pero and tesseract datasets.
-All labelled datasets contains the same subset of directories entries by construction (section 3).
-Kraken performs quite poorly, which hinders the projection of labels from the ground truth and drastically reduces the size of the subset of labelled entries.
+> [Reviewer 2] Reason for not reporting Kraken's performance on downstream NER
 
-[Reviewer 2] missing Figure 1 reference, inconsistent reference to figures, Tesseract spelling
-Thanks for pointing these issues, [TODO] we fixed them.
-(Refs. to figures fixed)
-(About T/tesseract : is he refering to fig 7 ? If so, "tesseract" is the name of the test set from "NER-tesseract").
+We chose not to report Kraken's performance as it would have drastically reduced the size of the labelled sets "NER-x" and
+thus hampered the other evaluations on the reference, PERO and tesseract datasets. All labelled datasets contain the
+same subset of directories entries by construction (section 3). Kraken performs quite poorly, which hinders the
+projection of labels from the ground truth and drastically reduces the size of the subset of labelled entries.
 
-[Reviewer 2] Experiment with artificial OCR noise in future work
+> [Reviewer 2] Experiment with artificial OCR noise in future work
+
 We plan to perform such experiments following e.g. the techniques presented in https://doi.org/10.1007/978-3-030-86331-9_48
@@
 
-[Reviewer 3] Section 2.3 should be a separate section about the pipeline
-[TODO attention espace supplémentaire] We extracted this content as a new section as suggested.
+## Paper organization
 
-[Reviewer 3] What are "thumbnails" in page 7
-We refer to "thumbnails" as image crops, have the same resolution as the original image.
-We updated the text to clarify this. (TODO ?)
+> [Reviewer 3] Section 2.3 should be a separate section about the pipeline
 
-[Reviewer 3] Description of NER QA is confusing
-[TODO] we clarified the relevant section
+We have tried to extract this content as a new section as suggested. However, this change made the sections unbalanced,
+and the explication flow disrupted. Moreover, it breaks the 15-pages limits. As a consequence, the overall organization
+of the paper was left unchanged.
 
-[Reviewer 3] Figure 4 needs clarification
-[Bertrand] Détailler la légende de la figure 4 ? 
-[TODO] We improved the description of fig. 4.
+## Typos, inconsistencies and lack of clarity
+
+> [Reviewer 1] Figure 1 is never referred to in the text
+> [Reviewer 1] When referring to figures, you sometimes write "fig. n", "figure n", and "Figure n"
+> [Reviewer 1] You wrote Tesseract twice with a lowercase t
+
+Thanks for pointing these issues, we fixed them.
+
+> [Reviewer 3] What are "thumbnails" in page 7
+
+We refer to "thumbnails" as image crops, they have the same resolution as the original image. We updated the text to
+clarify this.
+
+> [Reviewer 3] Description of NER QA is confusing
+> [Reviewer 3] Figure 4 needs clarification
+
+We clarified the relevant section and fig.4's caption.

### 2022-02-28 rebuttal/rebuttal_response_max_500_words.txt
500 words exactly

@@
-We thank the reviewers for their comments. We summarize and respond to them in what follows.
+We thank the reviewers for their comments. We will do our best to update the final version accordingly.
 
@@
 
-> [Reviewer 1] ICDAR Contest in 2011 related to that topic
+> [Reviewer 1] ICDAR 2011 Contest related to that topic
 
-Thanks for pointing this out. They are essentially using a full-cross evaluation of all methods to perform end-to-end
-evaluation, just like us. Also, they do not report NER evaluation, unlike us. We updated the paper with these
-clarifications.
+Thanks for pointing this out. They are essentially using a cross evaluation of all methods to perform end-to-end evaluation, just like us, but do not report NER evaluation, unlike us.
 
@@
 
-Effects of the OCR noise are indeed non-uniform on NER results. The impact on structuring elements of the entries like
-punctuation symbols appears to be most significant, as they mark the separation between semantically homogeneous blocks
-(e.g. between PERS and ACT). We plan to carry out a more in-depth qualitative study in the near future.
+Effects of the OCR noise are indeed non-uniform on NER results.
+The impact on structuring elements of the entries like punctuation symbols appears to be most significant, as they mark the separation between semantically homogeneous blocks.
+This is under further study.
 
-> [Reviewer 1] The authors take great care to have a clean and reproducible way of projecting expected outcomes onto
-> noisy OCR results. The question arises to how useful this is and whether it is not introducing biases in various ways,
-> overestimating or underestimating the effects of the NER labeling
+> [Reviewer 1] how useful [projecting expected outcomes onto noisy OCR results] is 
+> and whether [this] is not introducing biases in various ways, overestimating or underestimating 
+> the effects of the NER labeling
 
-The metrics used are the ones commonly used in previous works. Projecting the ground-truth labels on a noisy text was a
-sensible way to assess whether NERs could help to correct OCR errors. Here OCR is either pretty good or very wrong, so
-alignments usually make sense or can be detected as degenerated cases. However, it is very true that some borderline
-cases may cause trouble. This requires more study, and we updated the paper to be more explicit about these limitations.
+We used standard metrics from previous works.
+Projecting the ground-truth labels on a noisy text was a sensible way to assess whether NERs could help to correct OCR errors.
+In our experiments, alignments usually make sense or can be detected as degenerated cases.
+However, it is very true that some borderline cases may cause trouble.
+We will be more explicit about these limitations.
 
@@
 
-We chose not to report Kraken's performance as it would have drastically reduced the size of the labelled sets "NER-x" and
-thus hampered the other evaluations on the reference, PERO and tesseract datasets. All labelled datasets contain the
-same subset of directories entries by construction (section 3). Kraken performs quite poorly, which hinders the
-projection of labels from the ground truth and drastically reduces the size of the subset of labelled entries.
+We chose not to report Kraken's performance as it would have drastically reduced the size of the labelled sets used to compare the performance of NER systems, because poor Kraken's performance hinders the projection of labels from the ground truth and creates degenerated cases which are skipped.
 
@@
 
-We plan to perform such experiments following e.g. the techniques presented in https://doi.org/10.1007/978-3-030-86331-9_48
-As it seems to be quite hard to generate realistic noise, we decided to first focus on real cases with no further assumptions.
+We plan to perform such experiments, but as it is hard to generate realistic noise, we decided to first focus on real cases with no further assumptions.
 
@@
 
-We have tried to extract this content as a new section as suggested. However, this change made the sections unbalanced,
-and the explication flow disrupted. Moreover, it breaks the 15-pages limits. As a consequence, the overall organization
-of the paper was left unchanged.
+We have tried to extract this content as a new section as suggested, but this unbalanced sections, disrupted the explication flow, and breaks the 15-pages limits.
+As a consequence, the overall organization of the paper was left unchanged.
 
@@
 
-> [Reviewer 1] Figure 1 is never referred to in the text
-> [Reviewer 1] When referring to figures, you sometimes write "fig. n", "figure n", and "Figure n"
-> [Reviewer 1] You wrote Tesseract twice with a lowercase t
+> [Reviewer 1] Missing Figure 1 reference
+> [Reviewer 1] you sometimes write "fig. n", "figure n", and "Figure n"
+> [Reviewer 1] Tesseract spelling
 
-Thanks for pointing these issues, we fixed them.
+Thanks, we fixed this.
 
@@
 
-We refer to "thumbnails" as image crops, they have the same resolution as the original image. We updated the text to
-clarify this.
+We refer to "thumbnails" as image crops, they have the same resolution as the original image.
 

### 2022-03-28 rebuttal/rebuttal_response_max_500_words.txt
Ajt TODOs pour camera ready

@@
 
+* [ ] JC rajouter une phrase
+
 > [Reviewer 1] Qualitative analysis of influence of OCR quality on NER
@@
 This is under further study.
+
+* [x] ~~?? mettre 1 phrase en future work ?~~
 
@@
 
+* [ ] BD détailler un peu les limitations, être plus explicites à ce sujet
+
 > [Reviewer 2] Comparing two runs of PERO OCR together, or two runs of Tesseract together
@@
 
+* [ ] EC ajouter une phrase ?? (dire qu'on n'avait même pas assez de données pour faire du fine-tuning)
+
 > [Reviewer 2] Reason for not reporting Kraken's performance on downstream NER
@@
 
+* [ ] BD revoir un peu la phrase concernée ?
+
 > [Reviewer 2] Experiment with artificial OCR noise in future work
@@
 We plan to perform such experiments, but as it is hard to generate realistic noise, we decided to first focus on real cases with no further assumptions.
+
+* [ ] BD Explain in future work section
 
@@
 As a consequence, the overall organization of the paper was left unchanged.
+
+* [x] ~~No change required~~
 
@@
 
+* [ ] EC
+
 > [Reviewer 3] What are "thumbnails" in page 7
@@
 We refer to "thumbnails" as image crops, they have the same resolution as the original image.
+
+* [ ] EC
 
@@
 We clarified the relevant section and fig.4's caption.
+
+* [ ] EC
+
+## AUTRES
+
+* [ ] NA passe de relecture
+* [ ] NA relecture readme dataset + critique structure
+* [ ] JC rajouter lien Zenodo dataset
+* [ ] ?? dépôt HAL + ArXiV

### 2022-03-28 rebuttal/rebuttal_response_max_500_words.txt
rebuttal points 2/4/5

@@
 
-* [ ] BD détailler un peu les limitations, être plus explicites à ce sujet
+* [X] BD détailler un peu les limitations, être plus explicites à ce sujet
 
@@
 
-* [ ] BD revoir un peu la phrase concernée ?
+* [X] BD revoir un peu la phrase concernée ?
 
@@
 
-* [ ] BD Explain in future work section
+* [X] BD Explain in future work section
 

### 2022-03-28 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
rebuttal points 2/4/5

@@
 \subsection{Training and evaluation protocol}
-Only CmBERT and CmBERT+ptrn are considered since the first experiment shows that SpaCy NER is outperformed by these two models for all sizes of training sets.
+Only CmBERT and CmBERT+ptrn are considered since the first experiment shows that SpaCy NER is outperformed for all sizes of training sets.
+We leave Kraken aside as its performance hinders the projection of NER labels from the ground truth, thus dramatically reducing the size of labelled sets.
+Because the NER metrics are entity-wise, they may underestimate the NER performances on stronly noised texts with fragmented entities.
 We leverage the labelled sets of entries NER-\emph{reference}, NER-\emph{pero} and NER-\emph{tesseract} created as explained in \cref{sec:dataset}.
@@
 The measured F1-score are given in \cref{fig:exp_2_eval_ner}.
-Results clearly show that models perform best when both the pre-training and the NER fine-tuning share the same characteristics (here, OCR noise) as the texts to be processed.
+Results clearly show that models perform best when both the pre-training and the NER fine-tuning are similar to the processed texts in term of OCR noise.
 

### 2022-03-29 src-latex/parts/20-related-work.tex
Fix minor typos while re-reading

--- 2022-01-21 src-latex/parts/20-related-work.tex
@@
 Rule based approaches look for portions of the text that match patterns like in \cite{bell2020automated,nouvel2011} or dictionary (gazetteers, author lists, etc.) entries like in \cite{mansouri2008,maurel2011}.
-Such kind of approaches achieve very good results when applied to a specialised domain corpus and when an exhaustive lexicon are available, but at high system engineering cost \cite{nadeau2007}.
+Such kind of approaches achieve very good results when applied to a specialised domain corpus and when an exhaustive lexicon is available, but at high system engineering cost \cite{nadeau2007}.
 
@@
 However, the NER performance drops significantly as the quality of the OCR decreases and is correlated with its decrease.
+% FIXME few training data -> little training data? / is correlated with its decrease -> redundant?
 
@@
 Most of the time, they have evaluated NER approaches based on deep learning architectures as they seem to adapt more easily to OCR errors than rule-based or more classical supervised approaches.
-\cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by SpaCy \cite{spacy} library to perform NER on a corpus made of many journal articles with different levels of OCR errors.
+\cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by the SpaCy \cite{spacy} library to perform NER on a corpus made of many journal articles with different levels of OCR errors.
 For each OCRed article, a ground truth text is available so that the Word Error Rate (WER) can be computed.
@@
 
-\textbf{\peroocr} is a very recent project (started in 2020) from the Brno University of Technology in Czech Republic.
+\textbf{\peroocr} is a very recent project (started in 2020) from the Brno University of Technology in the Czech Republic.
 Their authors used many state-of-the-art techniques to train it very efficiently.
@@
 % deep-sequoia and wikiner-fr, capable of named entity recognition.\textit{fr\_core\_news\_lg}\footnote{https://spacy.io/models/fr} trained two corpora in French: deep-sequoia and wikiner-fr, capable of named entity recognition.
-The global architecture of these pipelines have not been yet published but are explained by the developers on their website.
+The global architecture of these pipelines have not been published yet, but are explained by the developers on their website.
 Words are first encoded into local context-aware embeddings using a window-based CNN similar to~\cite{collobert2011}.
@@
 
-\textbf{CamemBERT} is well-known adaptation of the BERT transformer-based model for the French language\cite{vaswani2017attention,devlin2018bert,martin-etal-2020-camembert}.
-Such language models have become a new paradigm for NER\cite{li2020}.
-The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually referred as a "head".
-They can also be pre-trained in an unsupervised way on large amount of unlabeled texts for domain adaptation.
+\textbf{CamemBERT} is well-known adaptation of the BERT transformer-based model for the French language~\cite{vaswani2017attention,devlin2018bert,martin-etal-2020-camembert}.
+Such language models have become a new paradigm for NER~\cite{li2020}.
+The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually referred as a ``head''.
+They can also be pre-trained in an unsupervised way on large amount of unlabelled texts for domain adaptation.

### 2022-03-29 src-latex/parts/00-abstract.tex
Move fig. 1 ref from abstract to introduction

@@
 and show how Transformer-based NER can benefit from unsupervised pre-training 
-and supervised fine-tuning on noisy data (\cref{fig.pipeline-overview}).
+and supervised fine-tuning on noisy data.
 Results can be reproduced using resources available at \url{https://github.com/soduco/paper-ner-bench-das22}.

### 2022-03-29 src-latex/parts/30-dataset.tex
Minor changes while re-reading

@@
 	       {./images/DirectoryExcerpts2.png}}
-	  \caption{\label{fig:directories} Examples of directory layouts and contents: 1) Duverneuil et La Tynna 1806 - index by name; 2) Deflandre 1828 - index by activity ; 3) Bottin 1851 - index by street name}
+	  \caption{\label{fig:directories} Examples of directory layouts and contents.
+\emph{Top}: Duverneuil et La Tynna 1806 - index by name; \emph{Middle}: Deflandre 1828 - index by activity ; \emph{Bottom}: Bottin 1851 - index by street name.}
 \end{figure}
@@
     \caption{Data extraction pipeline with two quality control checkpoints. The
-    NER Q.A. checkpoint may either assess the NER system lonely (the dashed
-    \emph{groundtruth} path) or may evaluate the joint work of a NER
+    NER Q.A. checkpoint may either assess the NER system in isolation (using the dashed
+    \emph{groundtruth} path), or may evaluate the joint work of a NER
     system with an OCR system.}
@@
 Some glyphs used in this dataset might be unknown to the OCR, and some like
-\scalerel*{\includegraphics{./images/LH.png}}{H} do not even have a Unicode codepoint, and were annotated using Unicode Private User Area 1.
+\scalerel*{\includegraphics{./images/LH.png}}{H}, which do not even have a Unicode codepoint, were annotated using Unicode Private User Area 1.
 Furthermore, as some annotations guidelines where unclear to human annotators, some projections rules were applied: whitespace, dash, dots and a couple of commonly confused characters where projected to well-defined codepoints.
@@
 \textbf{NER Stage.} The second stage of the pipeline aims at extracting the named entities from a text with a NER
-system. This text originates from the OCR outputs in a real-word scenario but might be the groundtruth text in order to
+system. This text can either originate from the OCR outputs in a real-word scenario, or from groundtruth text in order to
 evaluate the NER performance independently. There are 5 types of entity to detect~(see.~\cref{tab.entities}).
@@
 \begin{tabular}{llp{8cm}}
-Entity & Count & Description\\    
+\toprule
+\textbf{Entity} & \textbf{Count} & \textbf{Description}\\    
+\midrule
 PER & 8788 & a person name or a business name, usually referred to as several person names.
@@
 FT & 43 & a geographic entity type, used to give more details on a location, e.g. \textit{boutique}, \textit{atelier}, \textit{fab.} or \textit{dépôt}. \\
+\bottomrule
 \end{tabular}

### 2022-03-29 src-latex/parts/20-related-work.tex
Update missing comments

@@
 
-The directory processing pipeline presented in \cite{bell2020automated} includes an OCR step, done with Tesseract, and a NER step to identify company names and addresses, performed using regular expressions.
+The directory processing pipeline presented in \cite{bell2020automated} and illustrated in \cref{fig.pipeline-overview}, includes an OCR step done with Tesseract, and a NER step to identify company names and addresses, performed using regular expressions.
 This section reviews existing OCR and NER approaches on historical texts and presents some works assessing the effects of OCR quality on the NER performance and the proposed solutions.

### 2022-03-29 src-latex/parts/30-dataset.tex
Add links and refs to dataset
Update missing comments

@@
 It also details the construction of the groundtruth dataset leveraged in our experiments and the metrics used to evaluate OCR and NER results.
+The resulting dataset is publicly available~\cite{abadie_dataset_22} under the permissive \emph{Creative Commons Attribution 4.0 International} licence.
 
@@
 They cover a large period of time and originate from several publishers.
-Therefore, their contents, indexes, layouts, methods of printing, etc. may vary a lot from one directory to another (see fig. \ref{fig:directories}).
+Therefore, their contents, indexes, layouts, methods of printing, etc. may vary a lot from one directory to another (see \cref{fig:directories}).
 We want our groundtruth dataset to be representative of the diversity of directories available in the period.
@@
     \includegraphics[width=\linewidth]{figs/eval-ocr-ner.pdf}
-    \caption{OCR and NER evaluation protocol example.}
+    \caption{OCR and NER evaluation protocol example. The OCR prediction is first aligned on the GOLD text to get the
+    OCR quality. The same alignment is used to project the GOLD tags on the OCR prediction. Projected tags enable
+    the NER system run on the OCR predicted text to be assessed.}
     \label{fig.eval-ocr-ner}
@@
 precision, recall, and f-measure (or f-score) are computed considering only the exact matches between entities of the
-\emph{NER-target} and those from the \emph{NER-prediction}, i.e. pairs of entries for which the type, start and end positions are exactly the same.
-Precision is the ratio of exact matches divided by the number of predicted entries,
-and recall is defined as the ratio of exact matches divided by the number of expected entries;
-the f-measure is their harmonic mean.
+\emph{NER-target} and those from the \emph{NER-prediction}, i.e. pairs of entries for which the type, start and end
+positions are exactly the same. Precision is the ratio of exact matches divided by the number of predicted entries, and
+recall is defined as the ratio of exact matches divided by the number of expected entries; the f-measure is their
+harmonic mean.
 

### 2022-03-29 src-latex/parts/50-xp-setup-BACKUP.tex
Update missing comments

--- 2022-01-20 src-latex/parts/50-xp-setup.tex
@@
 \label{subsection:experiment-2-setup}
-Noise introduced by OCR is known to impede the named entity recognition process (see section~\ref{subsection:stoa-ner-on-historical-texts}).
+Noise introduced by OCR is known to impede the named entity recognition process (see \cref{subsection:stoa-ner-on-historical-texts}).
 %The second experiment assesses the impact of the OCR on the quality of the NER results.

### 2022-03-29 src-latex/parts/60-results-BACKUP.tex
Update missing comments

@@
 \subsection{Experiment 1: NER sensibility to the number of training examples}
-Table \ref{tab:experiment-1-models-performances} and Figure \ref{fig:f1-vs-trainsize} show the mean F1 score, precision, and recall computed on the test dataset, for 5 fine tunings of each model and for each of the 8 training subsets.
+\Cref{tab:experiment-1-models-performances} and \cref{fig:f1-vs-trainsize} show the mean F1 score, precision, and recall computed on the test dataset, for 5 fine tunings of each model and for each of the 8 training subsets.
 CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.
@@
 	       {images/experiment-1-models-performances.pdf}}
-	  \caption{\label{fig:f1-vs-trainsize} Precision, recall and f1 score from table~\ref{tab:experiment-1-models-performances}}.
+	  \caption{\label{fig:f1-vs-trainsize} Precision, recall and f1 score from \cref{tab:experiment-1-models-performances}}.
 \end{figure}

### 2022-04-01 rebuttal/rebuttal_response_max_500_words.txt
Update TODO
Update missing comments
Up TODOs

@@
 
-* [ ] JC rajouter une phrase
+* [ ] ~~JC rajouter une phrase~~ ← Je ne vois pas où en parler, je trouve que cet article a très peu d'intérêt pour nous, je propose de ne pas le citer.
 
@@
 
-* [ ] EC ajouter une phrase ?? (dire qu'on n'avait même pas assez de données pour faire du fine-tuning)
+* [x] EC ajouter une phrase ?? (dire qu'on n'avait même pas assez de données pour faire du fine-tuning)
 
@@
 
-* [ ] EC
+* [x] EC
 
@@
 * [ ] NA relecture readme dataset + critique structure
-* [ ] JC rajouter lien Zenodo dataset
+* [x] JC rajouter lien Zenodo dataset
 * [ ] ?? dépôt HAL + ArXiV
+* [ ] ?? clean repo github
+* [ ] JC upload dataset sur zenodo

### 2022-04-01 src-latex/main-supplementary-material.tex
Supp. material: clean up

@@
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Dataset}
-
-\subsection{Content}
+\section{Dataset Content}
 
@@
 
-\subsubsection{Annotation tools}
+\clearpage
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%      ##    #    #  #    #   ####    #####    ##     #####     #     ####   #    #
+%     #  #   ##   #  ##   #  #    #     #     #  #      #       #    #    #  ##   #
+%    #    #  # #  #  # #  #  #    #     #    #    #     #       #    #    #  # #  #
+%    ######  #  # #  #  # #  #    #     #    ######     #       #    #    #  #  # #
+%    #    #  #   ##  #   ##  #    #     #    #    #     #       #    #    #  #   ##
+%    #    #  #    #  #    #   ####      #    #    #     #       #     ####   #    #
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Annotation tools}
 To build our ground truth corpus, we annotate 78 pages randomly sampled from each of the major directory series for which we have digitised copies, using a tool developed for our project (see Figure \ref{fig:annotator}).
@@
 
-\subsection{Entry segmentation}
+% \subsection{Entry segmentation}
 
-\subsection{Line detection}
-Pero OCR uses ParseNet as internal layout parser for line detection.
-Pero OCR uses an LSTM engine for text line recognition.
-Pero OCR generally works very well, as long as the bounding boxes of the regions to recognize are not too tightly adjusted.
-Trained on newspapers in European languages by its development team.
-Can output results in Latin, Greek and Cyrillic scripts, as well as some commonly-used typographic symbols.
-
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% #####   ######   ####   #    #  #        #####   ####
-% #    #  #       #       #    #  #          #    #
-% #    #  #####    ####   #    #  #          #     ####
-% #####   #            #  #    #  #          #         #
-% #   #   #       #    #  #    #  #          #    #    #
-% #    #  ######   ####    ####   ######     #     ####
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
+% \subsection{Line detection}
+% Pero OCR uses ParseNet as internal layout parser for line detection.
+% Pero OCR uses an LSTM engine for text line recognition.
+% Pero OCR generally works very well, as long as the bounding boxes of the regions to recognize are not too tightly adjusted.
+% Trained on newspapers in European languages by its development team.
+% Can output results in Latin, Greek and Cyrillic scripts, as well as some commonly-used typographic symbols.
 

### 2022-04-01 src-latex/parts/10-introduction.tex
Add section 3
WIP rebuttal
Ref to fig. 1 + mention to thumbnails
Move fig. 1 ref from abstract to introduction
Relecture globale et correction de coquilles

@@
     }
-    \label{<label>}
+    \label{fig.pipeline-overview}
 \end{figure}
@@
 Being active research topics, OCR and named entity recognition (NER) are still difficult tasks when applied to historical text documents.
-OCR approaches used for modern documents are likely to struggle even on printed historical documents due to multiple causes related to text readability (low resolution scans, inconsistent printing rules, artefacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts \& special glyphs) and the variability inherent to the great diversity of historical sources.
+OCR approaches used for modern documents are likely to struggle even on printed historical documents due to multiple causes related to text readability (low resolution scans, inconsistent printing rules, artefacts, show-through), document complexity (intricate and versatile page layout, use of ancient fonts and special glyphs) and the variability inherent to the great diversity of historical sources.
 On the other hand, the semantics of entities in NER approaches developed for modern texts may be different from those in ancient texts.
@@
 Trade directories have been leveraged in recent work to identify polluted urban soils \cite{bell2020automated} and locate all gas stations in the city of Providence over the last century.
-In an ongoing research project, we aim at producing structured spatio-temporal data from the entries of the Paris trade directories to study the dynamics of the fraction of the the XIX\textsuperscript{th} century Parisian society reachable through these sources.
+In an ongoing research project, we aim at producing structured spatio-temporal data from the entries of the Paris trade directories to study the dynamics of the fraction of the XIX\textsuperscript{th} century Parisian society reachable through these sources.
 %Because they originate from several publishers, their content is organised following different indexing methods (by name, by activity or by address), is printed in various layouts, and uses different fonts.
-Therefore, we investigate several state-of-the-art OCR and NER approaches to assess their usability to process the corpus.
+Therefore, we investigate several state-of-the-art OCR and NER approaches to assess their usability to process the corpus (\cref{fig.pipeline-overview}).
 
@@
     \item We evaluate modern NER approaches: their requirements in terms of training data, and the effects of pre-training (\cref{sec:ner-xp1}).
-    \item We show that Transformed-based NER can benefit from pre-training and fine-tuning to improve its performance on noisy OCR (\cref{sec:ner-xp2}).
+    \item We show that Transformer-based NER can benefit from pre-training and fine-tuning to improve its performance on noisy OCR (\cref{sec:ner-xp2}).
 \end{enumerate*}

### 2022-04-01 src-latex/parts/20-related-work.tex
Relecture globale et correction de coquilles

@@
 Most of the time, they have evaluated NER approaches based on deep learning architectures as they seem to adapt more easily to OCR errors than rule-based or more classical supervised approaches.
-\cite{van2020assessing} use the English model \textit{en-core-web-lg} provided by the SpaCy \cite{spacy} library to perform NER on a corpus made of many journal articles with different levels of OCR errors.
+\cite{van2020assessing} used the English model \textit{en-core-web-lg} provided by the SpaCy \cite{spacy} library to perform NER on a corpus made of many journal articles with different levels of OCR errors.
 For each OCRed article, a ground truth text is available so that the Word Error Rate (WER) can be computed.
@@
 Based on those works, we chose to test three OCR systems, namely, Tesseract, \peroocr, and Kraken.
-We also adopt two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that we can adapt to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
+We also adopted two deep-learning-based French language models, available in packaged software libraries, already trained for the NER task and that can be adapted to the domain of historical directories: SpaCy NLP pipelines and CamemBERT.
 In \cref{sec:dataset}, we will explain the evaluation protocol used to assess the combined performance of these OCR and NER systems.
 
-\textbf{Tesseract} is a long-living project, born as a closed-sourced OCR at Hewlett-Packard in the eighties, it was progressively modernized, then open-sourced in 2005.
+\textbf{Tesseract} is a long-living project, born as a closed-sourced OCR at Hewlett-Packard in the eighties; it was progressively modernized, then open-sourced in 2005.
 From 2006 until November 2018, it was developed by Google and is still very active.
-We used in our tests version 4.1.1, released Dec.
+We used in our tests the version 4.1.1, released Dec.
 26, 2019.
@@
 
-\textbf{CamemBERT} is well-known adaptation of the BERT transformer-based model for the French language~\cite{vaswani2017attention,devlin2018bert,martin-etal-2020-camembert}.
+\textbf{CamemBERT} is well-known adaptation of the BERT Transformer-based model for the French language~\cite{vaswani2017attention,devlin2018bert,martin-etal-2020-camembert}.
 Such language models have become a new paradigm for NER~\cite{li2020}.
-The learned embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually referred as a ``head''.
+The learned contextual embeddings can be used as distributed representations for input instead of traditional embeddings like Google Word2Vec, and they can be further fine-tuned for NER by adding an additional output layer, usually referred to as a ``head''.
 They can also be pre-trained in an unsupervised way on large amount of unlabelled texts for domain adaptation.

### 2022-04-01 src-latex/parts/40-ocr-bench.tex
rebuttal points 2/4/5
Update missing comments
Relecture globale et correction de coquilles

@@
 
-This section focuses on the evaluation of the performance of the three open-source OCR systems we selected, as described in \cref{sec:pipeline-summary}: Tesseract v4, Kraken and \peroocr.
-The dataset used to perform this evaluation is composed of the $8,765$ entries (containing $424,764$ characters) from the dataset we previously introduced.
-The single-column, cropped images of entries are used as input of each OCR system.
-As the pages were previously deskewed, the text is mostly horizontal except for a few cases.
-The expected output is the human transcription of these images provided in the dataset.
-Before computing the Character Error Rate (CER) for each entry, each text prediction is normalised with the same basic rules as the ones used to post-process human transcription: dashes, quotes and character codes for glyphs like stars or hands are normalised.
+This section focuses on the evaluation of the performance of the three open-source OCR systems we selected, as described
+in \cref{sec:pipeline-summary}: Tesseract v4, Kraken and \peroocr. The OCRs are used ``out-of-the-box'' with their
+default pre-trained models. No fine-tuning was performed as we did not get enough annotated data for such a task. The
+dataset used to perform this evaluation is composed of the $8,765$ entries (containing $424,764$ characters) from the
+dataset we previously introduced. The single-column, cropped images of entries are used as input of each OCR system. As
+the pages were previously deskewed, the text is mostly horizontal except for a few cases. The expected output is the
+human transcription of these images provided in the dataset. Before computing the Character Error Rate (CER) for each
+entry, each text prediction is normalised with the same basic rules as the ones used to post-process human
+transcription: dashes, quotes and character codes for glyphs like stars or hands are normalised.
 % Regarding the charset of each OCR, Tesseract predicts characters covering a large portion of Latin script, \peroocr generates characters with a similar range, sometimes predicting some Cyrillic characters, and Kraken is constrained by its English model to predict only ASCII characters.
@@
 not as good as the two first OCR. This is partially due to the fact that the closest available model is for English text
-and so it misses French specific symbols. On the other hand, even when using a French model trained on French 19th
+and so it misses French specific symbols. On the other hand, even when using a French model trained on French XIX\textsuperscript{th} century
 documents, the performance does not increase (and relaxing the character matching rules does not help either). Tesseract and
@@
 \peroocr and Tesseract could improve the overall recognition quality. 
-In addition, we will not consider Kraken in the following NER experiments because its recognition rate is already too low.
+%In addition, we will not consider Kraken in the following NER experiments because its recognition rate is already too low.

### 2022-04-01 src-latex/parts/50-ner-xp1-training-set-size-and-pretraining.tex
Update missing comments
Relecture globale et correction de coquilles

@@
 \item \textbf{CamemBERT pre-trained on Paris directories~[CmBERT+ptrn]}: 
-To evaluate whether adapting CamemBERT to the domain increases its performance, we do an unsupervised pre-training of CmBERT for next sentence prediction and masked language modeling, using approx.
+To evaluate whether adapting CamemBERT to the domain increases its performance, we do an unsupervised pre-training of CmBERT for next sentence prediction and masked language modelling, using approx.
 $845,000$ entries randomly sampled and OCRed with PERO.
@@
 To generate smaller training sets, we start from the initial training set and iteratively split it in half using the same stratified sampling strategy as for the train/dev split to maintain the relative frequency of directories.
-We stop if a directory has only one entry left, or if the current training subset contains less than 30 entries, maintaining the relative frequency of directories within it
+We stop if a directory has only one entry left, or if the current training subset contains less than 30 entries, maintaining the relative frequency of directories within it.
 Applying this procedure to the initial training set produces 8 training subsets containing 49, 99, 199, 398, 796, 1593, 3186, and 6373 entries.
@@
 \subsection{Results and discussion}
-Fig. \ref{fig:f1-vs-trainsize} displays the averaged precision, recall, and F1-score for all models on the 8 subsets created from the groundtruth.
+\Cref{fig:f1-vs-trainsize} displays the averaged precision, recall, and F1-score for all models on the 8 subsets created from the groundtruth.
 CmBERT, CmBERT+ptrn and SpaCy NER display the same behaviour: the performances increase dramatically with the number of training examples and rapidly reach an area of slower progress around 1000 examples.

### 2022-04-01 src-latex/parts/60-ner-xp2-pretraining-and-finetuning.tex
Relecture globale et correction de coquilles

@@
 We leave Kraken aside as its performance hinders the projection of NER labels from the ground truth, thus dramatically reducing the size of labelled sets.
-Because the NER metrics are entity-wise, they may underestimate the NER performances on stronly noised texts with fragmented entities.
+Because the NER metrics are entity-wise, they may underestimate the NER performances on very noisy texts with fragmented entities.
 We leverage the labelled sets of entries NER-\emph{reference}, NER-\emph{pero} and NER-\emph{tesseract} created as explained in \cref{sec:dataset}.
@@
 However, despite a loss of performance, the model pre-trained and fine-tuned on NER-pero still gives the best results.
-This is probably due to the fact that the texts produced by \peroocr feature characteristics intermediary between human transcriptions and Tesseract.
+This is probably due to the fact that the texts produced by \peroocr feature characteristics intermediary between human transcriptions and texts produced by Tesseract.
 This OCR tool removes the characters recognised with low confidence, which is probably a great help to the NER.
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{images/experiment_2_f1_with_noise_graph.pdf}
+    \includegraphics[width=0.90\textwidth]{images/experiment_2_f1_with_noise_graph.pdf}
     \caption{\label{fig:exp_2_eval_ner}NER F1-scores in presence of OCR noise in the training and testing sets, grouped by test set.

### 2022-05-24 src/ner/util/paths.py
Updated notebooks

@@ -2,27 +2,23 @@
     A srcipt to export the different paths used by the NER experiments: 
         - BASEDIR:    absolute path to the repository in the local environment
         - DATASETDIR: absolute path to the directory dataset/ that should exist in BASEDIR
-        - NERDIR:     absolute path to the directory src/ner/
-
-    If the notebooks are run in a Google Colab environment then this script will also mount 
-    your Google Drive to the current Google Colab environment. By default GDrive will be
-    mounted in /content/drive.
-    .
+        - NERDIR:     absolute path to the directory src/ner/    .
 """
 
-import dotenv
 import pathlib
-import os.path
+import os
 import sys
 
-_GDRIVE_MOUNT_POINT = "/content/drive"
-
+_GDRIVE_MOUNT_POINT = pathlib.Path("/content/drive")
+_GDRIVE_ROOT_FOLDER = pathlib.Path("MyDrive")
 
 def _mount_google_drive() -> str:
     """ Mount Google Drive to the Google Colab environment. Returns the mount point."""
     from google.colab import drive
-    drive.mount(_GDRIVE_MOUNT_POINT)
-    return pathlib.Path(_GDRIVE_MOUNT_POINT).absolute()
+    already_mounted = os.path.isdir(_GDRIVE_MOUNT_POINT / _GDRIVE_ROOT_FOLDER)
+    if not already_mounted:
+        drive.mount(str(_GDRIVE_MOUNT_POINT))
+    return _GDRIVE_MOUNT_POINT.absolute()
 
 
 def base_dir() -> pathlib.Path:
@@ -30,8 +26,8 @@ def base_dir() -> pathlib.Path:
     is_env_googlecolab = 'google.colab' in str(get_ipython())
 
     if is_env_googlecolab:
-        gdrive_path = _mount_google_drive()
-        base_dir = gcolab_mount_point / os.environ["GCOLAB_BASE_DIR"]
+        gcolab_mount_point = _mount_google_drive()
+        base_dir = gcolab_mount_point / _GDRIVE_ROOT_FOLDER / os.environ['GDRIVE_PAPER_FOLDER']
     else:
         # If not on GColab, BASE will be the directory of this notebook
         this_notebook_location = os.path.dirname(

### 2022-05-24 src/ner/util/camembert_util.py
Ner pipeline (#17)
Updated ner-pipeline notebooks

@@ -2,7 +2,7 @@ import numpy as np
 import nltk
 from xml.dom.minidom import parseString
 from datasets import load_metric, Dataset
-from config import logger
+from util.config import logger
 from transformers import (
     AutoModelForTokenClassification,
     AutoTokenizer,
@@ -19,20 +19,20 @@ LABELS_ID = {
     "I-PER": 2,      # A person, like "Baboulinet (Vincent)"
     "I-MISC": 3,     # Not used but present in the base model
     "I-ORG": 4,      # Not used but present in the base model
-    "I-CARDINAL": 5, # Not used but present in the base model
+    "I-CARDINAL": 5,  # Not used but present in the base model
     "I-ACT": 6,      # An activity, like "plombier-devin"
     "I-TITRE": 7,    # A person's encoded title, like "O. ::LH::" for "Officier de la Légion d'Honneur"
-    "I-FT": 8,       # A feature type, like "fabrique" or "dépot" in front of addresses.
+    # A feature type, like "fabrique" or "dépot" in front of addresses.
+    "I-FT": 8,
 }
-
 
 
 # Entry point
 def init_model(model_name, training_config):
     logger.info(f"Model {model_name}")
     tokenizer = AutoTokenizer.from_pretrained(model_name)
-    
-    #output_path = save_model_path or "/tmp/bert-model"    
+
+    #output_path = save_model_path or "/tmp/bert-model"
     training_args = TrainingArguments(**training_config)
 
     # Load the model
@@ -46,6 +46,8 @@ def init_model(model_name, training_config):
     return model, tokenizer, training_args
 
 # Main loop
+
+
 def train_eval_loop(model, training_args, tokenizer, train, dev, test, patience=3):
     data_collator = DataCollatorForTokenClassification(tokenizer)
     trainer = Trainer(
@@ -64,6 +66,8 @@ def train_eval_loop(model, training_args, tokenizer, train, dev, test, patience=
     return trainer.evaluate(test), trainer.evaluate()
 
 # Metrics
+
+
 def compute_metrics(p):
     predictions, labels = p
     predictions = np.argmax(predictions, axis=2)
@@ -78,19 +82,22 @@ def compute_metrics(p):
         for prediction, label in zip(predictions, labels)
     ]
     metric = load_metric("seqeval")
-    results = metric.compute(predictions=true_predictions, references=true_labels)
+    results = metric.compute(
+        predictions=true_predictions, references=true_labels)
     return {
[…]

### 2022-05-24 src/ner/util/config.py

@@ -1,0 +1,68 @@
+"""
+ Holds the global configuration.
+"""
+
+import util.paths
+import os
+import logging
+from pathlib import Path
+
+# =============================================================================
+# region ~ Configurations
+
+# If true, text versions of the spacy datasets will be saved along with the .spacy files.
+DEBUG = os.getenv("DEBUG", "False").lower() in ("true", "1", "t")
+
+# Random seed used in train/dev/test. Do not change it if you want to recreate the paper results.
+SPLIT_SEED = 42
+
+# endregion
+
+
+# =============================================================================
+# region ~ Logging
+
+logger = logging.getLogger("nerlogger")
+LOG_LEVEL = logging.getLevelName("DEBUG" if DEBUG else "INFO")
+logger.setLevel(LOG_LEVEL)
+logformatter = logging.Formatter(
+    "%(asctime)s ; %(levelname)s ; %(message)s", datefmt="%d/%m/%Y %I:%M:%S"
+)
+loghandler = logging.StreamHandler()
+loghandler.setFormatter(logformatter)
+logger.addHandler(loghandler)
+
+
+# endregion
+
+# =============================================================================
+# region ~ Paths
+
+
+BASEDIR = util.paths.BASEDIR
+DATASETDIR = util.paths.DATASETDIR
+NERDIR = util.paths.NERDIR
+
+# endregion
+
+
+# =============================================================================
+# region ~ Repeatability
+
+AS_IN_THE_PAPER = os.getenv(
+    "AS_IN_THE_PAPER", "True").lower() in ("true", "1", "t")
+
+# endregion
+
+
+def show() -> None:
+    logger.info("======= CONFIGURATION =======")
+    logger.info(f"BASEDIR: {BASEDIR}")
+    logger.info(
+        f"Input datasets will be loaded from DATASETDIR {DATASETDIR}")
+    logger.info(
+        f"Training data and models will be saved to NERDIR {NERDIR}")
+    logger.info(f"Debug mode is {'ON' if DEBUG else 'OFF.'}")
+    logger.info(f"Random seed: {SPLIT_SEED}")
+    logger.info(f"Enable reproducibility checks: {AS_IN_THE_PAPER}")
+    logger.info("============================")

### 2022-05-24 src/ner/util/dataset_util.py

@@ -2,16 +2,19 @@ import numpy as np
 from sklearn.model_selection import GroupKFold
 from sklearn.model_selection import train_test_split
 from datasets.dataset_dict import DatasetDict
-from spacy_util import create_spacy_dataset
-from camembert_util import create_huggingface_dataset
-import config
+from util.spacy_util import create_spacy_dataset
+from util.camembert_util import create_huggingface_dataset
+import util.config as config
+
 
 def unwrap(list_of_tuples2):
     return tuple(zip(*list_of_tuples2))
 
+
 def file_name(*parts, separator="_"):
     parts_str = [str(p) for p in parts]
-    return separator.join(filter(None,parts_str))
+    return separator.join(filter(None, parts_str))
+
 
 def train_dev_test_split(gold):
     """Splits the gold dataset into two subsets where entries from the same directory is garanteed
@@ -30,7 +33,8 @@ def train_dev_test_split(gold):
     #   - Duverneuil_et_La_Tynna_1806 (186 entries)
 
     _, groups = unwrap(gold)
-    index_tmp, index_test = list(GroupKFold(n_splits=5).split(gold, groups=groups))[0]
+    index_tmp, index_test = list(GroupKFold(
+        n_splits=5).split(gold, groups=groups))[0]
     subset_tmp, test = gold[index_tmp], gold[index_test]
 
     # Split subset_tmp into train (~90%) and dev (~10%) stratified on directories names
@@ -46,12 +50,12 @@ def train_dev_test_split(gold):
 def save_dataset(output_dir, datasets, names, suffix=None):
     """Export all datasets in Spacy and Huggingface native formats."""
     assert len(datasets) == len(names)
-    
+
     hf_dic = {}
     for ds, name in zip(datasets, names):
 
         examples, _ = unwrap(ds)
-        
+
         # Spacy
         spacy_file = output_dir / file_name("spacy", name, suffix)
         spacy_file = spacy_file.with_suffix(".spacy")
@@ -69,4 +73,3 @@ def save_dataset(output_dir, datasets, names, suffix=None):
     bert_ds = DatasetDict(hf_dic)
     hf_file = output_dir / file_name("huggingface", suffix)
     bert_ds.save_to_disk(hf_file)
-

### 2022-05-24 src/ner/util/paths.py

@@ -1,0 +1,45 @@
+"""
+    A srcipt to export the different paths used by the NER experiments: 
+        - BASEDIR:    absolute path to the repository in the local environment
+        - DATASETDIR: absolute path to the directory dataset/ that should exist in BASEDIR
+        - NERDIR:     absolute path to the directory src/ner/    .
+"""
+
+import pathlib
+import os
+import sys
+
+_GDRIVE_MOUNT_POINT = pathlib.Path("/content/drive")
+_GDRIVE_ROOT_FOLDER = pathlib.Path("MyDrive")
+
+def _mount_google_drive() -> str:
+    """ Mount Google Drive to the Google Colab environment. Returns the mount point."""
+    from google.colab import drive
+    already_mounted = os.path.isdir(_GDRIVE_MOUNT_POINT / _GDRIVE_ROOT_FOLDER)
+    if not already_mounted:
+        drive.mount(str(_GDRIVE_MOUNT_POINT))
+    return _GDRIVE_MOUNT_POINT.absolute()
+
+
+def base_dir() -> pathlib.Path:
+    """ Returns the absolute path to the repository in the local environment."""
+    is_env_googlecolab = 'google.colab' in str(get_ipython())
+
+    if is_env_googlecolab:
+        gcolab_mount_point = _mount_google_drive()
+        base_dir = gcolab_mount_point / _GDRIVE_ROOT_FOLDER / os.environ['GDRIVE_PAPER_FOLDER']
+    else:
+        # If not on GColab, BASE will be the directory of this notebook
+        this_notebook_location = os.path.dirname(
+            os.path.realpath("__file__"))
+        base_dir = pathlib.Path(this_notebook_location).parent.parent
+        # make base_dir absolute
+        base_dir = base_dir.absolute()
+
+    return base_dir
+
+
+BASEDIR = base_dir()
+DATASETDIR = base_dir() / "dataset"  # Path to the input datasets
+# Where to store the models & data produced by the NER experiments
+NERDIR = base_dir() / "src/ner"
